---
title: "02 netcoloc obs expect"
author: "Karissa Barthelson"
date: "2026-06-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Setup
```{r}
library(tidyverse)
library(magrittr)
library(readxl)
library(ggpubr)

theme_set(theme_light() +
            theme(plot.title = element_text(hjust = 0.5), 
                  strip.text = element_text(face = "bold", colour = "white"),
                  strip.background = element_rect(fill = "grey20")
            )
)

knitr::opts_chunk$set(
  message = F, echo = F, warning = F
)
```

## Helper functions
```{r}
# run KEGG ORA
runKEGG = function(geneset) {

  uv =   bitr(nodes, 
                fromType = "SYMBOL", toType = "ENTREZID", 
                OrgDb ="org.Hs.eg.db") %>% 
    .$ENTREZID
  
  bitr(geneset, 
       fromType = "SYMBOL", toType = "ENTREZID", 
     OrgDb ="org.Hs.eg.db") %>% 
  .$ENTREZID %>%
  enrichKEGG(
    organism = 'hsa',
    universe = uv,
    pvalueCutoff = 0.05)
}

```


# Pos & neg control colocs
## import data

Here, I will import the dfs needed for comparing our pos and neg control disease comparisons: 

- AD and PD 
- ALS and FTLD
- PD and DLB
- CD and brain malform (neg control)
- AD and glioma (neg control)
for glioma, I obtain Supp Data 1 from here: https://www.nature.com/articles/ng.3823#MOESM53

This is the list of SNPs (germline) which are associated with increased risk for glioma. 

I them took those SNPs and input into gProfiler::SNPense, which tells me which gene the SNPs are in. Only SNPs with a single protein coding gene were kept. 

```{r}
glioma.snps = read_xlsx("data/glioma/Melin etal 2017 STab1.xlsx", skip = 3) %>% 
  .$`...2` %>% 
  na.omit %>% as.vector()

glioma.genes = gprofiler2::gsnpense(glioma.snps) %>% 
  as_tibble() %>% 
  .$gene_names

# # write this out for netcoloc. 
# tibble(GENE = glioma.genes[glioma.genes %in% nodes] %>% unique) %>% 
#   write_csv("data/genesets/glioma.csv")

```

The input gene lists for the adult onset diseases will be the GWAS 1 hit per loci gene sets

```{r}
nodes <- readRDS("data/R_objects/network/nodes.rds")
 
enrichment_dfs = list.files("data/results/control contrasts", 
           full.names = T, recursive = T, 
           pattern = "network_enrichment.csv") %>% 
  set_names(basename(dirname(.))) %>% 
  map(read.csv) %>% 
  bind_rows(.id = "comparison") %>%
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>%
  mutate(
    FDR = p.adjust(empirical_p, "fdr") %>% signif(2)
    ) %>% 
  pivot_longer(
    cols = c(observed_overlap, expected_overlap_mean),
    names_to = "type",
    values_to = "value"
  ) %>%
  arrange(desc(FDR)) %>% 
  mutate(
    type = recode(
      type,
      "observed_overlap" = "observed",
      "expected_overlap_mean" = "expected"), 
    comparison = str_replace(comparison, "_", "\n&\n"), 
    comparison = fct_inorder(comparison)

  )
```
## obs vs expected ratio
```{r}
enrichment_dfs %>% 
 ggplot(
    aes(x = type, y = value, fill = type)
    ) +
  geom_bar(
    stat = "identity", position = "dodge", width = 0.8
  ) +
  geom_errorbar(
    data = . %>% 
      dplyr::filter(type == "expected"),
    aes(
      ymin = value - 2 * expected_overlap_std,
      ymax = value + 2 * expected_overlap_std),
    width = 0.25
  ) + 
  geom_label(
    data =  . %>%
      dplyr::filter(type == "expected"),
    aes(
      x = 1.5, y =600,
      label = paste("p =", FDR)),
    inherit.aes = FALSE,
    fill = NA,
    border.colour = NA,
    colour = "grey20",
    size = 5,
    vjust = 0.2,
    fontface = "bold",
    show.legend = FALSE
  )  +
  labs(
    y = "Number of genes", 
    x = ""
  ) +
  facet_wrap(~comparison, nrow = 1, scales = "free") +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("grey50", "skyblue"))+
   theme(
    strip.text = element_text(size = 12),
    legend.position = "none",
    axis.title.x = element_blank(),
    plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) 
```

# OR of top pathways. 
```{r}
# read in the genes in the subnets
subnet.genes = list.files("data/results/control contrasts", 
           full.names = T, recursive = T, 
           pattern = "overlapping.+.csv") %>% 
  set_names(basename(dirname(.))) %>% 
  map(read.csv) %>% 
  bind_rows(.id = "comparison") %>% 
  # only keep sig subnets
  dplyr::filter(
    grepl(comparison, pattern = "PD|ALS")
  ) %>% 
  split(f = .$comparison) %>% 
  lapply(pull, "X")

ORA.controls.KEGG.results = subnet.genes %>% 
  lapply(runKEGG)

p2 = ORA.controls.KEGG.results %>% 
  map(as_tibble) %>% 
  bind_rows(.id = "comparison") %>% 
  separate(GeneRatio,into = c("temp1", "temp2"), remove = F) %>% 
  separate(BgRatio, into = c("bg1", "bg2"), sep = "/", remove = FALSE) %>% 
   group_by(comparison) %>% 
  dplyr::slice(1:7) %>% 
  mutate(
    across(c(temp1, temp2, bg1, bg2), as.numeric),
    
    a = temp1 + 0.5,
    b = bg1 - temp1 + 0.5,
    c = temp2 - temp1 + 0.5,
    d = bg2 - bg1 - (temp2 - temp1) + 0.5,
    
    OR    = (a * d) / (b * c),
    log2OR = log2(OR),
    SE    = sqrt(1/a + 1/b + 1/c + 1/d),
    CI_lo = exp(log(OR) - 1.96 * SE),
    CI_hi = exp(log(OR) + 1.96 * SE),
    
  
    Description = reorder_within(Description, OR, comparison), 
    Description = str_wrap(Description, width = 30),
  
  
    GeneRatio = as.numeric(temp1)/as.numeric(temp2), 
    comparison = str_replace(comparison, pattern = "_", "\n&\n")
  ) %>% 
  
  ggplot(
    aes(
      y = reorder(Description, OR), x = log2OR
    )
  ) +
  geom_point(
    aes(
      colour = -log10(p.adjust), size = temp1
      )
  ) +
  geom_errorbarh(
    aes(xmin = log2(CI_lo), xmax = log2(CI_hi)),
    height = 0.3, linewidth = 0.6
    ) +
 #scale_x_continuous(limits = c(0,12))+
  scale_y_reordered() +  # This removes the `___comparison` suffix added by reorder_within()
  scale_colour_viridis_c() +
   geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  labs(
    x = "log(OR) ± 95% CI",
    y = NULL,
    size = "Gene count"
  ) +
  facet_wrap(~comparison, 
             scales = "free", nrow = 1)+
   theme(
    text = element_text(size = 14),
    legend.position = "top",
    #axis.title.x = element_blank(),
    # plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) 
```

# New figure 1
```{r}
p1 = enrichment_dfs %>% 
 ggplot(
    aes(x = type, y = value, fill = type)
    ) +
  geom_bar(
    stat = "identity", position = "dodge", width = 0.8
  ) +
  geom_errorbar(
    data = . %>% 
      dplyr::filter(type == "expected"),
    aes(
      ymin = value - 2 * expected_overlap_std,
      ymax = value + 2 * expected_overlap_std),
    width = 0.25
  ) + 
  geom_label(
    data =  . %>%
      dplyr::filter(type == "expected"),
    aes(
      x = 1.5, y =600,
      label = paste("p =", FDR)),
    inherit.aes = FALSE,
    fill = NA,
    border.colour = NA,
    colour = "grey20",
    size = 5,
    vjust = 0.8,
    fontface = "bold",
    show.legend = FALSE
  )  +
  labs(
    y = "Number of genes", 
    x = ""
  ) +
  facet_wrap(~comparison, nrow = 1, scales = "free") +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("grey50", "skyblue"))+
   theme(
    strip.text = element_text(size = 12),
    legend.position = "none",
    axis.title.x = element_blank(),
    plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) 

p2 = ORA.controls.KEGG.results %>% 
  map(as_tibble) %>% 
  bind_rows(.id = "comparison") %>% 
  separate(GeneRatio,into = c("temp1", "temp2"), remove = F) %>% 
  separate(BgRatio, into = c("bg1", "bg2"), sep = "/", remove = FALSE) %>% 
   group_by(comparison) %>% 
  dplyr::slice(1:7) %>% 
  mutate(
    across(c(temp1, temp2, bg1, bg2), as.numeric),
    
    a = temp1 + 0.5,
    b = bg1 - temp1 + 0.5,
    c = temp2 - temp1 + 0.5,
    d = bg2 - bg1 - (temp2 - temp1) + 0.5,
    
    OR    = (a * d) / (b * c),
    log2OR = log2(OR),
    SE    = sqrt(1/a + 1/b + 1/c + 1/d),
    CI_lo = exp(log(OR) - 1.96 * SE),
    CI_hi = exp(log(OR) + 1.96 * SE),
    
  
    Description = reorder_within(Description, OR, comparison), 
    Description = str_wrap(Description, width = 30),
  
  
    GeneRatio = as.numeric(temp1)/as.numeric(temp2), 
    comparison = str_replace(comparison, pattern = "_", "\n&\n")
  ) %>% 
  
  ggplot(
    aes(
      y = reorder(Description, OR), x = log2OR
    )
  ) +
  geom_point(
    aes(
      colour = -log10(p.adjust), size = temp1
      )
  ) +
  geom_errorbarh(
    aes(xmin = log2(CI_lo), xmax = log2(CI_hi)),
    height = 0.3, linewidth = 0.6
    ) +
 #scale_x_continuous(limits = c(0,12))+
  scale_y_reordered() +  # This removes the `___comparison` suffix added by reorder_within()
  scale_colour_viridis_c() +
   geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  labs(
    x = "log(OR) ± 95% CI",
    y = NULL,
    colour = "-log10(FDR)",
    size = "Gene count"
  ) +
  facet_wrap(~comparison, 
             scales = "free", nrow = 1)+
   theme(
    text = element_text(size = 14),
    legend.position = "bottom",
    #axis.title.x = element_blank(),
    # plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) 
```

```{r}
ggarrange(
  p1, p2, 
  nrow = 2, 
  labels = "AUTO",
  heights = c(0.8, 1.2)
) +
  ggsave("output/plots4publication/Fig1 control_contrasts.pdf",
         width = 6, height = 4, 
         scale = 2)
```

# export
```{r}
ORA.controls.KEGG.results %>% 
  lapply(as_tibble) %>% 
  writexl::write_xlsx(
    "output/spreadsheets/ORA_controls_KEGG_results.xlsx", format_headers = TRUE)
```


