---
title: "03 pan-neurodegen"
author: "Karissa Barthelson"
date: "2026-06-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r}
library(tidyverse)
library(magrittr)
library(readxl)
library(ggpubr)
library(tidytext) # for reorderwithin

library(clusterProfiler)
library(MPO.db)
library(homologene)

library(org.Hs.eg.db)
library(org.Mm.eg.db)

theme_set(theme_light() +
            theme(plot.title = element_text(hjust = 0.5), 
                  strip.text = element_text(face = "bold", colour = "white"),
                  strip.background = element_rect(fill = "grey20")
            )
)

knitr::opts_chunk$set(
  message = F, echo = F, warning = F
)
```

# CD all and adult all

Now I want to do a pan-neurodegenerative analssis - all the CD and adult genes combined. I looked at whether the adult and childhood-onset dementia genetics sig coloc in the network space. yes they do. 

## import data & helper functions
```{r}
# All nodes in PCnet
nodes <- readRDS("data/R_objects/network/nodes.rds")

# run KEGG ORA
runKEGG = function(geneset) {

  uv =   bitr(nodes, 
                fromType = "SYMBOL", toType = "ENTREZID", 
                OrgDb ="org.Hs.eg.db") %>% 
    .$ENTREZID
  
  bitr(geneset, 
       fromType = "SYMBOL", toType = "ENTREZID", 
     OrgDb ="org.Hs.eg.db") %>% 
  .$ENTREZID %>%
  enrichKEGG(
    organism = 'hsa',
    universe = uv,
    pvalueCutoff = 0.05)
}
```

```{r}
input.f2a = read_csv("data/results/pan neurodegen analysis - adult v2/CD_all_adult_allv2/network_enrichment.csv") %>% 
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>%
  pivot_longer(
    cols = c(observed_overlap, expected_overlap_mean),
    names_to = "type",
    values_to = "value"
  ) %>%
  mutate(
    type = recode(
      type,
      "observed_overlap" = "observed",
      "expected_overlap_mean" = "expected"), 
    comparison = "Childhood dementia\n&\nadult-onset\nneurodegen",
    FDR = case_when(
      empirical_p < 0.0001 ~ "p.FDR < 0.0001"
    )
  ) 
  
input.f2a %>% 
ggplot(
    aes(x = type, y = value, fill = type)
    ) +
  geom_bar(
    stat = "identity", position = "dodge", width = 0.8
  ) +
  geom_errorbar(
    data = . %>% 
      dplyr::filter(type == "expected"),
    aes(
      ymin = value - 2 * expected_overlap_std,
      ymax = value + 2 * expected_overlap_std),
    width = 0.25
  ) + 
  geom_label(
    data =  . %>%
      dplyr::filter(type == "expected"),
    aes(
      x = 1.5, y =800,
      label = FDR),
    inherit.aes = FALSE,
    fill = NA,
    border.colour = NA,
    colour = "grey20",
    size = 5,
    vjust = 0,
    fontface = "bold",
    show.legend = FALSE
  )  +
  labs(
    y = "Number of genes", 
    x = ""
  ) +
  facet_wrap(~comparison, nrow = 1, scales = "free") +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("grey50", "skyblue"))+
   theme(
    strip.text = element_text(size = 12),
    legend.position = "none",
    axis.title.x = element_blank(),
    plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) 
  
  
```

### ORA KEGG of pan-neurodegen
```{r}
# extract the genes in the pan-neurodegen subnetwork
network.genes.pan.neurodegen <-  "data/results/pan neurodegen analysis - adult v2/CD_all_adult_allv2/overlapping_nodes_metadata.csv" %>% 
  read.csv %>%
  mutate(
     comparison = "Childhood dementia\n&\nadult-onset\nneurodegen"
  ) %>% 
  as_tibble %>% 
  pull(X)

# Run ORA of KEGG pathways
ORA.network.genes.pan.peurodegen.results = runKEGG(network.genes.pan.neurodegen)
```

```{r}
ORA.network.genes.pan.peurodegen.results %>% 
  as_tibble %>% 
  mutate(
    comparison = "Childhood dementia\n&\nadult-onset\nneurodegen"
  ) %>% 
  separate(GeneRatio,into = c("temp1", "temp2")) %>% 
  mutate(GeneRatio = as.numeric(temp1)/as.numeric(temp2)) %>% 
  group_by(comparison) %>% 
  dplyr::slice(1:10) %>% 
  mutate(
    Description = reorder_within(Description, GeneRatio, comparison), 
    Description = str_wrap(Description, width = 22)
  ) %>%
  ggplot(
    aes(
      y = reorder(Description, GeneRatio), x = GeneRatio
    )
  ) +
  geom_col(
    aes(fill = -log10(p.adjust))
  ) +
  scale_y_reordered() +  # This removes the `___comparison` suffix added by reorder_within()
  scale_fill_viridis_c() +
  labs(
    y = "KEGG pathway"
  ) +
  facet_wrap(~comparison, 
             scales = "free_y", nrow = 1)
```

## ORA of MPO terms

We can also run mouse MPO enrichment analysis
```{r}
# this was obtained fromhttps://www.informatics.jax.org/downloads/reports/MGI_PhenoGenoMP.rpt

mm_symbol2mp_id = 
  read_delim(
    "data/MPO/HMD_HumanPhenotype.txt", delim = "\t") %>% 
  dplyr::select(1,4) %>% 
  set_colnames(
    c("mm_symbol", "MP_id")
  ) %>% 
  mutate(
    mm_symbol = str_remove(mm_symbol, pattern = "<.+")
  )

# mpid 2 mp name
mpo_mgi <- AnnotationDbi::select(
  MPO.db,
  keys    = keys(MPO.db, keytype = "mpid"),
  columns = c("mpid", "term"),
  keytype = "mpid"
) %>% 
  dplyr::select(
    MP_id = mpid, mp_full_name = term
  ) %>% 
  as_tibble

# mouse name to human name
mm2hs <- homologene(
  genes  = unique(mm_symbol2mp_id$mm_symbol),
  inTax  = 10090,  # mouse
  outTax = 9606    # human
) %>% 
  dplyr::select(
    mm_symbol = `10090`, hs_symbol = `9606`, 
    mm_entrez = `10090_ID`, hs_entrez = `9606_ID`
  ) %>% 
  mutate(across(everything(), as.character)) %>% 
  as_tibble

# join everything
mpo.df <- mm_symbol2mp_id %>% 
   left_join(mpo_mgi) %>% 
  left_join(mm2hs)

mpo.df.hs.entrez = 
  mpo.df %>% 
  dplyr::filter(hs_symbol %in% nodes) %>% 
  dplyr::select(mp_full_name, hs_entrez) %>% 
  arrange(mp_full_name)

```

```{r}
ORA.network.genes.pan.peurodegen.results.MPO <- enricher(
  gene          = bitr(network.genes.pan.neurodegen, 
                  fromType = "SYMBOL", toType = "ENTREZID", 
                  OrgDb ="org.Hs.eg.db") %>% .$ENTREZID,
  TERM2GENE     = mpo.df.hs.entrez,
  pvalueCutoff  = 0.05,
  minGSSize     = 5,
  maxGSSize     = 500,
  qvalueCutoff  = 0.05
)

ORA.network.genes.pan.peurodegen.results.MPO %>% 
  as_tibble %>% 
  separate(GeneRatio,into = c("temp1", "temp2")) %>% 
  mutate(GeneRatio = as.numeric(temp1)/as.numeric(temp2)) %>% 
  dplyr::slice(1:10) %>% 
  mutate(
    Description = str_wrap(Description, width = 22)
  ) %>%
  ggplot(
    aes(
      y = reorder(Description, GeneRatio), x = GeneRatio
    )
  ) +
  geom_col(
    aes(fill = -log10(p.adjust))
  ) +
  scale_fill_viridis_c() +
  labs(
    title = "MPO terms",
    y = "MPO term"
  )
```

### ORA of druggable biology
```{r}
drugbank <- readRDS("random/drugbank/joined_atc_df.rds")
fulldb.drugbank <- readRDS("random/drugbank/fulldb.rds")

# convert symbol to entrez using bit
drugbank.df.for.ORA = drugbank$identifier %>% 
  bitr(fromType = "SYMBOL", toType = "ENTREZID", 
                  OrgDb ="org.Hs.eg.db") %>% 
  as_tibble %>% 
  dplyr::rename(identifier = SYMBOL) %>% 
  left_join(drugbank, by = "identifier") %>% 
  dplyr::filter(identifier %in% nodes ) %>% 
  dplyr::select(level_1, ENTREZID) %>% 
  na.omit(level_1) %>% 
  unique


ORA.network.genes.pan.peurodegen.results.drugbank <- enricher(
  gene          = bitr(network.genes.pan.neurodegen, 
                  fromType = "SYMBOL", toType = "ENTREZID", 
                  OrgDb ="org.Hs.eg.db") %>% .$ENTREZID,
  TERM2GENE     = drugbank.df.for.ORA,
  pAdjustMethod = "fdr",
  minGSSize     = 2,
  maxGSSize     = 5000,
)

ORA.network.genes.pan.peurodegen.results.drugbank %>% 
  setReadable(OrgDb = org.Hs.eg.db, keyType = "ENTREZID") %>% 
  as_tibble %>% view

drugbank %>% 
  dplyr::filter(
    level_1 == "Digitalis glycosides"
) %>% 
  .$name.y %>% 
  unique

drugbank %>% 
  dplyr::filter(
    level_1 == "Strophanthus glycosides"
) %>% 
  .$name.y %>% 
  unique
```


