---
title: "03_hsDLB_snRNAseq_pseudobulk"
author: "Karissa Barthelson"
date: "2025-07-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# setup
```{r}
knitr::opts_chunk$set(
  message = F, echo = F, warning = F
)
```

```{r}
library(tidyverse)
library(magrittr)
library(readxl)
library(qs) # for fast save and reading of objects into memory (alternative to rds)
library(parallel) 
library(Matrix)

# annotation
library(org.Hs.eg.db)
library(KEGGREST)

# analysis
library(Seurat)
library(ngsReports)
library(edgeR)
library(clusterProfiler)

# vis tools
library(ggpubr)
library(scales)
library(patchwork)
library(cowplot)
library(viridisLite)
library(ggeasy)
library(ggfortify)
library(ggrepel)
library(tidyHeatmap)
library(ComplexHeatmap)
library(circlize)
library(grid)

theme_set(theme_light() +
            theme(plot.title = element_text(hjust = 0.5), 
                  strip.text = element_text(face = "bold"),
                  strip.background = element_rect(fill = "grey20")
            ) 
)

```

# Introduction
Here, I will perform the pseudobulk comparisons within cell type between DLB and contorl brains. 

The preprocessing of the data was performed previously and the seurat object was loaded. It was then converted to a PB object for use in edgeR.

Note that I have tried having each celltype have its own subtypes (e.g. microglia 1, microglia 2), but this results in not enough reads per celltype to do pseudo-bulk analyses. Therefore, I will combine each celltype. 

```{r}
# read in smaple metadata
meta <- 
  data.table::fread("data/public data/GSE178146_RAW/GSE178146_series_matrix.txt", 
                          skip = 27, sep = "\t", fill = T) %>% 
  mutate(
    `!Sample_title` = str_remove(`!Sample_title`, pattern = "\\!")
  ) %>% 
  t %>% 
  .[,c(1,7,9)] %>% 
  set_colnames(
    c("geo_acc", "brain_region", "disease")
  ) %>% 
  .[-1,] %>% 
  as.data.frame() %>% 
  mutate(
    sample = rownames(.), 
    disease = str_remove(disease, pattern ="disease state: ")
  ) %>% 
  as_tibble()

# seurat object
data.seurat <- qread("data/public data/GSE178146_RAW/R_objects/PD_DLB_data_seurat.qs")

# need to merge layers of the seurat object before converting to PB object
# Seurat::DimPlot(data.seurat, 
#         reduction = "umap", group.by = c("orig.ident", "celltype"), shuffle = TRUE)
# 
# Seurat::DimPlot(data.seurat, 
#         reduction = "umap", 
#         group.by = "celltype", 
#         split.by = "disease",
#         shuffle = TRUE)
        
data.seurat[["RNA"]] <- JoinLayers(data.seurat[["RNA"]])

y <- Seurat2PB(data.seurat, sample="orig.ident", cluster="celltype")

 # tidyup samples
y$samples %<>% 
  as_tibble(rownames = "sample_cluster") %>% 
  left_join(meta %>% dplyr::select(sample, disease)) %>% 
  mutate(
    disease = factor(disease, levels = c("Control", "DLB", "PD", "PDD")), 
    cluster = as.factor(cluster), 
    sample_id = rownames(.),
    sample_cluster = as.factor(sample_cluster),
    disease_cluster = paste(disease, cluster, sep = "_") %>% as.factor
  ) %>%  
  column_to_rownames("sample_cluster") %>% 
  mutate(sample_id = rownames(.))


# Get all Entrez Gene IDs
all_genes <- keys(org.Hs.eg.db, keytype = "SYMBOL")

# Select gene SYMBOL and CHR (chromosome) info
hs_chrs <- AnnotationDbi::select(org.Hs.eg.db,
                    keys = all_genes,
                    columns = c("SYMBOL", "CHR"),
                    keytype = "SYMBOL")
```

# work out sex
The sex of each donor was not provided by GEO. Here, i will look at this by assessing expression of genes on the Y chromosome. There appears to be reads are aligning to genes  on the Y chromosome, however, these are in the pseudoautosomal regiosns. Nevertheless, there are clear differences in expression of genes on the Y chr. 

```{r}
y %>% 
  cpm(log=T) %>% 
  as_tibble(rownames = "SYMBOL") %>% 
  left_join(hs_chrs) %>% 
  pivot_longer(names_to = "sample_cluster", 
               values_to = "counts", rownames(y$samples)) %>% 
  dplyr::filter(CHR == "Y") %>% 
  mutate(
    sample = str_remove(sample_cluster, pattern = "_cluster.+")
  ) %>% 
  ggplot(
    aes(x = sample, y = counts)
  ) +
  geom_boxplot()+
  facet_wrap(~CHR) +
  scale_y_continuous(limits = c(0,10))
```

Whle the sample-spefiic info was NA in the paper and in GEO, there was this in the paper: 

`proportion female: control = 1/7, PD = 5/7, PDD = 2/7, DLB = 0/7`

So the samples which have low y-specific chr expression are likely the female samples. This all matches up nicely. 

```{r}
female.samples = c(
  #control
  "PDC87", 
  #PD - 5 samples
  "PD413", "PD416", "PD523", "PD666", "PD683", 
  
  # PDD - 2 samples
  "PD366", "PD501"
  )


```

```{r}
y$counts %>% 
  cpm(log=T) %>% 
  as_tibble(rownames = "SYMBOL") %>% 
  left_join(hs_chrs) %>% 
  pivot_longer(names_to = "sample_cluster", 
               values_to = "counts", rownames(y$samples)) %>% 
  dplyr::filter(CHR == "Y") %>%
  # dplyr::filter(SYMBOL %in% c("ZFY", "PCDH11Y", "USP9Y", "UTY")) %>% 
  mutate(
    sample = str_remove(sample_cluster, pattern = "_cluster.+"), 
    sex = case_when(
      sample %in% female.samples ~ "female", 
      TRUE ~ "male"
    ) %>% as.factor()
  ) %>%
  left_join(meta) %>% 
  ggplot(
    aes(x = sample, y = counts, fill = sex)
  ) +
  geom_boxplot()+
  facet_wrap(~disease, scales = "free_x") +
  scale_y_continuous(limits = c(0,10)) +
  labs(
    subtitle = "proportion female: control = 1/7, PD = 5/7, PDD = 2/7, DLB = 0/7"
  )
```

```{r}
# add this info back into y$samples
y$samples %<>% 
  mutate(
    sex = case_when(
      sample %in% female.samples ~ "female", 
      TRUE ~ "male"
    ) %>% as.factor()
  ) 
```

# lib sizes
## plot number of cells per celltype

Generally there are less cells in the DLB cases vs rest
```{r}
data.seurat@meta.data %>%
  as_tibble(rownames = "umi") %>% 
  group_by(disease, celltype) %>% 
  summarise(n = n()) %>% 
  ggplot(
    aes(x = celltype, y = n)
  ) +
  geom_col(
    aes(fill = disease), 
    position = "dodge"
  ) +
  labs(
    x = "Cell type", 
    y = "Number of nuclei",
    title = "Total number of nuclei detected in the experiment"
  ) +
  scale_fill_manual(values = c("#5F676D", "#17BEBB", "#FE5448", "#A20C01")) 

  

y$samples %>% 
  as.data.frame() %>% 
  rownames_to_column("sample_cluster") %>% 
  arrange(disease) %>% 
  # makase sample cluster a factor to group within disease
  mutate(
    sample_cluster = str_remove(sample_cluster, "cluster"),
    sample = fct_inorder(sample)
    ) %>% 
  ggplot(
    aes(x = sample, y = lib.size, fill = disease)
  ) + 
  geom_col() +
  scale_y_continuous(labels = comma) +
  facet_wrap(~cluster, scales = "free_x", nrow = 2) +
  easy_rotate_x_labels(angle = -45) +
  labs(
    x = "Disease and Celltype", 
    y = "Library size", 
    title = "Number of reads per celltype"
  ) +
   scale_fill_manual(values = c("#5F676D", "#17BEBB", "#FE5448", "#A20C01")) 


# retain samsample# retain samples with more than 50k counts
keep.samples <- y$samples$lib.size > 5e4
y <- y[,keep.samples]

y$samples %>% 
  as.data.frame() %>% 
  rownames_to_column("sample_cluster") %>% 
  mutate(sample_cluster = str_remove(sample_cluster,pattern = "cluster")) %>% 
  ggplot(
    aes(x = sample_cluster, y = lib.size, fill = disease)
  ) + 
  geom_col() +
  scale_y_continuous(labels = comma) +
  facet_wrap(~cluster, scales = "free", nrow = 2) +
  easy_rotate_x_labels(angle = -45) +
  labs(
    title = "Library sizes after filtering"
  )
```

After filtering the small pseudobulk libraries (50k reads was the recommended by the edgeR manual),There are some celltypes n.d. in all groups. 

I will only retain clusters which are detected in all grousp. 

```{r}
clusters2remove = c("Endothelial", "VLMCs")

samples2keep <- y$samples %>% 
  dplyr::filter(!cluster %in% clusters2remove) %>% 
  rownames(.)

# make a copy of y before filtering and filter low expressed genes
y.before.removing.clusters = y
keep.genes <- filterByExpr(
  y.before.removing.clusters,  
  group=y.before.removing.clusters$samples$cluster)

y.before.removing.clusters <- y.before.removing.clusters[keep.genes, , keep=FALSE]

y <- y[,samples2keep]
```

Then, lowly expressible genes are filterd. the table below shows the number of genes retained. TMM normalization is performed to estimate effective library sizes. 

Lib sizes are much smaller in the DLB group. Is this biological or technical in nature?

```{r}
keep.genes <- filterByExpr(y, group=y$samples$cluster)
table(keep.genes)
y <- y[keep.genes, , keep=FALSE]

y <- normLibSizes(y, method = "TMM")

y$samples %>% 
  ggplot(
    aes(y = lib.size, x = cluster, colour = disease )
  ) +
  geom_boxplot(
    outlier.shape = NA
  ) +
  geom_point(position = position_jitterdodge()) +
  facet_wrap(~cluster, scales = "free_x") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Total number of nuclei per celltype and sample"
    
  )

y %>% 
  cpm(log = T) %>% 
  as.data.frame() %>% 
  pivot_longer(cols = everything(), 
               names_to = "sample", values_to = "logCPM") %>% 
  left_join(
    y$samples %>% 
              mutate(sample = rownames(y$samples))
    ) %>% 
  arrange(disease) %>% 
  mutate(
    sample = str_remove(sample, pattern = "cluster"), 
    sample = fct_inorder(sample)
    
  ) %>% 
  ggplot(
    aes(x = sample, y = logCPM)
    ) +
  geom_boxplot(
    aes(fill = disease),
    outlier.size = 0.5
               ) +
  facet_wrap(~cluster, scales = "free") +
  easy_rotate_x_labels(angle = -45) +
  labs(title = "Distribution of logCPM values per sample",
       x = "Sample", y = "log2 CPM")
```


# PCA
## all samples
```{r}
pca = y$counts %>% 
  cpm(log = TRUE) %>% 
  t %>% 
  prcomp 


pca %>%   
  autoplot(
    data = tibble(
      sample = rownames(.$x)
    ) %>% 
      left_join(y$samples %>% 
                  mutate(sample = paste0(sample, "_cluster", cluster)
                )
                ),
    colour = "cluster", 
    shape = "disease",
    alpha = 0.5,
    size = 6
  ) +
  scale_color_viridis_d(option = "turbo")
```

## PCA by celltype
```{r}
plot.pca.bycelltype = function(dge.object, celltype.vector) {
  
  dge.object$counts %>% 
  as.data.frame() %>% 
  .[,grepl(colnames(y), pattern = celltype.vector)] %>% 
  cpm(log = T) %>% 
  t %>% 
  prcomp %>% 
  autoplot(
    data = tibble(
      sample = rownames(.$x)
    ) %>% 
      left_join(y$samples %>% 
                  mutate(sample = paste0(sample, "_cluster", cluster)
                )
                ),
    colour = "disease", 
    shape = "sex",
    alpha = 0.5,
    size = 6
  ) +
  scale_color_manual(values = c("#5F676D", "#17BEBB", "#FE5448", "#A20C01")) +
    ggtitle(celltype.vector)
}

plot.pca.bycelltype(dge.object = y, celltype.vector = "Astrocyte")
plot.pca.bycelltype(dge.object = y, celltype.vector = "Microglia")
plot.pca.bycelltype(dge.object = y, celltype.vector = "Neuron_Ex")
plot.pca.bycelltype(dge.object = y, celltype.vector = "Neuron_In")
plot.pca.bycelltype(dge.object = y, celltype.vector = "Oligodendrocyte")
plot.pca.bycelltype(dge.object = y, celltype.vector = "OPC")
```

# DE analysis

DE gene analysis was performed only on autosomal genes (to account for sex differences) using edgeR. 

Design and contrasts matrices were used to be able to perform pairwise contrasts between control and disease celltypes. 

```{r}
# create a design matrix
design <- model.matrix(~0 + disease_cluster, 
                       data = y$samples %>% 
                         droplevels()) 

colnames(design) <- str_remove(colnames(design), 
                               "geno_cluster")

# extract out autosome genes
autosomal.genes <- y$genes %>% 
  left_join(hs_chrs %>% 
              set_colnames(c("gene", "chr"))) %>% 
  dplyr::filter(!chr %in% c("X", "Y")) %>% 
  .$gene
y <- y[autosomal.genes,]

# estimae dispersion
y <- estimateDisp(y, robust = TRUE)
plotBCV(y)

fit <- glmQLFit(y, design, robust = TRUE)
plotQLDisp(fit)

# Extract cell types shared by all diseases
design_cols <- colnames(design)
celltypes <- gsub("^disease_cluster(Control|DLB|PD|PDD)_", "", design_cols)
shared_celltypes <- names(table(celltypes))[table(celltypes) == 4]

# Build contrast strings: Control_celltype - DLB_celltype
contrast_strings <- 
  c(
    paste0("disease_clusterDLB_", 
           shared_celltypes, 
           " - disease_clusterControl_", 
           shared_celltypes
           ), 
    paste0("disease_clusterPD_", 
         shared_celltypes, 
         " - disease_clusterControl_", 
         shared_celltypes
         ), 
    
     paste0("disease_clusterPDD_", 
         shared_celltypes, 
         " - disease_clusterControl_", 
         shared_celltypes
         )
  )
    

# Generate contrast matrix
contrast_matrix <- makeContrasts(contrasts = contrast_strings, levels = design)

# rename columns for readability
colnames(contrast_matrix) <- gsub("disease_cluster", "", colnames(contrast_matrix))

toptables <- colnames(contrast_matrix) %>% 
  sapply(function(x) {
      glmQLFTest(fit, design, contrast=contrast_matrix[,x]) %>% 
      topTags(n = Inf, sort.by = "PValue") %>% 
      .$table %>% 
      as_tibble %>% 
      dplyr::select(
        gene, logCPM, logFC, PValue, FDR, `F`
      ) %>% 
      mutate(
        coef = x, 
        DE = FDR < 0.05, 
        coef = str_remove(coef, pattern = " - Control.+$")
      )
  }, simplify = F)

```
## volcano plots
```{r}
toptables %>%
  bind_rows() -> volcano_data

# Count number of DE up/down per celltype (coef)
volcano_labels <- volcano_data %>%
  dplyr::filter(DE) %>%
  mutate(direction = ifelse(logFC > 0, "Up", "Down")) %>%
  group_by(coef, direction) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(
    x = ifelse(direction == "Up", 3, -3),  # approximate x positions
    y = max(-log10(volcano_data$PValue), na.rm = TRUE) * 0.9,  # high y
    label = paste0(direction, ": ", n)
  )

# Plot
volcano_data %>% 
  ggplot(
    aes(x = logFC, y = -log10(PValue))
    ) +
  geom_point(
    aes(colour = DE)
    ) +
  scale_color_manual(values = c("grey50", "red")) +
  facet_wrap(~coef, nrow = 3) +
  geom_text(
    data = volcano_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 4,
    hjust = 0.5
  ) +
  labs(
    x = "log2 Fold Change",
    y = "-log10(P value)"
  ) 
```

Some of the volcano plots havce some weird arm shapes. This usually comes from insufficient filtering. So i re-ran the analysis, while increasing the filtering. 


## MD plots
These plots have a bunch of low expressed genes - the MD plots dont have a nice triangle shape. I think i need to increase the filtering. probs going to lose some celltypes. 

```{r}
toptables %>% 
  bind_rows() %>% 
  ggplot(
    aes(x = logCPM, y = logFC)
  ) +
  geom_point(
    aes(colour = DE ),
    alpha = 0.5
  ) +
  facet_wrap(~coef) +
  scale_color_manual(values = c("grey50", "red"))
```

# enrichmenmt andlysis 
## CD pathways
```{r}
cluster.gprofiler.results <- 
  list.files(path = "data/results/adultchildsubtypes v2", 
             pattern = "gprofiler", recursive = TRUE, full.names = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/gprofiler_analysis_results.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/adultchildsubtypes v2", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))

# extract only PD and DLB networks. 
cluster.gprofiler.results <- cluster.gprofiler.results[
  names(cluster.gprofiler.results) %>% 
  grepl(pattern = "lyso_[PD|DLB]")
]


network.genes.cluster <- cluster.gprofiler.results %>% 
  lapply(function(x) {
    x %>% 
      dplyr::select(Cluster, genes) %>% 
      mutate(genes = str_split(genes, " ")) %>% 
      deframe()
  })

network.genes.cluster$CD_1A_lyso_DLB %<>% 
  lapply(function(x) 
  {
    x[x %in% rownames(y)] %>% 
      .[length(.) > 2] 
  }
  ) %>% 
  compact() # omit any empty slots

network.genes.cluster$CD_1A_lyso_PD %<>% 
  lapply(function(x) 
  {
    x[x %in% rownames(y)] %>% 
      .[length(.) > 2]
  }
  ) %>% 
  compact()

  
### obtain the genes in the subnetworks
node_metadata <- 
  list.files(path = "data/results/adultchildsubtypes v2", 
             pattern = "overlapping_nodes_metadata", recursive = TRUE, full.names = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/overlapping_nodes_metadata.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/adultchildsubtypes v2", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))
```

# how many network genes are DE?
```{r}
## how many network genes are detected
n.net.genes.detect.DLB = toptables$`DLB_Microglia - Control_Microglia`$gene %in% node_metadata$CD_1A_lyso_DLB$X %>% summary %>% .[3] 

n.net.genes.detect.PD = toptables$`PD_Microglia - Control_Microglia`$gene %in% node_metadata$CD_1A_lyso_PD$X %>% summary %>% .[3] 




toptables[grepl(names(toptables), pattern = "DLB")] %>% 
  bind_rows() %>% 
  dplyr::filter(
    DE == T, 
    gene %in% node_metadata$CD_1A_lyso_DLB$X
    ) %>% 
  dplyr::select(!starts_with("vf")) %>% 
  
  ggplot(
    aes(y = coef)
  ) +
  geom_bar() +
  labs(
    title = "Number Lyso_DLB subnetwork genes which are DE in DLB brain cells"
  ) +

toptables[grepl(names(toptables), pattern = "PD")] %>% 
  bind_rows() %>% 
  dplyr::filter(
    DE == T, 
    gene %in% node_metadata$CD_1A_lyso_PD$X
    ) %>% 
  dplyr::select(!starts_with("vf")) %>% 
  ggplot(
    aes(y = coef)
  ) +
  geom_bar(
  ) +
  labs(
    title = "Number Lyso_PD subnetwork genes which are DE in PD brain cells"
  )
```

Two particularly important genes in the network, NPC1 and GBA, have high absolute logFC and high centrality (z_both) in inhibitory neurons in DLB vs control brains.

```{r}
toptables$`DLB_Neuron_In - Control_Neuron_In` %>% 
  dplyr::filter(
    gene %in% node_metadata$CD_1A_lyso_DLB$X, 
    DE == TRUE
    ) %>% 
  left_join( node_metadata$CD_1A_lyso_DLB %>% 
               dplyr::rename(gene = X)) %>% 
  ggplot(
    aes(x = z_both, y = abs(logFC))
  ) +
  geom_point(
    aes(fill = -log10(PValue))
  ) +
  geom_label_repel(
    aes(label = gene), 
    # data = . %>% 
    #   dplyr::filter(z_both > 10)
  )  +
  scale_fill_viridis_c() +
  labs(
    title= "LysoDLB FC vs network proximity in NeuronIn"
  )
```

There are much less DE network genes in the PD and PDD nuclei transcriptomes No clear trends. 

```{r}
toptables$`PDD_Microglia - Control_Microglia` %>% 
  bind_rows(toptables$`PD_Microglia - Control_Microglia`) %>% 
  dplyr::filter(
    gene %in% node_metadata$CD_1A_lyso_PD$X, 
    DE == TRUE
    ) %>% 
  left_join( node_metadata$CD_1A_lyso_PD %>% 
               dplyr::rename(gene = X)) %>% 
  ggplot(
    aes(x = z_both, y = abs(logFC))
  ) +
  geom_point(
    aes(colour =-log10(PValue))
  ) +
  geom_label_repel(
    aes(label = gene)
  ) +
  facet_wrap(~coef) +
  scale_colour_viridis_c()
```


# Over representation analysis


#### number of DE genes per cluster
```{r}
#ora.DEG.in.clusters <-
  toptables %>%
  lapply(function(x) {
    genes <- x %>%
      dplyr::filter(DE == TRUE) %>%
      pull(gene)

    clusterProfiler::enricher(
      gene = genes,
      TERM2GENE = network.genes.cluster$CD_1A_lyso_DLB %>%
        enframe(name = "TERM", value = "GENE") %>%
        unnest(cols = "GENE"),
      universe = x$gene
    )
  })
  purrr::compact()

ora.DEG.in.clusters %>%
  lapply(as_tibble) %>%
  bind_rows(.id = "comparison")

keep = toptables %>%
  bind_rows() %>%
  dplyr::filter(
    gene %in% network.genes.cluster$C438
  ) %>%
  dplyr::filter(DE == T) %>%
  .$gene

  # Build the FDR matrix (for asterisk overlay)
  FDR_mat <- toptables %>%
    bind_rows() %>%
    dplyr::filter(gene %in% network.genes$X) %>%
    dplyr::select(coef, gene, FDR) %>%
    pivot_wider(names_from = gene, values_from = FDR) %>%
    column_to_rownames("coef") %>%
    as.matrix()

Heatmap(
  matrix = toptables %>%
    bind_rows() %>%
    dplyr::filter(gene %in% network.genes$X) %>%
    dplyr::select(coef, gene, logFC) %>%
    pivot_wider(names_from = gene, values_from = logFC) %>%
    column_to_rownames("coef") %>%
    as.matrix(),
  name = "logFC",
  col = brewer_pal(palette = "RdBu")(7),

  cell_fun =function(j, i, x, y, width, height, fill) {
      # j = column index, i = row index
      # x, y = coordinates of the center of the cell
      # width, height = size of the cell
      # fill = background color (not used here)

      # Check if the FDR value for this cell exists and is below 0.05
      if (!is.na(FDR_mat[i, j]) && FDR_mat[i, j] < 0.05) {
        # If yes, overlay a text label "*" on that cell
        grid.text("*", x, y, gp = gpar(fontsize = 15))
      }
    },
  column_names_gp = gpar(fontsize = 10),
  show_row_names = TRUE,
  show_column_names = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE
)




<!-- ```{r} -->
<!-- #plot number of DE genes per cluster -->
<!-- network.genes.cluster %>%  -->
<!--   lapply(function(x) { -->
<!--     tibble( -->
<!--       gene = x -->
<!--     ) %>%  -->
<!--       left_join(toptables %>%  -->
<!--                   bind_rows() %>%  -->
<!--                   dplyr::select(gene, coef, FDR) %>%  -->
<!--                   mutate(DE = FDR < 0.05),  -->
<!--                 by = "gene" -->
<!--       ) -->
<!--   }) %>%  -->
<!--   bind_rows(.id = "cluster") %>%  -->
<!--   dplyr::filter(DE == T) %>%  -->
<!--   mutate( -->
<!--     celltype = str_extract( -->
<!--       coef, -->
<!--       pattern = "(?<=DLB_).+(?= - Control_)")) %>%  -->
<!--  count(cluster, celltype, name = "num_DE_genes") %>% -->
<!--   ggplot(aes(x = cluster, y = num_DE_genes, fill = celltype)) + -->
<!--   geom_col(position = "dodge") + -->
<!--   labs( -->
<!--     title = "Number of DE genes (FDR < 0.05)", -->
<!--     x = "Network cluster", -->
<!--     y = "# DE genes" -->
<!--   ) + -->
<!--   scale_y_continuous(breaks = 0:12) + -->
<!--   theme(axis.text.x = element_text(angle = 45, hjust = 1)) -->
<!-- ``` -->



<!-- ```{r} -->


<!-- } -->
<!-- ``` -->

<!-- ```{r} -->
<!-- plot.cnet( -->
<!--   toptable = toptables$`DLB_astrocyte - Control_astrocyte`, -->
<!--   ORA.cluster.object = ORA.clusters$C453,  -->
<!--   plot.title = "Astrocytes") -->

<!-- plot.cnet( -->
<!--   toptable = toptables$`DLB_microglia_immune - Control_microglia_immune`, -->
<!--   ORA.cluster.object = ORA.clusters$C439,  -->
<!--   plot.title = "Microglia") -->

<!-- plot.cnet( -->
<!--   toptable = toptables$`DLB_microglia_immune - Control_microglia_immune`, -->
<!--   ORA.cluster.object = ORA.clusters$C448,  -->
<!--   plot.title = "Microglia") -->

<!-- plot.cnet( -->
<!--   toptable = toptables$`DLB_microglia_immune - Control_microglia_immune`, -->
<!--   ORA.cluster.object = ORA.clusters$C438,  -->
<!--   plot.title = "Microglia") -->
<!-- ``` -->





<!-- ```{r} -->
<!-- plot.cnet( -->
<!--   toptable = toptables$`DLB_neuron_inhib - Control_neuron_inhib`, -->
<!--   ORA.cluster.object = ORA.clusters$C441,  -->
<!--   plot.title = "inhib neurons") -->

<!-- plot.cnet( -->
<!--   toptable = toptables$`DLB_neuron_inhib - Control_neuron_inhib`, -->
<!--   ORA.cluster.object = ORA.clusters$C440,  -->
<!--   plot.title = "inhib neurons") -->
<!-- ``` -->

<!-- ## ORA of KEGG pathways -->
<!-- ```{r} -->
<!-- ora.DEG.KEGG <- toptables %>% -->
<!--   lapply(function(x) { -->
<!--     genes <- x %>% -->
<!--       dplyr::filter(DE == TRUE) %>% -->
<!--       pull(gene) -->

<!--     if (length(genes) < 5) return(NULL)  # skip if too few genes -->

<!--     clusterProfiler::enrichKEGG( -->
<!--       gene = bitr(genes,  -->
<!--                   OrgDb = org.Hs.eg.db, -->
<!--                   fromType="SYMBOL", toType="ENTREZID") %>% .$ENTREZID, -->
<!--       universe = bitr(x$gene,  -->
<!--                   OrgDb = org.Hs.eg.db, -->
<!--                   fromType="SYMBOL", toType="ENTREZID") %>% .$ENTREZID, -->
<!--       keyType = "ncbi-geneid",  -->
<!--       organism = "hsa")  -->
<!--   }) %>% -->
<!--   purrr::compact()  -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ora.DEG.KEGG %>% -->
<!--   lapply(as_tibble) %>%  -->
<!--   bind_rows(.id = "coef") %>%  -->
<!--   dplyr::filter(Count > 1) %>%  -->
<!--   ggplot( -->
<!--     aes(x = -log10(pvalue), y = Description) -->
<!--   ) +  -->
<!--   geom_col() + -->
<!--   facet_wrap(~coef, scales = "free_y") -->
<!-- ``` -->
```


```{r}
# fry gene set tests
#### Obtain KEGG gene sets
### obtain the pathway to entrez geneid link
path2entrezgene  <-
  keggLink("pathway", "hsa") %>%
  tibble(
    pathway = .,
    entrez_gene_id = sub("hsa:", "", names(.))
    )

# obtain the pathway names
pathways <-
  keggList("pathway", "hsa") %>%
  tibble(pathway = names(.), description = .)

# Use org.eg.db to obtain the conversion of entrez to gene symbol
hsa_kegg_anno <-
  path2entrezgene %>%
  mutate(
    symbol = mapIds(org.Hs.eg.db , entrez_gene_id,
                    "SYMBOL", "ENTREZID")
  ) %>%
  mutate(pathway = str_remove(pathway, pattern = "path:")) %>%
  left_join(pathways) %>%
  dplyr::select(-entrez_gene_id) %>%
  dplyr::filter(
    symbol %in% (y$genes$gene)) %>%
  mutate(
    description = str_remove_all(description, pattern = " \\- Homo .+"))


# make the KEGG object for gene set testing
KEGG <- hsa_kegg_anno %>%
  na.omit %>%
  distinct(pathway, symbol, .keep_all = TRUE) %>%
  split(f = .$description) %>%
  lapply(function(x) {x$symbol %>% unname()})

# only retain pathways with more than 3 genes
KEGG <- KEGG[lapply(KEGG, length) > 3 ]

```

#### Run FRY on network clusters

I next want to assess the communities identified and whether they show significant changes to expression as a group.

```{r}
run.fry = function(disease, fry.index, cluster.gprofiler.results.object)
{
colnames(contrast_matrix) %>% 
  grep(pattern = disease, value = TRUE) %>% 
  sapply(function(x) {
   cpm(y, log = T)  %>% 
      fry(
        index = fry.index,
        design = design, 
        contrast = contrast_matrix[,x], 
        sort = "mixed"
      ) %>% 
      rownames_to_column("pathway") %>% 
      as_tibble() %>% 
      mutate(coef = x) %>% 
      left_join(
        cluster.gprofiler.results.object %>% 
          dplyr::select(pathway = Cluster, top.pathway.gprofiler = Top.pathway)
      )
  }, simplify = FALSE)
}
```

```{r}
fry.results = list(
  DLB = run.fry(
    disease = "DLB", 
    fry.index = network.genes.cluster$CD_1A_lyso_DLB,
    cluster.gprofiler.results.object = cluster.gprofiler.results$CD_1A_lyso_DLB), 
  
  PD = run.fry(
    disease = "PD", 
    fry.index = network.genes.cluster$CD_1A_lyso_PD,
    cluster.gprofiler.results.object = cluster.gprofiler.results$CD_1A_lyso_PD)
)
```

### Which clusters are sig?
```{r}
sig.clusters.DLB <- fry.results$DLB %>% 
  bind_rows() %>% 
  dplyr::filter(FDR.Mixed < 0.05) %>% 
  .$pathway %>% 
  unique

fry.results$DLB %>% 
  bind_rows() %>% 
  dplyr::filter(pathway %in% sig.clusters.DLB) %>% 
  pivot_longer(
    cols = c(PValue, PValue.Mixed, FDR, FDR.Mixed),
    names_to = c(".value", "Hypothesis"),
    names_sep = "\\." # Splits 'PValue.Mixed' into 'PValue' and 'Mixed'
  ) %>%  
  mutate(
    Hypothesis = replace_na(Hypothesis, replace = "Directional"), 
    coef = str_remove(coef, pattern = " - Contro.+")
  ) %>% 
  ggplot(
    aes(x = Hypothesis, y = pathway)
  ) +
  geom_tile(
    aes(fill = -log10(PValue), 
        alpha = FDR < 0.05), 
    ) +
  geom_label(
    aes(label = signif(FDR, 2)), 
    data = . %>% 
      dplyr::filter(FDR < 0.05)
  ) +
  facet_wrap(~coef, nrow = 1) +
  scale_fill_viridis_c()
```

```{r}
fry.results$DLB %>% 
  bind_rows() %>% 
  dplyr::filter(pathway %in% sig.clusters.DLB) %>% 

  mutate(
    coef = str_remove(coef, pattern = " - Contro.+")
  ) %>% 
  ggplot(
    aes(x = -log10(PValue.Mixed), y = pathway)
  ) +
  geom_col(
    aes(fill = Direction,
        alpha = FDR.Mixed < 0.05), 
    ) +
  facet_wrap(~coef, nrow = 1) +
  scale_fill_viridis_d()
```

```{r}
fry.results$DLB %>% 
  bind_rows() %>% 
  mutate(
    coef = str_remove(coef, pattern = " - Contro.+")
  ) %>% 
  ggplot(
    aes(
      x = coef,
      y = -log10(PValue.Mixed))
  ) +
  geom_point(
    aes(
      size = NGenes, 
      colour = Direction)
  ) +
  geom_hline(yintercept = -log10(0.05)) +
  geom_label_repel(
    aes(label = pathway), 
    data = . %>% 
      dplyr::filter(FDR.Mixed < 0.05))
```

```{r}
Idents(data.seurat) = data.seurat@meta.data$celltype
```


```{r}
data.seurat %>% 
  VlnPlot(features = network.genes.cluster$CD_1A_lyso_DLB$C441)

```


## DLB
### Cluster 436
#### expression of C436
```{r}
y[
  # subset genes
  network.genes.cluster$CD_1A_lyso_DLB$C436,
  # subset samples
  (y$samples %>% 
     dplyr::filter(
       disease %in% c("Control", "DLB")
       ) %>% 
     rownames)
  ] %>% 
  cpm(log=T) %>% 
  Heatmap(
    name = "logCPM",
    top_annotation = HeatmapAnnotation(
      disease = y$samples[colnames(.), "disease"] %>% droplevels(),
      col = list(
        disease = c("Control" = "grey50", "DLB" = "red")
      )
    ),
    col = viridis_pal()(10),
    column_split = y$samples[colnames(.), "cluster"] %>% droplevels(),
    show_column_names = F,
    show_row_names = T,
    row_names_gp = gpar(fontsize = 5),
    cluster_rows = TRUE,
    cluster_columns = TRUE,  
  )  %>% 
  draw(column_title = "Lyso_DLB C436", 
       column_title_gp = gpar(fontsize = 18, fontface = "bold"))
  
```
#### lpofFC of genes in C436
```{r}

toptables[names(toptables) %>%
  grep(
    pattern = "DLB", 
    value = T
  )] %>% 
  bind_rows() %>% 
  dplyr::filter(
    gene %in%  network.genes.cluster$CD_1A_lyso_DLB$C436
  ) %>% 
  dplyr::select(gene, coef, logFC) %>% 
  mutate(coef = str_remove(coef, "DLB_")) %>% 
  pivot_wider(
    names_from = coef,
    values_from = logFC
  ) %>% 
  column_to_rownames("gene") %>% 
  as.matrix() %>% 
  Heatmap(
    name = "logFC",
    col = colorRamp2(c(-10, 0, 3), c("blue4", "white", "red3")),
    show_row_dend = F, 
    show_column_dend =  F, 
    show_row_names = F,
    show_column_names = TRUE,
    cluster_rows = TRUE,
    cluster_columns = TRUE, 
    width = unit(4, "cm"), 
    border_gp = gpar("black")
    
  ) %>% 
   draw(column_title = "Lyso_DLB C436", 
       column_title_gp = gpar(fontsize = 18, fontface = "bold"))
```

There are some genes with a low logFC in inhib nurons. 

I wnt to have a closer look at these
```{r}
toptables[names(toptables) %>%
  grep(
    pattern = "DLB", 
    value = T
  )] %>% 
  bind_rows() %>% 
  dplyr::filter(
    gene %in%  network.genes.cluster$CD_1A_lyso_DLB$C436
  ) %>% 
  dplyr::select(gene, coef, logFC) %>% 
  mutate(coef = str_remove(coef, "DLB_")) %>% 
  pivot_wider(
    names_from = coef,
    values_from = logFC
  ) %>% 
  column_to_rownames("gene") %>% 
  arrange(Neuron_In) %>% 
  as.matrix() %>% 
  Heatmap(
    name = "logFC",
    col = colorRamp2(c(-10, 0, 5), c("blue4", "white", "red3")),
    show_row_names = TRUE,
    show_column_names = TRUE,
    cluster_rows = TRUE,
    cluster_columns = TRUE, 
    width = unit(4, "cm")
    
  )   %>% 
  .[1:46,]
   draw(column_title = "Lyso_DLB C436", 
       column_title_gp = gpar(fontsize = 18, fontface = "bold"))
```

Many of these are lysosomal enzymes.. 
GBA, NAGLU, IDUA, MAN2B1. AGA , ACP2, HYAL1

#### PD

```{r}
sig.clusters.PD <- fry.results$PD %>% 
  bind_rows() %>% 
  dplyr::filter(FDR.Mixed < 0.05) %>% 
  .$pathway %>% 
  unique

fry.results$PD %>% 
  bind_rows() %>% 
  dplyr::filter(pathway %in% sig.clusters.PD) %>% 
  mutate(
    coef = str_remove(coef, pattern = " - Contro.+")
  ) %>% 
  ggplot(
    aes(x = -log10(PValue.Mixed), 
        y = pathway)
  ) +
  geom_col(
    aes(fill = -log10(PValue.Mixed), 
        alpha = FDR.Mixed < 0.05)
  ) +
  facet_wrap(~coef, nrow = 2) +
  scale_fill_viridis_c()
```
#### expression of 289
```{r}
y[
  # subset genes
  network.genes.cluster$CD_1A_lyso_PD$C289,
  # subset samples
  (y$samples %>% 
     dplyr::filter(
       disease %in% c("Control", "PD", "PDD")
       ) %>% 
     rownames)
  ] %>% 
  cpm(log=T) %>% 
  Heatmap(
    name = "logCPM",
    top_annotation = HeatmapAnnotation(
      disease = y$samples[colnames(.), "disease"] %>% droplevels(),
      col = list(
        disease = c("Control" = "grey50", "PD" = "red", PDD = "red4")
      )
    ),
    col = viridis_pal()(10),
    column_split = y$samples[colnames(.), "cluster"] %>% droplevels(),
    show_column_names = F,
    show_row_names = T,
    row_names_gp = gpar(fontsize = 5),
    cluster_rows = TRUE,
    cluster_columns = TRUE,  
  )  %>% 
  draw(column_title = "Lyso_PD C289", 
       column_title_gp = gpar(fontsize = 18, fontface = "bold"))
  
```
#### logFC of genes in C290
```{r}
toptables[names(toptables) %>%
  grep(
    pattern = "PD", 
    value = T
  )] %>% 
  bind_rows() %>% 
  dplyr::filter(
    gene %in%  network.genes.cluster$CD_1A_lyso_PD$C290
  ) %>% 
  dplyr::select(gene, coef, logFC) %>% 
  mutate(coef = str_remove(coef, "DLB_")) %>% 
  pivot_wider(
    names_from = coef,
    values_from = logFC
  ) %>% 
  column_to_rownames("gene") %>% 
  as.matrix() %>%
  Heatmap(
    name = "logFC",
    
    col = colorRamp2(c(-2, 0, 2), c("blue4", "white", "red3")),
    
    top_annotation = HeatmapAnnotation(
      disease =   colnames(.) %>% 
        as.data.frame() %>% 
        set_colnames("coef") %>% 
        mutate(disease = str_extract(coef, pattern = "PDD|PD")) %>% 
        column_to_rownames("coef") %>% 
        .$disease, 
       col = list(
        disease = c("PD" = "red", PDD = "red4")
      ), 
      show_legend = F
      ),
    
    show_row_names = T,
    show_column_names = TRUE,
    show_column_dend = F, 
    cluster_rows = TRUE,
    cluster_columns = TRUE, 
    
    row_names_gp= gpar(fontsize = 8),
    width = unit(4, "cm")
    
  ) %>% 
   draw(column_title = "Lyso_PD C251", 
       column_title_gp = gpar(fontsize = 18, fontface = "bold"))
```

## shared genes across the 2 similar clusters

PD_C251 is the most simialr to DLB_C436. 

```{r}
toptables[names(toptables) %>%
  grep(
    pattern = "PD|DLB", 
    value = T
  )] %>% 
  bind_rows() %>% 
  dplyr::filter(
    gene %in%  intersect(
      network.genes.cluster$CD_1A_lyso_PD$C251,
      network.genes.cluster$CD_1A_lyso_DLB$C436
      )
  ) %>% 
  dplyr::select(gene, coef, logFC) %>% 
  # mutate(
  #   coef = str_remove(coef, "DLB_"), 
  #   coef = str_remove(coef, "PD._")
  # ) 
  pivot_wider(
    names_from = coef,
    values_from = logFC
  ) %>% 
  column_to_rownames("gene") %>% 
  as.matrix() %>%
  Heatmap(
    name = "logFC",
    
    col = colorRamp2(c(-2, 0, 2), c("blue4", "white", "red3")),
    
    top_annotation = HeatmapAnnotation(
      disease =   colnames(.) %>% 
        as.data.frame() %>% 
        set_colnames("coef") %>% 
        mutate(disease = str_extract(coef, pattern = "PDD|PD|DLB")) %>% 
        column_to_rownames("coef") %>% 
        .$disease, 
       col = list(
        disease = c("PD" = "red", PDD = "red4", DLB = "pink")
      ), 
      show_legend = T
      ),
    
    column_split = colnames(.) %>% 
        as.data.frame() %>% 
        set_colnames("coef") %>% 
        mutate(celltype = str_remove(coef, pattern = "^.+_")) %>% 
        column_to_rownames("coef") %>% 
        .$celltype,
    
    show_row_names = T,
    show_column_names = TRUE,
    show_column_dend = F, 
    cluster_rows = TRUE,
    cluster_columns = F, 
    
    row_names_gp= gpar(fontsize = 8),
    width = unit(20, "cm")
    
  ) %>% 
  draw(column_title = "shared C436 and C251 genes", 
       column_title_gp = gpar(fontsize = 18, fontface = "bold"))
```


# data export
```{r}
y %>% 
  qsave("data/public data/GSE178146_RAW/R_objects/pseudobulk_dgeObject.qs")


fry.results %>% 
  qsave("data/public data/GSE178146_RAW/R_objects/fry_results.qs")

toptables %>% 
  qsave("data/public data/GSE178146_RAW/R_objects/toptables.qs")
```


# figs
## seurat 
```{r}
# SET THE IDEMTS TO CELLTYPE
Idents(data.seurat) = data.seurat@meta.data$celltype
# extract the network genems
network.genes <- node_metadata[
  node_metadata %>% names %>% grepl(pattern = "CD_1A")
] %>% 
  lapply(function(x) {
    tibble(
      gene_name = x$X
    )
  }) %>% 
  set_names(str_remove(names(.), pattern = "CD_1A_lyso_"))

# cell type marker genes
genes2plot = c(
  # inhib neurons
  "GAD1","GAD2", 
  # exc neurons
  "CAMK2A","NRGN", "SLC17A7", 
  # astrocytes
  "GFAP","AQP4", 
  # microglia
  "C3","CD74", "CSF1R", "AIF1",
  # OPCs
  "OLIG1","OLIG2",
  # oligodend
  "MBP","MOBP", "PLP1", 
  # VLMCs
  "COL1A2","DCN", "IGFBP7",
  # endo
  "CLDN5","FLT1"
  )

sig.clusters.DLB <- fry.results$DLB %>% 
  bind_rows() %>% 
  dplyr::filter(FDR.Mixed < 0.05) %>% 
  .$pathway %>% 
  unique


# assemble th eplot  
fig4a <- Seurat::DimPlot(data.seurat,
        reduction = "umap",
        label = TRUE,
        repel = TRUE,
        group.by = "celltype",
        split.by = "disease",
        shuffle = TRUE
        ) +
  theme(
    legend.position = "none", 
    aspect.ratio = 1
  ) 


fig4b <-  
  Seurat::DotPlot(
    data.seurat,
    features = genes2plot,
    cols = c("#440154FF", "#22A884FF"), 
    scale = TRUE, 
    dot.scale = 10
  ) +
  labs(
    y = "Cell type", 
    x = "Genes"
  ) +
  scale_fill_viridis_c()


fig.4c <- fry.results$DLB %>% 
  bind_rows() %>% 
  dplyr::filter(pathway %in% sig.clusters.DLB) %>% 
  mutate(
    coef = str_remove(coef, pattern = " - Con.+$"),
    coef = str_remove(coef, pattern = "DLB_")
  ) %>% 
  ggplot(
    aes( 
      x = coef, 
      y = pathway)
  ) +
  geom_tile(
    aes(fill = -log10(PValue.Mixed)), 
    width = 1
  ) + 
  geom_label(
    aes(label = signif(FDR.Mixed, 2))
  ) +
  scale_fill_viridis_c() +
  scale_alpha_manual(
    values = c(0.2, 1)
  ) +
  labs(
    y = "LysoDLB community",
    x = "Cell type"
  ) 

fig4a + fig4b + fig.4c +
  theme(
    text = element_text(size = 14)
  )
  
  ggsave("output/figs/paper/fig4abc.png",
         width = 19, height = 11, units = "cm", 
         scale = 2.5)
  
  
```

## logFC of genes in clusters
```{r}
gene_annot <- tibble(
  gene =  colnames(t(mat)),
  C437 = gene %in% network.genes.cluster$CD_1A_lyso_DLB$C437,
  C438 = gene %in% network.genes.cluster$CD_1A_lyso_DLB$C438
)

mat = 
  toptables[names(toptables) %>%
  grep(
    pattern = "DLB", 
    value = T
  )] %>% 
  bind_rows() %>% 
  dplyr::filter(
    gene %in%  network.genes.cluster$CD_1A_lyso_DLB$C436
  ) %>% 
  dplyr::select(gene, coef, logFC) %>% 
  mutate(coef = str_remove(coef, "DLB_")) %>% 
  pivot_wider(
    names_from = coef,
    values_from = logFC
  ) %>% 
  column_to_rownames("gene") %>% 
  as.matrix() %>% 
  t 

mat %>% 
  t %>%
  Heatmap(
    name = "logFC",
    col = colorRamp2(c(-4, 0, 4), c("blue4", "white", "red3")),
    show_row_dend = F, 
    show_column_dend =  F, 
    show_row_names = F,
    show_column_names = T,
    column_names_gp = gpar(fontsize = 12),
    cluster_rows = TRUE,
    cluster_columns = TRUE, 
    width = unit(4, "cm"), 
    border_gp = gpar("black"), 
    
    left_annotation =  rowAnnotation(
      # C437 = colnames(mat) %in%
      #   network.genes.cluster$CD_1A_lyso_DLB$C437 ,
      # C438 = colnames(mat) %in%
      #   network.genes.cluster$CD_1A_lyso_DLB$C438,
      lysosome = colnames(mat) %in%
        KEGG$Lysosome,
      sphingolipid = colnames(mat) %in%
        KEGG$`Sphingolipid metabolism`,
      col = list(
        lysosome = c("TRUE" = "red", "FALSE" = "grey70"),
        sphingolipid = c("TRUE" = "red", "FALSE" = "grey70")
      ),
      annotation_name_gp = gpar(fontsize = 10)
    )
  ) 
  .[307:250,]

C436_clusterprofiler <- network.genes.cluster$CD_1A_lyso_DLB$C436 %>% 
  bitr(toType = "ENTREZID", fromType = "SYMBOL", OrgDb = org.Hs.eg.db) %>% 
  .$ENTREZID %>% 
  enrichKEGG(
    organism = "hsa", 
    universe = rownames(y) %>% 
       bitr(toType = "ENTREZID", fromType = "SYMBOL", OrgDb = org.Hs.eg.db)
    )%>% 
  setReadable(OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

C436_clusterprofiler %>% 
  as_tibble() %>% view
  .$Description


cnetplot(
  C436_clusterprofiler, 
  showCategory = 28, 
  foldChange = 
    toptables$`DLB_Microglia - Control_Microglia`%>% 
    dplyr::select(gene, logFC) %>% 
    dplyr::distinct(gene,.keep_all = T) %>%
    { setNames(.$logFC, .$gene) }, 
  color_item = "white"
) +
  scale_color_gradient2(
    name='associated data', low='darkblue', high='firebrick', na.value = "white")



hm.437 <- toptables[names(toptables) %>%
  grep(
    pattern = "DLB", 
    value = T
  )] %>% 
  bind_rows() %>% 
  dplyr::filter(
    gene %in%  network.genes.cluster$CD_1A_lyso_DLB$C437
  ) %>% 
  dplyr::select(gene, coef, logFC) %>% 
  mutate(coef = str_remove(coef, "DLB_")) %>% 
  pivot_wider(
    names_from = coef,
    values_from = logFC
  ) %>% 
  column_to_rownames("gene") %>% 
  as.matrix() %>% 
  t %>% 
  Heatmap(
    name = "logFC",
    col = colorRamp2(c(-5, 0, 5), c("blue4", "white", "red3")),
    show_row_dend = F, 
    show_column_dend =  F, 
    show_row_names = T,
    show_column_names = F,
    cluster_rows = TRUE,
    cluster_columns = TRUE, 
    height = unit(4, "cm"), 
    border_gp = gpar("black")
    
  )

hm.438 <- toptables[names(toptables) %>%
  grep(
    pattern = "DLB", 
    value = T
  )] %>% 
  bind_rows() %>% 
  dplyr::filter(
    gene %in%  network.genes.cluster$CD_1A_lyso_DLB$C438
  ) %>% 
  dplyr::select(gene, coef, logFC) %>% 
  mutate(coef = str_remove(coef, "DLB_")) %>% 
  pivot_wider(
    names_from = coef,
    values_from = logFC
  ) %>% 
  column_to_rownames("gene") %>% 
  as.matrix() %>% 
  t
  Heatmap(
    name = "logFC",
    col = colorRamp2(c(-5, 0, 5), c("blue4", "white", "red3")),
    show_row_dend = F, 
    show_column_dend =  F, 
    show_row_names = F,
    show_column_names = TRUE,
    cluster_rows = TRUE,
    cluster_columns = TRUE, 
    width = unit(4, "cm"), 
    border_gp = gpar("black")
    
  )

hm.436 %v% hm.437 %v% hm.438

draw(column_title = "Lyso_DLB C437", 
       column_title_gp = gpar(fontsize = 18, fontface = "bold"))
```

```{r}

```



```{r}
# violin plot to show the expression of network genes across celltypes
plot.network.genes.across.celltypes = function(
  network.genes, 
  disease.object,
  ggtitle
  ){
y %>% 
  cpm(log = F) %>% 
  as_tibble(rownames = "gene_name") %>% 
  pivot_longer(
    names_to = "sample_id", 
    values_to = "CPM", 
    colnames(y)
  ) %>% 
  left_join(y$samples) %>% 
  dplyr::filter(
    gene_name %in% network.genes, 
    disease %in% disease.object
  ) %>% 
  ggplot(
    aes(y = CPM, x = disease, colour = disease)
  ) +
  geom_violin() +
  geom_boxplot(
    width = 0.2, outliers = F,
    position = position_dodge(width = 1)
  ) +
  facet_wrap(~cluster, scales = "free_x", nrow = 1) +
    scale_y_continuous(
      labels = comma
    ) +
  labs(
    title = ggtitle
  ) +
    theme(legend.position = "none")
}

ggarrange(
plot.network.genes.across.celltypes(
  network.genes = network.genes$DLB$gene_name,
  disease.object = c("Control", "DLB"), 
  ggtitle = "Expression of lysosomal childhood dementias & DLB subnetwork genes"
) +
  scale_color_manual(
    values = viridis_pal(option = "magma", end = 0.8)(2)
  ),
plot.network.genes.across.celltypes(
  network.genes = network.genes$PD$gene_name,
  disease.object = c("Control", "PD", "PDD"), 
  ggtitle = "Expression of lysosomal childhood dementias & PD subnetwork genes"
  ) +
  scale_color_manual(
    values = viridis_pal(begin = 0.1, end = 0.5)(3)
    ) , 

ncol = 1, align = "h"
) 

## now plot the proportion of the network genes expressed
y$samples %>% 
  dplyr::filter(
    disease %in% c("Control", "DLB")
  ) %>% 
  left_join(
    y.before.removing.clusters %>% 
      cpm(log = F) %>% 
      as_tibble(rownames = "gene_name") %>% 
      pivot_longer(
        names_to = "sample_id", 
        values_to = "CPM", 
        colnames(y)
      ) %>% 
      dplyr::filter(
        gene_name %in% network.genes$DLB$gene_name
      ) %>% 
    mutate(
      expressed = CPM > 0
    ) %>% 
      group_by(sample_id, expressed) %>% 
      summarise(
        num_genes = n(),
        .groups = "drop"
      ) %>% 
      pivot_wider(
        names_from = expressed,
        values_from = num_genes,
        names_prefix = "expressed_"
      ) %>% 
      mutate(
        expressed_Fraction = expressed_TRUE / (expressed_TRUE + expressed_FALSE)
      ),
    by = "sample_id"
  ) %>% 
  group_by(cluster, disease) %>% 
  summarise(
    mean_expressed_Fraction = mean(expressed_Fraction, na.rm = T),
    .groups = "drop"
      ) %>% 
  ggplot(
    aes(x = cluster, y = mean_expressed_Fraction, fill = disease)
  ) +
  geom_col(position = "dodge") +
  labs(
    title = "Proportion of network genes expressed (CPM > 1)",
    y = "Proportion of genes expressed",
    x = "Cell type"
  ) +
  scale_y_continuous(
    limits = c(0,1),
    labels = scales::percent_format(accuracy = 1)) +
  scale_fill_viridis_d() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```


```{r}
ggplot(volcano_data, aes(x = logFC, y = -log10(PValue))) +
  geom_point(aes(colour = DE)) +
  scale_color_manual(values = c("grey50", "red")) +
  facet_wrap(~coef) +
  geom_text(
    data = volcano_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 4,
    hjust = 0.5
  ) +
  labs(
    x = "log2 Fold Change",
    y = "-log10(P value)"
  ) +
  ggsave(
    "output/DLB_snRNAseq/figs/talk/celltype_volcanoplots.png", 
    width = 12, height = 8, units = "cm",
    scale = 2.5)
```
```{r}
network.genes.cluster %>% 
  lapply(function(x) {
    tibble(
      gene = x
    ) %>% 
      left_join(toptables %>% 
                  bind_rows() %>% 
                  dplyr::select(gene, coef, FDR) %>% 
                  mutate(DE = FDR < 0.05), 
                by = "gene"
      )
  }) %>% 
  bind_rows(.id = "cluster") %>% 
  dplyr::filter(DE == T) %>% 
  mutate(
    celltype = str_extract(
      coef,
      pattern = "(?<=DLB_).+(?= - Control_)")) %>% 
 count(cluster, celltype, name = "num_DE_genes") %>%
  ggplot(aes(x = cluster, y = num_DE_genes, fill = celltype)) +
  geom_col(position = "dodge", colour = "black") +
  labs(
    title = "Number of DE genes (FDR < 0.05)",
    x = "Network cluster",
    y = "# DE genes"
  ) +
  scale_y_continuous(breaks = 0:9) +
  scale_fill_viridis_d() +
  coord_flip() 
  ggsave(
    "output/DLB_snRNAseq/figs/talk/numDEGsperclusterpercelltype.png", 
    width = 12, height = 8, units = "cm",
    scale = 2.5)


a <- network.genes.cluster %>% 
  lapply(function(x) {
    tibble(
      gene = x
    ) %>% 
      left_join(toptables %>% 
                  bind_rows() %>% 
                  dplyr::select(gene, coef, FDR) %>% 
                  mutate(DE = FDR < 0.05), 
                by = "gene"
      )
  }) %>% 
  bind_rows(.id = "cluster") %>% 
  dplyr::filter(DE == T) %>% 
  mutate(
    celltype = str_extract(
      coef,
      pattern = "(?<=DLB_).+(?= - Control_)")) %>% 
 count(cluster, celltype, name = "num_DE_genes") %>%
  ggplot(aes(x = cluster, y = num_DE_genes, fill = celltype)) +
  geom_col(position = "dodge", colour = "black") +
  labs(
    title = "Number of DE genes (FDR < 0.05)",
    x = "Network cluster",
    y = "# DE genes"
  ) +
  scale_y_continuous(breaks = 0:9) +
  scale_fill_viridis_d() +
  coord_flip() +
   theme(legend.position = "bottom")

b <- fry %>% 
  bind_rows() %>% 
   mutate(
    coef = str_extract(coef, 
      pattern = "(?<=DLB_).+(?= - Control_)"
    )) %>% 
  dplyr::filter(FDR.Mixed < 0.05) %>% 
  ggplot(
    aes(y = pathway, x = -log10(PValue.Mixed), fill = Direction)
  ) +
  geom_col() +
  facet_wrap(~coef, scales = "free_y") +
  labs(
    y = "Cluster from network", 
    title = "Ranked-list gene set testing (limma::fry)"
  ) +
  theme(legend.position = "bottom")

a + b +
  plot_layout(
    widths = c(1,2),
    heights = c(1,2)
    ) +
  
ggsave(
    "output/DLB_snRNAseq/figs/talk/numDEGsperclusterpercelltype.png", 
    width = 12, height = 8, units = "cm",
    scale = 2.5)
```

## Pld3
```{r}
toptables %>% 
  bind_rows() %>% 
  dplyr::filter(gene == "GABARAP") %>% 
  arrange(FDR)
```









