---
title: "03_hsDLB_snRNAseq_pseudobulk"
author: "Karissa Barthelson"
date: "2025-07-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# setup
```{r}
knitr::opts_chunk$set(
  message = F, echo = F, warning = F
)
```

```{r}
library(tidyverse)
library(magrittr)
library(readxl)
library(qs) # for fast save and reading of objects into memory (alternative to rds)
library(parallel) 
library(Matrix)

# annotation
library(org.Hs.eg.db)
library(KEGGREST)

# analysis
library(Seurat)
library(ngsReports)
library(edgeR)
library(clusterProfiler)

# vis tools
library(ggpubr)
library(scales)
library(patchwork)
library(cowplot)
library(viridisLite)
library(ggeasy)
library(ggfortify)
library(ggrepel)
library(tidyHeatmap)
library(ComplexHeatmap)
library(grid)

theme_set(theme_light() +
            theme(plot.title = element_text(hjust = 0.5), 
                  strip.text = element_text(face = "bold"),
                  strip.background = element_rect(fill = "grey20")
            )
)

```

# Introduction
Here, I will perform the pseudobulk comparisons within cell type between DLB and contorl brains. 

The preprocessing of the data was performed previously and the seurat object was loaded. It was then converted to a PB object for use in edgeR.
```{r}
# read in smaple metadata
meta <- 
  data.table::fread("data/public data/GSE178146_RAW/GSE178146_series_matrix.txt", 
                          skip = 27, sep = "\t", fill = T) %>% 
  mutate(
    `!Sample_title` = str_remove(`!Sample_title`, pattern = "\\!")
  ) %>% 
  t %>% 
  .[,c(1,7,9)] %>% 
  set_colnames(
    c("geo_acc", "brain_region", "disease")
  ) %>% 
  .[-1,] %>% 
  as.data.frame() %>% 
  mutate(
    sample = rownames(.), 
    disease = str_remove(disease, pattern ="disease state: ")
  ) %>% 
  as_tibble()

# seurat object
data.seurat <- qread("data/public data/GSE178146_RAW/R_objects/PD_DLB_data_seurat.qs")

# need to merge layers of the seurat object before converting to PB object
# Seurat::DimPlot(data.seurat, 
#         reduction = "umap", group.by = c("orig.ident", "celltype"), shuffle = TRUE)
# 
# Seurat::DimPlot(data.seurat, 
#         reduction = "umap", 
#         group.by = "celltype", 
#         split.by = "disease",
#         shuffle = TRUE)
        
data.seurat[["RNA"]] <- JoinLayers(data.seurat[["RNA"]])

y <- Seurat2PB(data.seurat, sample="orig.ident", cluster="celltype")

 # tidyup samples
y$samples %<>% 
  as_tibble(rownames = "sample_cluster") %>% 
  left_join(meta %>% dplyr::select(sample, disease)) %>% 
  mutate(
    disease = factor(disease, levels = c("Control", "DLB", "PD", "PDD")), 
    cluster = as.factor(cluster), 
    sample_id = rownames(.),
    sample_cluster = as.factor(sample_cluster),
    disease_cluster = paste(disease, cluster, sep = "_") %>% as.factor
  ) %>%  
  column_to_rownames("sample_cluster") %>% 
  mutate(sample_id = rownames(.))


# Get all Entrez Gene IDs
all_genes <- keys(org.Hs.eg.db, keytype = "SYMBOL")

# Select gene SYMBOL and CHR (chromosome) info
hs_chrs <- AnnotationDbi::select(org.Hs.eg.db,
                    keys = all_genes,
                    columns = c("SYMBOL", "CHR"),
                    keytype = "SYMBOL")
```

# work out sex
The sex of each donor was not provided by GEO. Here, i will look at this by assessing expression of genes on the Y chromosome. There appears to be reads are aligning to genes  on the Y chromosome, however, these are in the pseudoautosomal regiosns. Nevertheless, there are clear differences in expression of genes on the Y chr. 

```{r}
y %>% 
  cpm(log=T) %>% 
  as_tibble(rownames = "SYMBOL") %>% 
  left_join(hs_chrs) %>% 
  pivot_longer(names_to = "sample_cluster", 
               values_to = "counts", rownames(y$samples)) %>% 
  dplyr::filter(CHR == "Y") %>% 
  mutate(
    sample = str_remove(sample_cluster, pattern = "_cluster.+")
  ) %>% 
  ggplot(
    aes(x = sample, y = counts)
  ) +
  geom_boxplot()+
  facet_wrap(~CHR) +
  scale_y_continuous(limits = c(0,10))
```

Whle the sample-spefiic info was NA in the paper and in GEO, there was this in the paper: 

`proportion female: control = 1/7, PD = 5/7, PDD = 2/7, DLB = 0/7`

So the samples which have low y-specific chr expression are likely the female samples. This all matches up nicely. 

```{r}
female.samples = c(
  #control
  "PDC87", 
  #PD - 5 samples
  "PD413", "PD416", "PD523", "PD666", "PD683", 
  
  # PDD - 2 samples
  "PD366", "PD501"
  )


```

```{r}
y$counts %>% 
  cpm(log=T) %>% 
  as_tibble(rownames = "SYMBOL") %>% 
  left_join(hs_chrs) %>% 
  pivot_longer(names_to = "sample_cluster", 
               values_to = "counts", rownames(y$samples)) %>% 
  dplyr::filter(CHR == "Y") %>%
  # dplyr::filter(SYMBOL %in% c("ZFY", "PCDH11Y", "USP9Y", "UTY")) %>% 
  mutate(
    sample = str_remove(sample_cluster, pattern = "_cluster.+"), 
    sex = case_when(
      sample %in% female.samples ~ "female", 
      TRUE ~ "male"
    ) %>% as.factor()
  ) %>%
  left_join(meta) %>% 
  ggplot(
    aes(x = sample, y = counts, fill = sex)
  ) +
  geom_boxplot()+
  facet_wrap(~disease, scales = "free_x") +
  scale_y_continuous(limits = c(0,10)) +
  labs(
    subtitle = "proportion female: control = 1/7, PD = 5/7, PDD = 2/7, DLB = 0/7"
  )
```

```{r}
# add this info back into y$samples
y$samples %<>% 
  mutate(
    sex = case_when(
      sample %in% female.samples ~ "female", 
      TRUE ~ "male"
    ) %>% as.factor()
  ) 
```

## lib sizes
```{r}
y$samples %>% 
  as.data.frame() %>% 
  rownames_to_column("sample_cluster") %>% 
  ggplot(
    aes(x = sample_cluster, y = lib.size, fill = disease)
  ) + 
  geom_col() +
  scale_y_continuous(labels = comma) +
  facet_wrap(~cluster, scales = "free_x", nrow = 2) +
  easy_rotate_x_labels(angle = -45)


# retain samples with more than 50k counts
keep.samples <- y$samples$lib.size > 5e4
y <- y[,keep.samples]

y$samples %>% 
  as.data.frame() %>% 
  rownames_to_column("sample_cluster") %>% 
  mutate(sample_cluster = str_remove(sample_cluster,pattern = "cluster")) %>% 
  ggplot(
    aes(x = sample_cluster, y = lib.size, fill = disease)
  ) + 
  geom_col() +
  scale_y_continuous(labels = comma) +
  facet_wrap(~cluster, scales = "free", nrow = 2) +
  easy_rotate_x_labels(angle = -45) +
  labs(
    title = "Library sizes after filtering"
  )
```

After filtering the small pseudobulk libraries (50k reads was the recommended by the edgeR manual, but this left out too many libraries),There are some celltypes n.d. in all groups. 

I will only retain clusters which are detected in all grousp. 


```{r}
clusters2remove = c("Endothelial", "VLMCs", "Neuron_Ex3", "Neuron_Ex4", "Neuron_Ex5", "Neuron_In1", "Neuron_In2")

samples2keep <- y$samples %>% 
  dplyr::filter(!cluster %in% clusters2remove) %>% 
  rownames(.)

# make a copy of y before filtering and filter low expressed genes
y.before.removing.clusters = y
keep.genes <- filterByExpr(
  y.before.removing.clusters,  
  group=y.before.removing.clusters$samples$cluster)

y.before.removing.clusters <- y.before.removing.clusters[keep.genes, , keep=FALSE]

y <- y[,samples2keep]
```

Then, lowly expressible genes are filterd. the table below shows the number of genes retained. TMM normalization is performed to estimate effective library sizes. 

Lib sizes are much smaller in the DLB group. Is this biological or technical in nature?

```{r}
keep.genes <- filterByExpr(y, group=y$samples$cluster)
table(keep.genes)
y <- y[keep.genes, , keep=FALSE]

y <- normLibSizes(y, method = "TMM")

y$samples %>% 
  ggplot(
    aes(y = lib.size, x = cluster, colour = disease )
  ) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge()) +
  stat_compare_means() +
  facet_wrap(~cluster, scales = "free_x")

y %>% 
  cpm(log = T) %>% 
  as.data.frame() %>% 
  pivot_longer(cols = everything(), 
               names_to = "sample", values_to = "logCPM") %>% 
  left_join(y$samples %>% 
              mutate(sample = rownames(y$samples))) %>% 
  ggplot(
    aes(x = sample, y = logCPM)
    ) +
  geom_boxplot(
    aes(fill = disease),
    outlier.size = 0.5
               ) +
  facet_wrap(~cluster, scales = "free")
  labs(title = "Distribution of logCPM values per sample",
       x = "Sample", y = "log2 CPM")
```


# PCA
## all samples
```{r}
pca = y$counts %>% 
  cpm(log = TRUE) %>% 
  t %>% 
  prcomp 


pca %>%   
  autoplot(
    data = tibble(
      sample = rownames(.$x)
    ) %>% 
      left_join(y$samples %>% 
                  mutate(sample = paste0(sample, "_cluster", cluster)
                )
                ),
    colour = "cluster", 
    shape = "disease",
    alpha = 0.5,
    size = 6
  ) +
  scale_color_viridis_d(option = "turbo")
```

## PCA by celltype
```{r}
plot.pca.bycelltype = function(dge.object, celltype.vector) {
  
  dge.object$counts %>% 
  as.data.frame() %>% 
  .[,grepl(colnames(y), pattern = celltype.vector)] %>% 
  cpm(log = T) %>% 
  t %>% 
  prcomp %>% 
  autoplot(
    data = tibble(
      sample = rownames(.$x)
    ) %>% 
      left_join(y$samples %>% 
                  mutate(sample = paste0(sample, "_cluster", cluster)
                )
                ),
    colour = "disease", 
    shape = "sex",
    alpha = 0.5,
    size = 6
  ) +
  scale_color_manual(values = c("#5F676D", "#17BEBB", "#FE5448", "#A20C01")) +
    ggtitle(celltype.vector)
}

plot.pca.bycelltype(dge.object = y, celltype.vector = "Astrocyte1")
plot.pca.bycelltype(dge.object = y, celltype.vector = "Astrocyte2")
plot.pca.bycelltype(dge.object = y, celltype.vector = "Microglia2")
plot.pca.bycelltype(dge.object = y, celltype.vector = "Microglia1")
plot.pca.bycelltype(dge.object = y, celltype.vector = "Neuron_Ex1")
plot.pca.bycelltype(dge.object = y, celltype.vector = "Neuron_Ex2")
plot.pca.bycelltype(dge.object = y, celltype.vector = "Oligodendrocyte")
```





# DE analysis

DE gene analysis was performed only on autosomal genes (to account for sex differences)
```{r}
# create a design matrix
design <- model.matrix(~0 + disease_cluster, 
                       data = y$samples %>% 
                         droplevels()) 

colnames(design) <- str_remove(colnames(design), 
                               "geno_cluster")

# extract out autosome genes
autosomal.genes <- y$genes %>% 
  left_join(hs_chrs %>% 
              set_colnames(c("gene", "chr"))) %>% 
  dplyr::filter(!chr %in% c("X", "Y")) %>% 
  .$gene
y <- y[autosomal.genes,]

# estimae dispersion
y <- estimateDisp(y, robust = TRUE)
plotBCV(y)

fit <- glmQLFit(y, design, robust = TRUE)
plotQLDisp(fit)

# Extract cell types shared by all diseases
design_cols <- colnames(design)
celltypes <- gsub("^disease_cluster(Control|DLB|PD|PDD)_", "", design_cols)
shared_celltypes <- names(table(celltypes))[table(celltypes) == 4]

# Build contrast strings: Control_celltype - DLB_celltype
contrast_strings <- 
  c(
    paste0("disease_clusterDLB_", 
           shared_celltypes, 
           " - disease_clusterControl_", 
           shared_celltypes
           ), 
    paste0("disease_clusterPD_", 
         shared_celltypes, 
         " - disease_clusterControl_", 
         shared_celltypes
         ), 
    
     paste0("disease_clusterPDD_", 
         shared_celltypes, 
         " - disease_clusterControl_", 
         shared_celltypes
         )
  )
    

# Generate contrast matrix
contrast_matrix <- makeContrasts(contrasts = contrast_strings, levels = design)

# rename columns for readability
colnames(contrast_matrix) <- gsub("disease_cluster", "", colnames(contrast_matrix))
```

## toptables
```{r}
toptables <- colnames(contrast_matrix) %>% 
  sapply(function(x) {
      glmQLFTest(fit, design, contrast=contrast_matrix[,x]) %>% 
      topTags(n = Inf, sort.by = "PValue") %>% 
      .$table %>% 
      as_tibble %>% 
      dplyr::select(
        gene, logCPM, logFC, PValue, FDR, `F`
      ) %>% 
      mutate(
        coef = x, 
        DE = FDR < 0.05, 
        coef = str_remove(coef, pattern = " - Control.+$")
      )
  }, simplify = F)

```
## volcano plots
```{r}
toptables %>%
  bind_rows() -> volcano_data

# Count number of DE up/down per celltype (coef)
volcano_labels <- volcano_data %>%
  dplyr::filter(DE) %>%
  mutate(direction = ifelse(logFC > 0, "Up", "Down")) %>%
  group_by(coef, direction) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(
    x = ifelse(direction == "Up", 3, -3),  # approximate x positions
    y = max(-log10(volcano_data$PValue), na.rm = TRUE) * 0.9,  # high y
    label = paste0(direction, ": ", n)
  )

# Plot
volcano_data %>% 
  ggplot(
    aes(x = logFC, y = -log10(PValue))
    ) +
  geom_point(
    aes(colour = DE)
    ) +
  scale_color_manual(values = c("grey50", "red")) +
  facet_wrap(~coef) +
  geom_text(
    data = volcano_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 4,
    hjust = 0.5
  ) +
  labs(
    x = "log2 Fold Change",
    y = "-log10(P value)"
  ) 
```

Some of the volcano plots havce some weird arm shapes. This usually comes from insufficient filtering. So i re-ran the analysis, while increasing the filtering. 


## MD plots
These plots have a bunch of low expressed genes - the MD plots dont have a nice triangle shape. I think i need to increase the filtering. probs going to lose some celltypes. 

```{r}
toptables %>% 
  bind_rows() %>% 
  ggplot(
    aes(x = logCPM, y = logFC)
  ) +
  geom_point(
    aes(colour = DE ),
    alpha = 0.5
  ) +
  facet_wrap(~coef) +
  scale_color_manual(values = c("grey50", "red"))
```

This hasnt worked. 

# enrichmenmt andlysis 
## CD pathways
```{r}
cluster.gprofiler.results <- 
  list.files(path = "data/results/adultchildsubtypes", 
             pattern = "gprofiler", recursive = TRUE, full.names = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/gprofiler_analysis_results.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/adultchildsubtypes", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))

network.genes.cluster <- cluster.gprofiler.results$CD_1A_lyso_DLB %>% 
  dplyr::select(Cluster, genes) %>% 
  mutate(genes = str_split(genes, " ")) %>% 
  deframe()

network.genes.cluster %<>% 
  lapply(function(x) {
    x[x %in% rownames(y)]
  })

# only retain clusters if they have more than 2 genes
network.genes.cluster <- 
  network.genes.cluster[lapply(network.genes.cluster, length) > 2]
  
### obtain the genes in the subnetworks
node_metadata <- 
  list.files(path = "data/results/adultchildsubtypes", 
             pattern = "overlapping_nodes_metadata", recursive = TRUE, full.names = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/overlapping_nodes_metadata.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/adultchildsubtypes", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))

network.genes <- node_metadata$CD_1A_lyso_DLB
```

# how many network genes are DE?
```{r}
toptables[grepl(names(toptables), pattern = "DLB")] %>% 
  bind_rows() %>% 
  dplyr::filter(
    DE == T, 
    gene %in% node_metadata$CD_1A_lyso_DLB$X
    ) %>% 
  dplyr::select(!starts_with("vf")) %>% 
  ggplot(
    aes(y = coef)
  ) +
  geom_bar() +
  labs(
    title = "Number of DE genes in DLB network"
  ) +

toptables[grepl(names(toptables), pattern = "PD")] %>% 
  bind_rows() %>% 
  dplyr::filter(
    DE == T, 
    gene %in% node_metadata$CD_1A_lyso_PD$X
    ) %>% 
  dplyr::select(!starts_with("vf")) %>% 
  ggplot(
    aes(y = coef)
  ) +
  geom_bar() +
  labs(
    title = "Number of DE genes in PD network"
  )

```


```{r}
net.genes.DE <- toptables[grepl(names(toptables), pattern = "DLB")] %>% 
  bind_rows() %>% 
  dplyr::filter(DE == T) %>% 
  dplyr::filter(gene %in% network.genes$X) %>% 
  .$gene %>% unique

toptables %>% 
  bind_rows() %>% 
  dplyr::filter(
    gene %in% net.genes.DE, 
    grepl(coef, pattern = "DLB")
    ) %>% 
  tidyHeatmap::heatmap(
    .row = gene, 
    .column = coef, 
    .value = logFC,
    
    palette_value = circlize::colorRamp2(
      seq(-5, 5, length.out = 11), 
      rev(RColorBrewer::brewer.pal(11, "RdBu"))
    )
  ) %>% 
  layer_text(
    FDR < 0.05, .value = "*"
  )
  
y %>% 
  cpm(log=T) %>% 
  .[net.genes.DE,] %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  pivot_longer(
    names_to = "sample_cluster", 
    values_to = "logCPM", 
    colnames(y)
    ) %>% 
  left_join(rownames_to_column(y$samples, 
                               "sample_cluster")) %>% 
  dplyr::filter(disease %in% c("Control", "DLB")) %>% 
  ggplot(
    aes(x = cluster, y = logCPM, fill = disease)
  ) +
  #geom_jitter() +
  geom_boxplot() +
  facet_wrap(~gene, scales = "free_y") +
  easy_rotate_x_labels(angle = -45) +
  theme(
    text = element_text(size = 18) 
    )

```


# Over representation analysis
## ORA of KEGG pathways within clusters
```{r}
ORA.clusters <- network.genes.cluster %>% 
  lapply(function(x) {
    clusterProfiler::enrichKEGG(
      gene = bitr(x, 
                  OrgDb = org.Hs.eg.db,
                  fromType="SYMBOL", toType="ENTREZID") %>% .$ENTREZID,
      keyType = "ncbi-geneid", 
      organism = "hsa"
    ) 
    })
```

#### number of DE genes per cluster
```{r}
ora.DEG.in.clusters <- toptables %>%
  lapply(function(x) {
    genes <- x %>%
      dplyr::filter(DE == TRUE) %>%
      pull(gene)
    
    if (length(genes) < 5) return(NULL)  # skip if too few genes
    
    clusterProfiler::enricher(
      gene = genes,
      TERM2GENE = network.genes.cluster %>%
        enframe(name = "TERM", value = "GENE") %>%
        unnest(cols = "GENE"), 
      universe = x$gene
    )
  }) %>%
  purrr::compact() 

ora.DEG.in.clusters %>% 
  lapply(as_tibble) %>% 
  bind_rows(.id = "comparison")

keep = toptables %>% 
  bind_rows() %>% 
  dplyr::filter(
    gene %in% network.genes.cluster$C438
  ) %>% 
  dplyr::filter(DE == T) %>% 
  .$gene

  # Build the FDR matrix (for asterisk overlay)
  FDR_mat <- toptables %>%
    bind_rows() %>%
    dplyr::filter(gene %in% network.genes$X) %>%
    dplyr::select(coef, gene, FDR) %>%
    pivot_wider(names_from = gene, values_from = FDR) %>%
    column_to_rownames("coef") %>%
    as.matrix()

Heatmap(
  matrix = toptables %>%
    bind_rows() %>%
    dplyr::filter(gene %in% network.genes$X) %>%
    dplyr::select(coef, gene, logFC) %>%
    pivot_wider(names_from = gene, values_from = logFC) %>%
    column_to_rownames("coef") %>%
    as.matrix(),
  name = "logFC",
  col = brewer_pal(palette = "RdBu")(7), 
  
  cell_fun =function(j, i, x, y, width, height, fill) {
      # j = column index, i = row index
      # x, y = coordinates of the center of the cell
      # width, height = size of the cell
      # fill = background color (not used here)

      # Check if the FDR value for this cell exists and is below 0.05
      if (!is.na(FDR_mat[i, j]) && FDR_mat[i, j] < 0.05) {
        # If yes, overlay a text label "*" on that cell
        grid.text("*", x, y, gp = gpar(fontsize = 15))
      }
    },
  column_names_gp = gpar(fontsize = 10),
  show_row_names = TRUE,
  show_column_names = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE
)


```



```{r}
#plot number of DE genes per cluster
network.genes.cluster %>% 
  lapply(function(x) {
    tibble(
      gene = x
    ) %>% 
      left_join(toptables %>% 
                  bind_rows() %>% 
                  dplyr::select(gene, coef, FDR) %>% 
                  mutate(DE = FDR < 0.05), 
                by = "gene"
      )
  }) %>% 
  bind_rows(.id = "cluster") %>% 
  dplyr::filter(DE == T) %>% 
  mutate(
    celltype = str_extract(
      coef,
      pattern = "(?<=DLB_).+(?= - Control_)")) %>% 
 count(cluster, celltype, name = "num_DE_genes") %>%
  ggplot(aes(x = cluster, y = num_DE_genes, fill = celltype)) +
  geom_col(position = "dodge") +
  labs(
    title = "Number of DE genes (FDR < 0.05)",
    x = "Network cluster",
    y = "# DE genes"
  ) +
  scale_y_continuous(breaks = 0:12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r}
plot.cnet = function(toptable, plot.title, ORA.cluster.object) {
  # 1. Get toptable for the cluster you're plotting
top.genes <- toptable %>%
  # dplyr::filter(PValue < 0.05) %>% 
  dplyr::select(gene, logFC)

# 2. Map gene symbols to ENTREZID
fc.entrez <- bitr(
  top.genes$gene,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
) %>%
  left_join(top.genes, by = c("SYMBOL" = "gene")) %>%
  dplyr::distinct(ENTREZID, .keep_all = TRUE) %>%
  dplyr::select(-SYMBOL) %>% 
  deframe()  # named vector: names are ENTREZID, values are logFC

cnetplot(
  x = setReadable(ORA.cluster.object, 
                  OrgDb = org.Hs.eg.db, keyType = "ENTREZID"),
  foldChange = fc.entrez,
  showCategory = 20,
  circular = FALSE,
  color_category='grey20', size_category = 0.5,
  color_item = "grey90", 
  colorEdge = F
) +
  scale_color_gradient2(name='logFC', low='blue4', high='firebrick4') + 
  ggtitle(plot.title)
  
}
```

```{r}
plot.cnet(
  toptable = toptables$`DLB_astrocyte - Control_astrocyte`,
  ORA.cluster.object = ORA.clusters$C453, 
  plot.title = "Astrocytes")

plot.cnet(
  toptable = toptables$`DLB_microglia_immune - Control_microglia_immune`,
  ORA.cluster.object = ORA.clusters$C439, 
  plot.title = "Microglia")

plot.cnet(
  toptable = toptables$`DLB_microglia_immune - Control_microglia_immune`,
  ORA.cluster.object = ORA.clusters$C448, 
  plot.title = "Microglia")

plot.cnet(
  toptable = toptables$`DLB_microglia_immune - Control_microglia_immune`,
  ORA.cluster.object = ORA.clusters$C438, 
  plot.title = "Microglia")
```





```{r}
plot.cnet(
  toptable = toptables$`DLB_neuron_inhib - Control_neuron_inhib`,
  ORA.cluster.object = ORA.clusters$C441, 
  plot.title = "inhib neurons")

plot.cnet(
  toptable = toptables$`DLB_neuron_inhib - Control_neuron_inhib`,
  ORA.cluster.object = ORA.clusters$C440, 
  plot.title = "inhib neurons")
```

## ORA of KEGG pathways
```{r}
ora.DEG.KEGG <- toptables %>%
  lapply(function(x) {
    genes <- x %>%
      dplyr::filter(DE == TRUE) %>%
      pull(gene)
    
    if (length(genes) < 5) return(NULL)  # skip if too few genes
    
    clusterProfiler::enrichKEGG(
      gene = bitr(genes, 
                  OrgDb = org.Hs.eg.db,
                  fromType="SYMBOL", toType="ENTREZID") %>% .$ENTREZID,
      universe = bitr(x$gene, 
                  OrgDb = org.Hs.eg.db,
                  fromType="SYMBOL", toType="ENTREZID") %>% .$ENTREZID,
      keyType = "ncbi-geneid", 
      organism = "hsa") 
  }) %>%
  purrr::compact() 
```

```{r}
ora.DEG.KEGG %>%
  lapply(as_tibble) %>% 
  bind_rows(.id = "coef") %>% 
  dplyr::filter(Count > 1) %>% 
  ggplot(
    aes(x = -log10(pvalue), y = Description)
  ) + 
  geom_col() +
  facet_wrap(~coef, scales = "free_y")
```



### fry gene set tests
#### Obtain KEGG gene sets
```{r}
library(KEGGREST)

### obtain the pathway to entrez geneid link
path2entrezgene  <- 
  keggLink("pathway", "hsa") %>% 
  tibble(
    pathway = ., 
    entrez_gene_id = sub("hsa:", "", names(.))
    )

# obtain the pathway names
pathways <-
  keggList("pathway", "hsa") %>%
  tibble(pathway = names(.), description = .)

# Use org.eg.db to obtain the conversion of entrez to gene symbol
hsa_kegg_anno <- 
  path2entrezgene %>%
  mutate(
    symbol = mapIds(org.Hs.eg.db , entrez_gene_id, 
                    "SYMBOL", "ENTREZID")
  ) %>% 
  mutate(pathway = str_remove(pathway, pattern = "path:")) %>%
  left_join(pathways) %>%
  dplyr::select(-entrez_gene_id) %>% 
  dplyr::filter(
    symbol %in% (y$genes$gene)) %>%
  mutate(
    description = str_remove_all(description, pattern = " \\- Homo .+"))


# make the KEGG object for gene set testing
KEGG <- hsa_kegg_anno %>%
  na.omit %>%
  distinct(pathway, symbol, .keep_all = TRUE) %>%
  split(f = .$description) %>%
  lapply(function(x) {x$symbol %>% unname()})

# only retain pathways with more than 3 genes
KEGG <- KEGG[lapply(KEGG, length) > 3 ]

```

#### Run FRY on network clusters
```{r}
fry <- colnames(contrast_matrix) %>% 
  sapply(function(x) {
   cpm(y, log = T)  %>% 
      fry(
        index = network.genes.cluster,
        design = design, 
        contrast = contrast_matrix[,x], 
        sort = "directional"
      ) %>% 
      rownames_to_column("pathway") %>% 
      as_tibble() %>% 
      mutate(coef = x) %>% 
      left_join(
        cluster.gprofiler.results$CD_1A_lyso_DLB %>% 
          dplyr::select(pathway = Cluster, top.pathway.gprofiler = Top.pathway)
      )
  }, simplify = FALSE)
```

#### run FRY on KEGG pws
```{r}
fry.KEGG <- colnames(contrast_matrix) %>% 
  sapply(function(x) {
   cpm(y, log = T)  %>% 
      fry(
        index = KEGG,
        design = design, 
        contrast = contrast_matrix[,x], 
        sort = "mixed"
      ) %>% 
      rownames_to_column("pathway") %>% 
      as_tibble() %>% 
      mutate(coef = x)
      
  }, simplify = FALSE)
```

```{r}
fry.KEGG %>% 
  bind_rows() %>% 
  dplyr::filter(FDR.Mixed < 0.05) %>% 
  ggplot(
    aes(x = coef, y = pathway, fill = -log10(PValue.Mixed))
  ) +
  geom_tile() 
  facet_wrap(~coef, scales = "free")
```

```{r}

pw.interest = c("Lysosome", "Sphingolipid metabolism", "Other glycan degradation" , "Glycosaminoglycan degradation","Phagosome", "Ferroptosis", "Mineral absorption", "Antigen processing and presentation", "Cholesterol metabolism" , "Starch and sucrose metabolism", "Sphingolipid signaling pathway", "Rheumatoid arthritis", "Synaptic vesicle cycle", "Galactose metabolism", "Allograft rejection", "Carbohydrate digestion and absorption", "Tuberculosis", "Porphyrin metabolism", "Type I diabetes mellitus", "Graft-versus-host disease", "Protein processing in endoplasmic reticulum", "Amino sugar and nucleotide sugar metabolism", "Autoimmune thyroid disease", "Viral myocarditis", "PPAR signaling pathway", "Salivary secretion")

fry.KEGG %>% 
  bind_rows() %>% 
  dplyr::filter(pathway %in% pw.interest) %>% 
  ggplot(
    aes(x = -log10(FDR), y = pathway,
        fill = Direction,
        alpha = FDR < 0.05)
  ) +
  geom_col() + 
  facet_wrap(~coef, nrow = 1) +
  theme(
    legend.position = "top"
  )

```



### vis 
#### which clusters are sig
```{r}
fry %>% 
  bind_rows() %>% 
  pivot_longer(
    cols = c(PValue, PValue.Mixed),
    names_to = "pval_type",
    values_to = "pval"
  ) %>%
  mutate(
    coef = str_extract(coef, 
                         pattern = "(?<=DLB_).+(?= - Control_)"),
    neglog10_p = -log10(pval),
    pval_type = recode(pval_type, "PValue" = "Directional", "PValue.Mixed" = "Mixed"),
    fdr = case_when(
      pval_type == "Directional" ~ FDR,
      pval_type == "Mixed" ~ FDR.Mixed
    ),
    sig = fdr < 0.05,
    alpha_val = if_else(sig, 1, 0.3),
    pathway = factor(pathway, levels = unique(pathway))
  ) %>% 
  ggplot(aes(x = pval_type,
             y = pathway, fill = neglog10_p, alpha = alpha_val)) +
  geom_tile() +
  facet_wrap(~coef, nrow = 1) +
  scale_fill_viridis_c(option = "plasma", name = "-log10(P)") +
  scale_alpha(range = c(0.3, 1), guide = "none") +
  labs(
    x = "Hypothesis test",
    y = "Netcoloc cluster"
  ) 
```

#### network
##### netcoloc cluster 436
```{r}
plot.cnet(
  toptable = toptables$`DLB_neuron_inhib - Control_neuron_inhib`, 
  ORA.cluster.object = ORA.clusters$C436, 
  plot.title = "Cluster C436 in inhib neurons"
  )

plot.cnet(
  toptable = toptables$`DLB_vascular - Control_vascular`, 
  ORA.cluster.object = ORA.clusters$C436, 
  plot.title = "Cluster C436 in vascular cells"
  )


```


##### neurons
```{r}
# 1. Get toptable for the cluster you're plotting
top.genes <- toptables$`DLB_neuron_inhib - Control_neuron_inhib` %>%
  dplyr::select(gene, logFC)

# 2. Map gene symbols to ENTREZID
fc.entrez <- bitr(
  top.genes$gene,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
) %>%
  left_join(top.genes, by = c("SYMBOL" = "gene")) %>%
  dplyr::distinct(ENTREZID, .keep_all = TRUE) %>%
  dplyr::select(-SYMBOL) %>% 
  deframe()  # named vector: names are ENTREZID, values are logFC

cnetplot(
  x = setReadable(ORA.clusters$C436, OrgDb = org.Hs.eg.db, keyType = "ENTREZID"),
  foldChange = fc.entrez,
  showCategory = 10,
  circular = FALSE,
  color_category='grey80', 
  color_item = "grey90", 
  colorEdge = TRUE
) +
  scale_color_gradient2(name='logFC', low='blue4', high='firebrick4') + 
  ggtitle("Neuron network")
```

#### logCPM
```{r}
plot.logCPM = function(cluster.list, plot.title) {
  
    logcpm_mat <- y %>%
    cpm(log = TRUE) %>%
    .[cluster.list, , drop = FALSE]
  
    sample_meta <- y$samples[colnames(logcpm_mat), , drop = FALSE]
    
 Heatmap(
    matrix = logcpm_mat,
    name = "logCPM",
    top_annotation = HeatmapAnnotation(
      celltype = sample_meta$cluster %>% droplevels(),
      disease = sample_meta$disease %>% droplevels(),
      col = list(
        disease = c("Control" = "grey10", "DLB" = "red"), 
        celltype = viridis_pal(option = "turbo")(8) %>% 
          set_names(sample_meta$cluster %>% unique)
      )
    ),
    
    col = viridis_pal()(10),
    column_split = sample_meta$cluster,
    column_title = plot.title,
    show_column_names = F,
    show_row_names = T,
    cluster_rows = TRUE,
    cluster_columns = TRUE, 
  )
}

plot.logCPM(cluster.list = network.genes.cluster$C453, 
            plot.title = "C501: lysosomal lumen")
```

#### logFC of DEGs
```{r}

keep = toptables %>% 
  bind_rows() %>% 
  dplyr::filter(
    gene %in% network.genes.cluster$C453
  ) %>% 
  dplyr::filter(DE == T) %>% 
  .$gene

  # Build the FDR matrix (for asterisk overlay)
  FDR_mat <- toptables %>%
    bind_rows() %>%
    dplyr::filter(gene %in% net.genes.DE) %>%
    dplyr::select(coef, gene, FDR) %>%
    pivot_wider(names_from = gene, values_from = FDR) %>%
    column_to_rownames("coef") %>%
    as.matrix()

Heatmap(
  matrix = toptables %>%
    bind_rows() %>%
    dplyr::filter(gene %in% net.genes.DE) %>%
    dplyr::select(coef, gene, logFC) %>%
    pivot_wider(names_from = gene, values_from = logFC) %>%
    column_to_rownames("coef") %>%
    as.matrix(),
  name = "logFC",
  col = brewer_pal(palette = "RdBu")(7), 
  
  cell_fun =function(j, i, x, y, width, height, fill) {
      # j = column index, i = row index
      # x, y = coordinates of the center of the cell
      # width, height = size of the cell
      # fill = background color (not used here)

      # Check if the FDR value for this cell exists and is below 0.05
      if (!is.na(FDR_mat[i, j]) && FDR_mat[i, j] < 0.05) {
        # If yes, overlay a text label "*" on that cell
        grid.text("*", x, y, gp = gpar(fontsize = 15))
      }
    },
  column_names_gp = gpar(fontsize = 12),
  show_row_names = TRUE,
  show_column_names = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE
)
```


```{r}
toptables$`DLB_microglia_immune - Control_microglia_immune` %>%
  dplyr::filter(gene %in% network.genes.cluster$C436) %>% 
  # dplyr::filter(FDR < 0.05) %>% 
  dplyr::select(
    gene, logFC, PValue, FDR, coef
  ) %>% 
  left_join(network.genes %>% 
              dplyr::select(gene = X, z_d2, z_both)) %>% 
  ggplot(
    aes(x = -log10(PValue), y = z_d2)
  ) +
  geom_point() +
  geom_label_repel(
    aes(label= gene), 
    data = . %>% 
      dplyr::filter(
        FDR < 0.1 & z_both > 5
      )
  )

toptables$`DLB_microglia_immune - Control_microglia_immune` %>%
  dplyr::filter(gene %in% network.genes.cluster$C436) %>% 
  # dplyr::filter(FDR < 0.05) %>% 
  dplyr::select(
    gene, logFC, PValue, FDR, coef
  ) %>% 
  left_join(network.genes %>% 
              dplyr::select(gene = X, z_d2, z_both)) %>% 
  ggplot(
    aes(x = logFC, y = z_both, colour = -log10(PValue))
  ) +
  geom_point() +
  geom_label_repel(
    aes(label= gene), 
    data = . %>% 
      dplyr::filter(
        FDR < 0.05 & z_both > 5
      )
  )
```

# data export
```{r}


y %>% 
  qsave("data/public data/GSE178146_RAW/R_objects/pseudobulk_dgeObject.qs")

ORA.clusters %>% 
  qsave("data/public data/GSE178146_RAW/R_objects/ORA_clusters.qs")

ora.DEG.in.clusters %>% 
  qsave("data/public data/GSE178146_RAW/R_objects/ORA_DEGs_clusters.qs")

fry %>% 
  qsave("data/public data/GSE178146_RAW/R_objects/fry_results.qs")

toptables %>% 
  qsave("data/public data/GSE178146_RAW/R_objects/toptables.qs")
```


# figs

## Which cell types are the network genes expressed in?
```{r}
network.genes <- node_metadata[
  node_metadata %>% names %>% grepl(pattern = "CD_1A")
] %>% 
  lapply(function(x) {
    tibble(
      gene_name = x$X
    )
  }) %>% 
  set_names(str_remove(names(.), pattern = "CD_1A_lyso_"))


# violin plot to show the expression of network genes across celltypes
plot.network.genes.across.celltypes = function(
  network.genes, 
  disease.object,
  ggtitle
  ){
y %>% 
  cpm(log = T) %>% 
  as_tibble(rownames = "gene_name") %>% 
  pivot_longer(
    names_to = "sample_id", 
    values_to = "logCPM", 
    colnames(y)
  ) %>% 
  left_join(y$samples) %>% 
  dplyr::filter(
    gene_name %in% network.genes, 
    disease %in% disease.object
  ) %>% 
  ggplot(
    aes(y = logCPM, x = disease, colour = disease)
  ) +
  geom_violin() +
  geom_boxplot(
    width = 0.2, 
    position = position_dodge(width = 1)
  ) +
  facet_wrap(~cluster, scales = "free_x", nrow = 1) +
  labs(
    title = ggtitle
  ) +
    theme(legend.position = "none")
}

ggarrange(
plot.network.genes.across.celltypes(
  network.genes = network.genes$DLB$gene_name,
  disease.object = c("Control", "DLB"), 
  ggtitle = "Expression of lysosomal childhood dementias & DLB subnetwork genes"
) +
  scale_color_manual(
    values = viridis_pal(option = "magma", end = 0.8)(2)
  ),
plot.network.genes.across.celltypes(
  network.genes = network.genes$PD$gene_name,
  disease.object = c("Control", "PD", "PDD"), 
  ggtitle = "Expression of lysosomal childhood dementias & PD subnetwork genes"
  ) +
  scale_color_manual(
    values = viridis_pal(begin = 0.1, end = 0.5)(3)
    ) , 

ncol = 1
)

```


```{r}
ggplot(volcano_data, aes(x = logFC, y = -log10(PValue))) +
  geom_point(aes(colour = DE)) +
  scale_color_manual(values = c("grey50", "red")) +
  facet_wrap(~coef) +
  geom_text(
    data = volcano_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 4,
    hjust = 0.5
  ) +
  labs(
    x = "log2 Fold Change",
    y = "-log10(P value)"
  ) +
  ggsave(
    "output/DLB_snRNAseq/figs/talk/celltype_volcanoplots.png", 
    width = 12, height = 8, units = "cm",
    scale = 2.5)
```
```{r}
network.genes.cluster %>% 
  lapply(function(x) {
    tibble(
      gene = x
    ) %>% 
      left_join(toptables %>% 
                  bind_rows() %>% 
                  dplyr::select(gene, coef, FDR) %>% 
                  mutate(DE = FDR < 0.05), 
                by = "gene"
      )
  }) %>% 
  bind_rows(.id = "cluster") %>% 
  dplyr::filter(DE == T) %>% 
  mutate(
    celltype = str_extract(
      coef,
      pattern = "(?<=DLB_).+(?= - Control_)")) %>% 
 count(cluster, celltype, name = "num_DE_genes") %>%
  ggplot(aes(x = cluster, y = num_DE_genes, fill = celltype)) +
  geom_col(position = "dodge", colour = "black") +
  labs(
    title = "Number of DE genes (FDR < 0.05)",
    x = "Network cluster",
    y = "# DE genes"
  ) +
  scale_y_continuous(breaks = 0:9) +
  scale_fill_viridis_d() +
  coord_flip() 
  ggsave(
    "output/DLB_snRNAseq/figs/talk/numDEGsperclusterpercelltype.png", 
    width = 12, height = 8, units = "cm",
    scale = 2.5)


a <- network.genes.cluster %>% 
  lapply(function(x) {
    tibble(
      gene = x
    ) %>% 
      left_join(toptables %>% 
                  bind_rows() %>% 
                  dplyr::select(gene, coef, FDR) %>% 
                  mutate(DE = FDR < 0.05), 
                by = "gene"
      )
  }) %>% 
  bind_rows(.id = "cluster") %>% 
  dplyr::filter(DE == T) %>% 
  mutate(
    celltype = str_extract(
      coef,
      pattern = "(?<=DLB_).+(?= - Control_)")) %>% 
 count(cluster, celltype, name = "num_DE_genes") %>%
  ggplot(aes(x = cluster, y = num_DE_genes, fill = celltype)) +
  geom_col(position = "dodge", colour = "black") +
  labs(
    title = "Number of DE genes (FDR < 0.05)",
    x = "Network cluster",
    y = "# DE genes"
  ) +
  scale_y_continuous(breaks = 0:9) +
  scale_fill_viridis_d() +
  coord_flip() +
   theme(legend.position = "bottom")

b <- fry %>% 
  bind_rows() %>% 
   mutate(
    coef = str_extract(coef, 
      pattern = "(?<=DLB_).+(?= - Control_)"
    )) %>% 
  dplyr::filter(FDR.Mixed < 0.05) %>% 
  ggplot(
    aes(y = pathway, x = -log10(PValue.Mixed), fill = Direction)
  ) +
  geom_col() +
  facet_wrap(~coef, scales = "free_y") +
  labs(
    y = "Cluster from network", 
    title = "Ranked-list gene set testing (limma::fry)"
  ) +
  theme(legend.position = "bottom")

a + b +
  plot_layout(
    widths = c(1,2),
    heights = c(1,2)
    ) +
  
ggsave(
    "output/DLB_snRNAseq/figs/talk/numDEGsperclusterpercelltype.png", 
    width = 12, height = 8, units = "cm",
    scale = 2.5)
```










