---
title: "04_npc_mouse_snRNAseq"
author: "Karissa Barthelson"
date: "2025-07-09"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# Introduction
```{r}
library(tidyverse)
library(forcats)
library(magrittr)
library(readxl)
library(qs) # for fast save and reading of objects into memory (alternative to rds)
library(parallel) 
library(Matrix)

library(Seurat)
library(org.Mm.eg.db)

# vis
library(pander)
library(ggpubr)

theme_set(theme_light() +
            theme(plot.title = element_text(hjust = 0.5), 
                  strip.text = element_text(face = "bold"),
                  strip.background = element_rect(fill = "grey20")
            )
)

```

This is the analysis of NPC1 single cell data. Sample metadata was obtained from the [SRA run table](https://www.ncbi.nlm.nih.gov/Traces/study/?uids=10122620%2C10122619%2C10122618%2C10122617%2C10122616%2C10122615%2C10122614%2C10122613&o=acc_s%3Aa&s=SRR11106929,SRR11106930,SRR11106931,SRR11106932,SRR11106933,SRR11106934,SRR11106935,SRR11106936). The single cell data was downloaded from SRA and processed using cellranger. 
```{r}
meta <- read_csv(
  "data/public data/PRJNA606815_NPC_mousebrain/SraRunTable.csv"
) %>% 
  # only selec relevant cols
  dplyr::select(age = AGE, genotype, sample = `Library Name`) %>% 
  # tidy up genotype and age
  mutate(
    genotype = case_when(
      genotype == "Npc1+/+" ~ "WT",
      genotype == "Npc1-/-" ~ "Npc1"
    ) %>% fct_rev(), 
    age = str_remove(age, " ") %>% as.factor()
  )

cellranger.parent.dir = "data/public data/PRJNA606815_NPC_mousebrain/cellranger"

# create named objects of the path and sample id for each filtered and raw object. 
sample.paths.filt = 
  grep(list.dirs(cellranger.parent.dir), pattern = "filter", value = TRUE)

sample.ids <- str_extract(sample.paths.filt, pattern = "(?<=/)S[0-9]+(?=/)")
sample.paths.filt <- setNames(sample.paths.filt, sample.ids)

# read in filtered data and convert to seurat object
data.10x.filt <-  Read10X(
  # get the filtered data from cellranger  
    data.dir = sample.paths.filt, 
  )

data.seurat <- CreateSeuratObject(
  counts = data.10x.filt, 
  project = "NPC1", 
  min.features = 200, # include nuclei with at least this many genes. 
  min.cells = 3 #include genes expressed in at least this many nuclei
  )

# join the metadata to the eurat object
data.seurat@meta.data %<>% 
  as_tibble(rownames = "umi") %>% 
  left_join(
    meta %>% 
      dplyr::rename(orig.ident = sample)
  ) %>% 
  column_to_rownames("umi")
```

# QC
## histogram of lib sizes

```{r}
data.seurat@meta.data %>% 
  ggplot(
    aes(x = nCount_RNA, fill = orig.ident)
  ) +
  geom_histogram(binwidth = 10) +
  coord_cartesian(xlim = c(0,10000)) + 
  facet_wrap(~orig.ident+genotype) +
  scale_y_log10() +
  labs(
    y = "log10(counts)"
  )
```

## percent mt transcripts

When a cell has a low number of detected genes, a low count depth and a high fraction of mitochondrial counts it might have a broken membrane which can indicate a dying cell. As these cells are usually not the main target of our analysis and might distort our downstream analysis, we are removing them during quality control. 

```{r}
data.seurat[["percent.mt"]] <- 
  PercentageFeatureSet(data.seurat, pattern = "^mt-")

# Visualize QC metrics as a violin plot
VlnPlot(data.seurat, 
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), 
        split.by = "genotype",
        ncol = 3, alpha = 0.25) +
  theme(aspect.ratio = 1)
```

```{r}
FeatureScatter(data.seurat, 
               feature1 = "nCount_RNA", feature2 = "percent.mt", 
               shuffle = TRUE
               )

FeatureScatter(data.seurat, 
               feature1 = "nFeature_RNA", feature2 = "percent.mt", 
               shuffle = TRUE
               ) + 
  geom_hline(yintercept = 35)


FeatureScatter(data.seurat, 
               feature1 = "nCount_RNA", feature2 = "nFeature_RNA", 
               shuffle = TRUE
               ) 

```

# initial clustering

I first want to suss the clustering before doing any filtering - how pure are the main neural cell clusters of the brain?

```{r}
# temp <- data.seurat %>% 
#   NormalizeData() %>% 
#   FindVariableFeatures() %>% 
#   ScaleData() %>% 
#   RunPCA()
# ElbowPlot(temp, ndims = 30)
remove(temp)

data.seurat %<>% 
  NormalizeData() %>% 
  FindVariableFeatures() %>% 
  ScaleData() %>% 
  RunPCA() %>% 
  FindNeighbors(dims = 1:30) %>% 
  FindClusters(resolution = 0.2) %>% # less clusters
  RunUMAP(dims = 1:30)
  
DimPlot(data.seurat, 
         group.by = "genotype", 
        split.by = "age",
        shuffle = TRUE
         ) +
  ggtitle("UMAP of all cells by genotype and age") +
  theme_light()

DimPlot(data.seurat, 
        group.by = "seurat_clusters",
        shuffle = TRUE) 
```
## assess marker genes

```{r}
markers.1 <-FindAllMarkers(data.seurat, only.pos = TRUE)
markers.1 %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 2)
```

```{r}
runFeatureplot = function(seurat.object, genes, celltype) {
  DimPlot(seurat.object,label = T) +
  FeaturePlot(
    seurat.object,
    features= genes
  ) + 
  # add the title to the plot
    plot_annotation(
    title = celltype,  
    theme = theme(plot.title = element_text(size = 18, face = "bold"))
    )
}
```

I will be using the marker genes defined in a previous snRNAseq experiment of the mouse cerebellum: [Kozareva et al 2021](https://www.nature.com/articles/s41586-021-03220-z). They have nice marker genes for the main cell types in the cerebellum with literature evidence (experimental) 

```{r}
runFeatureplot(
  seurat.object = data.seurat, 
  genes = c("Gfap","Aqp4"), 
  celltype = "Astrocytes")

runFeatureplot(
  seurat.object = data.seurat, 
  genes = c("Ppp1r17"), 
  celltype = "Purkinje cells")

runFeatureplot(
  seurat.object = data.seurat, 
  genes = c("Gabra6"), 
  celltype = "Granule cells")

runFeatureplot(
  seurat.object = data.seurat, 
  genes = c("Slc6a5", "Grm2", "Sst"), 
  celltype = "Golgi cells")

runFeatureplot(
  seurat.object = data.seurat, 
  genes = c("Prkcd", "Sorcs3", "Ptprk"), 
  celltype = "MLI1")

runFeatureplot(
  seurat.object = data.seurat, 
  genes = c("Prkcd", "Nxph1", "Cdh22"), 
  celltype = "MLI2")

runFeatureplot(
  seurat.object = data.seurat, 
  genes = c("Lypd6", "Ppfibp1"), 
  celltype = "OPC")

runFeatureplot(
  seurat.object = data.seurat, 
  genes = c("Mobp", "Mbp"), 
  celltype = "ODC")

runFeatureplot(
  seurat.object = data.seurat, 
  genes = c("Dcn"), 
  celltype = "Fibroblast")

runFeatureplot(
  seurat.object = data.seurat, 
  genes = c("Mrc1", "C1qa"), 
  celltype = "Macrophage & Microglia")

runFeatureplot(
  seurat.object = data.seurat, 
  genes = c("Flt1"), 
  celltype = "Endo-stalk")

runFeatureplot(
  seurat.object = data.seurat, 
  genes = c("Foxj1"), 
  celltype = "Ependynal")
```





## filtering
I will filter the data based on the number of features (genes) detected per cell, the number of counts per cell and the percentage of mitochondrial genes expressed.
```{r}

data.seurat.filt <- data.seurat %>% 
  subset(
    subset = nFeature_RNA > 200 & nFeature_RNA < 3000 & percent.mt < 20
  ) %>% 
  NormalizeData() %>% 
  FindVariableFeatures() %>% 
  ScaleData() %>% 
  RunPCA() %>% 
  FindNeighbors(dims = 1:20) %>% 
  FindClusters(resolution = 0.3) %>% # less clusters
  RunUMAP(dims = 1:20)

ElbowPlot(data.seurat.filt)

DimPlot(data.seurat.filt)
DimPlot(data.seurat.filt, group.by = "genotype")
DimPlot(data.seurat.filt, split.by = "age") +
  theme_classic()

VlnPlot(data.seurat.filt, 
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), 
        split.by = "genotype",
        ncol = 3, alpha = 0.25) +
  theme(aspect.ratio = 1)
```

## assess marker genes
```{r}
markers.2 <-FindAllMarkers(data.seurat.filt, only.pos = TRUE) %>% 
  as.data.frame() %>% 
  split(f = .$cluster)
```

### Cluster 0 and 1
```{r}
runFeatureplot(
  seurat.object = data.seurat.filt, 
  genes = c("Gabra6", "Kcnd2", "Grik2", "Tenm1"), 
  celltype = "Granule cells")
```

### Cluster 2
```{r}
runFeatureplot(
  seurat.object = data.seurat.filt, 
  genes = c("Gria1","Slc1a3", "Gabrb1", "Luzp2"), 
  celltype = "Bergman Astrocytes")
```

### Cluster 3
```{r}
runFeatureplot(
  seurat.object = data.seurat.filt, 
  genes = c("Hexb", "C1qa", "Ctss", "Trem2"), 
  celltype = "Macrophage & Microglia")
```

### Cluster 4 
```{r}
runFeatureplot(
  seurat.object = data.seurat.filt, 
  genes = c("Mobp", "Mbp", "Prr5l", "Plp1"), 
  celltype = "Oligodendrocytes")
```


### Cluster 5
Endothelial stalk cells are specialized vascular cells vital in angiogenesis, a process through which new blood vessels are formed from pre-existing ones; specifically, endothelial stalk cells are involved in sprouting angiogenesis, where they help form the body of new blood vessels. 
```{r}
runFeatureplot(
  seurat.object = data.seurat.filt, 
  genes = c("Flt1", "Ly6c1", "Cldn5", "Slco1a4"), 
  celltype = "Endothelial stalk cells")

```


### Cluster 6
```{r}
runFeatureplot(
  seurat.object = data.seurat.filt, 
  genes = c("Slc4a4","Aqp4", "Gabrb1", "Npas3"), 
  celltype = "Astrocytes")

```

### Cluster 7
Mixed population. Top marker genes do not have clear expression across clusters in the cerebellum (see [here](https://singlecell.broadinstitute.org/single_cell/study/SCP795/a-transcriptomic-atlas-of-the-mouse-cerebellum?genes=Fabp7&tab=distribution))

Aldoc expression is observed in Astrocytes (but more so in purkinje cells) [Fugita et al. 2014](https://pmc.ncbi.nlm.nih.gov/articles/PMC3903578). 

This clister appears to mostly be present in 3 week cerebellums. 
Also has low RNA counts. I think this cluster should be omitted from the analysis

```{r}
VlnPlot(data.seurat.filt, features = c("nCount_RNA"), pt.size = 0)
DimPlot(data.seurat.filt, group.by = "genotype", split.by = "age", shuffle = TRUE)

runFeatureplot(
  seurat.object = data.seurat.filt, 
  genes = c("Tubb2b","Fabp7", "Aldoc", "Hopx"), 
  celltype = "unsure")
```

### Cluster 8
```{r}
runFeatureplot(
  seurat.object = data.seurat.filt, 
  genes = c("Mobp", "Bcas1", "Mbp", "Plekhb1"), 
  celltype = "Oligodendrocyte")
```

### Cluster 9
```{r}
runFeatureplot(
  seurat.object = data.seurat.filt, 
  genes = c("Rgs5", "Myl9", "Igfbp7", "Notch3"), 
  celltype = "Endothelial cell")
```

### Cluster 10
```{r}
runFeatureplot(
  seurat.object = data.seurat.filt, 
  genes = c("Gria3", "Kcnma1", "Pvalb", "Slc24a2"), 
  celltype = "Neuron (possibly purkinje)")
```

### Cluster 11

```{r}
runFeatureplot(
  seurat.object = data.seurat.filt, 
  genes = c("Calml4", "Rbp1", "Sostdc1", "Igfbp2"), 
  celltype = "ependymal cells")
```

### Cluster 12
```{r}
runFeatureplot(
  seurat.object = data.seurat.filt, 
  genes = c("Dcn", "Pdgfra", "Apod", "Col1a1"), 
  celltype = "Fibroblast")
```

### Cluster 13
```{r}
runFeatureplot(
  seurat.object = data.seurat.filt, 
  genes = c("Unc5d", "Sln", "Trpc3", "Lingo2"), 
  celltype = "Purkinje and excitory neurons")
```



```{r}
# Save the cell type info to the metadata
data.seurat.filt@meta.data %<>% 
  mutate(
    celltype = case_when(
      RNA_snn_res.0.3 == "0" ~ "GranuleCells", 
      RNA_snn_res.0.3 == "1" ~ "GranuleCells2", 
      RNA_snn_res.0.3 == "2" ~ "BergmannGlia",
      RNA_snn_res.0.3 == "3" ~ "Microglia",
      RNA_snn_res.0.3 == "4" ~ "Oligodendrocyte",
      RNA_snn_res.0.3 == "5" ~ "EndothelialStalkCells",
      RNA_snn_res.0.3 == "6" ~ "Astrocytes",
      RNA_snn_res.0.3 == "7" ~ "low quality",
      RNA_snn_res.0.3 == "8" ~ "Oligodendrocyte2",
      RNA_snn_res.0.3 == "9" ~ "EndothelialCells",
      RNA_snn_res.0.3 == "10" ~ "NeuronPossiblePurkinjeCells",
      RNA_snn_res.0.3 == "11" ~ "EpendymalCells",
      RNA_snn_res.0.3 == "12" ~ "Fibroblast",
      RNA_snn_res.0.3 == "13" ~ "PurkinjeExcitoryNeurons"
      )
    )
```

# Pseudobulk
## prepare DGE object
```{r}
y <- Seurat2PB(data.seurat.filt, sample="orig.ident", cluster="celltype")

y$samples %<>% 
  as_tibble(rownames = "sample_cluster") %>% 
  left_join(meta) %>% 
  mutate(
    sample_cluster = str_remove(sample_cluster, "cluster"), 
    celltype = str_remove(sample_cluster, pattern = "S\\d_"),
    group = paste0(celltype, "_", age, "_", genotype) %>% as.factor(),
    age_geno = paste0(age, "_", genotype) %>% as_factor()
  ) %>% 
  column_to_rownames("sample_cluster")

colnames(y$counts) %<>% str_remove("cluster")
```

## lib sizes
```{r}
y$samples %>% 
  ggplot(
    aes(x = sample, y = lib.size, fill = age_geno, alpha = lib.size > 20000)
  ) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  facet_wrap(~celltype, scales = "free")
```

## filtering
### samples
Only retained cell type clusters per age if the cluster contained at least 20k reads, and had two samples per genotype. 

```{r}
# retain samples with more than 20k counts
keep.samples <- y$samples$lib.size > 2e4
y <- y[,keep.samples]

# check which samples only one sample per genotype age
y$samples %>% 
  ggplot(
    aes(x = sample, y = lib.size, fill = age_geno)
  ) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  facet_wrap(~celltype, scales = "free")

y$samples %>% 
  group_by(age, genotype, celltype) %>% 
  summarise(n = n()) %>% 
  dplyr::filter(n < 2) 

# get the sample clusers to keep (still has two per group)
keep.2 <- y$samples %>% 
  dplyr::filter(
    !(age == "3week" & cluster == "EndothelialCells"),
    !(age == "3week" & cluster == "EpendymalCells"),
    !(age == "3week" & cluster == "NeuronPossiblePurkinjeCells"),
    !(age == "3week" & cluster == "Fibroblast"),
    !(age == "3week" & cluster == "EndothelialStalkCells"),
    !(age == "7week" & cluster == "PurkinjeExcitoryNeurons"),
     !(cluster == "low quality"),
  ) %>% 
  rownames()

keep.2 = rownames(y$samples) %in% keep.2 

y <- y[,keep.2]

y$samples %>% 
  ggplot(
    aes(x = sample, y = lib.size, fill = age_geno)
  ) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  facet_wrap(~celltype, scales = "free")
```

### genes
```{r}
design <- model.matrix(~0 + group, data = y$samples %>% 
                         droplevels()) 
colnames(design) <- str_remove(colnames(design), 
                               "group")
```

```{r}
keep.genes <- filterByExpr(y, group=y$samples$cluster, design = design)
table(keep.genes)

y <- y[keep.genes, , keep=FALSE]

y <- normLibSizes(y, method = "TMM")

y$samples %>% 
  ggplot(
    aes(y = lib.size*norm.factors, x = cluster, colour = age_geno )
  ) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge()) +
  stat_compare_means() +
  facet_wrap(~cluster, scales = "free_x")

y %>% 
  cpm(log = T) %>% 
  as.data.frame() %>% 
  pivot_longer(cols = everything(), 
               names_to = "sample", values_to = "logCPM") %>% 
  left_join(y$samples %>% 
              mutate(sample = rownames(y$samples))) %>% 
  ggplot(
    aes(x = sample, y = logCPM)
    ) +
  geom_boxplot(
    aes(fill = age_geno),
    outlier.size = 0.5
               ) +
  facet_wrap(~cluster, scales = "free") +
  easy_rotate_x_labels(angle = -45) +
  labs(title = "Distribution of logCPM values per sample",
       x = "Sample", y = "log2 CPM")
```

## PCA
```{r}
y$counts %>% 
  cpm(log = TRUE) %>% 
  t %>% 
  prcomp %>% 
  autoplot(
    data = tibble(
      sample = rownames(.$x)
    ) %>% 
      left_join(y$samples %>% 
              mutate(sample = rownames(y$samples))
              ),
    colour = "cluster", 
    shape = "age_geno",
    alpha = 0.5,
    size = 6
  ) +
  theme(aspect.ratio = 1) +
  facet_wrap(~cluster, scales = "free") +
  scale_color_viridis_d(option = "turbo")
```

## DE analysis
### design matrix
```{r}
design <- model.matrix(~0 + group, data = y$samples %>% 
                         droplevels()) 
colnames(design) <- str_remove(colnames(design), 
                               "group")
```

### dispersion estimate
```{r}
y <- estimateDisp(y, design, robust = TRUE)
plotBCV(y)
```

## glmFit
```{r}
fit <- glmQLFit(y, design, robust = TRUE)
plotQLDisp(fit)
```

## contrasts matrix
```{r}
# Extract cell types shared by both Control and DLB
design_cols <- colnames(design)

design_df <- tibble(colname = colnames(design)) %>%
  separate(colname, into = c("celltype", "age", "genotype"), sep = "_") %>%
  unite("group_id", celltype, age, remove = FALSE)

# Filter to groups that contain both WT and Npc_hom
contrast_df <- 
  design_df %>%
  mutate(colname = paste0(group_id, "_", genotype)) %>% 
  group_by(group_id) %>% 
  pivot_wider(names_from = genotype, values_from = colname) %>% 
  mutate(contrast = paste0(Npc1, " - ", WT))

# Build contrast matrix
contrast_matrix <- makeContrasts(contrasts = contrast_df$contrast, levels = design)

# rename columns for readability
colnames(contrast_matrix) <- gsub("group", "", colnames(contrast_matrix))

```

## toptables
```{r}
toptables <- colnames(contrast_matrix) %>% 
  sapply(function(x) {
      glmQLFTest(fit, design, contrast=contrast_matrix[,x]) %>% 
      topTags(n = Inf, sort.by = "PValue") %>% 
      .$table %>% 
      as_tibble %>% 
      dplyr::select(
        gene, logCPM, logFC, PValue, FDR, everything()
      ) %>% 
      mutate(
        coef = x, 
        DE = FDR < 0.05
      )
  }, simplify = F)
```
## volcanoplots
```{r}
volcano_data <- 
  toptables %>%
  bind_rows() %>% 
  mutate(
  coef = str_remove(coef, pattern = "_Npc1.+_WT$"),
    DE = FDR < 0.05
  ) 

# Count number of DE up/down per celltype (coef)
volcano_labels <- volcano_data %>%
  dplyr::filter(DE) %>%
  mutate(direction = ifelse(logFC > 0, "Up", "Down")) %>%
  group_by(coef, direction) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(
    x = ifelse(direction == "Up", 3, -3),  # approximate x positions
    y = max(-log10(volcano_data$PValue), na.rm = TRUE) * 0.9,  # high y
    label = paste0(direction, ": ", n)
  )

# Plot
ggplot(volcano_data, aes(x = logFC, y = -log10(PValue))) +
  geom_point(aes(colour = DE)) +
  scale_color_manual(values = c("grey50", "red")) +
  facet_wrap(~coef) +
  geom_text(
    data = volcano_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 4,
    hjust = 0.5
  ) +
  labs(
    x = "log2 Fold Change",
    y = "-log10(P value)"
  ) 
```

## MD plots

```{r}
toptables %>% 
  bind_rows() %>% 
  ggplot(
    aes(x = logCPM, y = logFC)
  ) +
  geom_point(
    aes(colour = DE )
  ) +
  facet_wrap(~coef) +
  scale_color_manual(values = c("grey50", "red"))
```

# enrichmenmt andlysis 
## CD pathways
Here, I need to convert the human gene id to mouse id. Network genes were only retained if they were detected in any cell type and had a 1 to 1 mouse orthologue. 

```{r}
# import clusters
cluster.gprofiler.results <- 
  list.files(path = "data/results/adultchildsubtypes", 
             pattern = "gprofiler", recursive = TRUE, full.names = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/gprofiler_analysis_results.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/adultchildsubtypes", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))

network.genes.cluster <- 
  cluster.gprofiler.results$CD_1A_lyso_DLB %>% 
  dplyr::select(Cluster, genes) %>% 
  mutate(genes = str_split(genes, " ")) %>% 
  deframe() %>% 
  lapply(function(x) {
    x %>% 
      as_tibble %>% 
      set_colnames("human")
  })

# download the mouse2human id conversion
mouse_human_genes = read.csv("http://www.informatics.jax.org/downloads/reports/HOM_MouseHumanSequence.rpt",sep="\t")

convert_mouse_to_human <- function(gene_list){

# Filter to mouse genes in gene_list
  mouse_subset <- mouse_human_genes %>%
    dplyr::filter(Common.Organism.Name == "mouse, laboratory", Symbol %in% gene_list)

  # Join with human genes via DB.Class.Key
  result <- mouse_subset %>%
    left_join(
      mouse_human_genes %>%
        dplyr::filter(Common.Organism.Name == "human") %>%
        dplyr::select(DB.Class.Key, human_symbol = Symbol),
      by = "DB.Class.Key"
    ) %>%
    dplyr::select(mouse = Symbol, human = human_symbol) %>%
    dplyr::distinct() %>% 
    as_tibble

  return(result)
}

network.genes.cluster %<>% 
  lapply(function(x) {
    x %>% 
      left_join(convert_mouse_to_human(rownames(y))) %>% 
      na.omit() %>% 
      .$mouse %>% 
      unique()
  })

# for gene set testing, only retain clusters with more than 3 genes
network.genes.cluster <- network.genes.cluster[lapply(network.genes.cluster, length) > 3]

  
### obtain the genes in the subnetworks
node_metadata <- 
  list.files(path = "data/results/adultchildsubtypes", 
             pattern = "overlapping_nodes_metadata", recursive = TRUE, full.names = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/overlapping_nodes_metadata.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/adultchildsubtypes", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))

network.genes <- node_metadata$CD_1A_lyso_DLB

network.genes %<>% 
  dplyr::select(human = X, everything()) %>% 
  left_join(convert_mouse_to_human(rownames(y))) %>% 
  na.omit %>% 
  dplyr::select(gene_name = mouse, everything())
```




### ORA

#### ORA of pathways in the clusters
first, I perfored ORA analysis on the clusters themselves. this is to give biological context to the genes in each cluster. 
```{r}
ORA.clusters <- network.genes.cluster %>% 
  lapply(function(x) {
    clusterProfiler::enrichKEGG(
      gene = bitr(x, 
                  OrgDb = org.Mm.eg.db,
                  fromType="SYMBOL", toType="ENTREZID") %>% .$ENTREZID,
      keyType = "ncbi-geneid", 
      organism = "mmu"
    ) 
    })
```
```{r}
ORA.clusters %>% 
  lapply(function(x) {
    x %>% 
      as_tibble 
  }) %>% 
  bind_rows(.id = "cluster") %>% 
  ggplot(
    aes(x = -log10(p.adjust), y = Description, fill = cluster)
  ) +
  geom_col() +
  facet_wrap(~cluster, scales = "free_y")
```


#### number of DE genes per cluster

I then tested whether each of the netcoloc clusters are significantly overrepresented with DE genes within each cell type and age. 
```{r}
ora.DEG.in.clusters <- toptables %>%
  lapply(function(x) {
    genes <- x %>%
      dplyr::filter(DE == TRUE) %>%
      pull(gene)
    
    if (length(genes) < 3) return(NULL)  # skip if too few genes
    
    clusterProfiler::enricher(
      gene = genes,
      TERM2GENE = network.genes.cluster %>%
        enframe(name = "TERM", value = "GENE") %>%
        unnest(cols = "GENE"), 
      universe = x$gene
    )
  }) %>%
  purrr::compact()

```

#### pvals
```{r}
ora.DEG.in.clusters %>% 
  lapply(as_tibble) %>% 
  bind_rows(.id = "comparison") %>% 
  mutate(
    comparison = str_extract(comparison, 
                             pattern = "^([A-Za-z0-9]+)_([0-9]+week)")
  ) %>% 
  ggplot(
    aes(x = -log10(p.adjust), y = ID, fill = comparison)
  ) +
  geom_col(position = "dodge")
```

Clusters 438 and 439 are significantly enriched with DEGs in microglia. Both ages. 
```{r}
keep = toptables %>% 
  bind_rows() %>% 
  dplyr::filter(
    gene %in% network.genes.cluster$C439
  ) %>% 
  dplyr::filter(DE == T) %>% 
  .$gene

  # Build the FDR matrix (for asterisk overlay)
  FDR_mat <- toptables %>%
    bind_rows() %>%
    dplyr::filter(gene %in% keep) %>%
    dplyr::select(coef, gene, FDR) %>%
    mutate(
      coef = str_extract(coef, 
                             pattern = "^([A-Za-z0-9]+)_([0-9]+week)")
    ) %>% 
    pivot_wider(names_from = gene, values_from = FDR) %>%
    column_to_rownames("coef") %>%
    as.matrix()

Heatmap(
  matrix = toptables %>%
    bind_rows() %>%
    dplyr::filter(gene %in% keep) %>%
    dplyr::select(coef, gene, logFC) %>%
      mutate(
      coef = str_extract(coef, 
                             pattern = "^([A-Za-z0-9]+)_([0-9]+week)")
    ) %>% 
    pivot_wider(names_from = gene, values_from = logFC) %>%
    column_to_rownames("coef") %>%
    as.matrix(),
  name = "logFC",
  col = brewer_pal(palette = "RdBu")(9), 
  
  cell_fun =function(j, i, x, y, width, height, fill) {
      # j = column index, i = row index
      # x, y = coordinates of the center of the cell
      # width, height = size of the cell
      # fill = background color (not used here)

      # Check if the FDR value for this cell exists and is below 0.05
      if (!is.na(FDR_mat[i, j]) && FDR_mat[i, j] < 0.05) {
        # If yes, overlay a text label "*" on that cell
        grid.text("*", x, y, gp = gpar(fontsize = 15))
      }
    },
  column_names_gp = gpar(fontsize = 10),
  border = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE
)

```



```{r}
plot.cnet = function(toptable, plot.title, ORA.cluster.object) {
  # 1. Get toptable for the cluster you're plotting
top.genes <- toptable %>%
   dplyr::filter(PValue < 0.05) %>% 
  dplyr::select(gene, logFC)

# 2. Map gene symbols to ENTREZID
fc.entrez <- bitr(
  top.genes$gene,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
) %>%
  left_join(top.genes, by = c("SYMBOL" = "gene")) %>%
  dplyr::distinct(ENTREZID, .keep_all = TRUE) %>%
  dplyr::select(-SYMBOL) %>% 
  deframe()  # named vector: names are ENTREZID, values are logFC

cnetplot(
  x = setReadable(ORA.cluster.object, 
                  OrgDb = org.Mm.eg.db, keyType = "ENTREZID"),
  foldChange = fc.entrez,
  showCategory = 20,
  circular = FALSE, 
  color_category='grey20', 
  size_category = 0.5,
  color_item = "grey90", 
  colorEdge = F
) +
  scale_color_gradient2(name='logFC', low='blue4', high='firebrick4') + 
  ggtitle(plot.title)
  
}

```

#### Cluster 438 in microglia
```{r}
plot.cnet(
  toptable = toptables$`Microglia_3week_Npc1 - Microglia_3week_WT`, 
  plot.title = "Microglia 3 week", 
  ORA.cluster.object = ORA.clusters$C438)

plot.cnet(
  toptable = toptables$`Microglia_7week_Npc1 - Microglia_7week_WT`, 
  plot.title = "Microglia 7 week", 
  ORA.cluster.object = ORA.clusters$C438) +
  ggsave(
    "output/Npc1mousebrain_snRNAseq/figs/talk/C438microgliaCNET.png", 
    width = 20, height = 10, units = "cm", 
    scale = 1.75
  )
```

#### Cluster 439 in microglia
```{r}
plot.cnet(
  toptable = toptables$`Microglia_3week_Npc1 - Microglia_3week_WT`, 
  plot.title = "Microglia 3 week", 
  ORA.cluster.object = ORA.clusters$C439)

plot.cnet(
  toptable = toptables$`Microglia_7week_Npc1 - Microglia_7week_WT`, 
  plot.title = "Microglia 7 week", 
  ORA.cluster.object = ORA.clusters$C439)

```

#### Cluster 438 in Bergmann Glia
```{r}
plot.cnet(
  toptable = toptables$`BergmannGlia_3week_Npc1 - BergmannGlia_3week_WT`,
  plot.title = "Bergmann 3 week,", 
  ORA.cluster.object = ORA.clusters$C438)

plot.cnet(
  toptable = toptables$`BergmannGlia_3week_Npc1 - BergmannGlia_3week_WT`,
  plot.title = "Bergmann 7 week", 
  ORA.cluster.object = ORA.clusters$C438)
```

#### Cluster 439 in Granule cells

```{r}
plot.cnet(
  toptable = toptables$`GranuleCells_7week_Npc1 - GranuleCells_7week_WT`,
  plot.title = "Granule cells 7 week", 
  ORA.cluster.object = ORA.clusters$C439)
```

## fry gene set testing
```{r}
fry <- colnames(contrast_matrix) %>% 
  sapply(function(x) {
   cpm(y, log = T)  %>% 
      fry(
        index = network.genes.cluster,
        design = design, 
        contrast = contrast_matrix[,x], 
        sort = "directional"
      ) %>% 
      rownames_to_column("pathway") %>% 
      as_tibble() %>% 
      mutate(
        coef = x, 
        coef = str_extract(coef, "^([A-Za-z0-9]+)_([0-9]+week)")
        )
  }, simplify = FALSE)

fry
sig.clusters.fry <- fry %>% 
  bind_rows() %>% 
  dplyr::filter(
    FDR < 0.05 | FDR.Mixed < 0.05,
    NGenes > 1
  ) %>% 
  .$pathway %>% 
  unique
  
fry %>% 
  bind_rows() %>% 
  dplyr::filter(pathway %in% sig.clusters.fry) %>% 
  pivot_longer(
    cols = c(PValue, PValue.Mixed),
    names_to = "pval_type",
    values_to = "pval"
  ) %>%
  mutate(
    neglog10_p = -log10(pval),
    pval_type = recode(pval_type, "PValue" = "Directional", "PValue.Mixed" = "Mixed"),
    fdr = case_when(
      pval_type == "Directional" ~ FDR,
      pval_type == "Mixed" ~ FDR.Mixed
    ),
    sig = fdr < 0.05,
    alpha_val = if_else(sig, 1, 0.3),
    pathway = factor(pathway, levels = unique(pathway))
  ) %>% 
  ggplot(aes(x = pval_type,
             y = pathway, fill = neglog10_p, alpha = alpha_val)) +
  geom_tile() +
  facet_wrap(~coef) +
  scale_fill_viridis_c(option = "plasma", name = "-log10(P)") +
  scale_alpha(range = c(0.3, 1), guide = "none") +
  labs(
    x = "Hypothesis test",
    y = "Netcoloc cluster"
  ) 
```

```{r}
sig.clusters.fry <- fry %>% 
  bind_rows() %>% 
  dplyr::filter(
    grepl(coef, pattern = "^Microglia"),
    FDR.Mixed < 0.05
  ) %>% 
  .$pathway %>% 
  unique

toptables$`Microglia_3week_Npc1 - Microglia_3week_WT` %>% 
  dplyr::filter(
    gene %in% network.genes.cluster$C439
  ) %>% 
  dplyr::select(gene, logFC, FDR) %>%
  mutate(age = "3week") %>% 
  rbind(
    toptables$`Microglia_7week_Npc1 - Microglia_7week_WT` %>% 
      dplyr::filter(
        gene %in% network.genes.cluster$C439
      ) %>% 
      dplyr::select(gene, logFC, FDR) %>%
      mutate(age = "7week") 
  ) %>% 
  ggplot(
    aes(x = age, y = reorder(gene, logFC), fill = logFC)
  ) + 
  geom_tile() +
  geom_label(
    aes(label = signif(FDR,1)), 
    data = . %>% 
      dplyr::filter(FDR < 0.05)
    ) +
  scale_fill_distiller(palette  = "RdBu") +
  labs(
    y = "gene", 
    title = "Netcoloc cluster 439"
  )
```





# export data
```{r}
data.seurat.filt %>% 
  qsave("data/public data/PRJNA606815_NPC_mousebrain/R_objects/dataseurat.qs")

y %>% 
  qsave("data/public data/PRJNA606815_NPC_mousebrain/R_objects/pseudobulk_dgeObject.qs")

ORA.clusters %>% 
  qsave("data/public data/PRJNA606815_NPC_mousebrain/R_objects/ORA_clusters.qs")

ora.DEG.in.clusters %>% 
  qsave("data/public data/PRJNA606815_NPC_mousebrain/R_objects/ORA_DEGs_clusters.qs")

fry %>% 
  qsave("data/public data/PRJNA606815_NPC_mousebrain/R_objects/fry_results.qs")

toptables %>% 
  qsave("data/public data/PRJNA606815_NPC_mousebrain/R_objects/toptables.qs")
```









