---
title: "DLB bulk RNASeq"
author: "Karissa Barthelson"
date: "2026-06-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
```{r}
library(tidyverse)
library(magrittr)
library(readxl)

# vis
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
library(UpSetR)
library(pander)
library(ggeasy)
library(ggfortify)
library(ggpubr)
library(ggrepel)
library(grid)
library(gridExtra)
library(gridGraphics)
library(viridis)
library(tidytext)
library(pheatmap)
library(patchwork)

library(tidygraph)
library(ggraph)
library(ggforce)

# analysis
library(edgeR)
library(limma)

# annotations
library(org.Hs.eg.db)

# enrichemnt
library(msigdbr)
library(clusterProfiler)
library(gprofiler2)
library(ReactomePA)
library(DOSE)


theme_set(theme_light() +
            theme(plot.title = element_text(hjust = 0.5), 
                  strip.text = element_text(face = "bold", colour = "white"),
                  strip.background = element_rect(fill = "grey20")
            )
)

knitr::opts_chunk$set(
  message = F, echo = F, warning = F
)
```

# import data
Paper: https://academic.oup.com/brain/article/148/1/69/7698452#499652267
synapse ID: https://www.synapse.org/Synapse:syn59213647

raw counts were downloaded. 

```{r}
files = list.files("data/public data/DLB bulk RNAseq", full.names = T)

samples_paper = read_xlsx(files[1])

counts = read_delim(files[2])
genes = read_delim(files[3])
samples = read_delim(files[4])
```

# Tidy data and build the DGEList
```{r}
# counts matrix, genes as rows, samples as columns
counts_mat <- counts %>% 
  column_to_rownames("...1") %>% 
  as.matrix()

# gene annotation, matched to the count rows
geneAnno <- genes %>% 
  dplyr::select(-...1) %>% 
  dplyr::select(gene_id, seqnames, start, end, strand, 
                gene_type, gene_name,everything()) %>% 
  dplyr::filter(gene_id %in% rownames(counts_mat))


# obtain sample metadata.
# fix up colnames
samples_tidy <- 
  samples %>% 
  dplyr::rename(sample = ...1) %>% 
  mutate(
    type = case_when(TYPE == "LBD" ~ "DLB", TRUE ~ TYPE),
    sex = factor(sex_inferred, levels = c("female", "male"))
  ) %>% 
  dplyr::select(-TYPE) %>% 
  dplyr::select(sample, type, everything()) %>% 
  left_join(
    # need the batch, but this was only included in the Supp table from the paper
    samples_paper %>% 
      dplyr::select(NPID, Flowcell)
            ) %>% 
  mutate(
    batch = as.factor(Flowcell)
  ) %>% 
  dplyr::select(-Flowcell)
```

## gene filtering

Following what they did in the paper. 

Raw counts were normalized to adjust for sample library size differences using the edgeR CPM (counts per million) function.

Counts were filtered to remove mitochondrial genes and keep only expressed protein-coding genes. 

A gene is expressed if at least 70% of the smallest group size has a minimum CPM count of 1. 
Before filtering, there were 60 649 genes, and post-filtering retained 15 208 expressed protein-coding genes. 

Following filtering, the trimmed mean of values method was used to convert raw library sizes into effective ones.

```{r}
keep.genes.protein_coding = genes %>% 
  dplyr::filter(gene_type == "protein_coding") %>% 
  dplyr::filter(seqnames != "chrM")

x <- counts_mat %>% 
  DGEList(
    samples = tibble(sample = colnames(.)) %>% 
      left_join(samples_tidy, by = "sample") %>% 
      as_tibble(),
    genes = tibble(
      gene_id = rownames(.)) %>% 
      left_join(geneAnno) %>% 
      as_tibble
  ) %>% 
  calcNormFactors()

x$samples %<>% as_tibble
x$genes %<>% as_tibble
```

```{r}
# before filtering
L <- mean(x$samples$lib.size) * 1e-6
M <- median(x$samples$lib.size) * 1e-6
c(L, M)

# first filter by expression
dim(x)
keep.expr <-
  filterByExpr(
    x,
    group = x$samples$type, # by disease groups
    min.count = M, # min count of the median library size, which should keep genes with at least 1 CPM in 30% of the smallest group
    min.prop = 0.3 # at least 30% of the smallest group 
  )
x <- x[keep.expr, , keep.lib.sizes = FALSE]

# remove mt genes
removeMT <- x$genes$seqnames != "chrM"  # true when NOT MT
x <- x[removeMT,,keep.lib.sizes = FALSE]

keepProtein <- x$genes$gene_type == "protein_coding"
x <- x[keepProtein,,keep.lib.sizes = FALSE]
x <- calcNormFactors(x, method = "TMM")

dim(x)
```



# PCA
Not particularly informative. just a big blob
```{r}
x %>% 
  cpm(log = TRUE) %>% 
  t() %>% 
  prcomp() %>% 
  autoplot(
    data = tibble(sample = rownames(.$x)) %>% 
      left_join(x$samples, by = "sample"),
    colour = "type",
    shape = "sex",
    size = 4
  ) +
  theme(aspect.ratio = 1) +
  labs(colour = "Clinical diagnosis", shape = "Sex") +
  facet_wrap(~sex)
```

# DE analysis

```{r}
# create tyope_sex as per originbal analysis
x$samples %<>% 
  mutate(
    type_sex = paste0(type, "_", sex_inferred) %>% 
      factor(levels = c(
        "CONTROL_male",
        "CONTROL_female",
        "AD_male",
        "AD_female",
        "PA_male",
        "PA_female",
        "DLB_male",
        "DLB_female"
      ))
  ) 
```

```{r biomarker_expression}
biomarkers <- c("ENO2", "GFAP", "OLIG2", "CD34", "P2RY12")
for (i in biomarkers) {
  biomarker <- subset(genes, gene_name == i) # gene of interest 
  biomarker_counts <- subset(cpm(x, log = T), rownames(cpm(x, log = T)) %in% biomarker)
  biomarker_melt <- reshape2::melt(biomarker_counts) # reshape data 
  # rename columns to merge with metadata 
  names(biomarker_melt)[names(biomarker_melt) == "value"] <- i 
  # rename columns to merge with metadata 
  names(biomarker_melt)[names(biomarker_melt) == "Var2"] <- "NPID" 
  names(biomarker_melt)[names(biomarker_melt) == "Var1"] <- "gene_id"
  biomarker_melt$gene_id <- NULL
  assign(paste0(i),biomarker_melt)
}

# add gene expression values into one dataframe 
# put all data frames into list
df_list <- list(ENO2, OLIG2, CD34, P2RY12, GFAP)

cell_biomarker_lcpm = ENO2 %>% 
  full_join(OLIG2) %>% 
  full_join(CD34) %>% 
  full_join(P2RY12) %>% 
  full_join(GFAP)
```

# Scale data
rescaling a predictor in a regression has absolutely no effect on the magnitude of the relation being studied—the slope itself will not change its steepness, nor will the p-values or variance explained be changed. Rescaling is merely a means of communicating the nature of the regression line in different, hopefully more intuitive language.
```{r scale}
df <- merge(cell_biomarker_lcpm, x$samples, by = "NPID")

cols_to_scale <- c(
  "Race_numeric", "RIN", "Age", "PCT_CODING_BASES", "PCT_INTERGENIC_BASES",
  "PCT_INTRONIC_BASES", "APOE_E4_allele_count", "ENO2", "GFAP", "OLIG2",
  "CD34", "P2RY12", "Brain.wt"
)

info_with_scale <- cell_biomarker_lcpm %>% 
  left_join(x$samples, by = "NPID") %>% 
  mutate(
    Race_numeric = as.numeric(as.factor(Race)),
    APOE_E4_allele_count = str_count(APOE, "4")
  ) %>% 
  mutate(across(all_of(cols_to_scale), ~ as.numeric(scale(.))))

all.equal(x$samples$NPID, as.character(info_with_scale$NPID))
x$samples <- info_with_scale
```

```{r}
design <- model.matrix(
  ~ 0 + type_sex + batch + RIN + Age + PCT_CODING_BASES +
    PCT_INTERGENIC_BASES + PCT_INTRONIC_BASES + ENO2,
  data = x$samples
) %>%
  set_colnames(gsub("type_sex", "", colnames(.)))

form <- (
  ~ 0 +
      type_sex + 
      batch + 
      RIN +
      PCT_CODING_BASES +
      PCT_INTERGENIC_BASES + 
      PCT_INTRONIC_BASES +
      ENO2
)

contrasts <- makeContrasts(
  # disease against control, averaged over sex
  DLBvsControl = (DLB_male + DLB_female)/2 - (CONTROL_male + CONTROL_female)/2,
  ADvsControl  = (AD_male + AD_female)/2 - (CONTROL_male + CONTROL_female)/2,
  PAvsControl  = (PA_male + PA_female)/2 - (CONTROL_male + CONTROL_female)/2,
  DLBvsAD      = (DLB_male + DLB_female)/2 - (AD_male + AD_female)/2,
  DLBvsPA      = (DLB_male + DLB_female)/2 - (PA_male + PA_female)/2,
  ADvsPA       = (AD_male + AD_female)/2 - (PA_male + PA_female)/2,

  # disease against control within males
  male_DLBvsControl = DLB_male - CONTROL_male,
  male_ADvsControl  = AD_male - CONTROL_male,
  male_PAvsControl  = PA_male - CONTROL_male,

  # disease against control within females
  female_DLBvsControl = DLB_female - CONTROL_female,
  female_ADvsControl  = AD_female - CONTROL_female,
  female_PAvsControl  = PA_female - CONTROL_female,

  # effect of sex within each group
  sex_diff_Control = CONTROL_female - CONTROL_male,
  sex_diff_AD      = AD_female - AD_male,
  sex_diff_PA      = PA_female - PA_male,
  sex_diff_DLB     = DLB_female - DLB_male,

  # whether the disease effect differs between the sexes
  DLBvsControl_sex_interaction = (DLB_female - CONTROL_female) - (DLB_male - CONTROL_male),
  ADvsControl_sex_interaction  = (AD_female - CONTROL_female) - (AD_male - CONTROL_male),
  PAvsControl_sex_interaction  = (PA_female - CONTROL_female) - (PA_male - CONTROL_male),
  DLBvsAD_sex_interaction      = (DLB_female - AD_female) - (DLB_male - AD_male),
  DLBvsPA_sex_interaction      = (DLB_female - PA_female) - (DLB_male - PA_male),
  ADvsPA_sex_interaction       = (AD_female - PA_female) - (AD_male - PA_male),

  levels = colnames(design)
)
```

```{r}
voom_cov <-
  variancePartition::voomWithDreamWeights(
    counts = x$counts,
    formula = form,
    data = x$samples,
    BPPARAM = BiocParallel::SnowParam(cores),
    plot = TRUE
  )

fit <- lmFit(voom_cov, design)
vfit <- contrasts.fit(fit, contrasts = contrasts)
veBayesFit <- eBayes(vfit, trend = TRUE, robust=TRUE)

pval <- 0.05
# lfc.cutoff <- 0.25

sumTable <- 
  summary(decideTests(
    veBayesFit,  # object
    adjust.method = "BH", # by default the method = "separate"
    p.value = pval,
    #lfc = lfc.cutoff  # numeric, minimum absolute log2-fold change required
  ))

allComparisons <- colnames(contrasts)

toptables <- allComparisons %>%
  sapply(function(coef){
    topTable(
      veBayesFit,
      coef = coef,
      n = Inf,
      sort.by = "P",
      genelist = x$genes,
      confint = TRUE
    ) %>%
      as_tibble() %>%
      mutate(
        coef = coef,
        DE = adj.P.Val < 0.05 
      ) %>%
      dplyr::select(
        gene_name, logFC, AveExpr, P.Value, FDR = adj.P.Val, DE, everything()
      )
  }, simplify = FALSE)

toptables$DLBvsControl
```

# assess LysoDLB genes
```{r}
lysoDLB = read_csv("data/results/adultchildsubtypes v2/CD_1A_lyso_DLB/overlapping_nodes_metadata.csv") %>% 
  dplyr::rename("gene_name" = 1)

check.prop = function(netcoloc.nodes, toptable) {
  genes.detected = intersect(netcoloc.nodes$gene_name, x$genes$gene_name)
  numDE = intersect(toptable %>%  dplyr::filter(DE == T) %>% .$gene_name, 
                    genes.detected)
  length(numDE)/length(genes.detected) %>% 
  signif(2)
}
check.prop(netcoloc.nodes = lysoDLB, toptable = toptables$DLBvsControl)
check.prop(netcoloc.nodes = lysoDLB, toptable = toptables$female_DLBvsControl)
check.prop(netcoloc.nodes = lysoDLB, toptable = toptables$male_DLBvsControl)
```

```{r}
toptables$DLBvsControl %>% 
  dplyr::filter(
    gene_name %in% lysoDLB$gene_name, 
    FDR < 0.05
    ) 
  
```


```{r}
lysoDLB %>% 
  left_join(toptables$DLBvsControl) %>% 
  dplyr::select(gene_name, FDR, z_both, DE, logFC) %>% 
  na.omit %>% 
  ggplot(
    aes(
      x = z_both, y = -log10(FDR)
    )
  ) +
  geom_point(
    aes(colour = logFC), 
    size = 2
  )  +
  geom_label_repel(
    aes(label = gene_name), 
    data = . %>% 
      dplyr::filter(z_both > 15 & FDR < 0.05)
  ) +
  geom_hline(yintercept = -log10(0.05)) +
  annotate(
    geom = "text", 
    label = "FDR = 0.05", 
    y = 1.5, x = 58) +
  scale_alpha_manual(values = c(0.4, 1)) +
  scale_colour_continuous(palette = "RdBu") +
  scale_x_continuous(limits = c(0,60)) +
  labs(
    x = "Network dual proximity (z)"
  ) 
  ggsave("output/plots4publication/Lyso_DLB_DLB_bulk_genezvFDR.pdf", 
         width =6, height = 5)
```


# Gene set testing
## Netcoloc communities
```{r}
cluster.gprofiler.results <- 
  list.files(path = "data/results/adultchildsubtypes v2", 
             pattern = "gprofiler", recursive = TRUE, full.names = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/gprofiler_analysis_results.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/adultchildsubtypes v2", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))

network.genes.cluster.DLB <- 
  cluster.gprofiler.results$CD_1A_lyso_DLB %>% 
  dplyr::select(Cluster, genes) %>% 
  mutate(genes = str_split(genes, " ")) %>% 
  deframe()
```

```{r}
logCPM = cpm(x, log=T)
logCPM.gene_name = logCPM %>% 
  as.data.frame() %>% 
  rownames_to_column("gene_id") %>% 
  left_join(x$genes %>% 
              dplyr::select(gene_id, gene_name)
            ) %>% 
  distinct(gene_name, .keep_all = T) %>% 
  dplyr::select(-gene_id) %>% 
  column_to_rownames("gene_name")

remove(logCPM)

fry.DLB <- colnames(contrasts) %>% grep(pattern = "DLB", value = T) %>% 
  sapply(function(y) {
   logCPM.gene_name %>% 
      fry(
        index = network.genes.cluster.DLB,
        design = design, 
        contrast = contrasts[,y], 
        sort = "directional"
      ) %>% 
      rownames_to_column("pathway") %>% 
      as_tibble() %>% 
      mutate(
        coef = y)
  }, simplify = FALSE)

keep = fry.DLB$DLBvsControl %>% 
  dplyr::filter(FDR < 0.05) %>% .$pathway


res = cluster.gprofiler.results$CD_1A_lyso_DLB %>% 
  dplyr::filter(Cluster %in% keep) 

# cluster to gene edges, one row per membership
edges <- res %>%
  dplyr::select(Cluster, genes) %>%
  tidyr::separate_rows(genes, sep = " ") %>%
  dplyr::filter(genes != "") %>%
  dplyr::rename(from = Cluster, to = genes)

# nodes, with seed status. clusters are their own category.
seed_status <- lysoDLB %>%
  transmute(
    gene_name,
    seed = case_when(
      d1_seeds == 1 & d2_seeds == 1 ~ "both",
      d1_seeds == 1                 ~ "d1",
      d2_seeds == 1                 ~ "d2",
      TRUE                          ~ "none"
    )
  )


nodes <- bind_rows(
  res %>% distinct(name = Cluster) %>% mutate(node_type = "cluster"),
  edges %>% distinct(name = to) %>% mutate(node_type = "gene")
) %>%
  mutate(
    seed = case_when(
      node_type == "cluster"                  ~ "cluster",
      name %in% seed_status$both              ~ "both",
      name %in% seed_status$d1                ~ "d1",
      name %in% seed_status$d2                ~ "d2",
      TRUE                                    ~ "none"
    )
  )

library(igraph)

# edges, carrying the cluster id so it can drive grouping in Cytoscape
edge_tbl <- res %>%
  dplyr::select(Cluster, genes) %>%
  tidyr::separate_rows(genes, sep = " ") %>%
  dplyr::filter(genes != "") %>%
  dplyr::rename(from = Cluster, to = genes) %>%
  mutate(cluster = from)

# nodes, with seed status, type, label and the top pathway on cluster nodes
node_tbl <- bind_rows(
  res %>%
    transmute(
      name = Cluster,
      node_type = "cluster",
      Top.pathway = if_else(Top.pathway == Cluster, "unannotated", Top.pathway)
    ),
  edge_tbl %>% distinct(name = to) %>% mutate(node_type = "gene")
) %>%
  left_join(seed_status, by = c("name" = "gene_name")) %>%
  mutate(
    seed = case_when(
      node_type == "cluster" ~ "cluster",
      is.na(seed)            ~ "none",
      TRUE                   ~ seed
    ),
    label = name
  ) %>% 
  left_join(
    toptables$DLBvsControl %>% 
      dplyr::select(name = gene_name, logFC, FDR)
  )

g <- graph_from_data_frame(d = edge_tbl, vertices = node_tbl, directed = FALSE)

write_graph(g, "output/lysoDLB_network.graphml", format = "graphml")
```




<!-- # assess LysoAD genes -->
<!-- ```{r} -->
<!-- lysoAD = read_csv("data/results/adultchildsubtypes v2/CD_1A_lyso_AD/overlapping_nodes_metadata.csv") %>%  -->
<!--   dplyr::rename("gene_name" = 1) -->

<!-- # number of LysoDLB genes detected in the bulk RNAseq exp -->
<!-- lysoAD.genes.detected = intersect(lysoAD$gene_name, x$genes$gene_name) -->
<!-- length(lysoAD.genes.detected) -->

<!-- # number of lysoDLB genes DE -->
<!-- lysoAD.genes.DE.AD = intersect( -->
<!--   toptables$ADvsControl %>%  -->
<!--     dplyr::filter(DE == T) %>%  -->
<!--     .$gene_name,  -->
<!--   lysoAD.genes.detected -->
<!--   )  -->
<!-- length(lysoAD.genes.DE.AD) -->

<!-- # prop of lysoDLB genes DE -->
<!-- # ~ 37% -->
<!-- length(lysoAD.genes.DE.AD)/length(lysoAD.genes.detected) %>%  -->
<!--   signif(1) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- lysoAD %>%  -->
<!--   left_join(toptables$ADvsControl) %>%  -->
<!--   ggplot( -->
<!--     aes( -->
<!--       x = z_both, y = -log10(FDR) -->
<!--     ) -->
<!--   ) + -->
<!--   geom_point( -->
<!--     aes(colour = DE) -->
<!--   )  -->
<!-- ``` -->


# export data for pub
```{r}
lysoDLB %>% 
  left_join(toptables$DLBvsControl) %>% 
  dplyr::select(gene_name, FDR, z_both, DE, logFC) %>% 
  na.omit %>% 
  write_csv("output/bulkseq/lysoDLB_genes.csv")
```


# Figs
```{r}
lysoDLB %>% 
  left_join(toptables$DLBvsControl) %>% 
  dplyr::select(gene_name, FDR, z_both, DE, logFC) %>% 
  na.omit %>% 
  ggplot(
    aes(
      x = z_both, y = -log10(FDR)
    )
  ) +
  geom_point(
    aes(colour = logFC), 
    size = 2
  )  +
  geom_label_repel(
    aes(label = gene_name), 
    data = . %>% 
      dplyr::filter(z_both > 15 & FDR < 0.05)
  ) +
  geom_hline(yintercept = -log10(0.05)) +
  scale_alpha_manual(values = c(0.4, 1)) +
  scale_colour_continuous(palette = "RdBu") +
  scale_x_continuous(limits = c(0,55)) +
  labs(
    x = "Network dual proximity (z)"
  ) +

fry.DLB$DLBvsControl %>% 
  ggplot(
    aes(
      x = -log10(PValue), 
      y = reorder(pathway, -PValue), 
      fill = FDR < 0.05
      )
  ) +
  geom_col() +
  scale_fill_manual(values = c("grey50", "red")) +
  labs(
    y = "Netcoloc community"
  ) &
  theme(
    text = element_text(size = 18)
  ) 

  ggsave("output/plots4publication/Lyso_DLB_DLB_bulk_genezvFDR.pdf", 
         width =13, height = 6)
```

