---
title: "06_ALS_FTD_scRNAseq_v1"
author: "Karissa Barthelson"
date: "2025-12-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r}

library(tidyverse)
library(forcats)
library(magrittr)
library(readxl)
library(qs) # for fast save and reading of objects into memory (alternative to rds)
library(org.Hs.eg.db)

# vis
library(pander)
library(ggpubr)
```

# introduction

here, i will assess the differential expression of network genes in the ALS/FTD dataset from [Pineda et al. 2024 (cell)](https://www.cell.com/cell/fulltext/S0092-8674(24)00234-4)

In the interest of time and effort, I'm just going to use the DE genes they defined. 

## data import
```{r}
files <- list.files("data/public data/ALS_FTD_scRNAseq/20251205", 
           full.names = T)

files %>% 
  grep(pattern = ".+20251205/\\d", value = T) %>% 
  .[1] %>% 
  excel_sheets() %>% 
  map_dfr(
    read_excel,
    path = files %>% 
      grep(pattern = ".+20251205/\\d", value = T) %>% 
      .[1] 
      )

read_datas <- function(file_path) {
  
  # Get all sheet names in the current Excel file
  sheet_names <- excel_sheets(file_path)
  print(sheet_names)
  
  # Read all sheets using the sheet names as input
  sheet_list <- sheet_names %>%
    set_names(sheet_names) %>% # Use the sheet names to name the list elements
    map(~ read_excel(
      path = file_path,
      sheet = .x
  ))
  
  
}

toptables <- files %>%
  # the sheets which start with 6 are the cell-type specific DE gen4e
  grep(pattern = ".+20251205/6", value = T) %>%
  set_names(basename(.)) %>%  # Optional: Name the outer list by file name
  map(read_datas)


nodes = readRDS("data/R_objects/network/nodes.rds")
### obtain the genes in the subnetworks
node_metadata <- 
  list.files(
    path = "data/results/adultchildsubtypes", 
    pattern = "overlapping_nodes_metadata", 
    recursive = TRUE, full.names = TRUE
  ) %>% 
  set_names(
    str_replace(
      ., 
      pattern = "/overlapping_nodes_metadata.csv",
      replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/adultchildsubtypes", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))
```

```{r}
# tidy up toptables names
names(toptables) %<>% 
  str_remove("6_.+deg_res_") %>% 
  str_remove(".xlsx")

# add a column which specifies the cell type and disease group
toptables %<>% 
# Outer map2: .x = inner list, .y = disease_group name
  map2(.y = names(.), .f = ~ {
    disease_name <- .y # Capture the disease group name explicitly

    # Inner map2: .x = tibble, .y = celltype name
    .x %>%
      map2(.y = names(.x), .f = ~ {
        .x %>%
          mutate(
            celltype = .y,          # .y is the celltype name (e.g., Vasc.Mural.Pericyte)
            disease_group = disease_name # Use the captured disease_name (e.g., BA9_C9FTLD)
          )
      })
  }) 
```

## assess differential expression of network genes
I now will assess the expression of the network genes in ALS nad FTLD, (both familial and sporadic forms)

```{r}
toptables %>% 
  list_c() %>% 
  list_rbind() %>% 
  dplyr::filter(
    Gene %in% node_metadata$CD_1A_lyso_ALS$X,
    padj < 0.05,
    grepl(disease_group, pattern = "ALS")
    ) %>% 
  group_by(disease_group, celltype) %>% 
  summarise(n = n()) %>% 
  ungroup %>% 
  tidyHeatmap::heatmap(
    .row = celltype, 
    .column = disease_group,
    .value = n, 
   
    cluster_columns = FALSE
  ) 
```

```{r}
toptables %>% 
  list_c() %>% 
  list_rbind() %>% 
  dplyr::filter(
    Gene %in% node_metadata$CD_1A_lyso_FTLD$X,
    padj < 0.05,
    grepl(disease_group, pattern = "FTLD")
    ) %>% 
  group_by(disease_group, celltype) %>% 
  summarise(n = n()) %>% 
  ungroup %>% 
  tidyHeatmap::heatmap(
    .row = celltype, 
    .column = disease_group,
    .value = n, 
   
    cluster_columns = FALSE, 
    cluster_rows = F
  ) 
```

```{r}

```








