---
title: "CDsubtypes"
author: "Karissa Barthelson"
date: "2025-05-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
```{r}
knitr::opts_chunk$set(
  message = F, echo = T, warning = F
)
```

```{r}
library(tidyverse)
library(magrittr)
library(readxl)

# vis
library(ComplexUpset)
library(UpSetR)
library(pander)
library(ggeasy)
library(ggfortify)
library(ggpubr)
library(ggrepel)
library(plotly)
library(tidyHeatmap)
library(grid)
library(gridExtra)
library(gridGraphics)
library(viridis)
library(tidytext)

# annotations
library(org.Hs.eg.db)
library(ndexr)

# enrichemnt
library(clusterProfiler)
library(gprofiler2)
library(ReactomePA)

theme_set(theme_light() +
            theme(plot.title = element_text(hjust = 0.5), 
                  strip.text = element_text(face = "bold"),
                  strip.background = element_rect(fill = "grey20")
            )
)

knitr::opts_chunk$set(
  message = F, echo = F, warning = F
)
```

# import 
## nodes

```{r}
### obtain the genes in the subnetworks
node_metadata <- 
  list.files(
    path = "data/results/adultchildsubtypes", 
    pattern = "overlapping_nodes_metadata", 
    recursive = TRUE, full.names = TRUE
  ) %>% 
  set_names(
    str_replace(
      ., 
      pattern = "/overlapping_nodes_metadata.csv",
      replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/adultchildsubtypes", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))
```
## gene sets

```{r}
genesets <- list.files(
  path = "data/genesets/adultchildsubtypes",
  pattern = ".csv",
  full.names = TRUE
) %>% 
  sapply(function(x) {
    x %>% 
      read_csv() %>% 
      .$GENE
  }, simplify = F) %>% 
  setNames(names(.) %>% 
             str_remove(pattern = "data/genesets/adultchildsubtypes/") %>% 
             str_remove(pattern = ".csv"))

# get these for help later
adult = c("AD", "PD", "ALS", "FTLD", "DLB")
childhood = grep(names(genesets), pattern = "^CD", value = TRUE)
```

tony curated these genesets from clingen. here is the number of genes per gene set.  

```{r}
genesets %>% 
  unlist %>% 
  as.data.frame(col.names = "gene") %>% 
  rownames_to_column("geneset") %>%
  set_colnames(c("geneset", "gene")) %>% 
  mutate(
    geneset = str_remove(geneset, pattern = "\\d{1,3}$"), 
    geneset = fct_infreq(geneset), # Reorder by frequency, 
    type = case_when(
      grepl(geneset, pattern = "^CD") ~ "childhood", 
      TRUE ~ "adult"
    )
   
   ) %>% 
  ggplot(
    aes(x = geneset)
  ) +
  geom_bar() +
  easy_rotate_x_labels(angle = -45) +
  labs(
    title = "Number of genes per gene set", 
    y = "# genes"
  ) +
  facet_wrap(~type, nrow = 2, scales = "free_x", space = "free_x") +
  theme(
    plot.margin = margin(r = 2, unit = "cm")
  )
```


## overlap of genes across gene sets
There is minimal overlap of genes across the adult and childhod dementia gene sets. Just 1 or 2 in each comparison

the 10 gene overlap is between ALS and FTLD

```{r}
genesets %>% 
  fromList %>% 
  UpSetR::upset(nsets = length(.))
```

# over-representation analysis of gene sets

Here I will perform pathway/enrichment analysis on these gene sets to see what pathways are enriched, and how much overlap there is. 

the background gene set will be all genes found in the PCNet, as these are the ones I am analysing. The node info was obtained from NDex. 
```{r}
# ~~ only run this on the first time ~~#
#
# login to the NDEx server - run this in console
# ndexcon <- ndex_connect(
#  "karissa.barthelson@flinders.edu.au", "password - capital and 21")

# search the networks for 'pcNet v1.3'
# pcnet1.3 = "4de852d9-9908-11e9-bcaf-0ac135e8bacf"

# obtain the nodes
# nodes = ndex_network_get_aspect(ndexcon, pcnet1.3, 'nodes') %>% .$name

# save this out for reimporting later
# saveRDS(nodes, "data/R_objects/network/nodes.rds")

nodes <- readRDS("data/R_objects/network/nodes.rds")
```

<!-- ## GO terms -->

<!-- Here, i performed over-represetation analysis (hypergeometric test) of GO terms (all ontologies) of each gene set using clusterprofiler -->

<!-- ```{r} -->
<!-- runGO = function(geneset) { -->
<!--   enrichGO( -->
<!--     gene = geneset,  -->
<!--     universe = nodes,  -->
<!--     OrgDb = org.Hs.eg.db,  -->
<!--     keyType = "SYMBOL",  -->
<!--     ont = "ALL" -->
<!--   ) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r} -->
<!-- GO <- genesets %>%  -->
<!--   sapply(runGO, simplify = F) %>%  -->
<!--   sapply(as_tibble, simplify = F) -->

<!-- GO %>%  -->
<!--   lapply(function(x) { -->
<!--     x$Description -->
<!--   }) %>%  -->
<!--   fromList() %>%  -->
<!--   UpSetR::upset( -->
<!--     nsets = length(.),  -->
<!--     order.by = "freq" -->
<!--   ) -->

<!-- GO.terms.pergeneset <- GO %>%  -->
<!--   lapply(function(x) { -->
<!--     x$Description -->
<!--   })  -->
<!-- ``` -->

## KEGG pathways

```{r}
runKEGG = function(geneset) {

  uv =   bitr(nodes, 
                fromType = "SYMBOL", toType = "ENTREZID", 
                OrgDb ="org.Hs.eg.db") %>% 
    .$ENTREZID
  
  bitr(geneset, 
       fromType = "SYMBOL", toType = "ENTREZID", 
     OrgDb ="org.Hs.eg.db") %>% 
  .$ENTREZID %>%
  enrichKEGG(
    organism = 'hsa',
    universe = uv,
    pvalueCutoff = 0.05) %>% 
  as_tibble 
  
}
```

```{r}
kegg <- genesets %>% 
  sapply(function(x) {runKEGG(x)}, simplify = F) 

kegg %>% 
  lapply(function(x) {x$Description}) %>% 
  fromList() %>% 
  UpSetR::upset()
```


<!-- ## reactome pathways  -->
<!-- ```{r} -->
<!-- runReactome = function(geneset, genesetname) { -->

<!--   enrichPathway( -->
<!--     gene = bitr(geneset,  -->
<!--                 fromType = "SYMBOL", toType = "ENTREZID",  -->
<!--                 OrgDb ="org.Hs.eg.db") %>%  -->
<!--       .$ENTREZID, -->
<!--     universe = bitr(nodes, -->
<!--                     fromType = "SYMBOL", toType = "ENTREZID", -->
<!--                     OrgDb= "org.Hs.eg.db") %>%  -->
<!--       .$ENTREZID, # all genes in pcnet -->
<!--     readable = TRUE,  -->
<!--     pvalueCutoff = 0.05 -->
<!--   ) %>%  -->
<!--   as_tibble  -->
<!-- } -->

<!-- # perform hypergeometic test for significant overrep of Reactome db -->
<!-- reactome <-  -->
<!--   genesets %>%  -->
<!--   sapply(runReactome, "reactome", simplify = F) -->

<!-- reactome %>%  -->
<!--   lapply(function(x) {x$Description}) %>%  -->
<!--   fromList() %>%  -->
<!--   UpSetR::upset( -->
<!--     order.by = "freq",  -->
<!--     nsets = length(.) -->
<!--   ) -->
<!-- ``` -->

# Is there significant colocalisation?

Netcoloc was performed and below shows the expected vs observed ratio of the genes identified by dual network propogation. 

```{r}
# Define path
base_path <- "data/results/adultchildsubtypes/"

# Read enrichment files
enrichment_dfs <- 
  list.files(base_path, recursive = TRUE, full.names = TRUE) %>%
  grep(pattern = "network_colocalization_enrichment", 
       value = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/network_colocalization_enrichment.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0(base_path, "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y)) %>% 
  bind_rows() %>%
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>%
  pivot_longer(
    cols = c(observed_overlap, expected_overlap_mean),
    names_to = "type",
    values_to = "value"
  ) %>%
  mutate(
    type = recode(type,
                  "observed_overlap" = "observed",
                  "expected_overlap_mean" = "expected")
  )
```

## plot observed v expected ratio

```{r}
enrichment_dfs %>% 
  bind_rows() %>% 
  as_tibble %>% 
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>% 
 mutate(
    # use the childhood vector to extract the CD seed gene name
    CD = purrr::map_chr(comparison, ~ {
      match <- childhood[startsWith(.x, childhood)]
      if (length(match) > 0) match[1] else NA_character_
    }),
    CD = case_when(
      grepl(CD, pattern = "1A") ~ "Lysosomal disorders", 
      grepl(CD, pattern = "1B") ~ "Other disorders of lipid metabolism & transport", 
      grepl(CD, pattern = "1C") ~ "Disorders of amino acid & other organic acid metabolism", 
      grepl(CD, pattern = "1D") ~ "Vitamin-responsive inborn errors of metabolism", 
      grepl(CD, pattern = "1E") ~ "Disorders of mineral absorption & transport", 
      grepl(CD, pattern = "1F") ~ "Peroxisomal diseases", 
      grepl(CD, pattern = "1G") ~ "Mitochondrial disorders", 
      grepl(CD, pattern = "1H") ~ "Other inborn errors of metabolism", 
      grepl(CD, pattern = "2_") ~ "Leukodystrophies not notherwise categorised", 
      grepl(CD, pattern = "3_") ~ "Neurodegeneration with brain iron accumulation", 
      grepl(CD, pattern = "4_") ~ "Diseases not otherwise categorised"
    ) %>% 
      str_wrap(width = 22),
    adult = str_remove(comparison, pattern = "CD_.+_.+_"),
    adult = case_when(
      adult == "malform" ~ "brain_malform", 
      TRUE ~ adult
    ),
    .after = obs_exp, 
    FDR = p.adjust(empirical_p, "fdr")
  ) %>% 
  dplyr::filter(type == "observed") %>% 
  ggplot(
    aes(x = obs_exp, y = reorder(adult, obs_exp))
  ) +
  geom_col(
    aes(fill = -log10(FDR), alpha = FDR < 0.05), 
    colour = "grey50"
  ) +
  scale_fill_viridis_c(option = "magma") +
  scale_alpha_manual(values = c(0.1, 1)) +
  labs(
    x = "observed / expected", 
    y = "Disease"
  ) +
  facet_wrap(~CD, ncol = 3) 
  # ggsave(
  #   "output/figs/paper/figure2.svg", 
  #   width = 14, height = 12, units = "cm", scale = 1.5
  #   
  # )
```

# Are the subnetworks biologically meaningful?

Here, I will only consider the subnetworks which are significant

```{r}
sig.networks <- enrichment_dfs %>% 
  bind_rows() %>% 
  as_tibble %>% 
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>% 
  mutate(
    FDR = p.adjust(empirical_p, "fdr")
  ) %>% 
  dplyr::filter(FDR < 0.05) %>% 
  .$comparison %>% 
  unique

top.sig.networks <- 
  enrichment_dfs %>% 
  bind_rows() %>% 
  as_tibble %>% 
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>% 
  mutate(
    CD = purrr::map_chr(comparison, ~ {
      match <- childhood[startsWith(.x, childhood)]
      if (length(match) > 0) match[1] else NA_character_
    }),
    FDR = p.adjust(empirical_p, "fdr")
  ) %>% 
  group_by(CD) %>% 
  slice_max(order_by = obs_exp, n = 1) %>%
  dplyr::filter(FDR < 0.05) %>% 
  .$comparison %>% 
  unique
```


## check the genes in the subnetwork are enriched with mouse brain phenotypes

I tested the enrichment of genes associated with mouse brain/neural phenotypes in the NetColoc subnetwork as a *whole*. I also included here a negative control phenotype -  abnormal digit morphology. This is not expected to be associated with the subnetworks. 

```{r}
mouse.variant.wholesubnetworks <- list.files(
    "data/results/adultchildsubtypes", recursive = TRUE, full.names = TRUE) %>% 
  grep(pattern = "/mousevariant_KO_enrichment.csv", 
       value = TRUE) %>% 
  set_names(
    str_replace(., 
               pattern = "/mousevariant_KO_enrichment.csv",
               replacement = ""
  ) %>% 
    str_remove(
      pattern = paste0("data/results/adultchildsubtypes", "/")
    )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y)) %>% 
  bind_rows() %>% 
  dplyr::filter(comparison %in% sig.networks) %>%
  as_tibble() %>%
  group_by(comparison) %>%
  mutate(
        # use the childhood vector to extract the CD seed gene name
    CD = purrr::map_chr(comparison, ~ {
      match <- childhood[startsWith(.x, childhood)]
      if (length(match) > 0) match[1] else NA_character_
    }),

    adult = str_remove(comparison, pattern = "CD_.+_.+_"),
    adult = case_when(
      adult == "malform" ~ "brain_malform", 
      TRUE ~ adult
    ),
    FDR = p.adjust(OR_p),
    .after = OR_p
  )
# a function for plotting by CD subgtype:
plot.mouse.variant.subtype.forestplot.by.CD.subtype <- function(data, CD_subtype) {
  data %>%
    dplyr::filter(CD == CD_subtype) %>% 
    group_by(comparison, MPO_term_type) %>%
    slice_min(order_by = OR_p, n = 10) %>%
    ungroup() %>%
    mutate(
      MP_description = reorder_within(MP_description, log_OR, comparison)
    ) %>%
    ggplot(
      aes(x = log_OR, y = MP_description)
    ) +
    geom_errorbarh(
      aes(xmin = log_OR_CI_lower, xmax = log_OR_CI_upper),
      height = 0.2, color = "black"
    ) +
    geom_point(
      aes(size = num_genes_in_term, color = MPO_term_type),
      alpha = 0.6
    ) +
    geom_vline(xintercept = 0, lty = 2, color = "grey50") +
    geom_label(
      aes(label = signif(OR_p, 2)),
      nudge_y = 0.4
    ) +
    scale_y_reordered() +  # This removes the `___comparison` suffix added by reorder_within()
    labs(
      x = "log(OR) ± 95% CI",
      y = NULL,
      size = "# Genes in MPO term",
      color = "Term Type"
    ) +
  facet_wrap(~comparison, nrow = 1
             #scales = "free"
             )
}

mouse.variants.plots <- childhood %>% 
  lapply(
    function(x) {
      plot.mouse.variant.subtype.forestplot.by.CD.subtype(
        CD_subtype = x, data = mouse.variant.wholesubnetworks)
    })

mouse.variants.plots[1]
mouse.variants.plots[2]
mouse.variants.plots[3]
mouse.variants.plots[4]
mouse.variants.plots[5]
mouse.variants.plots[6]
mouse.variants.plots[7]
mouse.variants.plots[8]
```

## are the genes in the subnetwork human brain expressed?

I next want to see whether the genes in the subnetwork are expressed in the brain. To do this, I will assess it in the GTEx dataset. I will use the GTEx v10 data, which is the most recent version. 

```{r}
# plot number of genes per subnetworks
node_metadata %>% 
  bind_rows() %>% 
  dplyr::filter(comparison %in% sig.networks) %>% 
  dplyr::rename(gene = X) %>% 
  as_tibble() %>%
  mutate(
    seed_type = case_when(
      d1_seeds == 1 & d2_seeds == 1 ~ "both",
      d1_seeds == 1 & d2_seeds == 0 ~ "D1 - childhood",
      d1_seeds == 0 & d2_seeds == 1 ~ "D2 - adult",
      d1_seeds == 0 & d2_seeds == 0 ~ "network proximal"
    )
  ) %>% 
  ggplot(
    aes(y = comparison, fill = seed_type)
  ) +
  geom_bar() +
  scale_fill_manual(values = c(
    "#7570b3", "#1b9e77", "#d95f02", "grey70"
  )) +
  labs(
    x = "Number of genes in subnetwork", 
    y = "Seed gene sets", 
    title = "Number of genes in the subnetworks"
  ) 
```

## assessing which genes are driving the DLB/Lyso
about half of the DLB_lyso gene set is unique to that comparsion
```{r}
list(
  DLB_lyso_genes = node_metadata$CD_1A_lyso_DLB$X, 
  AD_lyso_genes = node_metadata$CD_1A_lyso_AD$X,
  PD_lyso_genes = node_metadata$CD_1A_lyso_PD$X,
  ALS_lyso_genes = node_metadata$CD_1A_lyso_ALS$X,
  FTLD_lyso_genes = node_metadata$CD_1A_lyso_FTLD$X) %>% 
  fromList() %>% 
  upset()
```

```{r}
# what genes are the unique ones
unique_genes <- node_metadata$CD_1A_lyso_DLB %>% 
  dplyr::filter(!(X %in% node_metadata$CD_1A_lyso_AD$X)) %>% 
  dplyr::filter(!(X %in% node_metadata$CD_1A_lyso_PD$X)) %>% 
  dplyr::filter(!(X %in% node_metadata$CD_1A_lyso_ALS$X)) %>% 
  dplyr::filter(!(X %in% node_metadata$CD_1A_lyso_FTLD$X)) %>% 
  .$X

node_metadata$CD_1A_lyso_DLB %>% 
  dplyr::filter(X %in% unique_genes) %>% 
  .$X %>% 
  enrichGO(OrgDb = org.Hs.eg.db, keyType = "SYMBOL") %>% 
  as_tibble 
```




### obtain GTEx expression data
This file was downloaded from here :
https://www.gtexportal.org/home/downloads/adult-gtex/bulk_tissue_expression

The file contains the median gene-level TPM by tissue. 

```{r}
# this was downloaded from gtex websiote
# version 10 
gtex <- read.delim("data/gtex/GTEx_Analysis_v10_RNASeQCv2.4.2_gene_median_tpm.gct.gz", skip =2)

gtex %<>% 
  pivot_longer(names_to = "tissue", 
               cols = 3:56, 
               values_to = "tpm"
  ) %>% 
  dplyr::select(
    gene = Description, tissue, tpm
  ) %>% 
  mutate(
    logtpm = log(tpm))
```

```{r}
# define a threshold for expressed or not  
gtex.threshold = 0.5

# Fix: safe join and summarize per gene per tissue
node_metadata %>%
  bind_rows(.id = "comparison") %>%  # Add comparison info
  dplyr::rename(gene = X) %>%
  as_tibble() %>%
  dplyr::select(comparison, gene) %>% 
  left_join(
    gtex %>% 
      dplyr::filter(grepl("Brain", tissue)) %>% 
      dplyr::distinct(gene, tissue, logtpm)
  ) %>% 
  mutate(
    expressed = ifelse(logtpm > gtex.threshold, 1, 0)
    ) %>% 
  group_by(tissue, comparison) %>%
  dplyr::filter(
    grepl("Brain", tissue), 
    comparison %in% sig.networks
    ) %>% 
  mutate(
    n_genes = n(),
    n_expressed = sum(expressed),
    proportion_expressed = n_expressed / n_genes, 
           # use the childhood vector to extract the CD seed gene name
    CD = purrr::map_chr(comparison, ~ {
      match <- childhood[startsWith(.x, childhood)]
      if (length(match) > 0) match[1] else NA_character_
    }), 
    
    adult = str_remove(comparison, pattern = "CD_.+_.+_"),
  ) %>% 
  ungroup %>% 
  group_by(CD) %>% 
   tidyHeatmap::heatmap(
    .row = tissue,
    .column = comparison,
    .value = proportion_expressed, 
    
    # turn off clustering
    cluster_rows = FALSE, 
    column_dend_height = unit(2, "cm"),
    
   ) %>% 
  annotation_tile(CD)
```

## what cell types are they primarily expressed in human brains?

I have requested access for a human snRNAseq dataset from ROSMAP [Mathys et al.](https://www.nature.com/articles/s41586-024-07606-7) synapse ID 	
[syn52293442](https://www.synapse.org/Synapse:syn52293442)

Also getting a dementia with lewy body dataset from GEO: 
https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE178146
https://pmc.ncbi.nlm.nih.gov/articles/PMC8357687/

A LSD dataset: NPC1 single cell mouse brain:
sra: PRJNA606815
https://pmc.ncbi.nlm.nih.gov/articles/PMC7432835/
only n = 2 tho



# what pathways are in the subnetwork

To interrogate the biological context of the sub networks, I will perform some enrichment analyses. I will tackle this using two approaches: 

1- Perform hypergeometric tests using KEGG pathways on the entire subnetwork (background is the list of genes in PCNet)
2- use the annotations of the clusters within the subnetworks (performed using gprofiler) of KEGG, Reactome and GO terms. 

## 1- whole subnetowrks
```{r}
enrichment.wholesubnet.kegg <- 
  node_metadata[sig.networks] %>% 
  lapply(magrittr::extract2, "X") %>% 
  sapply(runKEGG, simplify = F)

enrichment.wholesubnet.kegg %>% 
  lapply(function(x) {
    x$Description
  }) %>% 
  fromList() %>% 
  UpSetR::upset(
    nsets = length(.), 
    order.by = "freq"
  )
```


### using gprofiler
```{r}
# this was provided by William
# the latest canonical set from MSigDB
upload_GMT_file(gmtfile = "data/genesets/c2.cp.v2025.1.Hs.symbols.gmt")

gprofiler.wholenet.genesetsv1 <- node_metadata[sig.networks] %>% 
  lapply(magrittr::extract2, "X") %>% 
  sapply(function(x) {
    gost(query = x, 
         organism = "gp__XoLn_Cg43_BEg", 
         ordered_query = FALSE, 
         significant = TRUE, # only return sig results
         user_threshold = 0.05,
         correction_method = "fdr",
         domain_scope = "annotated", # only analyse genes with the annotation. 
         evcodes = TRUE,
         custom_bg = nodes) %>% 
      .$result %>% 
      as_tibble
  }, simplify = F)

gprofiler.wholenet.genesetsv1 %>% 
  imap(~ mutate(.x, comparison = .y)) %>% 
    bind_rows() %>% 
  group_by(term_id) %>% 
  summarise(n = n()) %>% 
  arrange(-n)

gprofiler.wholenet.genesetsv1 %>% 
  lapply(extract2, "term_id") %>% 
  fromList %>% 
  upset(
    nsets = length(.), 
    order.by = "freq", 
    mb.ratio = c(0.5,0.5)
  )

gprofiler.wholenet.genesetsv1[
  grepl(names(gprofiler.wholenet.genesetsv1), pattern = "lyso")
  ] %>% 
  lapply(function(x) {
    x$term_name
  }) %>%
  fromList %>% 
  upset(
    nsets = length(.), 
    order.by = "freq", 
    mb.ratio = c(0.5,0.5)
  )
```

```{r}
gprofiler.wholenet.genesetsv1[
  grepl(names(gprofiler.wholenet.genesetsv1), pattern = "lyso")
  ] %>% 
  sapply(extract2, "term_id", simplify = F) %>% 
  ggVennDiagram::ggVennDiagram(force_upset = T)
  
```




## 2- cluster annotations
I will here show the top pathways identified in the clusters within the subnetworks. 

```{r}
cluster.gprofiler.results <- 
  list.files(path = "data/results/adultchildsubtypes", 
             pattern = "gprofiler", recursive = TRUE, full.names = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/gprofiler_analysis_results.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/adultchildsubtypes", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))

# only keep the signigicant subnets
cluster.gprofiler.results <- cluster.gprofiler.results[sig.networks]
```




```{r}
cluster.gprofiler.results %>% 
  lapply(function(x) {
    x %>% 
      mutate(
        FDR = p.adjust(p_value, method = "fdr"), 
        CD = purrr::map_chr(comparison, ~ {
          match <- childhood[startsWith(.x, childhood)]
          if (length(match) > 0) match[1] else NA_character_
        }) 
      ) %>% 
      dplyr::filter(p_value != "NA") %>% 
      dplyr::distinct(Top.pathway, .keep_all = TRUE) %>% 
      head(10) %>% 
      mutate(
        Top.pathway = str_wrap(Top.pathway, width = 30)
      )
  }) %>% 
  bind_rows() %>% 
  ggplot(
    aes(
      x = -log10(FDR),
      y = reorder(Top.pathway, -p_value),
      fill = )
  ) +
  geom_col() +
  facet_wrap(~comparison, scales = "free_y")
 

plot_top_pathways_for_CD <- function(cluster_results, CD_subtype, top_n = 10) {
  cluster_results %>%
    lapply(function(x) {
      x %>%
        mutate(
          FDR = p.adjust(p_value, method = "fdr"),
          CD = purrr::map_chr(comparison, ~ {
            match <- childhood[startsWith(.x, childhood)]
            if (length(match) > 0) match[1] else NA_character_
          })
        ) %>%
        dplyr::filter(!is.na(p_value), CD == CD_subtype) %>%
        group_by(comparison) %>%
        slice_min(order_by = p_value, n = top_n) %>%
        ungroup() %>%
        distinct(Top.pathway, comparison, .keep_all = TRUE) %>%
        mutate(Top.pathway = str_wrap(Top.pathway, width = 30))
    }) %>%
    bind_rows() %>%
      mutate(
        Top.pathway = reorder_within(Top.pathway, -FDR, comparison)
        ) %>%
    ggplot(aes(
      x = -log10(FDR),
      y = reorder(Top.pathway, -FDR)
    )) +
    geom_col() +
     scale_y_reordered() +
    facet_wrap(~comparison, scales = "free_y") +
    labs(
      x = "-log10(FDR)",
      y = NULL,
      #title = paste0("Top ", top_n, " Pathways for ", CD_subtype)
    ) +
    theme(strip.text = element_text(face = "bold"))
}

plots_list <- lapply(childhood, function(cd) {
  plot_top_pathways_for_CD(cluster.gprofiler.results, CD_subtype = cd)
})


for (p in plots_list) {
  print(p)
}

plots_list[1]
plots_list[2]
plots_list[3]
plots_list[4]
plots_list[5]
plots_list[6]
plots_list[7]
plots_list[8]
plots_list[9]
plots_list[10]
plots_list[11]
```

```{r}
plots_list2 <- lapply(childhood, function(cd) {
  plot_top_pathways_for_CD(cluster.gprofiler.results[top.sig.networks], CD_subtype = cd)
})

ggarrange(plotlist = plots_list2)
```

```{r}
cluster.gprofiler.results[top.sig.networks] %>% 
  bind_rows() %>% 
  group_by(  Top.pathway) %>% 
  summarise(n = n()) %>% 
  arrange(-n) %>% view
```



# what genes are in the subnetworks which are drug targetable?

Here, i obtained the drug targets data from [omicsynth](https://doi.org/10.1016/j.ajhg.2023.12.006) to see what genes in the subnetwork have a known drug target. 

The plots below show the network proximity score (higher means more proximal to network) vs the evidence of drug interaction (higher interaction = higher score)

```{r}
# the genes in the subnetworks
gene.proximitiy.scores <- node_metadata %>% 
  lapply(function(x) {
    x %>% 
      dplyr::select(gene_name = X, z_both) %>% 
      as_tibble    
  })
  

omicsynth <- read.csv("data/omicsynth/drug_genome_dgidb.csv") 

temp <- gene.proximitiy.scores %>% 
  lapply(function(x) {
    omicsynth %>% # has the drug info 
      # only reatin genes in subnet
      dplyr::filter(gene_name %in% x$gene_name) %>% 
      as_tibble %>% 
      left_join(x)
  }) %>% 
  bind_rows(.id = "comparison") %>% 
  mutate(
    CD = purrr::map_chr(comparison, ~ {
      match <- childhood[startsWith(.x, childhood)]
      if (length(match) > 0) match[1] else NA_character_
    }), 
    .after = comparison
  ) %>% 
  dplyr::filter(comparison %in% sig.networks) 

temp %>% 
  dplyr::filter( comparison == "CD_1A_lyso_AD")

plot_drug_vs_proximity_by_CD <- function(gene_proximity_list, omicsynth, CD_subtype, sig.networks, childhood) {
  gene_proximity_list %>% 
    lapply(function(x) {
      omicsynth %>%
        dplyr::filter(gene_name %in% x$gene_name) %>%
        as_tibble() %>%
        left_join(x)
    }) %>%
    bind_rows(.id = "comparison") %>%
    mutate(
      CD = purrr::map_chr(comparison, ~ {
        match <- childhood[startsWith(.x, childhood)]
        if (length(match) > 0) match[1] else NA_character_
      }),
      .after = comparison
    ) %>%
    dplyr::filter(CD == CD_subtype, comparison %in% sig.networks) %>%
    ggplot(aes(x = interaction_group_score, y = z_both)) +
    geom_point() +
    geom_label_repel(
      aes(label = drug_claim_name),
      data = . %>% dplyr::filter(z_both > 5 & interaction_group_score > 5)
    ) +
    labs(
      x = "Interaction group score", 
      y = "Network proximity z score",
      title = paste("Drug proximity scores -", CD_subtype)
    ) +
    facet_wrap(~comparison, scales = "free")
}

for (cd in childhood) {
  p <- plot_drug_vs_proximity_by_CD(
    gene_proximity_list = gene.proximitiy.scores,
    omicsynth = omicsynth,
    CD_subtype = cd,
    sig.networks = sig.networks,
    childhood = childhood
  )
  print(p)
}
```

# LSDs and adult onset
## heatmap of z scores
note i converted NA to 0

```{r}
library(RColorBrewer)
library(pheatmap)
# heatmap on the genes in the subnetworks
node_metadata[grepl(names(node_metadata), pattern = "CD_1A")] %>% 
  bind_rows(.id = "comparison") %>% 
  dplyr::filter(comparison != "CD_1A_lyso_brain_malform") %>% 
  dplyr::rename(gene = X) %>% 
  as_tibble() %>%
  dplyr::select(comparison, gene, z_both) %>% 
  pivot_wider(values_from = z_both, names_from = comparison) %>% 
  column_to_rownames("gene") %>% 
  # set NA values to 0
  mutate(across(everything(), ~ replace_na(., 0))) %>% 
  as.matrix() %>% 
  pheatmap(
    color = colorRampPalette(rev(viridis_pal(option = "turbo")(100)))(100),
  )
```

## PCA

The z-scores of the genes are quite different 
```{r}
node_metadata[grepl(names(node_metadata), pattern = "CD_1A")] %>% 
  bind_rows(.id = "comparison") %>% 
  dplyr::filter(comparison != "CD_1A_lyso_brain_malform") %>% 
  dplyr::rename(gene = X) %>% 
  as_tibble() %>%
  dplyr::select(comparison, gene, z_both) %>% 
  pivot_wider(values_from = z_both, names_from = comparison) %>% 
  column_to_rownames("gene") %>% 
  # set NA values to 0
  mutate(across(everything(), ~ replace_na(., 0)))  %>% 
  t %>% 
  prcomp %>% 
  autoplot(
    data = tibble(
      comparison = rownames(.$x)
    ),
    alpha = 0.5,
    size = 6
  ) +
  geom_label_repel(
    aes(label = comparison)
  )
```


DLB has a unique lysosomal disease overlap, potentially suggesting stronger involvement of lysosomal pathways than in AD/FTLD/ALS.

CD_1A_lyso_PD is in the bottom-left, far from DLB and the AD/FTLD/ALS group. Suggests its LSD colocalization profile is distinct from DLB, despite both being synucleinopathies.

What genes are driving this?

```{r}
node_metadata[grepl(names(node_metadata), pattern = "CD_1A")] %>% 
  bind_rows(.id = "comparison") %>% 
  dplyr::filter(comparison %in% c("CD_1A_lyso_AD", "CD_1A_lyso_DLB")) %>% 
  dplyr::rename(gene = X) %>% 
  as_tibble() %>% 
  dplyr::select(gene, comparison, d2_name, d1_seeds, d2_seeds, z_both) %>% 
  arrange(z_both) %>% 
  mutate(
    AD_lyso_seed_both = case_when(
      comparison == "CD_1A_lyso_AD" & d1_seeds == 1 & d2_seeds == 1 ~ 1,
      TRUE ~ 0
    ),
    AD_lyso_seed_AD = case_when(
      comparison == "CD_1A_lyso_AD" & d1_seeds == 0 & d2_seeds == 1 ~ 1,
      TRUE ~ 0),
    
    AD_lyso_seed_lyso = case_when(
      comparison == "CD_1A_lyso_AD" & d1_seeds == 1 & d2_seeds == 0 ~ 1,
      TRUE ~ 0),
    AD_lyso_noseed = case_when(
      comparison == "CD_1A_lyso_AD" & d1_seeds == 0 & d2_seeds == 0 ~ 1,
      TRUE ~ 0),
    
    DLB_lyso_seed_both = case_when(
      comparison == "CD_1A_lyso_DLB" & d1_seeds == 1 & d2_seeds == 1 ~ 1,
      TRUE ~ 0
    ),
    DLB_lyso_seed_AD = case_when(
      comparison == "CD_1A_lyso_DLB" & d1_seeds == 0 & d2_seeds == 1 ~ 1,
      TRUE ~ 0),
    
    DLB_lyso_seed_lyso = case_when(
      comparison == "CD_1A_lyso_DLB" & d1_seeds == 1 & d2_seeds == 0 ~ 1,
      TRUE ~ 0),
    
    DLB_lyso_noseed = case_when(
      comparison == "CD_1A_lyso_DLB" & d1_seeds == 0 & d2_seeds == 0 ~ 1,
      TRUE ~ 0)
  ) %>% 
  dplyr::select(shared_name = gene, everything()) 
  #write.csv("output/adult child subtypes/lyso_AD_DLB_genes.csv")
```

```{r}

temp <- runKEGG(geneset = node_metadata$CD_1A_lyso_DLB$X)

temp %>% 
  dplyr::filter(Count > 5) %>% .$Description
  ggplot(
    aes(
      x = -log10(p.adjust), 
      y = reorder(Description, -pvalue),
      fill = Count)
  ) +
  geom_col() +
  scale_fill_viridis_c() +
  labs(
    y = "KEGG pathway", 
    fill = "Number of genes"
  ) 
  ggsave(
    "output/adult child subtypes/figs/talk/overrep_KEGG.svg", 
    width = 10, height = 5,
    scale = 0.8
    )
  
```

# New genesets

Netcoloc was performed on new genesets identified by William. below shows the expected vs observed ratio of the genes identified by dual network propogation. 

## number of genes per geneset
```{r}
genesets_v2 <- 
  list.files("data/genesets/Alternate_gene_lists_for_NetColoc_senstivity_analyses", 
             full.names = T, pattern = "_.*txt") %>% 
  sapply(function(x) {
    x %>% 
      read.delim() %>% 
      .$Gene
  }, simplify = F) %>% 
  set_names(
    str_replace(
      list.files("data/genesets/Alternate_gene_lists_for_NetColoc_senstivity_analyses", 
                 full.names = F, pattern = "_.*txt"), 
      pattern = ".txt", 
      replacement = ""
    )
  ) %>% 
  set_names(
    str_remove(names(.), pattern = "_v2_list")
  )

genesets %>% 
  unlist %>% 
  as.data.frame(col.names = "gene") %>% 
  rownames_to_column("geneset") %>%
  set_colnames(c("geneset", "gene")) %>% 
  mutate(
    geneset = str_remove(geneset, pattern = "\\d{1,3}$"), 
    type = case_when(
      grepl(geneset, pattern = "^CD") ~ "childhood", 
      TRUE ~ "adult"
    )
   ) %>% 
  dplyr::filter(type == "childhood") %>% 
  bind_rows(
    genesets_v2 %>% 
      unlist %>% 
      as.data.frame(col.names = "gene") %>% 
      rownames_to_column("geneset") %>%
      set_colnames(c("geneset", "gene")) %>% 
      mutate(
        geneset = str_remove(geneset, pattern = "\\d{1,3}$"), 

        type = case_when(
          grepl(geneset, pattern = "^CD") ~ "childhood", 
          TRUE ~ "adult"
    )
   ) 
  ) %>% 
  mutate(
      geneset = case_when(
      grepl(geneset, pattern = "1A") ~ "Lysosomal disorders", 
      grepl(geneset, pattern = "1B") ~ "Other disorders of lipid metabolism & transport", 
      grepl(geneset, pattern = "1C") ~ "Disorders of amino acid & other organic acid metabolism", 
      grepl(geneset, pattern = "1D") ~ "Vitamin-responsive inborn errors of metabolism", 
      grepl(geneset, pattern = "1E") ~ "Disorders of mineral absorption & transport", 
      grepl(geneset, pattern = "1F") ~ "Peroxisomal diseases", 
      grepl(geneset, pattern = "1G") ~ "Mitochondrial disorders", 
      grepl(geneset, pattern = "1H") ~ "Other inborn errors of metabolism", 
      grepl(geneset, pattern = "2_") ~ "Leukodystrophies not notherwise categorised", 
      grepl(geneset, pattern = "3_") ~ "Neurodegeneration with brain iron accumulation", 
      grepl(geneset, pattern = "4_") ~ "Diseases not otherwise categorised",
      TRUE ~ geneset
      
    ) %>% 
      str_wrap(width = 30),
      geneset = fct_infreq(geneset),  # Reorder by frequency, 
  ) %>% 
  ggplot(
    aes(x = geneset, fill = type)
    )+
  geom_bar(size = 1) +
  #easy_rotate_x_labels(angle =-90) +
  labs(
    y = "# genes"
  ) +
  facet_wrap(
    ~type, 
    ncol = 1, 
    scales = "free_y",
    space = "free_y", 
    strip.position = "right") +
  coord_flip() +
  scale_fill_manual(
    values = c("#FF6666","#3399FF")) 
  # theme(
  #   plot.margin = margin(r = 2, unit = "cm"), 
  #   legend.position = "none",
  #   panel.grid = element_blank(),
  # 
  #   text = element_text(colour = "white"),
  #   axis.title = element_text(colour = "white"),
  #   axis.text = element_text(colour = "white"),
  #   
  #   #strip.background = element_blank(), 
  #   plot.background = element_rect(fill = "#25283D"),
  #   plot.background = element_blank(),
  #   panel.grid.minor = element_blank(),
  #   panel.background = element_blank(), 
  # )
  # ggsave("output/plots4poster/numGenes.svg", 
  #        width = 20, height = 18, units = "cm", scale= 0.8)

```


```{r}
# Define path
base_path <- "data/results/results_3"

# Read enrichment files
enrichment_dfs.2 <- 
  list.files(base_path, recursive = TRUE, full.names = TRUE) %>%
  grep(pattern = "network_colocalization_enrichment", 
       value = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/network_colocalization_enrichment.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0(base_path, "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y)) %>% 
  bind_rows() %>%
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>%
  pivot_longer(
    cols = c(observed_overlap, expected_overlap_mean),
    names_to = "type",
    values_to = "value"
  ) %>%
  mutate(
    type = recode(type,
                  "observed_overlap" = "observed",
                  "expected_overlap_mean" = "expected")
  ) %>% 
  mutate(source = "GWAS only (no parental proxy)")
```

## plot observed v expected ratio

```{r}
enrichment_dfs.2 %>% 
  bind_rows(enrichment_dfs %>% mutate(source = "Clingen, NDKP, familial")) %>% 
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>% 
  mutate(
    # use the childhood vector to extract the CD seed gene name
    CD = purrr::map_chr(comparison, ~ {
      match <- childhood[startsWith(.x, childhood)]
      if (length(match) > 0) match[1] else NA_character_
    }),

    adult = str_remove(comparison, pattern = "CD_.+_.+_"),
    adult = case_when(
      adult == "malform" ~ "brain_malform", 
      TRUE ~ adult
    ),
    .after = obs_exp, 
    FDR = p.adjust(empirical_p, "fdr")
  ) %>% 
    mutate(
      CD = case_when(
      grepl(CD, pattern = "1A") ~ "Lysosomal disorders", 
      grepl(CD, pattern = "1B") ~ "Other disorders of lipid metabolism & transport", 
      grepl(CD, pattern = "1C") ~ "Disorders of amino acid & other organic acid metabolism", 
      grepl(CD, pattern = "1D") ~ "Vitamin-responsive inborn errors of metabolism", 
      grepl(CD, pattern = "1E") ~ "Disorders of mineral absorption & transport", 
      grepl(CD, pattern = "1F") ~ "Peroxisomal diseases", 
      grepl(CD, pattern = "1G") ~ "Mitochondrial disorders", 
      grepl(CD, pattern = "1H") ~ "Other inborn errors of metabolism", 
      grepl(CD, pattern = "2_") ~ "Leukodystrophies not notherwise categorised", 
      grepl(CD, pattern = "3_") ~ "Neurodegeneration with brain iron accumulation", 
      grepl(CD, pattern = "4_") ~ "Diseases not otherwise categorised",
      TRUE ~ CD) %>% str_wrap(width = 20)
    ) %>%  
  dplyr::filter(type == "observed") %>% 
  
  ggplot(
    aes(x = obs_exp, y = reorder(adult, obs_exp))
  ) +
  geom_col(
    aes(fill = source, alpha = FDR < 0.05), 
    colour = "grey50", position = "dodge", 
    width = 0.75
  ) +
  scale_fill_viridis_d() +
  scale_alpha_manual(values = c(0.1, 1)) +
  labs(
    x = "observed / expected", 
    y = "Disease"
  ) +
  facet_wrap(~CD, ncol = 3) 
  # theme(legend.position = "top") +
  # ggsave(
  #   "output/figs/paper/networkcoloc_v2.svg",
  #   width = 20, height = 20, units = "cm",
  #   scale = 1
  )
```

```{r}
tony2keep = c("DLB", "brain_malform")

enrichment_dfs.2 %>% 
  bind_rows(
    enrichment_dfs %>% 
      mutate(source = "Tony") %>% 
      dplyr::filter(
        grepl(comparison, pattern = "DLB|brain_malf")
      )
  ) %>% 
  dplyr::filter(
    z_comb == 3, z_12 == 1.5
  ) %>% 
  mutate(
    # use the childhood vector to extract the CD seed gene name
    CD = purrr::map_chr(comparison, ~ {
      match <- childhood[startsWith(.x, childhood)]
      if (length(match) > 0) match[1] else NA_character_
    }),
    CD = case_when(
      grepl(CD, pattern = "1A") ~ "Lysosomal disorders", 
      grepl(CD, pattern = "1B") ~ "Other disorders of lipid metabolism & transport", 
      grepl(CD, pattern = "1C") ~ "Disorders of amino acid & other organic acid metabolism", 
      grepl(CD, pattern = "1D") ~ "Vitamin-responsive inborn errors of metabolism", 
      grepl(CD, pattern = "1E") ~ "Disorders of mineral absorption & transport", 
      grepl(CD, pattern = "1F") ~ "Peroxisomal diseases", 
      grepl(CD, pattern = "1G") ~ "Mitochondrial disorders", 
      grepl(CD, pattern = "1H") ~ "Other inborn errors of metabolism", 
      grepl(CD, pattern = "2_") ~ "Leukodystrophies not notherwise categorised", 
      grepl(CD, pattern = "3_") ~ "Neurodegeneration with brain iron accumulation", 
      grepl(CD, pattern = "4_") ~ "Diseases not otherwise categorised"
    ) %>% 
      str_wrap(width = 22),
    adult = str_remove(comparison, pattern = "CD_.+_.+_"),
    adult = case_when(
      adult == "malform" ~ "brain_malform", 
      TRUE ~ adult
    ),
    .after = obs_exp, 
    FDR = p.adjust(empirical_p, "fdr")
  ) %>% 
  dplyr::filter(
    type == "observed",
    CD != "Other inborn errors\nof metabolism"
  ) %>% 
  ggplot(
    aes(x = obs_exp, y = reorder(adult, obs_exp))
  ) +
  geom_col(
    aes(fill = -log10(FDR), alpha = FDR < 0.05), 
    colour = "grey50", 
    position = "dodge", width = 0.7
  ) +
  scale_fill_viridis_c(option = "viridis", begin = 0.6) +
  scale_alpha_manual(values = c(0.1, 1)) +
  labs(
    x = "observed / expected", 
    y = "adult-onset NDD", 
  ) +
  facet_wrap(~CD, nrow = 3) +
  theme(
    
    legend.position = "top",
    
    text = element_text(colour = "white"),
    axis.title = element_text(colour = "white"),
    axis.text = element_text(colour = "white"),
    
    strip.background = element_blank(), 
    legend.background = element_blank(),
    plot.background = element_rect(fill = "#25283D"),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
    
  ) 
  # ggsave("output/plots4poster/networkcoloc_v2.svg", 
  #        width = 45, height = 40, units = "cm", 
  #        scale = 0.4)
  # 
```


## pathways
```{r}
sig.networks.2 <- 
  enrichment_dfs.2 %>% 
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>% 
  mutate(FDR = p.adjust(empirical_p, method ="fdr")) %>% 
  dplyr::filter(FDR < 0.05) %>% 
  .$comparison %>% 
  unique
```

### whole subnetworks
```{r}
node_metadata.2 <- 
  list.files(path = "data/results/results_3", 
             pattern = "overlapping_nodes_metadata", recursive = TRUE, full.names = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/overlapping_nodes_metadata.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/results_3", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))

enrichment.wholesubnet.kegg.2 <- 
  node_metadata.2[sig.networks.2] %>% 
  lapply(magrittr::extract2, "X") %>% 
  sapply(runKEGG, simplify = F)

enrichment.wholesubnet.kegg.2 %>% 
  lapply(function(x) {
    x$Description
  }) %>% 
  fromList() %>% 
  UpSetR::upset(
    nsets = length(.), 
    order.by = "freq"
  )
```

```{r}
  gprofiler.wholenet.genesetsv2 <- node_metadata.2[sig.networks.2] %>% 
  lapply(magrittr::extract2, "X") %>% 
  sapply(function(x) {
    gost(query = x, 
         organism = "gp__XoLn_Cg43_BEg", 
         ordered_query = FALSE, 
         significant = TRUE, # only return sig results
         user_threshold = 0.05,
         correction_method = "fdr",
         domain_scope = "annotated", # only analyse genes with the annotation. 
         evcodes = TRUE,
         custom_bg = nodes) %>% 
      .$result %>% 
      as_tibble
  }, simplify = F)

gprofiler.wholenet.genesetsv2 %>% 
  imap(~ mutate(.x, comparison = .y)) %>% 
  bind_rows() %>% 
  group_by(term_id) %>% 
  summarise(
    n = n()
    ) %>% 
  arrange(-n)

gprofiler.wholenet.genesetsv2 %>% 
  lapply(extract2, "term_id") %>% 
  fromList %>% 
  upset(
    nsets = length(.), 
    order.by = "freq", 
    mb.ratio = c(0.5,0.5)
  )
```


### clusters
```{r}
cluster.gprofiler.results.2 <- 
  list.files(path = "data/results/results_3", 
             pattern = "gprofiler", recursive = TRUE, full.names = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/gprofiler_analysis_results.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/results_3", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))

# only keep the significant subnets
cluster.gprofiler.results.2 <- cluster.gprofiler.results.2[sig.networks.2]


```

# output
## output files some for william
```{r}
cluster.gprofiler.results %>% 
  lapply(function(x) {
    x %>% 
      write_csv(paste0("output/network_pathways/genesets_v1/clusters/", unique(x$comparison), ".csv"))
  })

cluster.gprofiler.results.2 %>% 
  lapply(function(x) {
    x %>% 
      write_csv(paste0("output/network_pathways/genesets_v2/clusters/", unique(x$comparison), ".csv"))
  })

enrichment.wholesubnet.kegg %>% 
   imap(~ mutate(.x, comparison = .y)) %>% 
    lapply(function(x) {
    x %>% 
      write_csv(paste0("output/network_pathways/genesets_v1/wholenetwork/", unique(x$comparison), ".csv"))
    })

enrichment.wholesubnet.kegg.2 %>% 
   imap(~ mutate(.x, comparison = .y)) %>% 
    lapply(function(x) {
    x %>% 
      write_csv(paste0("output/network_pathways/genesets_v2/wholenetwork/", unique(x$comparison), ".csv"))
    })

gprofiler.wholenet.genesetsv1 %>% 
   imap(~ mutate(.x, comparison = .y)) %>% 
    lapply(function(x) {
    x %>% 
      write_csv(paste0("output/network_pathways/genesets_v1/wholenetwork_gprofiler/", unique(x$comparison), ".csv"))
    })

gprofiler.wholenet.genesetsv2 %>% 
   imap(~ mutate(.x, comparison = .y)) %>% 
    lapply(function(x) {
    x %>% 
      write_csv(paste0("output/network_pathways/genesets_v2/wholenetwork_gprofiler/", unique(x$comparison), ".csv"))
    })
      
```
## Figs
### n genes per gs
```{r}
genesets %>% 
  unlist %>% 
  as.data.frame(col.names = "gene") %>% 
  rownames_to_column("geneset") %>%
  set_colnames(c("geneset", "gene")) %>% 
  mutate(
    geneset = str_remove(geneset, pattern = "\\d{1,3}$"), 
    geneset = fct_infreq(geneset), # Reorder by frequency, 
    type = case_when(
      grepl(geneset, pattern = "^CD") ~ "childhood", 
      TRUE ~ "adult"
    )
   
   ) %>% 
  ggplot(
    aes(x = geneset)
  ) +
  geom_bar() +
  easy_rotate_x_labels(angle = -45) +
  labs(
    title = "Number of genes per gene set", 
    y = "# genes"
  ) +
  facet_wrap(~type, scales = "free_x") +
  theme(
    plot.margin = margin(r = 2, unit = "cm")
  ) +
    ggsave(
    "output/adult child subtypes/figs/talk/numgenespergs.svg", 
    width = 8, height = 4, units = "cm", 
    scale = 2
  )
```

### netcoloc
```{r}
enrichment_dfs %>% 
  bind_rows() %>% 
  as_tibble %>% 
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>% 
  mutate(
    # use the childhood vector to extract the CD seed gene name
    CD = purrr::map_chr(comparison, ~ {
      match <- childhood[startsWith(.x, childhood)]
      if (length(match) > 0) match[1] else NA_character_
    }),

    adult = str_remove(comparison, pattern = "CD_.+_.+_"),
    adult = case_when(
      adult == "malform" ~ "brain_malform", 
      TRUE ~ adult
    ),
    .after = obs_exp, 
    FDR = p.adjust(empirical_p, "fdr")
  ) %>% 
  dplyr::filter(type == "observed") %>% 
  dplyr::filter(
    CD != "CD_4_notcateg", 
    CD != "CD_1H_other_IEMs"
    ) %>%
  
  # make new names
  
  mutate(
    CD = case_when(
      grepl(CD, pattern = "1A") ~ "Lysosomal disorders", 
      grepl(CD, pattern = "1B") ~ "Other disorders of lipid metabolism & transport", 
      grepl(CD, pattern = "1C") ~ "Disorders of amino acid & other organic acid metabolism", 
      grepl(CD, pattern = "1D") ~ "Vitamin-responsive inborn errors of metabolism", 
      grepl(CD, pattern = "1E") ~ "Disorders of mineral absorption & transport", 
      grepl(CD, pattern = "1F") ~ "Peroxisomal diseases", 
      grepl(CD, pattern = "1G") ~ "Mitochondrial disorders", 
      grepl(CD, pattern = "1H") ~ "Other inborn errors of metabolism", 
      grepl(CD, pattern = "2_") ~ "Leukodystrophies not otherwise categorised", 
      grepl(CD, pattern = "3_") ~ "Neurodegeneration with brain iron accumulation", 
      grepl(CD, pattern = "4_") ~ "Diseases not otherwise categorised"
    ) %>% 
      str_wrap(width = 22),
  ) %>% 
  
  ggplot(
    aes(x = obs_exp, y = reorder(adult, obs_exp))
  ) +
  geom_col(
    aes(
      fill = -log10(FDR), 
      alpha = FDR < 0.05
      ), 
    colour = "grey50"
  ) +
  scale_fill_viridis_c() +
  scale_alpha_manual(values = c(0.1, 1)) +
  labs(
    x = "observed / expected", 
    y = "adult-onset NDD", 
    title = "pairwise contrasts between CD & adult-onset NDD"
  )  +
  facet_wrap(~CD, ncol = 3) +
    ggsave(
    "output/adult child subtypes/figs/talk/networkcoloc.svg", 
    width = 20, height = 14, units = "cm", 
    scale = 1
  )
```

##
