---
title: "CDsubtypes"
author: "Karissa Barthelson"
date: "2025-05-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
```{r}
knitr::opts_chunk$set(
  message = F, echo = T, warning = F
)
```

```{r}
library(tidyverse)
library(magrittr)
library(readxl)

# vis
library(ComplexUpset)
library(UpSetR)
library(pander)
library(ggeasy)
library(ggpubr)
library(ggrepel)
library(plotly)
library(tidyHeatmap)
library(grid)
library(gridExtra)
library(gridGraphics)
library(viridis)
library(tidytext)

# annotations
library(org.Hs.eg.db)
library(ndexr)

# enrichemnt
library(clusterProfiler)
library(ReactomePA)

theme_set(theme_light() +
            theme(plot.title = element_text(hjust = 0.5), 
                  strip.text = element_text(face = "bold"),
                  strip.background = element_rect(fill = "grey20")
            )
)

knitr::opts_chunk$set(
  message = F, echo = F, warning = F
)
```

# import gene sets

```{r}
genesets <- list.files(
  path = "data/genesets/adultchildsubtypes",
  pattern = ".csv",
  full.names = TRUE
) %>% 
  sapply(function(x) {
    x %>% 
      read_csv() %>% 
      .$GENE
  }, simplify = F) %>% 
  setNames(names(.) %>% 
             str_remove(pattern = "data/genesets/adultchildsubtypes/") %>% 
             str_remove(pattern = ".csv"))

# get these for help later
adult = c("AD", "PD", "ALS", "FTLD", "DLB")
childhood = grep(names(genesets), pattern = "^CD", value = TRUE)
```

tony curated these genesets from clingen. here is the number of genes per gene set.  

```{r}
genesets %>% 
  unlist %>% 
  as.data.frame(col.names = "gene") %>% 
  rownames_to_column("geneset") %>%
  set_colnames(c("geneset", "gene")) %>% 
  mutate(
   geneset = str_remove(geneset, pattern = "\\d{1,3}$"), 
    geneset = fct_infreq(geneset)  # Reorder by frequency
   ) %>% 
  ggplot(
    aes(x = geneset)
  ) +
  geom_bar() +
  easy_rotate_x_labels(angle = -45) +
  labs(
    title = "Number of genes per gene set", 
    y = "# genes"
  )
```


## overlap of genes across gene sets
There is minimal overlap of genes across the adult and childhod dementia gene sets. Just 1 or 2 in each comparison

the 10 gene overlap is between ALS and FTLD

```{r}
genesets %>% 
  fromList %>% 
  UpSetR::upset(nsets = length(.))
```

# over-representation analysis of gene sets

Here I will perform pathway/enrichment analysis on these gene sets to see what pathways are enriched, and how much overlap there is. 

the background gene set will be all genes found in the PCNet, as these are the ones I am analysing. The node info was obtained from NDex. 
```{r}
# ~~ only run this on the first time ~~#
#
# login to the NDEx server - run this in console
# ndexcon <- ndex_connect(
#  "karissa.barthelson@flinders.edu.au", "password - capital and 21")

# search the networks for 'pcNet v1.3'
# pcnet1.3 = "4de852d9-9908-11e9-bcaf-0ac135e8bacf"

# obtain the nodes
# nodes = ndex_network_get_aspect(ndexcon, pcnet1.3, 'nodes') %>% .$name

# save this out for reimporting later
# saveRDS(nodes, "data/R_objects/network/nodes.rds")

nodes <- readRDS("data/R_objects/network/nodes.rds")
```

## GO terms

Here, i performed over-represetation analysis (hypergeometric test) of GO terms (all ontologies) of each gene set using clusterprofiler

```{r}
runGO = function(geneset) {
  enrichGO(
    gene = geneset, 
    universe = nodes, 
    OrgDb = org.Hs.eg.db, 
    keyType = "SYMBOL", 
    ont = "ALL"
  )
}
```

```{r}
GO <- genesets %>% 
  sapply(runGO, simplify = F) %>% 
  sapply(as_tibble, simplify = F)

GO %>% 
  lapply(function(x) {
    x$Description
  }) %>% 
  fromList() %>% 
  UpSetR::upset(
    nsets = length(.), 
    order.by = "freq"
  )

GO.terms.pergeneset <- GO %>% 
  lapply(function(x) {
    x$Description
  }) 
```

## KEGG pathways

```{r}
runKEGG = function(geneset) {

  uv =   bitr(nodes, 
                fromType = "SYMBOL", toType = "ENTREZID", 
                OrgDb ="org.Hs.eg.db") %>% 
    .$ENTREZID
  
  bitr(geneset, 
       fromType = "SYMBOL", toType = "ENTREZID", 
     OrgDb ="org.Hs.eg.db") %>% 
  .$ENTREZID %>%
  enrichKEGG(
    organism = 'hsa',
    universe = uv,
    pvalueCutoff = 0.05) %>% 
  as_tibble 
  
}
```

```{r}
kegg <- genesets %>% 
  sapply(function(x) {runKEGG(x)}, simplify = F) 

kegg %>% 
  lapply(function(x) {x$Description}) %>% 
  fromList() %>% 
  UpSetR::upset()
```


## reactome pathways 
```{r}
runReactome = function(geneset, genesetname) {
  
  enrichPathway(
    gene = bitr(geneset, 
                fromType = "SYMBOL", toType = "ENTREZID", 
                OrgDb ="org.Hs.eg.db") %>% 
      .$ENTREZID,
    universe = bitr(nodes,
                    fromType = "SYMBOL", toType = "ENTREZID",
                    OrgDb= "org.Hs.eg.db") %>% 
      .$ENTREZID, # all genes in pcnet
    readable = TRUE, 
    pvalueCutoff = 0.05
  ) %>% 
  as_tibble 
}

# perform hypergeometic test for significant overrep of Reactome db
reactome <- 
  genesets %>% 
  sapply(runReactome, "reactome", simplify = F)

reactome %>% 
  lapply(function(x) {x$Description}) %>% 
  fromList() %>% 
  UpSetR::upset(
    order.by = "freq", 
    nsets = length(.)
  )
```

# Is there significant colocalisation?

Netcoloc was performed and below shows the expected vs observed ratio of the genes identified by dual network propogation. 

```{r}
# Define path
base_path <- "data/results/adultchildsubtypes/"

# Read enrichment files
enrichment_dfs <- 
  list.files(base_path, recursive = TRUE, full.names = TRUE) %>%
  grep(pattern = "network_colocalization_enrichment", 
       value = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/network_colocalization_enrichment.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0(base_path, "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y)) %>% 
  bind_rows() %>%
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>%
  pivot_longer(
    cols = c(observed_overlap, expected_overlap_mean),
    names_to = "type",
    values_to = "value"
  ) %>%
  mutate(
    type = recode(type,
                  "observed_overlap" = "observed",
                  "expected_overlap_mean" = "expected")
  )
```

## plot observed v expected per comparison
```{r}
enrichment_dfs %>% 
  ggplot(
    aes(x = type, y = value, fill = type)
    ) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  geom_errorbar(
    data = . %>% 
      dplyr::filter(type == "expected"),
    aes(ymin = value - 2 * expected_overlap_std,
        ymax = value + 2 * expected_overlap_std),
    width = 0.25
  ) +
  geom_label(
    data =  . %>% 
      dplyr::filter(type == "expected"),
    aes(x = 1.5, y = 10, 
        label = sprintf("p = %.2e", empirical_p)),
    inherit.aes = FALSE,
    vjust = 0,
    hjust = 0.5,
    fontface = "bold",
    show.legend = FALSE
  ) +
  labs(
    y = "Number of genes", 
    x = ""
  ) +
  facet_wrap(~comparison, scales = "free_x") +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("grey50", "skyblue"))
```

## plot observed v expected ratio

```{r}
enrichment_dfs %>% 
  bind_rows() %>% 
  as_tibble %>% 
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>% 
  mutate(
    # use the childhood vector to extract the CD seed gene name
    CD = purrr::map_chr(comparison, ~ {
      match <- childhood[startsWith(.x, childhood)]
      if (length(match) > 0) match[1] else NA_character_
    }),

    adult = str_remove(comparison, pattern = "CD_.+_.+_"),
    adult = case_when(
      adult == "malform" ~ "brain_malform", 
      TRUE ~ adult
    ),
    .after = obs_exp, 
    FDR = p.adjust(empirical_p, "fdr")
  ) %>% 
  dplyr::filter(type == "observed") %>% 
  ggplot(
    aes(x = obs_exp, y = reorder(adult, obs_exp))
  ) +
  geom_col(
    aes(fill = -log10(FDR), alpha = FDR < 0.05), 
    colour = "grey50"
  ) +
  scale_fill_viridis_c() +
  scale_alpha_manual(values = c(0.1, 1)) +
  labs(
    x = "observed / expected", 
    y = "adult-onset NDD", 
    title = "pairwise contrasts between CD & adult-onset NDD"
  ) +
  facet_wrap(~CD)
```

# Are the subnetworks biologically meaningful?

Here, I will only consider the subnetworks which are significant

```{r}
sig.networks <- enrichment_dfs %>% 
  bind_rows() %>% 
  as_tibble %>% 
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>% 
  mutate(
    FDR = p.adjust(empirical_p, "fdr")
  ) %>% 
  dplyr::filter(FDR < 0.05) %>% 
  .$comparison %>% 
  unique
```


## check the genes in the subnetwork are enriched with mouse brain phenotypes

I tested the enrichment of genes associated with mouse brain/neural phenotypes in the NetColoc subnetwork as a *whole*. I also included here a negative control phenotype -  abnormal digit morphology. This is not expected to be associated with the subnetworks. 

```{r}
mouse.variant.wholesubnetworks <- list.files(
    "data/results/adultchildsubtypes", recursive = TRUE, full.names = TRUE) %>% 
  grep(pattern = "/mousevariant_KO_enrichment.csv", 
       value = TRUE) %>% 
  set_names(
    str_replace(., 
               pattern = "/mousevariant_KO_enrichment.csv",
               replacement = ""
  ) %>% 
    str_remove(
      pattern = paste0("data/results/adultchildsubtypes", "/")
    )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y)) %>% 
  bind_rows() %>% 
  dplyr::filter(comparison %in% sig.networks) %>%
  as_tibble() %>%
  group_by(comparison) %>%
  mutate(
        # use the childhood vector to extract the CD seed gene name
    CD = purrr::map_chr(comparison, ~ {
      match <- childhood[startsWith(.x, childhood)]
      if (length(match) > 0) match[1] else NA_character_
    }),

    adult = str_remove(comparison, pattern = "CD_.+_.+_"),
    adult = case_when(
      adult == "malform" ~ "brain_malform", 
      TRUE ~ adult
    ),
    FDR = p.adjust(OR_p),
    .after = OR_p
  )
# a function for plotting by CD subgtype:
plot.mouse.variant.subtype.forestplot.by.CD.subtype <- function(data, CD_subtype) {
  data %>%
    dplyr::filter(CD == CD_subtype) %>% 
    group_by(comparison, MPO_term_type) %>%
    slice_min(order_by = OR_p, n = 10) %>%
    ungroup() %>%
    mutate(
      MP_description = reorder_within(MP_description, log_OR, comparison)
    ) %>%
    ggplot(
      aes(x = log_OR, y = MP_description)
    ) +
    geom_errorbarh(
      aes(xmin = log_OR_CI_lower, xmax = log_OR_CI_upper),
      height = 0.2, color = "black"
    ) +
    geom_point(
      aes(size = num_genes_in_term, color = MPO_term_type),
      alpha = 0.6
    ) +
    geom_vline(xintercept = 0, lty = 2, color = "grey50") +
    geom_label(
      aes(label = signif(OR_p, 2)),
      nudge_y = 0.4
    ) +
    scale_y_reordered() +  # This removes the `___comparison` suffix added by reorder_within()
    labs(
      x = "log(OR) ± 95% CI",
      y = NULL,
      size = "# Genes in MPO term",
      color = "Term Type"
    ) +
  facet_wrap(~comparison, scales = "free")
}

mouse.variants.plots <- childhood %>% 
  lapply(
    function(x) {
      plot.mouse.variant.subtype.forestplot.by.CD.subtype(
        CD_subtype = x, data = mouse.variant.wholesubnetworks)
    })

mouse.variants.plots[1]
mouse.variants.plots[2]
mouse.variants.plots[3]
mouse.variants.plots[4]
mouse.variants.plots[5]
mouse.variants.plots[6]
mouse.variants.plots[7]
mouse.variants.plots[8]
```

## are the genes in the subnetwork human brain expressed?

I next want to see whether the genes in the subnetwork are expressed in the brain. To do this, I will assess it in the GTEx dataset. I will use the GTEx v10 data, which is the most recent version. 

```{r}
### obtain the genes in the subnetworks
node_metadata <- 
  list.files(path = "data/results/adultchildsubtypes", 
             pattern = "overlapping_nodes_metadata", recursive = TRUE, full.names = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/overlapping_nodes_metadata.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/adultchildsubtypes", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))
```

```{r}
# plot number of genes per subnetworks
node_metadata %>% 
  bind_rows() %>% 
  dplyr::filter(comparison %in% sig.networks) %>% 
  dplyr::rename(gene = X) %>% 
  as_tibble() %>%
  mutate(
    seed_type = case_when(
      d1_seeds == 1 & d2_seeds == 1 ~ "both",
      d1_seeds == 1 & d2_seeds == 0 ~ "D1",
      d1_seeds == 0 & d2_seeds == 1 ~ "D2",
      d1_seeds == 0 & d2_seeds == 0 ~ "network.proximal"
    )
  ) %>% 
  ggplot(
    aes(y = comparison, fill = seed_type)
  ) +
  geom_bar() +
  scale_fill_manual(values = c(
    "#7570b3", "#1b9e77", "#d95f02", "grey70"
  )) +
  labs(
    x = "Number of genes in subnetwork", 
    y = "Seed gene sets", 
    title = "Number of genes in the subnetworks"
  ) 
```

### obtain GTEx expression data
This file was downloaded from here :
https://www.gtexportal.org/home/downloads/adult-gtex/bulk_tissue_expression

The file contains the median gene-level TPM by tissue. 

```{r}
# this was downloaded from gtex websiote
# version 10 
gtex <- read.delim("data/gtex/GTEx_Analysis_v10_RNASeQCv2.4.2_gene_median_tpm.gct.gz", skip =2)

gtex %<>% 
  pivot_longer(names_to = "tissue", 
               cols = 3:56, 
               values_to = "tpm"
  ) %>% 
  dplyr::select(
    gene = Description, tissue, tpm
  ) %>% 
  mutate(
    logtpm = log(tpm))
```

```{r}
# define a threshold for expressed or not  
gtex.threshold = 0.5

# Fix: safe join and summarize per gene per tissue
node_metadata %>%
  bind_rows(.id = "comparison") %>%  # Add comparison info
  dplyr::rename(gene = X) %>%
  as_tibble() %>%
  dplyr::select(comparison, gene) %>% 
  left_join(
    gtex %>% 
      dplyr::filter(grepl("Brain", tissue)) %>% 
      dplyr::distinct(gene, tissue, logtpm)
  ) %>% 
  mutate(
    expressed = ifelse(logtpm > gtex.threshold, 1, 0)
    ) %>% 
  group_by(tissue, comparison) %>%
  dplyr::filter(
    grepl("Brain", tissue), 
    comparison %in% sig.networks
    ) %>% 
  mutate(
    n_genes = n(),
    n_expressed = sum(expressed),
    proportion_expressed = n_expressed / n_genes, 
           # use the childhood vector to extract the CD seed gene name
    CD = purrr::map_chr(comparison, ~ {
      match <- childhood[startsWith(.x, childhood)]
      if (length(match) > 0) match[1] else NA_character_
    }), 
    
    adult = str_remove(comparison, pattern = "CD_.+_.+_"),
  ) %>% 
  ungroup %>% 
  group_by(CD) %>% 
   tidyHeatmap::heatmap(
    .row = tissue,
    .column = comparison,
    .value = proportion_expressed, 
    
    # turn off clustering
    cluster_rows = FALSE, 
    column_dend_height = unit(2, "cm"),
    
   ) %>% 
  annotation_tile(CD)
```

# what pathways are in the subnetwork

I will here show the top pathways identified in the clusters within the subnetworks. 

```{r}
cluster.gprofiler.results <- 
  list.files(path = "data/results/adultchildsubtypes", 
             pattern = "gprofiler", recursive = TRUE, full.names = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/gprofiler_analysis_results.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/adultchildsubtypes", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))

# only keep the signigicant subnets
cluster.gprofiler.results <- cluster.gprofiler.results[sig.networks]
```

```{r}
cluster.gprofiler.results %>% 
  lapply(function(x) {
    x %>% 
      mutate(
        FDR = p.adjust(p_value, method = "fdr"), 
        CD = purrr::map_chr(comparison, ~ {
          match <- childhood[startsWith(.x, childhood)]
          if (length(match) > 0) match[1] else NA_character_
        }) 
      ) %>% 
      dplyr::filter(p_value != "NA") %>% 
      dplyr::distinct(Top.pathway, .keep_all = TRUE) %>% 
      head(10) %>% 
      mutate(
        Top.pathway = str_wrap(Top.pathway, width = 30)
      )
  }) %>% 
  bind_rows() %>% 
  ggplot(
    aes(
      x = -log10(FDR),
      y = reorder(Top.pathway, -p_value),
      fill = )
  ) +
  geom_col() +
  facet_wrap(~comparison, scales = "free_y")
 

plot_top_pathways_for_CD <- function(cluster_results, CD_subtype, top_n = 10) {
  cluster_results %>%
    lapply(function(x) {
      x %>%
        mutate(
          FDR = p.adjust(p_value, method = "fdr"),
          CD = purrr::map_chr(comparison, ~ {
            match <- childhood[startsWith(.x, childhood)]
            if (length(match) > 0) match[1] else NA_character_
          })
        ) %>%
        filter(!is.na(p_value), CD == CD_subtype) %>%
        group_by(comparison) %>%
        slice_min(order_by = p_value, n = top_n) %>%
        ungroup() %>%
        distinct(Top.pathway, comparison, .keep_all = TRUE) %>%
        mutate(Top.pathway = str_wrap(Top.pathway, width = 30))
    }) %>%
    bind_rows() %>%
      mutate(Top.pathway = reorder_within(Top.pathway, -FDR, comparison)) %>%
    ggplot(aes(
      x = -log10(FDR),
      y = reorder(Top.pathway, -FDR)
    )) +
    geom_col() +
     scale_y_reordered() +
    facet_wrap(~comparison, scales = "free_y") +
    labs(
      x = "-log10(FDR)",
      y = NULL,
      title = paste0("Top ", top_n, " Pathways for ", CD_subtype)
    ) +
    theme(strip.text = element_text(face = "bold"))
}

plots_list <- lapply(childhood, function(cd) {
  plot_top_pathways_for_CD(cluster.gprofiler.results, CD_subtype = cd)
})

names(plots_list) <- childhood

for (p in plots_list) {
  print(p)
}
```

# what genes are in the subnetworks which are drug targetable?

Here, i obtained the drug targets data from [omicsynth](https://doi.org/10.1016/j.ajhg.2023.12.006) to see what genes in the subnetwork have a known drug target. 

The plots below show the network proximity score (higher means more proximal to network) vs the evidence of drug interaction (higher interaction = higher score)

```{r}
# the genes in the subnetworks
gene.proximitiy.scores <- node_metadata %>% 
  lapply(function(x) {
    x %>% 
      dplyr::select(gene_name = X, z_both) %>% 
      as_tibble    
  })
  

omicsynth <- read.csv("data/omicsynth/drug_genome_dgidb.csv") 

gene.proximitiy.scores %>% 
  lapply(function(x) {
    omicsynth %>% # has the drug info 
      # only reatin genes in subnet
      dplyr::filter(gene_name %in% x$gene_name) %>% 
      as_tibble %>% 
      left_join(x)
  }) %>% 
  bind_rows(.id = "comparison") %>% 
  mutate(
    CD = purrr::map_chr(comparison, ~ {
      match <- childhood[startsWith(.x, childhood)]
      if (length(match) > 0) match[1] else NA_character_
    }), 
    .after = comparison
  ) %>% 
  dplyr::filter(comparison %in% sig.networks) %>% 
  ggplot(
    aes(x = interaction_group_score, y = z_both)
  ) +
  geom_point() +
  geom_label_repel(
    aes(label = drug_claim_name), 
    data = . %>% 
      dplyr::filter(z_both > 5 & interaction_group_score > 5)
  ) +
  labs(
    x = "Interaction group score", 
    y = "Network proximity z score"
  ) + 
  facet_wrap(~comparison, scales = "free")

plot_drug_vs_proximity_by_CD <- function(gene_proximity_list, omicsynth, CD_subtype, sig.networks, childhood) {
  gene_proximity_list %>% 
    lapply(function(x) {
      omicsynth %>%
        dplyr::filter(gene_name %in% x$gene_name) %>%
        as_tibble() %>%
        left_join(x)
    }) %>%
    bind_rows(.id = "comparison") %>%
    mutate(
      CD = purrr::map_chr(comparison, ~ {
        match <- childhood[startsWith(.x, childhood)]
        if (length(match) > 0) match[1] else NA_character_
      }),
      .after = comparison
    ) %>%
    dplyr::filter(CD == CD_subtype, comparison %in% sig.networks) %>%
    ggplot(aes(x = interaction_group_score, y = z_both)) +
    geom_point() +
    geom_label_repel(
      aes(label = drug_claim_name),
      data = . %>% dplyr::filter(z_both > 5 & interaction_group_score > 5)
    ) +
    labs(
      x = "Interaction group score", 
      y = "Network proximity z score",
      title = paste("Drug proximity scores -", CD_subtype)
    ) +
    facet_wrap(~comparison, scales = "free")
}

for (cd in childhood) {
  p <- plot_drug_vs_proximity_by_CD(
    gene_proximity_list = gene.proximitiy.scores,
    omicsynth = omicsynth,
    CD_subtype = cd,
    sig.networks = sig.networks,
    childhood = childhood
  )
  print(p)
}
```













