---
title: "DLB_snRNAseq"
author: "Karissa Barthelson"
date: "2025-07-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

# setup
```{r}
knitr::opts_chunk$set(
  message = F, echo = F, warning = F
)
```

```{r}
library(tidyverse)
library(magrittr)
library(readxl)
library(qs) # for fast save and reading of objects into memory (alternative to rds)
library(parallel) 
library(Matrix)
library(GEOquery)

library(Seurat)
library(DropletUtils)
library(DoubletFinder)

# vis
library(pander)
library(ggpubr)

theme_set(theme_light() +
            theme(plot.title = element_text(hjust = 0.5), 
                  strip.text = element_text(face = "bold"),
                  strip.background = element_rect(fill = "grey20")
            )
)

```

# Introduction
Here, I will reanalyse the snRNAseq data generated by [Feleke et al. (2021)](https://doi.org/10.1007/s00401-021-02343-x): single-nucleus RNA-sequencing data of  anterior cingulate cortex samples derived from 28 individuals, including healthy controls, PD, PDD and DLB cases (n = 7 per group). 

The raw cellranger output data was downloaded from GEO (GSE178146) for all samples. 

# sample meta from GEO
sample metadata was obtained from GEO. Only disease state was recorded. No sex, age etc. This information is noted in the paper, but which sample is which was not reported... 

```{r}
# read in smaple metadata
gse = getGEO(filename="data/public data/GSE178146_RAW/GSE178146_series_matrix.txt")

meta <- pData(gse) %>% 
  as_tibble %>% 
  dplyr::select(
    sample = title, 
    geo_accession,
    region = source_name_ch1, 
    disease = characteristics_ch1
  ) %>% 
  mutate(disease = str_remove(disease, pattern = "disease state: "))
  

pander(meta, caption = "sample metadata")    

meta %>% 
  group_by(disease) %>% 
  summarise(n = n()) %>% 
  pander(caption = "Number of samples per disease state")
```

# preprocessing
## filter empty droplets

I want to filter out the empty droplets in each sample. Since the ambient RNA profile may be different between samples (as the authors processed each brain tissue seperately), i want to filter each sample independently. 

To identify the empty droplets, I used the `emptyDrops` from `DroplitUtils.` `EmptyDrops()` uses Monte Carlo simulations to compute p-values for the multinomial sampling transcripts from the ambient pool. Here, I selected a background number of UMIs as 300, to match the params of Feleke et al. 

This was performed in the `DLB snRNAseq emptydrops.R` script. The output is imported below. 

```{r}

data.seurat <- 
  qread(
    "data/public data/GSE178146_RAW/R_objects/data_seurat_DLB_PD_snRNAseq_emptydrops.qs")


colnames(data.seurat) %<>% 
  str_remove("_raw_feature_bc_matrix")


samples.by.disease <- meta %>% 
  split(f = .$disease) %>% 
  lapply(function(x) {x$sample})

  
data.seurat@meta.data %<>% 
  mutate(disease = case_when(
    orig.ident %in% samples.by.disease$Control ~ "control",
    orig.ident %in% samples.by.disease$DLB ~ "DLB", 
    orig.ident %in% samples.by.disease$PD ~ "PD", 
    orig.ident %in% samples.by.disease$PDD ~ "PDD"
  ))

# tidyup env
gc()
```
## fitltering from mitochondrial transcript content
Since this is snRNAseq data, we should not see a lot of transcripts derived from mtDNA.
```{r}
# Calculate percent mit
data.seurat[["percent.mt"]] <- 
   PercentageFeatureSet(data.seurat, pattern = "^MT-")

```

```{r}
VlnPlot(data.seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

```{r}
data.seurat <- subset(
  data.seurat, 
  subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 1)
```

## seurat workflow
```{r}
# first, run the seurat wflw up to the PCA step
data.seurat %<>% 
  NormalizeData() %>% 
  FindVariableFeatures %>% 
  ScaleData() %>% 
  RunPCA 

# work out the number of PCs needed for dimension reduction. 
# ElbowPlot(data.seurat)
#DimPlot(data.seurat, reduction = "pca", shuffle = T) 
#DimHeatmap(data.seurat, dims = 1:20, cells = 500, balanced = TRUE)
# 
# temp <- data.seurat %>%
#    FindNeighbors(dims = 1:20) %>% 
#    FindClusters(resolution = c(0.1, 0.3, 0.5) ) 
#  
#  ggarrange(
#    DimPlot(temp, group.by = "RNA_snn_res.0.1", label = TRUE, shuffle = TRUE), 
#    DimPlot(temp, group.by = "RNA_snn_res.0.3", label = TRUE, shuffle = TRUE), 
#    DimPlot(temp, group.by = "RNA_snn_res.0.5", label = TRUE, shuffle = TRUE)
#  )
#remove(temp)

data.seurat %<>% 
  FindNeighbors(dims = 1:20) %>% 
  FindClusters(resolution = 0.2) %>% 
  RunUMAP(dims = 1:20)

Idents(data.seurat) <- "RNA_snn_res.0.2"

DimPlot(data.seurat, reduction = "umap", group.by = "orig.ident", shuffle = TRUE) + 
  ggtitle("individual") +
  DimPlot(data.seurat, reduction = "umap", group.by = "disease", shuffle = TRUE) + 
  DimPlot(data.seurat, reduction = "umap", shuffle = TRUE, group.by = "RNA_snn_res.0.2")
```

# Look at marker gennes
## define top marker genes per cluster
```{r}
markers_v1 <- data.seurat %>% 
  JoinLayers() %>% 
  FindAllMarkers()

markers_v1 %<>% 
  split(f = .$cluster) %>% 
  lapply(
    as_tibble
  )
```

### Cluster 0 - oligodend.
the top 10 marker genes in this cluster are below. I had a look in the literature and the HPA (https://www.proteinatlas.org/) to assess cell type.
```{r}
markers_v1$`0` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  .$gene

tribble(
  ~gene, ~celltype, ~ref, 
  "ST18", "oligodend", "https://doi.org/10.1016/j.neurobiolaging.2021.01.021",
  "MBP", "oligodend", "known",
  "PLP1", "oligodend", "known",
  "TMEM144", "oligodend", "HPA",
  "RNF220", "oligodend", "HPA"
)
```

Looked at some known markers of oligodend - def omstly in cluster 0. 
```{r}
DimPlot(data.seurat, reduction = "umap", label = TRUE)  +
  
  FeaturePlot(
    data.seurat,
    features= c("MBP","MOBP", "PLP1"),
    combine=FALSE
  )  

VlnPlot(data.seurat, features=c("MBP","MOBP", "PLP1"), ncol = 1, pt.size = 0)
```


### Cluster 1: astrocytes
```{r}
markers_v1$`1` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) 

tribble(
  ~gene, ~celltype, ~ref, 
  "ADGRV1", "astrocyte", "HPA",
  "SLC1A2", "astrocyte", "HPA",
  "RYR3", "astrocyes and neurons", "HPA",
  "PITPNC1", "astrocyte", "HPA",
  "GLIS3", "atrocyte and Ependymal cells ", "HPA"
)
```

### Cluster 2: Neuron - mostly excite
```{r}
markers_v1$`2` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  dplyr::select(gene, avg_log2FC)


tribble(
  ~gene, ~celltype, ~ref, 
  "CBLN2", "Ex.neuron", "Allen brain map",
  "KCNQ5", "Neuron - Ex and inhub", "Allen brain map",
  "DLGAP2", "both neurons", "Allen brain map",
  "CHRM3", "both", "ABM",
  "RALYL", "bpth ", "ABM", 
  "LB2", "Ex.neuron", "ABM"
)
```


```{r}
DimPlot(data.seurat, reduction = "umap", label = TRUE) + 
  
  FeaturePlot(
    data.seurat,
    features= c("GFAP","AQP4"),
    combine=FALSE
  )  

VlnPlot(data.seurat, features=c("GFAP","AQP4"), pt.size = 0)
```


### Cluster 3: OPCs
```{r}
markers_v1$`3` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  dplyr::select(gene, avg_log2FC)


tribble(
  ~gene, ~celltype, ~ref, 
  "LHFPL3", "OPC", "HPA",
  "LHFPL3", "OPC or some in.neuron", "ABM",
  "VCAN", "OPC", "HPA",
  "PCDH15", "OPC", "ABM",
  "PCDH15", "OPC", "HPA",
  "TNR", "OPC", "ABM", 
  "TNR", "OPC", "HPA", 
  "LUZP2", "OPC and some neurons", "ABM",
  "LUZP2", "OPC and some neurons", "HPA",
  "PTPRZ1", "OPC"
  
)
```

```{r}
VlnPlot(data.seurat, features= c("OLIG2","OLIG1"), pt.size = 0, ncol = 1) 
```

### Cluster 4:  Ex.neuron
```{r}
markers_v1$`4` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  dplyr::select(gene, avg_log2FC)
```

```{r}
VlnPlot(data.seurat, 
        features= c("CAMK2A","NRGN", "SLC17A7"), 
        pt.size = 0, 
        ncol = 1) 
```
Mostly exc. neurons according to ABM

### Cluster 5: Microglia
```{r}
markers_v1$`5` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  dplyr::select(gene, avg_log2FC)
```

```{r}
VlnPlot(data.seurat, 
       features= c("C3","CD74", "CSF1R", "AIF1"),
        pt.size = 0, 
        ncol = 1) 
```
Microglia

### Cluster 6: Neuron - mostly inhib. maybe mixed
```{r}
markers_v1$`6` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  dplyr::select(gene, avg_log2FC)
```

```{r}
VlnPlot(data.seurat, 
       features= c("GAD1","GAD2"),
        pt.size = 0, 
        ncol = 1) 
```

### Cluster 7: Astrocytes
```{r}
markers_v1$`7` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  dplyr::select(gene, avg_log2FC)
```

```{r}
VlnPlot(data.seurat, 
       features= c("GFAP","AQP4"),
        pt.size = 0, 
        ncol = 1) 
```

### Cluster 8: Mostly inhibt
```{r}
markers_v1$`8` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  dplyr::select(gene, avg_log2FC)
```

```{r}
VlnPlot(data.seurat, 
       features= c("GAD1","GAD2"),
        pt.size = 0, 
        ncol = 1) 
```

```{r}
DimPlot(data.seurat, reduction = "umap", label = TRUE, split.by = "disease",group.by = "celltype")
```


### Cluster 9: Mostly exc
```{r}
markers_v1$`9` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  dplyr::select(gene, avg_log2FC)
```

```{r}
VlnPlot(data.seurat, 
        features= c("CAMK2A","NRGN", "SLC17A7"), 
        pt.size = 0, 
        ncol = 1) 
```


### Cluster 10: OPC again
```{r}
markers_v1$`10` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  dplyr::select(gene, avg_log2FC)
```

```{r}
VlnPlot(data.seurat, features= c("OLIG2","OLIG1"), pt.size = 0, ncol = 1) 
```

### Cluster 11: Microglia
```{r}
markers_v1$`11` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  dplyr::select(gene, avg_log2FC)
```

```{r}
VlnPlot(data.seurat, 
        features= c("C3","CD74", "CSF1R", "AIF1"),
        pt.size = 0, 
        ncol = 1) 
```

### Cluster 12: Vascular leptomeningeal cells
```{r}
markers_v1$`12` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  dplyr::select(gene, avg_log2FC)
```

```{r}
VlnPlot(data.seurat, 
        features= c("COL1A2","COLEC12"),
        pt.size = 0, 
        ncol = 1) 
```

### Cluster 13: Exc.Neuron 
```{r}
markers_v1$`13` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  dplyr::select(gene, avg_log2FC)
```

### Cluster 14: Endothelial cells
```{r}
markers_v1$`14` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  dplyr::select(gene, avg_log2FC)
```

```{r}
VlnPlot(data.seurat, 
        features= c("CLDN5","FLT1"),
        pt.size = 0, 
        ncol = 1) 
```

### Cluster 14: Exc.Neuron
```{r}
markers_v1$`15` %>% 
  dplyr::filter(avg_log2FC >0) %>% 
  head(10) %>% 
  dplyr::select(gene, avg_log2FC)
```

## assign celltypes
```{r}
data.seurat@meta.data %<>% 
  mutate(
    celltype = case_when(
      RNA_snn_res.0.2 == "0" ~ "Oligodendrocyte",
      RNA_snn_res.0.2 == "1" ~ "Astrocyte",
      RNA_snn_res.0.2 == "2" ~ "Neuron_Ex",
      RNA_snn_res.0.2 == "3" ~ "OPC",
      RNA_snn_res.0.2 == "4" ~ "Neuron_Ex",
      RNA_snn_res.0.2 == "5" ~ "Microglia",
      RNA_snn_res.0.2 == "6" ~ "Neuron_In",
      RNA_snn_res.0.2 == "7" ~ "Astrocyte",
      RNA_snn_res.0.2 == "8" ~ "Neuron_In",
      RNA_snn_res.0.2 == "9" ~ "Neuron_Ex",
      RNA_snn_res.0.2 == "10" ~ "OPC",
      RNA_snn_res.0.2 == "11" ~ "Microglia",
      RNA_snn_res.0.2 == "12" ~ "VLMCs",
      RNA_snn_res.0.2 == "13" ~ "Neuron_Ex",
      RNA_snn_res.0.2 == "14" ~ "Endothelial",
      RNA_snn_res.0.2 == "15" ~ "Neuron_Ex"
    )
    )
  
DimPlot(data.seurat, reduction = "umap", group.by = "celltype",label = TRUE, label.box = T)
```

# export files
```{r}
data.seurat %>% 
  qsave("data/public data/GSE178146_RAW/R_objects/PD_DLB_data_seurat.qs", 
        preset = "fast", 
        nthreads = 4)
```

# remove doublets
To help cleanup the data, I tried removing doublets from the dataset. This needs to be run on each sample individually (since this is how the doublets would appear). i uses doubletdetector. as it is the most accurate algo as demonstrated by [Xi et al.](https://doi.org/10.1016/j.cels.2020.11.008). However, it did not seem to improve the clustering. Code is not shown here, but is in the source code of the doc. 

## figs
```{r}
DimPlot(data.seurat, reduction = "umap", 
        group.by = "celltype", repel = T,
        label = TRUE, label.box = T) +
  theme(aspect.ratio = 1)
  # ggsave(
  #   "output/DLB_snRNAseq/figs/talk/umap_bycelltype.png", 
  #   width = 10, height = 10, units = "cm",
  #   scale = 2)
```

```{r}
 FeaturePlot(
    seurat_objects$human_DLB,
    features= c("NRGN", "MAP2", "GFAP",
                "MBP","MOBP", "PLP1",
                "C3","CD74", "CSF1R", 
                "PDGFRA","VCAN", "FLT1"
                
                ), 
    combine=T, ncol = 3
  ) +
   ggsave(
    "output/DLB_snRNAseq/figs/talk/umap_bymarkergenes.png", 
    width = 10, height = 10, units = "cm",
    scale = 3)

```




<!--# remove doublets
To help cleanup the data, will now remove the doublets from the dataset. This needs to be run on each sample individually (since this is how the doublets would appear). Here, i will use doubletdetector. as it is the most accurate algo as demonstrated by [Xi et al.](https://doi.org/10.1016/j.cels.2020.11.008). 

The code below is adapted from [Kennedi Todd and Kim Olney tutorial](https://rpubs.com/kenneditodd/doublet_finder_example). 
```{r}
data.seurat.split <- SplitObject(data.seurat, split.by = "orig.ident") 

# loop through samples to find doublets
for (i in 1:length(data.seurat.split)) {
  # print the sample we are on
  print(paste0("Sample ",i))
  
  # Pre-process seurat object with standard seurat workflow
  brain.sample <- NormalizeData(data.seurat.split[[i]])
  brain.sample <- FindVariableFeatures(brain.sample)
  brain.sample <- ScaleData(brain.sample)
  brain.sample <- RunPCA(brain.sample, nfeatures.print = 10)
  
  # Find significant PCs
  stdv <- brain.sample[["pca"]]@stdev
  sum.stdv <- sum(brain.sample[["pca"]]@stdev)
  percent.stdv <- (stdv / sum.stdv) * 100
  cumulative <- cumsum(percent.stdv)
  co1 <- which(cumulative > 90 & percent.stdv < 5)[1]
  co2 <- sort(which((percent.stdv[1:length(percent.stdv) - 1] - 
                       percent.stdv[2:length(percent.stdv)]) > 0.1), 
              decreasing = T)[1] + 1
  min.pc <- min(co1, co2)
  min.pc
  
  # finish pre-processing
  brain.sample <- RunUMAP(brain.sample, dims = 1:min.pc)
  brain.sample <- FindNeighbors(object = brain.sample, dims = 1:min.pc)              
  brain.sample <- FindClusters(object = brain.sample, resolution = 0.1)
  
  # pK identification (no ground-truth)
  sweep.list <- paramSweep(brain.sample, PCs = 1:min.pc, num.cores = detectCores() - 1)
  sweep.stats <- summarizeSweep(sweep.list)
  bcmvn <- find.pK(sweep.stats)
  
  # Optimal pK is the max of the bomodality coefficent (BCmvn) distribution
  bcmvn.max <- bcmvn[which.max(bcmvn$BCmetric),]
  optimal.pk <- bcmvn.max$pK
  optimal.pk <- as.numeric(levels(optimal.pk))[optimal.pk]
  
  ## Homotypic doublet proportion estimate
  annotations <- brain.sample@meta.data$seurat_clusters
  homotypic.prop <- modelHomotypic(annotations) 
  nExp.poi <- round(optimal.pk * nrow(brain.sample@meta.data)) ## Assuming 7.5% doublet formation rate - tailor for your dataset
  nExp.poi.adj <- round(nExp.poi * (1 - homotypic.prop))
  
  # run DoubletFinder
  brain.sample <- doubletFinder(seu = brain.sample, 
                                   PCs = 1:min.pc, 
                                   pK = optimal.pk,
                                   nExp = nExp.poi.adj)
  metadata <- brain.sample@meta.data
  colnames(metadata)[grepl("^DF", colnames(brain.sample@meta.data))] <- "doublet_finder"
  brain.sample@meta.data <- metadata 
  
  
  # subset and save
  singlets <- subset(brain.sample, doublet_finder == "Singlet")
  data.seurat.split[[i]] <- singlets
  remove(singlets)
}

data.seurat.singlets <- merge(
  x = data.seurat.split[[1]], 
  y = data.seurat.split[-1], 
  add.cell.ids = names(data.seurat.split)
)
```

# Look at marker gennes
## excitory neurons
```{r}
ggarrange(
VlnPlot(data.seurat, features= c("CAMK2A","NRGN", "SLC17A7"), pt.size = 0) +
  ggtitle("before doublet"), 
VlnPlot(data.seurat.singlets, features= c("CAMK2A","NRGN", "SLC17A7"), pt.size = 0) +
  ggtitle("after doublet"), 
ncol = 1
)

```

## inhib neurons
```{r}
ggarrange(
VlnPlot(data.seurat, features= c("GAD1","GAD2"), pt.size = 0) +
  ggtitle("before doublet"), 
VlnPlot(data.seurat.singlets, features= c("GAD1","GAD2"), pt.size = 0) +
  ggtitle("after doublet"), 
ncol = 1
)

```

## astrocyte
```{r}
ggarrange(
VlnPlot(data.seurat, features= c("GFAP","AQP4"), pt.size = 0) +
  ggtitle("before doublet"), 
VlnPlot(data.seurat.singlets, features= c("GFAP","AQP4"), pt.size = 0) +
  ggtitle("after doublet"), 
ncol = 1
)
```

## Oligodend
```{r}

ggarrange(
VlnPlot(data.seurat, features=  c("MBP","MOBP", "PLP1"), pt.size = 0) +
  ggtitle("before doublet"), 
VlnPlot(data.seurat.singlets, features=  c("MBP","MOBP", "PLP1"), pt.size = 0) +
  ggtitle("after doublet"), 
ncol = 1
)
```

## Microglia
```{r}
ggarrange(
VlnPlot(data.seurat, features=  c("C3","CD74", "CSF1R", "AIF1"), pt.size = 0) +
  ggtitle("before doublet"), 
VlnPlot(data.seurat.singlets, features=   c("C3","CD74", "CSF1R", "AIF1"), pt.size = 0) +
  ggtitle("after doublet"), 
nrow = 2
)

```

## Vascular
```{r}
ggarrange(
VlnPlot(data.seurat, features= c("CLDN5","FLT1"), pt.size = 0) +
  ggtitle("before doublet"), 
VlnPlot(data.seurat.singlets, features=   c("CLDN5","FLT1"), pt.size = 0) +
  ggtitle("after doublet"), 
nrow = 2
)
```

I dont think the doublet removal improved the dataset. 

-->
