---
title: "Home"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

This website describes how we analysed the genetics of adult- and childhood-onset dementia using network propogation and colocalisation to identify drugs to reposition for these diseases. 

We first looked at the genetics of adult-onset dementia (AD, PD, ALS, FTLD, DLB) and the childhood dementia subgroups described in [Elvidge et al. Brain](https://academic.oup.com/brain/article/146/11/4446/7226999)
```{r}
knitr::opts_chunk$set(
  message = F, echo = F, warning = F
)
```

- [neurodegen network: adult all vs childhood all](adultvCD.html)
- [subgroup networks: adult vs childhood subgroups](adultvCDsubgroups.html)
- [DLB human snRNAseq analysis](DLB_snRNaseq.html)


## single cell seq datasets
These are the scRNAseq datasets we will/are using to dissect the network genes in human adult neurodegen.
```{r}
library(tidyverse)
tribble(
  ~disease, ~region, ~link,
  "DLB, PD", "anterior cingulate cortex (BA24,32,33)", "https://pubmed.ncbi.nlm.nih.gov/34309761/", 
  
  "FTLD, ALS", "primary motor (BA4) and dorsolateral prefrontal cortices (BA9)", "https://www.cell.com/cell/fulltext/S0092-8674(24)00234-4#sec-4-2-3", 
  
  "AD", "AG (BA39), middle temporal gyrus (BA22), PFC: frontal pole (BA10), EC: entorhinal cortex (BA28), Posterior hippocampus (HC), TH: thalamus ","https://www.nature.com/articles/s41586-024-07606-7#Sec10"
  
  
)

```

