---
title: "netcoloc_paper"
author: "Karissa Barthelson"
date: "2026-02-26"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
```{r}
library(tidyverse)
library(magrittr)
library(readxl)
library(qs)

# vis
library(ComplexUpset)
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
library(UpSetR)
library(pander)
library(ggeasy)
library(ggfortify)
library(ggpubr)
library(ggrepel)
library(plotly)
library(grid)
library(gridExtra)
library(gridGraphics)
library(viridis)
library(tidytext)
library(pheatmap)
library(patchwork)

# annotations
library(org.Hs.eg.db)
library(org.Mm.eg.db)
library(MPO.db)
library(homologene)
library(ndexr)


# enrichemnt
library(clusterProfiler)
library(gprofiler2)
library(ReactomePA)
library(DOSE)

# analysis
library(Seurat)

theme_set(theme_light() +
            theme(plot.title = element_text(hjust = 0.5), 
                  strip.text = element_text(face = "bold", colour = "white"),
                  strip.background = element_rect(fill = "grey20")
            )
)

knitr::opts_chunk$set(
  message = F, echo = F, warning = F
)
```

# Introduction
Here will be the analysis of netcoloc in the story of the paper. 
# gene sets


## Pos & neg control colocs
### import data

Here, I will import the dfs needed for comparing our pos and neg control disease comparisons: 

- AD and PD 
- lysosomal gene lists - neuronopathic and non-neuronopathic
- mitochondrial
- peroxisomal
- amino acidopahties

The input gene lists for the adult onset diseases will be the GWAS 1 hit per loci gene sets

```{r}
nodes <- readRDS("data/R_objects/network/nodes.rds")

paths.netcolc = c(
  AD_PD = "data/results/pan neurodegen analysis - adult v2/AD_PD/network_enrichment.csv",
  #AD_T1D = "data/results/AD only/AD_T1D/network_enrichment.csv",
  # CD.lyso_nonCD.lyso = "data/results/CD subtypes and non neurodegen IEMs/CD_1A_lyso_nonCD_lyso/network_colocalization_enrichment.csv", 
  # CD.amino_nonCD.amino = "data/results/CD subtypes and non neurodegen IEMs/CD_1C_aminoacids_aminoacidopathy/network_colocalization_enrichment.csv", 
  # CD.perox_nonCD.perox =  "data/results/CD subtypes and non neurodegen IEMs/CD_1F_perox_peroxisomal/network_colocalization_enrichment.csv",
  # CD.mito_nonCD.mito = "data/results/CD subtypes and non neurodegen IEMs/CD_1G_mito_mitochondrial/network_colocalization_enrichment.csv",
  CD.all_brain.malform = "data/results/pan neurodegen analysis - adult v2/CD_all_brain_malform/network_enrichment.csv")

enrichment_dfsv1 <- paths.netcolc %>% 
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y)) %>% 
  bind_rows() %>%
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>%
  pivot_longer(
    cols = c(observed_overlap, expected_overlap_mean),
    names_to = "type",
    values_to = "value"
  ) %>%
  mutate(
    type = recode(
      type,
      "observed_overlap" = "observed",
      "expected_overlap_mean" = "expected"), 
    comparison = str_replace(comparison, "_", "\n&\n")
  )
```

## plot

we then plotted the observed vs expected ratio of overlapping genes for each comparison, with the FDR of the enrichment test shown as a label on the expected bar. the neg control (CD vs brain malform) did not coloc. 

another idea could be to run migraine as another neg control. 

the pos controls of CD and their non neuronopathic forms did coloc. Also AD and PD coloc. 

```{r}
enrichment_dfsv1 %>%
  mutate(
    FDR = p.adjust(empirical_p), 
    control = case_when(
      comparison == "CD.all\n&\nbrain.malform" ~ "negative",
      TRUE ~ "positive"
    )
  ) %>% 
  arrange(control) %>% #.$comparison %>% unique
  mutate(
    comparison = str_replace(comparison, "CD.all", "All childhood dementias"),
    comparison = str_replace(comparison, "brain.malform", "Brain malformations"),
    comparison = str_replace(comparison, "adult.all", "All adult-onset neurodegen"),
    comparison = str_replace(comparison, "CD.lyso", "Lysosomal\nchildhood dementias"),
    comparison = str_replace(comparison, "nonCD.lyso", "Non-neuronopathic\nlysosomal diseases"),
    comparison = str_replace(comparison, "CD.amino", "Disorders of amino acid\n& other organic acid\nmetabolism"),
    comparison = str_replace(comparison, "nonCD.amino", "Non-neuronopathic\namino acid diseases"),
    comparison = str_replace(comparison, "CD.perox", "Peroxisomal\nchildhood dementias"),
    comparison = str_replace(comparison, "nonCD.perox", "Non-neuronopathic\nperoxisomal diseases"),
    comparison = str_replace(comparison, "CD.mito", "Mitochondrial\nchildhood dementias"),
    comparison = str_replace(comparison, "nonCD.mito", "Non-neuronopathic\nmitochondrial diseases"),
    
    FDR = case_when(
      FDR == 0 ~ "p.FDR < 0.000001", 
      FDR == 1 ~ "p.FDR = 1", 
      TRUE ~ FDR %>% sprintf("p.FDR = %.1e", .)
    ),
    comparison = fct_inorder(comparison)
  ) %>% 
  ggplot(
    aes(x = type, y = value, fill = type)
    ) +
  geom_bar(
    stat = "identity", position = "dodge", width = 0.8
  ) +
  geom_errorbar(
    data = . %>% 
      dplyr::filter(type == "expected"),
    aes(
      ymin = value - 2 * expected_overlap_std,
      ymax = value + 2 * expected_overlap_std),
    width = 0.25
  ) + 
  geom_label(
    data =  . %>%
      dplyr::filter(type == "expected"),
    aes(
      x = 1.5, y =800,
      label = FDR),
    inherit.aes = FALSE,
    fill = NA,
    border.colour = NA,
    colour = "grey20",
    size = 5,
    vjust = 0,
    fontface = "bold",
    show.legend = FALSE
  )  +
  labs(
    y = "Number of genes", 
    x = ""
  ) +
  facet_wrap(~comparison, nrow = 1, scales = "free") +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("grey50", "skyblue"))+
   theme(
    strip.text = element_text(size = 12),
    legend.position = "none",
    axis.title.x = element_blank(),
    plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) 
```

## enriched pathways in the positive controls

I next wanted to see that the genes in these subnetworks were enriched with the right kinds of genes. 

so i performed ORA on the subnetworks using the KEGG pathways. 

```{r}
## obtain the list of genes within each subnetwork. 

network.genes.pos.controls <- c(
  AD_PD = "data/results/pan neurodegen analysis - adult v2/AD_PD/overlapping_nodes_metadata.csv",
  CD.lyso_nonCD.lyso = "data/results/CD subtypes and non neurodegen IEMs/CD_1A_lyso_nonCD_lyso/overlapping_nodes_metadata.csv", 
  CD.amino_nonCD.amino = "data/results/CD subtypes and non neurodegen IEMs/CD_1C_aminoacids_aminoacidopathy/overlapping_nodes_metadata.csv", 
  CD.perox_nonCD.perox =  "data/results/CD subtypes and non neurodegen IEMs/CD_1F_perox_peroxisomal/overlapping_nodes_metadata.csv",
  CD.mito_nonCD.mito = "data/results/CD subtypes and non neurodegen IEMs/CD_1G_mito_mitochondrial/overlapping_nodes_metadata.csv"
  # CD.all_adult.all = "data/results/pan neurodegen analysis - adult v2/CD_all_adult_allv2/overlapping_nodes_metadata.csv"
  ) %>% 
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y)) %>% 
  lapply(function(x) {
    x %>% 
      as_tibble %>% 
      pull(X)
  })

# the background gene set will be all genes found in the PCNet, as these are the ones I am analysing. The node info was obtained from NDex. 

```



```{r}
ORA.network.genes.poscontrols.results <- network.genes.pos.controls %>% 
  lapply(runKEGG)
```

### vis results

Yes good - all what is expected
```{r}
ORA.network.genes.poscontrols.results %>% 
  map(as_tibble) %>% 
  bind_rows(.id = "comparison") %>% 
  separate(GeneRatio,into = c("temp1", "temp2")) %>% 
  mutate(GeneRatio = as.numeric(temp1)/as.numeric(temp2)) %>% 
  group_by(comparison) %>% 
  dplyr::slice(1:10) %>% 
  mutate(
    Description = reorder_within(Description, GeneRatio, comparison), 
    Description = str_wrap(Description, width = 22)
    ) %>%
  ggplot(
    aes(
      y = reorder(Description, GeneRatio), x = GeneRatio
    )
  ) +
  geom_col(
    aes(fill = -log10(p.adjust))
  ) +
  scale_y_reordered() +  # This removes the `___comparison` suffix added by reorder_within()
  scale_fill_viridis_c() +
  facet_wrap(~comparison, 
             scales = "free_y", nrow = 1)

```

# CD all and adult all

Now I want to do a pan-neurodegenerative analssis - all the CD and adult genes combined. I looked at whether the adult and childhood-onset dementia genetics sig coloc in the network space. yes they do. 

```{r}
input.f2a = read_csv("data/results/pan neurodegen analysis - adult v2/CD_all_adult_allv2/network_enrichment.csv") %>% 
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>%
  pivot_longer(
    cols = c(observed_overlap, expected_overlap_mean),
    names_to = "type",
    values_to = "value"
  ) %>%
  mutate(
    type = recode(
      type,
      "observed_overlap" = "observed",
      "expected_overlap_mean" = "expected"), 
    comparison = "Childhood dementia\n&\nadult-onset\nneurodegen",
    FDR = case_when(
      empirical_p < 0.0001 ~ "p.FDR < 0.0001"
    )
  ) 
  
input.f2a %>% 
ggplot(
    aes(x = type, y = value, fill = type)
    ) +
  geom_bar(
    stat = "identity", position = "dodge", width = 0.8
  ) +
  geom_errorbar(
    data = . %>% 
      dplyr::filter(type == "expected"),
    aes(
      ymin = value - 2 * expected_overlap_std,
      ymax = value + 2 * expected_overlap_std),
    width = 0.25
  ) + 
  geom_label(
    data =  . %>%
      dplyr::filter(type == "expected"),
    aes(
      x = 1.5, y =800,
      label = FDR),
    inherit.aes = FALSE,
    fill = NA,
    border.colour = NA,
    colour = "grey20",
    size = 5,
    vjust = 0,
    fontface = "bold",
    show.legend = FALSE
  )  +
  labs(
    y = "Number of genes", 
    x = ""
  ) +
  facet_wrap(~comparison, nrow = 1, scales = "free") +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("grey50", "skyblue"))+
   theme(
    strip.text = element_text(size = 12),
    legend.position = "none",
    axis.title.x = element_blank(),
    plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) 
  
  
```

### ORA KEGG of pan-neurodegen
```{r}
# extract the genes in the pan-neurodegen subnetwork
network.genes.pan.neurodegen <-  "data/results/pan neurodegen analysis - adult v2/CD_all_adult_allv2/overlapping_nodes_metadata.csv" %>% 
  read.csv %>%
  mutate(
     comparison = "Childhood dementia\n&\nadult-onset\nneurodegen"
  ) %>% 
  as_tibble %>% 
  pull(X)

# Run ORA of KEGG pathways
ORA.network.genes.pan.peurodegen.results = runKEGG(network.genes.pan.neurodegen)
```

```{r}
ORA.network.genes.pan.peurodegen.results %>% 
  as_tibble %>% 
  mutate(
    comparison = "Childhood dementia\n&\nadult-onset\nneurodegen"
  ) %>% 
  separate(GeneRatio,into = c("temp1", "temp2")) %>% 
  mutate(GeneRatio = as.numeric(temp1)/as.numeric(temp2)) %>% 
  group_by(comparison) %>% 
  dplyr::slice(1:10) %>% 
  mutate(
    Description = reorder_within(Description, GeneRatio, comparison), 
    Description = str_wrap(Description, width = 22)
  ) %>%
  ggplot(
    aes(
      y = reorder(Description, GeneRatio), x = GeneRatio
    )
  ) +
  geom_col(
    aes(fill = -log10(p.adjust))
  ) +
  scale_y_reordered() +  # This removes the `___comparison` suffix added by reorder_within()
  scale_fill_viridis_c() +
  labs(
    y = "KEGG pathway"
  ) +
  facet_wrap(~comparison, 
             scales = "free_y", nrow = 1)
```

## ORA of MPO terms

We can also run mouse MPO enrichment analysis
```{r}
# this was obtained fromhttps://www.informatics.jax.org/downloads/reports/MGI_PhenoGenoMP.rpt

mm_symbol2mp_id = 
  read_delim(
    "data/MPO/HMD_HumanPhenotype.txt", delim = "\t") %>% 
  dplyr::select(1,4) %>% 
  set_colnames(
    c("mm_symbol", "MP_id")
  ) %>% 
  mutate(
    mm_symbol = str_remove(mm_symbol, pattern = "<.+")
  )

# mpid 2 mp name
mpo_mgi <- AnnotationDbi::select(
  MPO.db,
  keys    = keys(MPO.db, keytype = "mpid"),
  columns = c("mpid", "term"),
  keytype = "mpid"
) %>% 
  dplyr::select(
    MP_id = mpid, mp_full_name = term
  ) %>% 
  as_tibble

# mouse name to human name
mm2hs <- homologene(
  genes  = unique(mm_symbol2mp_id$mm_symbol),
  inTax  = 10090,  # mouse
  outTax = 9606    # human
) %>% 
  dplyr::select(
    mm_symbol = `10090`, hs_symbol = `9606`, 
    mm_entrez = `10090_ID`, hs_entrez = `9606_ID`
  ) %>% 
  mutate(across(everything(), as.character)) %>% 
  as_tibble

# join everything
mpo.df <- mm_symbol2mp_id %>% 
   left_join(mpo_mgi) %>% 
  left_join(mm2hs)

mpo.df.hs.entrez = 
  mpo.df %>% 
  dplyr::filter(hs_symbol %in% nodes) %>% 
  dplyr::select(mp_full_name, hs_entrez) %>% 
  arrange(mp_full_name)

```

```{r}
ORA.network.genes.pan.peurodegen.results.MPO <- enricher(
  gene          = bitr(network.genes.pan.neurodegen, 
                  fromType = "SYMBOL", toType = "ENTREZID", 
                  OrgDb ="org.Hs.eg.db") %>% .$ENTREZID,
  TERM2GENE     = mpo.df.hs.entrez,
  pvalueCutoff  = 0.05,
  minGSSize     = 5,
  maxGSSize     = 500,
  qvalueCutoff  = 0.05
)

ORA.network.genes.pan.peurodegen.results.MPO %>% 
  as_tibble %>% 
  separate(GeneRatio,into = c("temp1", "temp2")) %>% 
  mutate(GeneRatio = as.numeric(temp1)/as.numeric(temp2)) %>% 
  dplyr::slice(1:10) %>% 
  mutate(
    Description = str_wrap(Description, width = 22)
  ) %>%
  ggplot(
    aes(
      y = reorder(Description, GeneRatio), x = GeneRatio
    )
  ) +
  geom_col(
    aes(fill = -log10(p.adjust))
  ) +
  scale_fill_viridis_c() +
  labs(
    title = "MPO terms",
    y = "MPO term"
  )
```

### ORA of druggable biology
```{r}
drugbank <- readRDS("random/drugbank/joined_atc_df.rds")
fulldb.drugbank <- readRDS("random/drugbank/fulldb.rds")

# convert symbol to entrez using bit
drugbank.df.for.ORA = drugbank$identifier %>% 
  bitr(fromType = "SYMBOL", toType = "ENTREZID", 
                  OrgDb ="org.Hs.eg.db") %>% 
  as_tibble %>% 
  dplyr::rename(identifier = SYMBOL) %>% 
  left_join(drugbank, by = "identifier") %>% 
  dplyr::filter(identifier %in% nodes ) %>% 
  dplyr::select(level_1, ENTREZID) %>% 
  na.omit(level_1) %>% 
  unique


ORA.network.genes.pan.peurodegen.results.drugbank <- enricher(
  gene          = bitr(network.genes.pan.neurodegen, 
                  fromType = "SYMBOL", toType = "ENTREZID", 
                  OrgDb ="org.Hs.eg.db") %>% .$ENTREZID,
  TERM2GENE     = drugbank.df.for.ORA,
  pAdjustMethod = "fdr",
  minGSSize     = 2,
  maxGSSize     = 5000,
)

ORA.network.genes.pan.peurodegen.results.drugbank %>% 
  setReadable(OrgDb = org.Hs.eg.db, keyType = "ENTREZID") %>% 
  as_tibble %>% view

drugbank %>% 
  dplyr::filter(
    level_1 == "Digitalis glycosides"
) %>% 
  .$name.y %>% 
  unique

drugbank %>% 
  dplyr::filter(
    level_1 == "Strophanth.+"
) %>% 
  .$name.y %>% 
  unique
```


# CD subtypes (GWAS only)
## import netcoloc output
```{r}
# Define path
base_path <- "data/results/adultchildsubtypes v2"

# Read enrichment files
enrichment_dfs.2 <- 
  list.files(base_path, recursive = TRUE, full.names = TRUE) %>%
  grep(pattern = "network_colocalization_enrichment", 
       value = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/network_colocalization_enrichment.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0(base_path, "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y)) %>% 
  bind_rows() %>%
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>%
  pivot_longer(
    cols = c(observed_overlap, expected_overlap_mean),
    names_to = "type",
    values_to = "value"
  ) %>%
  mutate(
    type = recode(type,
                  "observed_overlap" = "observed",
                  "expected_overlap_mean" = "expected")
  ) %>% 
  mutate(source = "GWAS only (no parental proxy)")


base_path <- "data/results/adultchildsubtypes v2"

# Read enrichment files
node_metadata <- 
  list.files(base_path, recursive = TRUE, full.names = TRUE) %>%
  grep(pattern = "nodes_meta", 
       value = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/overlapping_nodes_metadata.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0(base_path, "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))

genesets_v2 = list.files("data/genesets/v2/adultchildsubtypes", full.names = T) %>% 
  sapply(read.csv, simplify = F) %>% 
  set_names(
    names(.) %>%
      str_remove("data/genesets/v2/adultchildsubtypes/") %>% 
      str_remove(".csv") %>% 
      str_remove("_list.txt") %>% 
      str_remove("_v2")
  )
```

## plot the obs v exp ratio
```{r}
enrichment_dfs.2 %>% 
  dplyr::filter(
    z_comb == 3, z_12 == 1.5, 
    
    ) %>% 
  mutate(
    # use the childhood vector to extract the CD seed gene name
    CD = str_remove(comparison, "_(AD|ALS|PD|DLB|FTLD|brain_malform)$"), 
    adult = str_extract(comparison, "(AD|ALS|PD|DLB|FTLD|brain_malform)$"), 
    FDR = p.adjust(empirical_p, "fdr"),
    CD = case_when(
      grepl(CD, pattern = "1A") ~ "Lysosomal disorders", 
      grepl(CD, pattern = "1B") ~ "Other disorders of lipid metabolism & transport", 
      grepl(CD, pattern = "1C") ~ "Disorders of amino acid & other organic acid metabolism", 
      grepl(CD, pattern = "1D") ~ "Vitamin-responsive inborn errors of metabolism", 
      grepl(CD, pattern = "1E") ~ "Disorders of mineral absorption & transport", 
      grepl(CD, pattern = "1F") ~ "Peroxisomal diseases", 
      grepl(CD, pattern = "1G") ~ "Mitochondrial disorders", 
      grepl(CD, pattern = "1H") ~ "Other inborn errors of metabolism", 
      grepl(CD, pattern = "2_") ~ "Leukodystrophies not notherwise categorised", 
      grepl(CD, pattern = "3_") ~ "Neurodegeneration with brain iron accumulation", 
      grepl(CD, pattern = "4_") ~ "Diseases not otherwise categorised",
      TRUE ~ CD) %>% str_wrap(width = 30), 
    
    facet.order = case_when(
      grepl(comparison, pattern = "1A") ~ 1, 
      grepl(comparison, pattern = "1E") ~ 2, 
      TRUE ~ 9
    )
    
    ) %>%  
  dplyr::filter(type == "observed") %>% 
  
  ggplot(
    aes(x = obs_exp, y = reorder(adult, obs_exp))
  ) +
  geom_col(
    aes(fill = -log10(FDR), alpha = FDR < 0.05), 
    colour = "grey50", position = "dodge", 
    width = 0.75
  ) +
  scale_fill_viridis_c() +
  scale_alpha_manual(values = c(0.1, 1)) +
  labs(
    x = "observed / expected", 
    y = "Disease"
  ) +
  facet_wrap(~reorder(CD, facet.order), ncol = 3) 
```

## ORA KEGG on whole subnets:
```{r}
# define the significant colocs. 
keep.subnets = enrichment_dfs.2 %>% 
  dplyr::filter(
    z_comb == 3, z_12 == 1.5, 
) %>% 
  mutate(FDR = p.adjust(empirical_p)) %>% 
  dplyr::filter(FDR < 0.05) %>% 
  .$comparison %>% unique

ORA.results.whole.subnets.kegg <- node_metadata[keep.subnets] %>% 
  sapply(function(x) {
    x$X %>%
      runKEGG
  }, simplify = F)
```


```{r}
keep = ORA.results.whole.subnets.kegg %>% 
  lapply(as_tibble) %>% 
  bind_rows(.id = "subnet") %>% 
  group_by(Description) %>% 
  summarise(n = n()) %>% 
  arrange(-n) %>% 
  dplyr::filter(n >= 8) %>% 
  .$Description


mat <- 
  ORA.results.whole.subnets.kegg %>% 
  lapply(as_tibble) %>% 
  bind_rows(.id = "subnet") %>% 
  dplyr::filter(Description %in% keep) %>%
  mutate(
    score = -log10(p.adjust), 
    # child = str_extract(subnet, pattern = "CD_.{1,2}"),
    # subnet = paste0(adult, "_", child),
    # subnet = str_remove(subnet, pattern = "_DLB$|_AD$|_FTLD$|_PD$|_ALS$"),
    .after = subnet
    ) %>% 
  dplyr::select(subnet, Description, score) %>%
  distinct() %>%
  arrange(subnet) %>% 
  pivot_wider(names_from = subnet, values_from = score) %>% 
  column_to_rownames("Description") %>%
  as.matrix() 

# Parse subnet names into adult and child components
subnet.meta = colnames(mat) %>%
  as_tibble() %>%
  rename(subnet = value) %>%
  mutate(
    adult = str_extract(subnet, "AD|DLB|PD|ALS|FTLD"),
    CD = str_extract(subnet, "CD_[^_]+_[^_]+"),
    CD = case_when(
      grepl("1A", CD) ~ "Lysosomal disorders",
      grepl("1B", CD) ~ "Other disorders of lipid metabolism & transport",
      grepl("1C", CD) ~ "Disorders of amino acid & other organic acid metabolism",
      grepl("1D", CD) ~ "Vitamin-responsive inborn errors of metabolism",
      grepl("1E", CD) ~ "Disorders of mineral absorption & transport",
      grepl("1F", CD) ~ "Peroxisomal diseases",
      grepl("1G", CD) ~ "Mitochondrial disorders",
      grepl("1H", CD) ~ "Other inborn errors of metabolism",
      grepl("2_", CD) ~ "Leukodystrophies not otherwise categorised",
      grepl("3_", CD) ~ "Neurodegeneration with brain iron accumulation",
      grepl("4_", CD) ~ "Diseases not otherwise categorised"
    ) %>% str_wrap(width = 30)
    )

# Colour palettes
adult.cols = setNames(
  viridis_pal(option = "magma", begin = 0.5)(5),
  unique(subnet.meta$adult)
)

cd.cols = setNames(
  colorRampPalette(brewer.pal(8, "Pastel2"))(n_distinct(subnet.meta$CD)),
  unique(subnet.meta$CD)
)

col_ann = HeatmapAnnotation(
 `Childhood disease` = subnet.meta$CD,
  `Adult disease`     = anno_simple(
    subnet.meta$adult,
    col     = adult.cols[subnet.meta$adult],
    pch     = subnet.meta$adult,
    pt_size = unit(4, "mm"),
    gp      = gpar(fontsize = 8, fontface = "bold")
  ),
  col = list(
    `Childhood disease` = cd.cols
  ),
  annotation_name_gp  = gpar(fontsize = 12)
  # annotation_legend_param = list(
  #   `Childhood disease` = list(
  #     labels_gp = gpar(fontsize = 8)
  #   )
  # )
)

# replace NA (not enriched) with 0
mat[is.na(mat)] <- 0

#pdf("output/plots4paperv2/Fig3b.pdf", width = 18, height = 12)
Heatmap(
  mat,
  name = "-log10(FDR)", 
  border = "black",
  rect_gp = gpar(color = "black"),
  col = colorRamp2(
    c(0, 0.0001, seq(0.0001, max(mat), length.out = 7)),
    viridis(9)
  ),
  
  cell_fun = function(j, i, x, y, width, height, fill) {
  if (mat[i, j] == 0) {
    grid.rect(x, y, width, height, gp = gpar(fill = "white", col = "black"))
  }
},
  
  cluster_rows = TRUE,
  cluster_columns = T,
  
  top_annotation  = col_ann,
  column_split = subnet.meta$CD,
  
  show_row_names = TRUE,
  show_column_names = FALSE,
  
  row_names_gp = gpar(fontsize = 12),
  column_names_gp = gpar(fontsize = 12),
  
  column_title = "Coloc subnetworks",
  row_title = "KEGG pathways"
) %>% 
  draw(
    heatmap_legend_side = "left", annotation_legend_side = "bottom", 
    padding = unit(c(2, 2, 2, 50), "mm"))
#dev.off()
```



### ORA KEGG on communities
```{r}
base_path <- "data/results/adultchildsubtypes v2"

# Read enrichment files
communities <- 
  list.files(base_path, recursive = TRUE, full.names = TRUE) %>%
  grep(pattern = "gprofi", 
       value = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/gprofiler_analysis_results.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0(base_path, "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))


# communities$CD_1A_lyso_PD %>% 
#   split(f = .$Cluster) %>% 
#   sapply(function(x) {
#     x$genes %>% 
#       str_split(pattern = " ", simplify = F) %>% 
#       unlist 
#       #runKEGG
#   })


ORA.results.genes.in.communities <- 
  list(
    lysoPD = communities$CD_1A_lyso_PD %>% 
      split(f = .$Cluster) %>% 
      sapply(function(x) {
        x$genes %>% 
          str_split(pattern = " ", simplify = F) %>% 
          unlist %>% 
          runKEGG
      }), 
      
      lysoDLB = communities$CD_1A_lyso_DLB %>% 
        split(f = .$Cluster) %>% 
        sapply(function(x) {
          x$genes %>% 
            str_split(pattern = " ", simplify = F) %>% 
            unlist %>% 
            runKEGG
        }) 
      )
```

## LysoDLB: C437 
```{r}
ORA.results.genes.in.communities$lysoDLB$C437 %>% 
  setReadable(OrgDb = org.Hs.eg.db, keyType = "ENTREZID") %>% 
  cnetplot()

ORA.results.genes.in.communities$lysoDLB$C438 %>% 
  setReadable(OrgDb = org.Hs.eg.db, keyType = "ENTREZID") %>% 
  cnetplot(showCategory = 31)


ORA.results.genes.in.communities$lysoDLB$C438 %>% 
  setReadable(OrgDb = org.Hs.eg.db, keyType = "ENTREZID") %>% 
  cnetplot(showCategory = c("Mineral absorption", 
                            "Antigen processing and presentation", 
                            "Lysosome biogenesis", 
                            "Leishmaniasis",                  
                            "Ferroptosis",
                            "Phagosome",
                            "Rheumatoid arthritis",
                            "Cell adhesion molecule (CAM) interaction",
                            "Cholesterol metabolism",
                            "Human T-cell leukemia virus 1 infection", 
                            "Hematopoietic cell lineage",
                            "Porphyrin metabolism",
                            "PPAR signaling pathway",                      
                            "Complement and coagulation cascades",         
                            "Osteoclast differentiation"                  
))


```

## ORA drugbank on communiities

```{r}
# take the communities in the sig subnets and perform ORA
test = communities[keep.subnets] %>% 
  lapply(function(x) {
    x %>% 
      split(f = .$Cluster) %>% # split into clusters
      sapply(function(y) {
        y$genes %>% 
          str_split(pattern = " ", simplify = F) %>% 
          unlist 
      }) %>% 
      sapply(function(z) {
        enricher(
          gene          = bitr(z, 
                               fromType = "SYMBOL", toType = "ENTREZID", 
                               OrgDb ="org.Hs.eg.db") %>% .$ENTREZID,
          TERM2GENE     = drugbank.df.for.ORA,
          pAdjustMethod = "fdr",
          minGSSize     = 1,
          maxGSSize     = 5000,
        ) 
      })
  })

test2 = test %>% 
  lapply(function(x) {
    x %>% 
      Filter(function(y) !is.null(y) && nrow(y@result) > 0, .) %>%
      sapply(function(z) {
        z %>% 
          setReadable(OrgDb = "org.Hs.eg.db", keyType = "ENTREZID") %>% 
          as_tibble()
      }, simplify = F) %>% 
      bind_rows(.id = "community") 
  })

test2 %>% bind_rows(.id = "subnet") 

test2$CD_1A_lyso_PD %>% 
  dplyr::select(community, ID, geneID) %>% 
  mutate(geneID = str_split(geneID, pattern = "/")) %>% 
  .$geneID %>% 
  unlist %>% 
  unique -> PD.genes

drugbank %>% 
  dplyr::filter(identifier %in% PD.genes) %>% 
  .$drugbank_id %>% unique %>% length
```


```{r}
test$CD_1A_lyso_AD %>% 
  
  bind_rows()

```


### how many drug targets for lysosomes
I first want to assess how many drugs in drugbank would act on the lysosome, and other lysosomal pathways
```{r}
fulldb.drugbank <- readRDS("random/drugbank/fulldb.rds")

search.patten = "lysosom|autophagy|mTOR|TFEB|sphingolipid|glycosaminoglycan|mucopolysaccharid|endosom|vacuol"

# search the db pathway name 
lyso.drugs.pathway.general.information <- 
  fulldb.drugbank$drugs$pathway$general_information %>% 
  dplyr::filter(
    grepl(
      name, 
      pattern = search.patten, 
      ignore.case = T)
  ) %>% 
  .$drugbank_id %>% 
  unique

## esarch the db pathway name 
lyso.drugs.pharmacology.MoA <- fulldb.drugbank$drugs$pharmacology %>% 
  dplyr::filter(
    grepl(mechanism_of_action, pattern =search.patten, ignore.case = T)
  ) %>% 
  .$drugbank_id %>% 
  unique 

## search the GO of the targets
bp_terms <- c(
  "lysosom", "autophagy", "autophagic", "vacuo.+acidification",
  "endolysosomal", "phagolysosome", "lysosomal exocytosis",
  "lysosomal biogenesis", "lysosomal storage"
)

lyso.targets = fulldb.drugbank$cett$targets$polypeptides$go %>% 
  dplyr::filter(
    grepl(description, pattern = paste(bp_terms, collapse = "|"), 
                                        ignore.case = TRUE)
    ) %>% 
  pull(target_id)
  
lyso.drugs.targets = fulldb.drugbank$cett$targets$general_information %>% 
  dplyr::filter(target_id %in% lyso.targets) %>% 
  pull(drugbank_id) %>% 
  unique
  

lyso.drugs = c(lyso.drugs.pathway.general.information,   # GO cellular component, targets
    lyso.drugs.pharmacology.MoA,  # subcellular location text, targets
    lyso.drugs.targets     # routes A+B across all CETT nodes
    # drug_ids_route_d,   # GO biological process
    # lyso_pathways,      # pathway membership
    # lyso_freetext,      # free-text mechanism/description fields
    # drug_ids_route_g,   # ATC codes
    # lyso_disease_drugs  # disease/category annotations
  ) %>% unique()

lyso.drugs %>% length
```

#### pull the targets from the lyso drugs
```{r}
lyso.drug.targetids = fulldb.drugbank$cett$targets$general_information %>% 
  dplyr::filter(drugbank_id %in% lyso.drugs) %>% 
  .$target_id

lyso.drug.target.name = fulldb.drugbank$cett$targets$polypeptides$general_information %>% 
  dplyr::filter(target_id %in% lyso.drug.targetids) %>% 
  .$gene_name %>% 
  unique

length(lyso.drug.target.name)
```


```{r}


druggable.lysogenes = node_metadata[names(node_metadata) %>% grepl(pattern = "lyso")] %>% 
  bind_rows() %>% 
  .$X %>% 
  unique %>% 
  intersect(lyso.drug.target.name) 

node_metadata[names(node_metadata) %>% grepl(pattern = "lyso")] %>% 
  lapply(function(x) {
    x$X %>% 
      unique %>% 
      intersect(lyso.drug.target.name)
  }) %>% 
  lapply(length)

druggable.genes = node_metadata[names(node_metadata) %>% grepl(pattern = "lyso")] %>% 
  bind_rows() %>% 
  .$X %>% 
  unique %>% 
  intersect(drugbank$identifier)



node_metadata[names(node_metadata) %>% grepl(pattern = "lyso")] %>% 
  .[keep.subnets] %>% 
  bind_rows() %>% 
  as_tibble() %>% 
  dplyr::select(comparison, X, z_both, everything()) %>% 
  mutate(
    adult = str_extract(comparison, pattern = "AD|PD|DLB|FTLD|ALS"), 
    CD = str_remove(comparison, pattern = "CD_.{1,2}_"), 
    CD = str_remove(CD, pattern = "_.+"), 
    comparison = paste0(CD, "_", adult),
    druggable = case_when(
      X %in% druggable.genes ~ "Druggable",
      TRUE ~ "No drugs in DrugBank"
    ),
    druggable = case_when(
      X %in% druggable.lysogenes ~ "Druggable & lysosomal",
      TRUE ~ druggable
    ) %>% as.factor,
    druggable = fct_rev(druggable),
    .after = comparison
  ) %>% 
  count(comparison, druggable) %>%
  ggplot(
    aes(y = comparison, x = n, fill = druggable)
    ) +
  geom_col(width = 0.7, color = "black", linewidth = 0.3) +
  scale_fill_manual(
    values = c("grey50", "green", "green4")
    ) +
  labs(
    x = "Number of genes in colocalised subnetwork", 
    y = NULL, 
    fill = "Druggable?") +
  # theme_minimal() +
  theme(
    panel.grid = element_blank(),
    text = element_text(size = 19)
  ) +
  ggsave("output/plots4ideasgrant/durggablebio.pdf", 
         height = 3, width = 6)

lyso.drug.targets = test2 %>% 
  bind_rows(.id = "subnet") %>% 
  mutate(
    adult = str_extract(subnet, pattern = "AD|PD|DLB|FTLD|ALS"), 
    CD = str_remove(subnet, pattern = "CD_.{1,2}_"), 
    CD = str_remove(CD, pattern = "_.+"), 
    .after = subnet
  ) %>% 
  dplyr::filter(CD == "lyso") %>% 
  .$geneID %>% 
  str_split(pattern = "/") %>%
  unlist %>% 
  unique %>% 
  sort
  
  
  ggplot(
    aes(y = Description, x = community, fill = -log10(pvalue))
  ) +
  geom_tile() +
  facet_wrap(~adult, scales = "free",space = "fixed")


```



# Adult-onset & Non-neuropathic IEMs
## import the coloc
```{r}
enrichment_dfs.non.neuro.IEMS <- 
  list.files(
    "data/results/adult and non-neurodegen IEMs", 
    recursive = TRUE, full.names = TRUE) %>%
  grep(pattern = "network_colocalization_enrichment", 
       value = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/network_colocalization_enrichment.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0(
          "data/results/adult and non-neurodegen IEMs", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y)) %>% 
  bind_rows() %>%
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>%
  pivot_longer(
    cols = c(observed_overlap, expected_overlap_mean),
    names_to = "type",
    values_to = "value"
  ) %>%
  mutate(
    type = recode(type,
                  "observed_overlap" = "observed",
                  "expected_overlap_mean" = "expected")
  )
```
## plot obs v exp ratio
```{r}
enrichment_dfs.non.neuro.IEMS %>% 
  dplyr::filter(
    z_comb == 3, z_12 == 1.5, 
    ) %>% 
  mutate(
    # use the childhood vector to extract the CD seed gene name
    adult = str_extract(comparison, "(AD|ALS|PD|DLB|FTLD|brain_malform)"), 
    IEM = str_remove(comparison, pattern = "^(AD|ALS|PD|DLB|FTLD|brain_malform)_"), 
    IEM = case_when(
      grepl(IEM, pattern = "aminoacidopathy") ~  "Aminoacidopathies", 
      grepl(IEM, pattern = "FA_ox") ~  "Fatty acid oxidation disorders", 
      grepl(IEM, pattern = "generalIEM") ~  "General inborn errors of metabolism", 
      grepl(IEM, pattern = "glycosy") ~  "Congenital disorders of glycosylation", 
      grepl(IEM, pattern = "mitochondrial") ~  "Mitochondrial disorders", 
      grepl(IEM, pattern = "nonCD_IEM") ~  "All non-neuronopathic IEMs", 
      grepl(IEM, pattern = "nonCD_lyso") ~  "Non-neurononpathic lysosomal disorders", 
      grepl(IEM, pattern = "peroxisomal") ~  "Non-neurononpathic peroxisomal disorders", 
    ) %>% str_wrap(width = 25),
    FDR = p.adjust(empirical_p, "fdr"),
    facet.order = case_when(
      grepl(comparison, pattern = "nonCD_IEM") ~ 1, 
      grepl(comparison, pattern = "lyso") ~ 2, 
      grepl(comparison, pattern = "general") ~ 9, 
      TRUE ~ 8
    )
    
    ) %>%  
  dplyr::filter(type == "observed") %>% 
  
  ggplot(
    aes(x = obs_exp, y = reorder(adult, obs_exp))
  ) +
  geom_col(
    aes(fill = -log10(FDR), alpha = FDR < 0.05), 
    colour = "grey50", position = "dodge", 
    width = 0.75
  ) +
  scale_fill_viridis_c() +
  scale_alpha_manual(values = c(0.1, 1)) +
  labs(
    x = "observed / expected", 
    y = "Disease"
  ) +
  facet_wrap(~reorder(IEM, facet.order), ncol = 3)
```

# DLB, PD and Lyso subnetworks

## ORA
```{r}
run.ORA.MPO = function(input) {
  enricher(
  gene          = bitr(input, 
                       fromType = "SYMBOL", toType = "ENTREZID", 
                       OrgDb ="org.Hs.eg.db") %>% .$ENTREZID,
  TERM2GENE     = mpo.df.hs.entrez,
  pvalueCutoff  = 0.05,
  minGSSize     = 5,
  maxGSSize     = 500,
  qvalueCutoff  = 0.05
)
}
```


```{r}
run.ORA.MPO(node_metadata$CD_1A_lyso_PD$X) %>% 
  as_tibble() %>% 
  mutate(network = "LysoPD") %>% 
  bind_rows(
    run.ORA.MPO(node_metadata$CD_1A_lyso_DLB$X) %>% 
      as_tibble %>% 
      mutate(
        network = "LysoDLB"
      )
  ) %>% 
  mutate(
    source = "MPO"
  ) %>% 
  bind_rows(
    runKEGG(node_metadata$CD_1A_lyso_PD$X) %>% 
      as_tibble() %>% 
      mutate(network = "LysoPD") %>% 
      bind_rows(
        runKEGG(node_metadata$CD_1A_lyso_DLB$X) %>% 
          as_tibble %>% 
          mutate(
            network = "LysoDLB"
          )
      ) %>% 
      mutate(source = "KEGG")
  ) %>% 
  group_by(Description, source) %>% 
  mutate(
    category = case_when(
      n_distinct(network) == 2 ~ 1,
      network == "LysoPD" ~ 2,
      network == "LysoDLB" ~ 3
    ), 
    source = factor(source, levels = c("MPO", "KEGG"))
  ) %>% 
  ungroup() %>% 
  ggplot(
    aes(
      x = network,
      y = reorder(Description, -category)
    )
  ) +
  geom_tile(
    aes(fill = FoldEnrichment)
  ) +
    geom_label(
          aes(
      label = p.adjust %>% signif(2)
      
      )
    ) +
  scale_fill_viridis_c() +
  facet_wrap(~source, 
             nrow = 1, ncol = 2, scales = "free_y") +
  labs(
    x = "Subnetwork",
    y = NULL
  ) +
   theme(
    strip.text = element_text(size = 12),
    text = element_text(size = 14),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom",
    #axis.title.x = element_blank(),
    plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) 
```

## is there sig overlap
```{r}
# Define your gene sets
lyso_DLB = node_metadata$CD_1A_lyso_DLB$X
lyso_PD  = node_metadata$CD_1A_lyso_PD$X

# Background = all genes in the interactome
background =  nodes

# Overlap
overlap = intersect(lyso_DLB, lyso_PD)

# Contingency table for Fisher's exact test
#               In PD    Not in PD
# In DLB        a        b
# Not in DLB    c        d

a = length(overlap)
b = length(lyso_DLB) - a
c = length(lyso_PD)  - a
d = length(background) - (a + b + c)

mat = matrix(c(a, b, c, d), nrow = 2,
             dimnames = list(
               c("In_DLB", "Not_in_DLB"),
               c("In_PD",  "Not_in_PD")
             ))
mat

fisher.test(mat, alternative = "greater")

phyper(q = 87, m = 289, n = N - 289, k = 436, lower.tail = FALSE)
```


## Community detection

Multiscvale community detection was performed using HiDef within my netcoloc pipeline. 

```{r}
communities.PD.DLB = communities[keep.PD.DLB]
```

### number of communities
```{r}
communities[keep.PD.DLB] %>% 
  bind_rows(.id = "network") %>% 
  ggplot(
    aes(x = network)
  ) +
  geom_bar() +
  labs(
    title = "number of communities"
  )
```

### jaccard simialrity

```{r}
# Define a helper to calculate Jaccard Similarity
calc_jaccard <- function(list_a, list_b) {
  intersection <- length(intersect(list_a, list_b))
  
  union <- length(list_a) + length(list_b) - intersection
  
  if (union == 0) return(0)
  return(intersection / union)
}

# extract DLB and PD clusters
dlb_clusters <- 
  communities.PD.DLB$CD_1A_lyso_DLB %>% 
  dplyr::select(Cluster, genes) %>% 
  split(f = .$Cluster) %>% 
  lapply(function(x) {
     x$genes %>% 
      str_split(" ") %>% 
      unlist()
  })

pd_clusters <- communities.PD.DLB$CD_1A_lyso_PD %>% 
  dplyr::select(Cluster, genes) %>% 
  split(f = .$Cluster) %>% 
  lapply(function(x) {
     x$genes %>% 
      str_split(" ") %>% 
      unlist
  })

# Create an empty matrix
sim_matrix <- matrix(NA, 
                     nrow = length(dlb_clusters), 
                     ncol = length(pd_clusters),
                     dimnames = list(names(dlb_clusters), names(pd_clusters)))

# Fill the matrix
for (d_name in names(dlb_clusters)) {
  for (p_name in names(pd_clusters)) {
    sim_matrix[d_name, p_name] <- 
      calc_jaccard(dlb_clusters[[d_name]], pd_clusters[[p_name]])
  }
}


# Filter rows and columns where the max value is >= 0.001
keep_rows <- apply(sim_matrix, 1, max) >= 0.1
keep_cols <- apply(sim_matrix, 2, max) >= 0.1

sim_filtered <- sim_matrix[keep_rows, keep_cols]

pheatmap(sim_filtered,
         main = "Jaccard index of LysoPD and LysoDLB network communities",
         color = colorRampPalette(c("white", "red"))(100),
         display_numbers = TRUE,
         number_format = "%.2f", 
         cluster_rows = TRUE,
         cluster_cols = TRUE)
```

There are 2 pairs of particularly similar communities

C310-C466 - shared genes involved in 
```{r}
intersect(
  communities.PD.DLB$CD_1A_lyso_DLB %>% dplyr::filter(Cluster == "C466") %>% .$genes %>% str_split(" ", simplify = F) %>% .[[1]], 
  
  communities.PD.DLB$CD_1A_lyso_PD %>% dplyr::filter(Cluster == "C310") %>% .$genes %>% str_split(" ", simplify = F) %>% .[[1]]
) %>% 
  as_tibble() %>% 
  set_colnames("gene") %>% 
  mutate(functio = case_when(
    gene == "HEXA" ~ "beta-hexosaminidase A, lyso enz. results in GM2 ganglio accu.", 
    gene == "HEXD" ~ "beta-hexosaminidase D, nucleaus/cytoplasm. ", 
    gene == "NEU4" ~ "sialidase, lyso enz ", 
  ))
```

C302-C456 - shared genes involved in

```{r}
C302 = communities.PD.DLB$CD_1A_lyso_PD %>% dplyr::filter(Cluster == "C302") %>% .$genes %>% str_split(" ", simplify = F) %>% .[[1]]

C456 = communities.PD.DLB$CD_1A_lyso_DLB %>% dplyr::filter(Cluster == "C456") %>% .$genes %>% str_split(" ", simplify = F) %>% .[[1]]

intersect(C302, C456)
```

## ORA on the communities

```{r}
## Extract the genes in the communities for hs
communities.genes.hs = list(
  lysoDLB = communities.PD.DLB$CD_1A_lyso_DLB %>% 
    split(f = .$Cluster) %>% 
    lapply(function(x) {
      x %>% 
        pull('genes') %>% 
        str_split(" ", simplify = F) %>% 
        .[[1]]
    }),
  lysoPD = communities.PD.DLB$CD_1A_lyso_PD %>% 
    split(f = .$Cluster) %>% 
    lapply(function(x) {
      x %>% 
        pull('genes') %>% 
        str_split(" ", simplify = F) %>% 
        .[[1]]
      })
)
# these are the communites by mouse gene name
communities.mm = qread("data/public data/PRJNA606815_NPC_mousebrain/R_objects/network.genes.cluster.mouseSymbol.qs")
```


```{r}
# perform ORA on the LysoPD communities 
## a function for running ORA on communities by species
runORA = function(community.object, org.db.object, org.symbol) {
  community.object %>%   
    lapply(function(x)
    {bitr(x, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.db.object) %>% 
        .$ENTREZID %>% 
          enrichKEGG(
            organism = org.symbol,
            pvalueCutoff = 0.05,
            qvalueCutoff = 0.05,
            minGSSize = 3,
            maxGSSize = 500
          )})  
}

## Run the ORA
ORA.communities = list(
  PD = list(
    hs = runORA(community.object = communities.genes.hs$lysoPD, org.db.object = org.Hs.eg.db, org.symbol = "hsa"), 
    mm = runORA(community.object = communities.mm$PD, org.db.object = org.Mm.eg.db, org.symbol = "mmu")
  ), 
  DLB = list(
    hs = runORA(community.object = communities.genes.hs$lysoDLB, org.db.object = org.Hs.eg.db, org.symbol = "hsa"), 
    mm = runORA(community.object = communities.mm$DLB, org.db.object = org.Mm.eg.db, org.symbol = "mmu")
  )
)
```

# DLB and NPC1 scRNA-seq
```{r}
# seurat.objects 
data.seurat.hs.DLB.PD = qread("data/public data/GSE178146_RAW/R_objects/PD_DLB_data_seurat.qs")
data.seurat.mm.Npc1 = qread("data/public data/PRJNA606815_NPC_mousebrain/R_objects/dataseurat.qs")

# pseudobulk objects
y.hs.DLB.PD = qread("data/public data/GSE178146_RAW/R_objects/pseudobulk_dgeObject.qs")
y.mm.Npc1 = qread("data/public data/PRJNA606815_NPC_mousebrain/R_objects/pseudobulk_dgeObject.qs")

toptables = list(
  hs.ACC = qread("data/public data/GSE178146_RAW/R_objects/toptables.qs"),
    npc1 = qread("data/public data/PRJNA606815_NPC_mousebrain/R_objects/toptables.qs")
)
```

## UMAP plots

```{r}
# perp an object to consistently plot celltypes the asme colour
cell2colour = tibble(
  celltype = 
    c(data.seurat.hs.DLB.PD@meta.data$celltype %>% unique, 
      data.seurat.mm.Npc1@meta.data$celltype %>% unique
    )%>% unique()
  ) %>% 
  mutate(
    colour = case_when(
      grepl(celltype, pattern = "Ex") ~ "#8DD3C7", 
      grepl(celltype, pattern = "Mix") ~ "#8DD3C7", 
      grepl(celltype, pattern = "In") ~ "#56B4E9", 
      grepl(celltype, pattern = "Micr") ~ "#BEBADA", 
      grepl(celltype, pattern = "Astro") ~ "#FB8072", 
      grepl(celltype, pattern = "Oligo") ~ "#009E73", 
      grepl(celltype, pattern = "OPC") ~ "#B3DE69", 
      grepl(celltype, pattern = "Endo") ~ "grey50", 
      grepl(celltype, pattern = "VL") ~ "grey50", 
      grepl(celltype, pattern = "Ependymal") ~ "grey50", 
      grepl(celltype, pattern = "Fib") ~ "grey50", 
    )
  ) %>% 
  pull(colour, name = celltype)

# plot the UMAP plots
data.seurat.hs.DLB.PD %>% 
  DimPlot(
    group.by = "celltype", 
    label = TRUE,
    label.color = "black",
    cols = cell2colour
  )  +
  labs(
    title = "Human anterior cingulate cortex", 
    subtitle = "Control, DLB, PD and PDD"
  ) +
   theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    legend.position = "none", 
    aspect.ratio = 1
   ) +

  (data.seurat.mm.Npc1 %>% 
  DimPlot(
    group.by = "celltype", 
    label = TRUE,
    label.color = "black",
    cols = cell2colour
  ) +
  labs(
    title = "Mouse cerebellum", 
    subtitle = "Npc1 -/- and WT"
  ) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    legend.position = "none", 
    aspect.ratio = 1
  ))


```

## celltype markers
```{r}

Idents(data.seurat.hs.DLB.PD) = data.seurat.hs.DLB.PD@meta.data$celltype
Idents(data.seurat.mm.Npc1) = data.seurat.mm.Npc1@meta.data$celltype

# cell type marker genes
markergenes2plot = list(
  hs = c(
    # inhib neurons
    "GAD1","GAD2", 
    # exc neurons
    "CAMK2A","NRGN",
    # astrocytes
    "GFAP","AQP4", 
    # microglia
    "C3","CD74",  "AIF1",
    # OPCs
    "OLIG1","OLIG2",
    # oligodend
    "MBP","MOBP", "PLP1", 
    # VLMCs
    "COL1A2","DCN", "IGFBP7",
    # endo
    "CLDN5","FLT1"
  ), 
  
  mm = c(
    # inhib neurons
    "Gad1","Gad2", 
    # exc neurons
   "Gabra6", "Kcnd2",
    # astrocytes
    "Slc4a4","Aqp4", "Gabrb1",
    # microglia
    "Csf1r", "Aif1",
    # oligodend
    "Mbp","Mobp", "Plp1", 
   # OPCs
   "Olig1", "Olig2",
    # VLMCs
    "Col1a2","Dcn", "Igfbp7",
    # endo
    "Cldn5","Flt1"
  )
)

ggarrange(
Seurat::DotPlot(
    data.seurat.hs.DLB.PD,
    features = markergenes2plot$hs,
    cols = c("#440154FF", "#22A884FF"), 
    scale = TRUE, 
    dot.scale = 10
  ) +
  labs(
    y = "Cell type", 
    x = "Genes"
  ) +
  scale_fill_viridis_c() +
  labs(
    title = "Marker genes in human ACC"
  ), 
  
Seurat::DotPlot(
  data.seurat.mm.Npc1,
  features = markergenes2plot$mm,
  cols = c("#440154FF", "#22A884FF"), 
  scale = TRUE, 
  dot.scale = 10
) +
  labs(
    y = "Cell type", 
    x = "Genes"
  ) +
  scale_fill_viridis_c() +
  labs(
    title = "Marker genes in mouse cerebellum"
  )
)

ggsave()
```


## DE network genes
### DLB
```{r}
net.genes.de.humanACC = toptables$hs.ACC %>% 
  bind_rows() %>% 
  dplyr::filter(
    grepl(coef, pattern = "^DLB"), 
    FDR < 0.05, 
    gene %in% node_metadata$CD_1A_lyso_DLB$X
  ) %>% 
  .$gene

# mouse name to human name
mm2hs <- homologene(
  genes  = toptables$npc1 %>% 
    bind_rows() %>% 
    .$gene %>% unique,
  inTax  = 10090,  # mouse
  outTax = 9606    # human
) %>% 
  dplyr::select(
    mm_symbol = `10090`, hs_symbol = `9606`, 
    mm_entrez = `10090_ID`, hs_entrez = `9606_ID`
  ) %>% 
  mutate(across(everything(), as.character)) %>% 
  as_tibble

net.genes.de.mmCerebel = 
  toptables$npc1 %>% 
  bind_rows() %>% 
  left_join(mm2hs, by = c("gene" = "mm_symbol")) %>% 
  dplyr::filter(
    FDR < 0.05, 
    hs_symbol %in% node_metadata$CD_1A_lyso_DLB$X
  ) %>% 
  .$hs_symbol

shared.de = intersect(net.genes.de.humanACC, net.genes.de.mmCerebel)
```

### PD
```{r}
net.genes.de.humanACC.PD = toptables$hs.ACC %>% 
  bind_rows() %>% 
  dplyr::filter(
    grepl(coef, pattern = "^PD"), 
    FDR < 0.05, 
    gene %in% node_metadata$CD_1A_lyso_PD$X
  ) %>% 
  .$gene

net.genes.de.mmCerebel.PD = 
  toptables$npc1 %>% 
  bind_rows() %>% 
  left_join(mm2hs, by = c("gene" = "mm_symbol")) %>% 
  dplyr::filter(
    FDR < 0.05, 
    hs_symbol %in% node_metadata$CD_1A_lyso_PD$X
  ) %>% 
  .$hs_symbol

shared.de.PD = intersect(net.genes.de.humanACC.PD, net.genes.de.mmCerebel.PD)
shared.de.PD
```


```{r}
# Build the FDR matrix (for asterisk overlay)
  FDR_mat <- toptables$hs.ACC %>%
    bind_rows() %>%
    dplyr::filter(
      gene %in% shared.de, 
      grepl(coef, pattern = "^DLB")
      ) %>%
    dplyr::select(coef, gene, FDR) %>%
    pivot_wider(names_from = gene, values_from = FDR) %>%
    column_to_rownames("coef") %>%
    as.matrix()

Heatmap(
  matrix = toptables$hs.ACC %>%
    bind_rows() %>%
    dplyr::filter(
      gene %in% shared.de, 
      grepl(coef, pattern = "^DLB")
    ) %>%
    dplyr::select(coef, gene, logFC) %>%

    pivot_wider(names_from = gene, values_from = logFC) %>%
    column_to_rownames("coef") %>%
    as.matrix(),
  
  name = "logFC",
  col = colorRamp2(c(-10, 0, 5), c("#2166AC", "#F7F7F7", "#B2182B")),
  
  cell_fun =function(j, i, x, y, width, height, fill) {
      # j = column index, i = row index
      # x, y = coordinates of the center of the cell
      # width, height = size of the cell
      # fill = background color (not used here)

      # Check if the FDR value for this cell exists and is below 0.05
      if (!is.na(FDR_mat[i, j]) && FDR_mat[i, j] < 0.05) {
        # If yes, overlay a text label "*" on that cell
        grid.text("*", x, y, gp = gpar(fontsize = 15))
      }
    },
  column_names_gp = gpar(fontsize = 10),
  border = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE
)
```



## DE network communities
### import
```{r}

fry.results = list(
  hs = qread("data/public data/GSE178146_RAW/R_objects/fry_results.qs"), 
  mm = qread("data/public data/PRJNA606815_NPC_mousebrain/R_objects/fry_results_DLB_PD_list.qs")
)

```
### plot.fry
#### DLB
```{r}
fry.input.DLB = fry.results$hs$DLB %>% 
  bind_rows() %>% 
  mutate(source = "human DLB") %>% 
  bind_rows(
    fry.results$mm$DLB %>% 
      bind_rows() %>% 
      mutate(source = "mouse Npc1") 
  ) 

keep.DLB.fry = fry.input.DLB %>% 
  dplyr::filter(FDR.Mixed < 0.05) %>% 
  .$pathway %>% 
  unique

fry.input.DLB %>% 
  dplyr::filter(pathway %in% keep.DLB.fry) %>%
  mutate(
    source = case_when(
      grepl(source, pattern = "human") ~ "Human DLB", 
      grepl(coef, pattern = "3week") ~ "Mouse Npc1\npre-symptomatic" ,
      grepl(coef, pattern = "7week") ~ "Mouse Npc1\nsymptomatic" ,
    ), 
    celltype = str_remove(coef, pattern = "^DLB_"), 
    celltype = str_remove(celltype, pattern = " - Contro.+"), 
    celltype = str_remove(celltype, pattern = "_[3|7]week"), 
    celltype = str_remove(celltype, pattern = "_"), 
  ) %>% 
  ggplot(
    aes(y = reorder(pathway, -PValue.Mixed), x = celltype)
  ) +
  geom_tile(
    aes(fill = -log10(PValue.Mixed))
  ) +
  geom_label(
    aes(label = signif(FDR.Mixed, 2)), 
    data = . %>% 
      dplyr::filter(FDR.Mixed < 0.1)
  ) +
  easy_rotate_x_labels(angle = -45) +
  scale_fill_viridis_c() +
  facet_wrap(~source, scales = "free_x", space = "free_x")
```

#### PD
```{r}
fry.input.PD = fry.results$hs$PD %>% 
  bind_rows() %>% 
  mutate(source = "human DLB") %>% 
  bind_rows(
    fry.results$mm$PD %>% 
      bind_rows() %>% 
      mutate(source = "mouse Npc1") 
  ) 

keep.PD.fry = fry.input.PD %>% 
  dplyr::filter(FDR.Mixed < 0.05) %>% 
  .$pathway %>% 
  unique

fry.input.PD %>% 
  dplyr::filter(pathway %in% keep.PD.fry) %>%
  mutate(
    source = case_when(
      grepl(coef, pattern = "PD_") ~ "Human PD", 
      grepl(coef, pattern = "PDD_") ~ "Human PDD", 
      grepl(coef, pattern = "3week") ~ "Mouse Npc1\npre-symptomatic" ,
      grepl(coef, pattern = "7week") ~ "Mouse Npc1\nsymptomatic" ,
    ),
    celltype = str_remove(coef, pattern = "^PDD_"),
    celltype = str_remove(celltype, pattern = "^PD_"),
    celltype = str_remove(celltype, pattern = " - Contro.+"), 
    celltype = str_remove(celltype, pattern = "_[3|7]week"), 
    celltype = str_remove_all(celltype, pattern = "_"), 
  ) %>% 
  ggplot(
    aes(y = reorder(pathway, -PValue.Mixed), x = celltype)
  ) +
  geom_tile(
    aes(fill = -log10(PValue.Mixed))
  ) +
  geom_label(
    aes(label = signif(FDR.Mixed, 2)), 
    data = . %>% 
      dplyr::filter(FDR.Mixed < 0.1)
  ) +
  easy_rotate_x_labels(angle = -45) +
  scale_fill_viridis_c() +
  facet_wrap(~source, scales = "free_x", space = "free_x")
```

### vis the results
I want to plot these in a network style, but the gene nodes are split and coloured by logFC from the snRNAseq data. 
this was crashing my R session. maybe come back to later.  

```{r}
# pathway_df = ORA.communities$PD$hs$C290 %>% 
#   setReadable(OrgDb = org.Hs.eg.db, keyType = "ENTREZID") %>% 
#   as_tibble() %>% 
#   dplyr::select(Description, geneID)
# 
# # Your two DE results
# de_hs  <- toptables$hs.ACC$`PDD_Microglia - Control_Microglia` 
# de_mmu <- toptables$npc1$`MicrogliaImmune_7week_Npc1 - MicrogliaImmune_7week_WT`  
# 
# de_mmu_mapped <- de_mmu %>%
#   left_join(mm2hs, by = c("gene" = "mm_symbol")) %>%
#   dplyr::filter(!is.na(hs_symbol)) %>%
#   group_by(hs_symbol) %>%
#   slice_min(FDR, n = 1, with_ties = FALSE) %>%
#   ungroup()
# 
# library(ggraph)
# library(igraph)
# library(scatterpie)
# 
# # build edge list
# edges <- pathway_df %>%
#   mutate(gene = strsplit(geneID, "/")) %>%
#   unnest(gene) %>%
#   select(from = Description, to = gene)
# 
# all_genes    <- unique(edges$to)
# all_pathways <- unique(edges$from)
# 
# # --- 2. Build pie data: one row per gene, cols = hs logFC, mmu logFC ---
# # --- Now build pie_data joining on human symbols ---
# pie_data <- tibble(gene = all_genes) %>%
#   left_join(de_hs  %>% select(gene, logFC) %>% rename(logFC_hs  = logFC),
#             by = "gene") %>%
#   left_join(de_mmu_mapped %>% select(hs_symbol, logFC) %>% rename(logFC_mmu = logFC),
#             by = c("gene" = "hs_symbol")) %>%
#   mutate(
#     logFC_hs  = replace_na(logFC_hs,  0),
#     logFC_mmu = replace_na(logFC_mmu, 0)
#   )
# 
# 
# # --- 3. Build igraph and layout ---
# g <- graph_from_data_frame(edges, directed = FALSE)
# 
# set.seed(42)
# layout <- create_layout(g, layout = "fr")
# 
# layout <- layout %>%
#   mutate(node_type = ifelse(name %in% all_genes, "gene", "category"))
# 
# # --- 4. Attach x, y to pie data ---
# pie_data <- pie_data %>%
#   left_join(layout %>% 
#               dplyr::filter(node_type == "gene") %>% 
#               dplyr::select(name, x, y),
#             by = c("gene" = "name"))
# 
# # --- 5. Scale for scatterpie (needs normalised radius) ---
# pie_data <- pie_data %>%
#   mutate(radius = 0.15)   # tune this to your layout scale
# 
# # --- 6. Plot ---
# ggraph(layout) +
#   
#   # Edges
#   geom_edge_link(colour = "grey80", width = 0.4, alpha = 0.6) 
#   
#   # Category nodes
#   geom_node_point(
#     data = . %>% filter(node_type == "category"),
#     aes(x, y), size = 12, shape = 21,
#     fill = "grey25", colour = "white", stroke = 0.5
#   ) +
#   geom_node_label(
#     data = . %>% filter(node_type == "category"),
#     aes(x, y, label = str_wrap(name, 20)),
#     size = 2.8, fontface = "bold", colour = "white",
#     fill = "grey25", label.size = 0, repel = TRUE
#   ) +
#   
#   # Gene nodes — split pie
#   geom_scatterpie(
#     data = pie_data,
#     aes(x = x, y = y, r = radius),
#     cols = c("logFC_hs", "logFC_mmu"),
#     colour = "white", linewidth = 0.3
#   ) +
#   
#   # Gene labels
#   geom_text_repel(
#     data = pie_data,
#     aes(x = x, y = y, label = gene),
#     size = 2.5, max.overlaps = 20,
#     segment.colour = "grey60", segment.size = 0.3
#   ) 
#   
#   scale_fill_gradient2(
#     low = "#4575B4", mid = "white", high = "#D73027",
#     midpoint = 0, name = "logFC",
#     limits = c(-2, 2), oob = scales::squish   # adjust limits to your data
#   ) +
#   
#   coord_fixed() +
#   theme_graph(base_size = 11) +
#   theme(legend.position = "right")
```

Here, I will just plot the CNET plot to show the relationship between genes within the subnets, then a logFC value

```{r}
# a function for plotting as a "CNET plot"
plot.cnet = function(plot.title, ORA.cluster.object, org.db.object) {

cnetplot(
  x = setReadable(ORA.cluster.object, 
                  OrgDb = org.db.object, keyType = "ENTREZID"),
  # foldChange = fc.entrez,
  showCategory = 20,
  size_item = 2,
  circular = FALSE, 
  color_category='grey20', 
  size_category = 0.5,
  color_item = "grey90", 
  colorEdge = F
) +
  scale_color_gradient2(name='logFC', low='blue4', high='firebrick') + 
  ggtitle(plot.title)
  
}
```

```{r}
plot.hm = function(toptable, genes2plot, species = "mouse", mm2hs = NULL) {
  
  # Helper to join and filter based on species
  prepare_mat = function(value.col) {
  df = toptable %>% 
    bind_rows()
  
  if (species == "mouse") {
    df = df %>%
      left_join(mm2hs, by = c("gene" = "mm_symbol")) %>%
      dplyr::filter(hs_symbol %in% genes2plot) %>%
      dplyr::distinct(coef, hs_symbol, .keep_all = TRUE)
    gene.col = "hs_symbol"
  } else {
    df = df %>%
      dplyr::filter(gene %in% genes2plot) %>%
      dplyr::distinct(coef, gene, .keep_all = TRUE)
    gene.col = "gene"
  }
  
  # Now filter AFTER the join, when logFC and FDR are guaranteed to exist
  sig.genes = df %>%
    dplyr::filter(FDR < 0.1) %>%
    dplyr::pull(!!sym(gene.col)) %>%
    unique()
  
  df %>%
    dplyr::filter(!!sym(gene.col) %in% sig.genes) %>%
    dplyr::select(coef, all_of(gene.col), all_of(value.col)) %>%
    pivot_wider(names_from = all_of(gene.col), values_from = all_of(value.col)) %>%
    dplyr::select(coef, any_of(sig.genes)) %>%
    column_to_rownames("coef") %>%
    as.matrix()
}
  logFC_mat = prepare_mat("logFC")
  FDR_mat   = prepare_mat("FDR")
  
  Heatmap(
    matrix = logFC_mat,
    name   = "logFC",
    col    = colorRamp2(c(-5, 0, 5), c("#2166AC", "#F7F7F7", "#B2182B")),
    
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (!is.na(FDR_mat[i, j]) && FDR_mat[i, j] < 0.05) {
        grid.text("*", x, y, gp = gpar(fontsize = 15))
      }
    },
    
    column_names_gp  = gpar(fontsize = 10),
    border           = TRUE,
    show_row_names   = TRUE,
    show_column_names = TRUE,
    cluster_rows     = TRUE,
    cluster_columns  = FALSE
  )
}


plot.hm(toptable = toptables$npc1, 
        genes2plot = communities.genes.hs$lysoPD$C290,  
        species = "mouse", mm2hs = mm2hs)
```


#### PD - C290
```{r}

plot.cnet(
          org.db.object = org.Hs.eg.db,
          plot.title = "  LysoPD C290", 
          ORA.cluster.object = ORA.communities$PD$hs$C290)
  

plot.hm(toptable = toptables$npc1, 
        genes2plot = communities.genes.hs$lysoPD$C290,  
        species = "mouse", mm2hs = mm2hs) 

plot.hm(toptable = toptables$hs.ACC[toptables$hs.ACC%>% names %>% grep(pattern = "^PD")], 
        genes2plot = communities.genes.hs$lysoPD$C290, species = "human", mm2hs = mm2hs)
```

#### DLB - C437
```{r}

plot.cnet(
          org.db.object = org.Hs.eg.db,
          plot.title = "  LysoDLB C437", 
          ORA.cluster.object = ORA.communities$DLB$hs$C437)
  

plot.hm(toptable = toptables$npc1, 
        genes2plot = communities.genes.hs$lysoDLB$C437,  
        species = "mouse", mm2hs = mm2hs) 

plot.hm(toptable = toptables$hs.ACC[toptables$hs.ACC%>% names %>% grep(pattern = "^DLB")], 
        genes2plot = communities.genes.hs$lysoDLB$C437, species = "human", mm2hs = mm2hs)
```

#### DLB - C438
```{r}

plot.cnet(
          org.db.object = org.Hs.eg.db,
          plot.title = "  LysoDLB C438", 
          ORA.cluster.object = ORA.communities$DLB$hs$C438)
  

plot.hm(toptable = toptables$npc1, 
        genes2plot = communities.genes.hs$lysoDLB$C438,  
        species = "mouse", mm2hs = mm2hs) 

plot.hm(toptable = toptables$hs.ACC[toptables$hs.ACC%>% names %>% grep(pattern = "^DLB")], 
        genes2plot = communities.genes.hs$lysoDLB$C438, species = "human", mm2hs = mm2hs)
```

## correlation between netcoloc genes & some clinical indicators
```{r}
clinical.info = read_excel("data/public data/GSE178146_RAW/sample meta401_2021_2343_MOESM2_ESM.xlsx", 
                           sheet = "Sample information", skip = 1) %>%
  dplyr::rename(sample = sample_id)

y.hs.DLB.PD$samples %<>%
  mutate(sample = case_when(
    str_detect(sample, "^C\\d+") ~ str_replace(sample, "^C(\\d+)", "C0\\1"),
    TRUE ~ sample
  ))
```

Some of the netcoloc genes correlate with disease duration
```{r}
y.hs.DLB.PD$samples %<>% 
  left_join(clinical.info, by = "sample")
```


```{r}
cpm(y.hs.DLB.PD, log = T) %>% 
  .["PLD3",] %>% 
  as.data.frame() %>% 
  rownames_to_column("sample") %>% 
  as_tibble() %>% 
  set_colnames(c("sample_id", "logCPM")) %>% 
  left_join(y.hs.DLB.PD$samples) %>% 
  dplyr::filter(disease_group != "Control") %>% 
  ggscatter(
    x = "logCPM", y = "aSN",
   # color = "disease_group",
    add = "reg.line",  # Add regressin line
   conf.int = TRUE, # Add confidence interval
   
  ) +
  stat_cor(
    aes(label = paste(..rr.label.., ..p.label.., sep = "~`,`~"), 
       # colour = disease_group
        ), 
    label.x = 1,
    method = "spearman"
    )+
  facet_wrap(~cluster)

cpm(y.hs.DLB.PD, log = T) %>% 
  .["HGSNAT",] %>% 
  as.data.frame() %>% 
  rownames_to_column("sample") %>% 
  as_tibble() %>% 
  set_colnames(c("sample_id", "logCPM")) %>% 
  left_join(y.hs.DLB.PD$samples) %>% 
  ggplot(
    aes(x = DD, y = logCPM, colour = disease)
    ) +
  geom_boxplot() +
  geom_jitter() +
  facet_wrap(~cluster)

```

```{r}
clinical.vars = c(
  "AoO", 
  "DD", 
  "thal AB",
  "aSN", 
  "TAU"
  )

# Extract logCPM and join clinical
expr.df = cpm(y.hs.DLB.PD, log = T) %>%
  .[shared.de, , drop = F] %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample_id") %>%
  left_join(y.hs.DLB.PD$samples, by = "sample_id") %>%
  as_tibble()

# Compute correlations
cor.long = expr.df %>%
  pivot_longer(all_of(genes), names_to = "gene", values_to = "expr") %>%
  group_by(cluster, gene) %>%
  summarise(
    map_dfr(clinical.vars, function(cv) {
      vals = cur_data()[[cv]]
      if (sum(!is.na(vals) & !is.na(expr)) < 5)
        return(tibble(clinical_var = cv, r = NA_real_, p = NA_real_))
      ct = cor.test(expr, vals, method = "spearman", use = "complete.obs")
      tibble(clinical_var = cv, r = as.numeric(ct$estimate), p = ct$p.value)
    }),
    .groups = "drop"
  )

r.mat = cor.long %>%
  mutate(row = paste(gene, cluster, sep = " | ")) %>%
  dplyr::select(row, clinical_var, r) %>%
  pivot_wider(names_from = clinical_var, values_from = r) %>%
  column_to_rownames("row") %>%
  as.matrix()

p.mat = cor.long %>%
  mutate(row = paste(gene, cluster, sep = " | ")) %>%
  dplyr::select(row, clinical_var, p) %>%
  pivot_wider(names_from = clinical_var, values_from = p) %>%
  column_to_rownames("row") %>%
  as.matrix()

Heatmap(
  matrix = r.mat,
  name   = "Spearman r",
    row_split = row.ann$Gene,
  col    = colorRamp2(c(-1, 0, 1), c("#2166AC", "#F7F7F7", "#B2182B")),
  
   na_col = "grey80",
  
  clustering_distance_rows = function(m) dist(m, method = "euclidean"),
  clustering_distance_columns = function(m) dist(m, method = "euclidean"),
  
  
  cell_fun = function(j, i, x, y, width, height, fill) {
    if (!is.na(p.mat[i, j]) && p.mat[i, j] < 0.05)
      grid.text("*", x, y, gp = gpar(fontsize = 12))
  },
  
  
  column_names_gp = gpar(fontsize = 12),
  row_names_gp    = gpar(fontsize = 12), 
  border          = TRUE,
  cluster_rows    = T,
  cluster_columns = T
)
```

```{r}
cpm(y.hs.DLB.PD, log = T) %>% 
  .["PLD3",] %>% 
  as.data.frame() %>% 
  rownames_to_column("sample") %>% 
  as_tibble() %>% 
  set_colnames(c("sample_id", "logCPM")) %>% 
  left_join(y.hs.DLB.PD$samples) %>% 
  mutate(aSN = as.numeric(aSN)) %>% 
  ggboxplot(
    x = "aSN",
    y = "logCPM",
    # add = "jitter",
    outlier.shape = NA,
    facet.by = "cluster"
  ) + 
  geom_jitter(
    aes(colour = disease), 
    size = 3
  ) +
  stat_compare_means(
    method = "kruskal.test",
    label = "p.format",
    hide.ns = TRUE,
    label.y.npc = 0.95          # overall p at top
  ) +
  stat_compare_means(
    ref.group = "0",
    vjust = 0.7, 
    size = 10,
    label = "p.signif",     # stars on bars vs stage 0
    hide.ns = TRUE,            # hide ns to keep it cleany
  )+
  labs(
    title = "PLD3"
  )

cpm(y.hs.DLB.PD, log = T) %>% 
  .["HGSNAT",] %>% 
  as.data.frame() %>% 
  rownames_to_column("sample") %>% 
  as_tibble() %>% 
  set_colnames(c("sample_id", "logCPM")) %>% 
  left_join(y.hs.DLB.PD$samples) %>% 
  dplyr::filter(disease != "Control") %>% 
  ggscatter(
    x = "DD",
    y = "logCPM",
    color = "disease",
    add    = "reg.line",
    facet.by = "cluster"
  ) + 
  stat_cor(
    method      = "spearman",
    aes(group = disease, colour = disease),  
    label.x.npc = 0.05,
    label.y.npc = 0.95
  ) +
  labs(title = "HGSNAT", x = "Disease duration")
  
```





# plots 4 paper
## Figure 1
### a - coloc pos/neg controls
```{r}
f1.a <- enrichment_dfsv1 %>%
  mutate(
    FDR = p.adjust(empirical_p), 
    control = case_when(
      comparison == "CD.all\n&\nbrain.malform" ~ "negative",
      TRUE ~ "positive"
    )
  ) %>% 
  arrange(control) %>% #.$comparison %>% unique
  mutate(
    comparison = str_replace(comparison, "CD.all", "All childhood dementias"),
    comparison = str_replace(comparison, "brain.malform", "Brain malformations"),
    comparison = str_replace(comparison, "adult.all", "All adult-onset neurodegen"),
    comparison = str_replace(comparison, "CD.lyso", "Lysosomal\nchildhood dementias"),
    comparison = str_replace(comparison, "nonCD.lyso", "Non-neuronopathic\nlysosomal diseases"),
    comparison = str_replace(comparison, "CD.amino", "Disorders of amino acid\n& other organic acid\nmetabolism"),
    comparison = str_replace(comparison, "nonCD.amino", "Non-neuronopathic\namino acid diseases"),
    comparison = str_replace(comparison, "CD.perox", "Peroxisomal\nchildhood dementias"),
    comparison = str_replace(comparison, "nonCD.perox", "Non-neuronopathic\nperoxisomal diseases"),
    comparison = str_replace(comparison, "CD.mito", "Mitochondrial\nchildhood dementias"),
    comparison = str_replace(comparison, "nonCD.mito", "Non-neuronopathic\nmitochondrial diseases"),
    
    FDR = case_when(
      FDR == 0 ~ "p.FDR < 0.000001", 
      FDR < 0.000001 ~ "p.FDR < 0.000001",
      FDR == 1 ~ "p.FDR = 1", 
      TRUE ~ FDR %>% sprintf("p.FDR = %.1e", .)
    ),
    comparison = fct_inorder(comparison)
  ) %>% 
  ggplot(
    aes(x = type, y = value, fill = type)
    ) +
  geom_bar(
    stat = "identity", position = "dodge", width = 0.8
  ) +
  geom_errorbar(
    data = . %>% 
      dplyr::filter(type == "expected"),
    aes(
      ymin = value - 2 * expected_overlap_std,
      ymax = value + 2 * expected_overlap_std),
    width = 0.25
  ) + 
  geom_label(
    data =  . %>%
      dplyr::filter(type == "expected"),
    aes(
      x = 1.5, y =800,
      label = FDR),
    inherit.aes = FALSE,
    fill = NA,
    border.colour = NA,
    colour = "grey20",
    size = 5,
    vjust = 1,
    fontface = "bold",
    show.legend = FALSE
  )  +
  labs(
    y = "Number of genes", 
    x = ""
  ) +
  facet_wrap(~comparison, nrow = 1, scales = "free") +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("grey50", "skyblue"))+
   theme(
   text = element_text(size = 14),
    legend.position = "none",
    axis.title.x = element_blank(),
    
    plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) 

f1.a
```
### b: ORA
```{r}

f1.b <- 
  ORA.network.genes.poscontrols.results %>% 
  map(as_tibble) %>% 
  bind_rows(.id = "comparison") %>% 
  separate(GeneRatio,into = c("temp1", "temp2"), remove = F) %>% 
  separate(BgRatio, into = c("bg1", "bg2"), sep = "/", remove = FALSE) %>% 
   group_by(comparison) %>% 
  dplyr::slice(1:7) %>% 
  mutate(
    across(c(temp1, temp2, bg1, bg2), as.numeric),
    
    a = temp1 + 0.5,
    b = bg1 - temp1 + 0.5,
    c = temp2 - temp1 + 0.5,
    d = bg2 - bg1 - (temp2 - temp1) + 0.5,
    
    OR    = (a * d) / (b * c),
    log2OR = log2(OR),
    SE    = sqrt(1/a + 1/b + 1/c + 1/d),
    CI_lo = exp(log(OR) - 1.96 * SE),
    CI_hi = exp(log(OR) + 1.96 * SE),
    
  
    Description = reorder_within(Description, OR, comparison), 
    Description = str_wrap(Description, width = 30),
  
  
    GeneRatio = as.numeric(temp1)/as.numeric(temp2), 
    comparison = str_replace(comparison, pattern = "_", "\n&\n"),
    comparison = str_replace(comparison, "CD.all", "All childhood dementias"),
    comparison = str_replace(comparison, "brain.malform", "Brain malformations"),
    comparison = str_replace(comparison, "adult.all", "All adult-onset neurodegen"),
    comparison = str_replace(comparison, "CD.lyso", "Lysosomal\nchildhood dementias"),
    comparison = str_replace(comparison, "nonCD.lyso", "Non-neuronopathic\nlysosomal diseases"),
    comparison = str_replace(comparison, "CD.amino", "Disorders of amino acid\n& other organic acid\nmetabolism"),
    comparison = str_replace(comparison, "nonCD.amino", "Non-neuronopathic\namino acid diseases"),
    comparison = str_replace(comparison, "CD.perox", "Peroxisomal\nchildhood dementias"),
    comparison = str_replace(comparison, "nonCD.perox", "Non-neuronopathic\nperoxisomal diseases"),
    comparison = str_replace(comparison, "CD.mito", "Mitochondrial\nchildhood dementias"),
    comparison = str_replace(comparison, "nonCD.mito", "Non-neuronopathic\nmitochondrial diseases"),
  ) %>% 
  
  ggplot(
    aes(
      y = reorder(Description, OR), x = log2OR
    )
    
    
  ) +
  geom_point(
    aes(
      colour = -log10(p.adjust), size = temp1
      )
  ) +
  geom_errorbarh(
    aes(xmin = log2(CI_lo), xmax = log2(CI_hi)),
    height = 0.3, linewidth = 0.6
    ) +
 #scale_x_continuous(limits = c(0,12))+
  scale_y_reordered() +  # This removes the `___comparison` suffix added by reorder_within()
  scale_colour_viridis_c() +
   geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  labs(
    x = "log(OR) ± 95% CI",
    y = NULL,
    size = "Gene count"
  ) +
  facet_wrap(~comparison, 
             scales = "free", nrow = 2)+
   theme(
    text = element_text(size = 14),
    legend.position = "top",
    #axis.title.x = element_blank(),
    # plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) 
f1.b
```

### final fig
```{r}
f1.a / f1.b +
  plot_layout(heights = c(1,3)) +
  plot_annotation(tag_levels = "A") & 
  theme(plot.tag = element_text(face = 'bold', size = 30)) +
  ggsave(
    #"output/plots4paperv2/Fig1.pdf", 
    "output/plots4paperv2/Fig1.png", 
    width = 25, height = 22, units = "cm", 
    dpi = 800,
    scale = 1.8
  )
```

## Fig.2 
### a - netcoloc pan neurodegen
```{r}
f2.a = input.f2a %>% 
  ggplot(
    aes(x = type, y = value, fill = type)
    ) +
  geom_bar(
    stat = "identity", 
    position = "dodge", 
    width = 0.5
  ) +
  geom_errorbar(
    data = . %>% 
      dplyr::filter(type == "expected"),
    aes(
      ymin = value - 2 * expected_overlap_std,
      ymax = value + 2 * expected_overlap_std),
    width = 0.25
  ) + 
  geom_label(
    data =  . %>%
      dplyr::filter(type == "expected"),
    aes(
      x = 1.5, y =800,
      label = FDR),
    inherit.aes = FALSE,
    fill = NA,
    border.colour = NA,
    colour = "grey20",
    size = 5,
    vjust = 1,
    fontface = "bold",
    show.legend = FALSE
  )  +
  labs(
    y = "Number of genes", 
    x = "", 
    title = "Pan-neurodegenerative disease genetics"
  ) +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("grey50", "skyblue")) +
   theme(
    strip.text = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "none",
    axis.title.x = element_blank(),
    plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) 
f2.a
```
### b - ORA whole subnet
```{r}
f2.b <- 
  ORA.network.genes.pan.peurodegen.results %>% 
  as_tibble %>% 
  mutate(
    source = "KEGG"
  ) %>% 
  bind_rows(
    ORA.network.genes.pan.peurodegen.results.MPO %>% 
      as_tibble %>% 
      mutate(
        source = "MPO"
      )
  ) %>% 
  separate(GeneRatio,into = c("temp1", "temp2"), remove = F) %>% 
  separate(BgRatio, into = c("bg1", "bg2"), sep = "/", remove = FALSE) %>% 
  mutate(
    across(c(temp1, temp2, bg1, bg2), as.numeric),
    
    a = temp1,
    b = bg1 - temp1,
    c = temp2 - temp1,
    d = bg2 - bg1 - (temp2 - temp1),
    
    OR    = (a * d) / (b * c),
    log2OR = log2(OR),
    SE    = sqrt(1/a + 1/b + 1/c + 1/d),
    CI_lo = exp(log(OR) - 1.96 * SE),
    CI_hi = exp(log(OR) + 1.96 * SE)
  ) %>% 
  mutate(
    Description = reorder_within(Description, OR, source), 
    #Description = str_wrap(Description, width = 40)
  ) %>% 
  ggplot(
    aes(
      y = reorder(Description, OR), x = log2OR
    )
  ) +
  geom_point(
    aes(
      colour = -log10(p.adjust), size = temp1
      )
  ) +
  geom_errorbarh(
    aes(xmin = log2(CI_lo), xmax = log2(CI_hi)),
    height = 0.3, linewidth = 0.6
    ) +
  scale_y_reordered() +  # This removes the `___comparison` suffix added by reorder_within()
  scale_colour_viridis_c() +
   geom_vline(xintercept = 1, linetype = "dashed", color = "grey50") +
  labs(
    x = "log(OR) ± 95% CI",
    y = NULL,
    size = "Gene count", 
    title = "Over-represented KEGG and MPO terms"
  ) +
  facet_wrap(~source, 
             scales = "free_y", nrow = 1)+
   theme(
    strip.text = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom",
    #axis.title.x = element_blank(),
    plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) 

f2.b +
  ggsave("output/plots4paperv2/Fig2b.pdf", 
         width = 30, height = 12, units = "cm", 
         scale = 1.2
         )
```

### final fig
```{r}
(f2.a + theme(aspect.ratio = 1) + 
   f2.b) +
  plot_layout(
    widths = c(1,3), 
  ) +
  plot_annotation(tag_levels = c("A", "C")) &
  theme(plot.tag = element_text(face = 'bold', colour = "black", size = 30))
  ggsave(
    "output/plots4paperv2/Fig2.pdf", 
    #"output/plots4paperv2/Fig2.png", 
    width = 39, height = 15, units = "cm", 
    #dpi = 800,
    scale = 1.4
  )

```





## Figure 3: 
```{r}
enrichment_dfs.2 %>% 
  dplyr::filter(
    z_comb == 3, z_12 == 1.5, 
    
    ) %>% 
  mutate(
    # use the childhood vector to extract the CD seed gene name
    CD = str_remove(comparison, "_(AD|ALS|PD|DLB|FTLD|brain_malform)$"), 
    adult = str_extract(comparison, "(AD|ALS|PD|DLB|FTLD|brain_malform)$"), 
    FDR = p.adjust(empirical_p, "fdr"),
    CD = case_when(
      grepl(CD, pattern = "1A") ~ "Lysosomal disorders", 
      grepl(CD, pattern = "1B") ~ "Other disorders of lipid metabolism & transport", 
      grepl(CD, pattern = "1C") ~ "Disorders of amino acid & other organic acid metabolism", 
      grepl(CD, pattern = "1D") ~ "Vitamin-responsive inborn errors of metabolism", 
      grepl(CD, pattern = "1E") ~ "Disorders of mineral absorption & transport", 
      grepl(CD, pattern = "1F") ~ "Peroxisomal diseases", 
      grepl(CD, pattern = "1G") ~ "Mitochondrial disorders", 
      grepl(CD, pattern = "1H") ~ "Other inborn errors of metabolism", 
      grepl(CD, pattern = "2_") ~ "Leukodystrophies not otherwise categorised", 
      grepl(CD, pattern = "3_") ~ "Neurodegeneration with brain iron accumulation", 
      grepl(CD, pattern = "4_") ~ "Diseases not otherwise categorised",
      TRUE ~ CD) %>% str_wrap(width = 30), 
    
    facet.order = case_when(
      grepl(comparison, pattern = "1A") ~ 1, 
      grepl(comparison, pattern = "1E") ~ 2, 
      grepl(comparison, pattern = "3") ~ 3, 
      grepl(comparison, pattern = "1B") ~ 4, 
      grepl(comparison, pattern = "1H") ~ 9, 
      grepl(comparison, pattern = "4") ~ 9, 
      TRUE ~ 8
    )
    
    ) %>%  
  dplyr::filter(type == "observed") %>% 
  
  ggplot(
    aes(x = obs_exp, y = reorder(adult, obs_exp))
  ) +
  geom_col(
    aes(fill = -log10(FDR), alpha = FDR < 0.05), 
    colour = "grey50", position = "dodge", 
    width = 0.75
  ) +
  scale_fill_viridis_c() +
  scale_alpha_manual(values = c(0.1, 1)) +
  labs(
    x = "observed / expected", 
    y = "Disease"
  ) +
  facet_wrap(~reorder(CD, facet.order), ncol = 3) +
  theme(
    text = element_text(size = 16)
  ) 

  ggsave(
    "output/plots4paperv2/Fig3.pdf", 
    #"output/plots4paperv2/Fig3.png", 
    width = 24, height = 12, units = "cm", 
    #dpi = 800,
    scale = 1.5
  )
```
## Figure 4: LysoDLB and LysoPD communities
A, B are from cytoscpae. 

```{r}
pdf("output/plots4paperv2/Fig4.pdf", width = 6, height = 5)
pheatmap(
  signif(sim_filtered,1),
  display_numbers = TRUE,
  main = "Jaccard index of LysoPD and LysoDLB network communities",
  color = colorRampPalette(c("white", "red"))(100)
)
dev.off()
```

## Fig.5 scRNAseq
### UMAP
```{r}
fig.5a =data.seurat.hs.DLB.PD %>% 
    DimPlot(
        group.by = "celltype", 
        label = TRUE,
        repel = T,
        label.color = "black",
        cols = cell2colour, 
    )  
fig.5a = fig.5a +
  scale_x_continuous(
    breaks = quantile(fig.5a[[1]]$data$umap_1, prob = 0.15), 
    guide = guide_axis(cap = 'upper')
    ) +
  scale_y_continuous(
    breaks = quantile(fig.5a[[1]]$data$umap_1, prob = 0.15), 
    guide = guide_axis(cap = 'upper')
    )  +
    coord_fixed() +
   labs(
    title = "Human anterior cingulate cortex", 
    subtitle = "Control, DLB, PD and PDD"
  ) +
    theme(
      axis.text = element_blank(), 
      plot.title = element_text(size = 14, face = "bold"),
      
      plot.subtitle = element_text(size = 14, hjust = 0.5),
      legend.position = "none", 
      aspect.ratio = 1,
      axis.ticks = element_blank(),
      axis.title = element_text(hjust = 0),
      axis.line = element_line(arrow = arrow(length = unit(0.4, "cm"), type="closed"))
    )

fig.5a

fig.5b = data.seurat.mm.Npc1 %>% 
    DimPlot(
        group.by = "celltype", 
        label = TRUE,
        repel = T,
        label.color = "black",
        cols = cell2colour, 
    )  

fig.5b = fig.5b +
  scale_x_continuous(
    breaks = quantile(fig.5b[[1]]$data$umap_1, prob = 0.1), 
    guide = guide_axis(cap = 'upper')
    ) +
  scale_y_continuous(
    breaks = quantile(fig.5b[[1]]$data$umap_2, prob = 0.1), 
    guide = guide_axis(cap = 'upper')
    )  +
  coord_fixed() +
  labs(
    title = "Mouse cerebellum", 
    subtitle = "Npc1 -/- and WT"
  ) +
  theme(
    axis.text = element_blank(), 
    plot.title = element_text(size = 14, face = "bold"),
    
    plot.subtitle = element_text(size = 14, hjust = 0.5),
    legend.position = "none", 
    aspect.ratio = 1,
    axis.ticks = element_blank(),
    axis.title = element_text(hjust = 0),
    axis.line = element_line(arrow = arrow(length = unit(0.4, "cm"), type="closed"))
  )

fig.5b
```

```{r}
fig.5a + fig.5b +
  ggsave(
    #"output/plots4paperv2/Fig5.pdf", 
    "output/plots4paperv2/fig5/Fig5.png",
    width = 15, height = 15, units = "cm", 
    #dpi = 800,
    scale = 1.4
  )
  
```

### B - DE network genes
```{r}
# Build the FDR matrix (for asterisk overlay)
FDR_mat <- toptables$hs.ACC %>%
  bind_rows() %>%
  dplyr::filter(
    gene %in% shared.de, 
    grepl(coef, pattern = "^DLB")
  ) %>%
  dplyr::select(coef, gene, FDR) %>%
  pivot_wider(names_from = gene, values_from = FDR) %>%
  dplyr::select(coef, shared.de) %>% 
  column_to_rownames("coef") %>%
  as.matrix()

fig.6a  = Heatmap(
  matrix = toptables$hs.ACC %>%
    bind_rows() %>%
    dplyr::filter(
      gene %in% shared.de, 
      grepl(coef, pattern = "^DLB")
    ) %>%
    dplyr::select(coef, gene, logFC) %>%

    pivot_wider(names_from = gene, values_from = logFC) %>%
    dplyr::select(coef, shared.de) %>% 
    column_to_rownames("coef") %>%
    as.matrix(),
  
  name = "logFC",
  col = colorRamp2(c(-5, 0, 5), c("#2166AC", "#F7F7F7", "#B2182B")),
  
  cell_fun =function(j, i, x, y, width, height, fill) {
      # j = column index, i = row index
      # x, y = coordinates of the center of the cell
      # width, height = size of the cell
      # fill = background color (not used here)

      # Check if the FDR value for this cell exists and is below 0.05
      if (!is.na(FDR_mat[i, j]) && FDR_mat[i, j] < 0.05) {
        # If yes, overlay a text label "*" on that cell
        grid.text("*", x, y, gp = gpar(fontsize = 15))
      }
    },
  column_names_gp = gpar(fontsize = 10),
  border = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  cluster_rows = TRUE,
  cluster_columns = F
)
```

```{r}
# Build the FDR matrix (for asterisk overlay)
  FDR_mat <- toptables$npc1 %>%
  bind_rows() %>%
  left_join(mm2hs, by = c("gene" = "mm_symbol")) %>% 
  dplyr::filter(
    hs_symbol %in% shared.de,
  ) %>%
  dplyr::distinct(coef, hs_symbol, .keep_all = T) %>% 
  dplyr::select(coef, hs_symbol, FDR) %>%
    pivot_wider(names_from = hs_symbol, values_from = FDR) %>%
  dplyr::select(coef, shared.de) %>% 
    column_to_rownames("coef") %>%
    as.matrix()

fig.6b = Heatmap(
  matrix = toptables$npc1 %>%
    bind_rows() %>%
    left_join(mm2hs, by = c("gene" = "mm_symbol")) %>% 
    dplyr::filter(
      hs_symbol %in% shared.de,
    ) %>%
    dplyr::distinct(coef, hs_symbol, .keep_all = T) %>% 
    dplyr::select(coef, hs_symbol, logFC) %>%
    pivot_wider(names_from = hs_symbol, values_from = logFC) %>%
    dplyr::select(coef, shared.de) %>% 
    column_to_rownames("coef") %>%
    as.matrix(),
  
  name = "logFC",
  col = colorRamp2(c(-5, 0, 5), c("#2166AC", "#F7F7F7", "#B2182B")),
  
  cell_fun =function(j, i, x, y, width, height, fill) {
      # j = column index, i = row index
      # x, y = coordinates of the center of the cell
      # width, height = size of the cell
      # fill = background color (not used here)

      # Check if the FDR value for this cell exists and is below 0.05
      if (!is.na(FDR_mat[i, j]) && FDR_mat[i, j] < 0.05) {
        # If yes, overlay a text label "*" on that cell
        grid.text("*", x, y, gp = gpar(fontsize = 15))
      }
    },
  column_names_gp = gpar(fontsize = 10),
  border = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  cluster_rows = T,
  cluster_columns = F
)
```

#### final fig
```{r}
pdf("output/plots4paperv2/fig5/Fig5c.pdf", width = 6, height = 5)
fig.6a %v% fig.6b
dev.off()
```

### correate with some clinical vars
```{r}
clinical.vars = c(
  "AoO", 
  "DD", 
  "aSN", 
  "TAU"
  )

# Extract logCPM and join clinical
expr.df = cpm(y.hs.DLB.PD, log = T) %>%
  .[shared.de, , drop = F] %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column("sample_id") %>%
  left_join(y.hs.DLB.PD$samples, by = "sample_id") %>%
  as_tibble()

# Compute correlations
cor.long = expr.df %>%
  pivot_longer(all_of(genes), names_to = "gene", values_to = "expr") %>%
  group_by(cluster, gene) %>%
  summarise(
    map_dfr(clinical.vars, function(cv) {
      vals = cur_data()[[cv]]
      if (sum(!is.na(vals) & !is.na(expr)) < 5)
        return(tibble(clinical_var = cv, r = NA_real_, p = NA_real_))
      ct = cor.test(expr, vals, method = "spearman", use = "complete.obs")
      tibble(clinical_var = cv, r = as.numeric(ct$estimate), p = ct$p.value)
    }),
    .groups = "drop"
  )

r.mat = cor.long %>%
  mutate(row = paste(gene, cluster, sep = " | ")) %>%
  dplyr::select(row, clinical_var, r) %>%
  pivot_wider(names_from = clinical_var, values_from = r) %>%
  column_to_rownames("row") %>%
  as.matrix()

p.mat = cor.long %>%
  mutate(row = paste(gene, cluster, sep = " | ")) %>%
  dplyr::select(row, clinical_var, p) %>%
  pivot_wider(names_from = clinical_var, values_from = p) %>%
  column_to_rownames("row") %>%
  as.matrix()
```


```{r}
pdf("output/plots4paperv2/fig5/Fig5d.pdf", width = 5, height = 10)
Heatmap(
  matrix = r.mat,
  name   = "Spearman r",
    row_split = row.ann$Gene,
  col    = colorRamp2(c(-1, 0, 1), c("#2166AC", "#F7F7F7", "#B2182B")),
  
   na_col = "grey80",
  
  clustering_distance_rows = function(m) dist(m, method = "euclidean"),
  clustering_distance_columns = function(m) dist(m, method = "euclidean"),
  
  
  cell_fun = function(j, i, x, y, width, height, fill) {
    if (!is.na(p.mat[i, j]) && p.mat[i, j] < 0.05)
      grid.text("*", x, y, gp = gpar(fontsize = 12))
  },
  
  
  column_names_gp = gpar(fontsize = 12),
  row_names_gp    = gpar(fontsize = 12), 
  border          = TRUE,
  cluster_rows    = T,
  cluster_columns = T
)
dev.off()
```



```{r}
#f5.e = 
  cpm(y.hs.DLB.PD, log = T) %>% 
  .["PLD3",] %>% 
  as.data.frame() %>% 
  rownames_to_column("sample") %>% 
  as_tibble() %>% 
  set_colnames(c("sample_id", "logCPM")) %>% 
  left_join(y.hs.DLB.PD$samples) %>% 
  mutate(aSN = as.numeric(aSN)) %>% 
  ggboxplot(
    x = "aSN",
    y = "logCPM",
    # add = "jitter",
    outlier.shape = NA,
  ) + 
  geom_jitter(
    aes(colour = disease, shape = sex), 
    size = 3
  ) +
  stat_compare_means(
    method = "kruskal.test",
    label = "p.format",
    hide.ns = TRUE,
    label.y.npc = 0.95          # overall p at top
  ) +
  stat_compare_means(
    ref.group = "0",
    vjust = 0.7, 
    size = 10,
    label = "p.signif",     # stars on bars vs stage 0
    hide.ns = TRUE,            # hide ns to keep it cleany
  )+
  labs(
    title = "PLD3", 
    x = "Braak a-synuclein stage"
  ) +
  facet_wrap(~cluster, ncol = 2) +
  scale_color_manual(values = c(
    Control = "#999999",
    DLB = "#1B9E77",
    PD = "#E69F00",
    PDD = "#A67000")) +
  theme(
    text = element_text(size = 16)
  ) +

#f5.f = 
  cpm(y.hs.DLB.PD, log = T) %>% 
  .["HGSNAT",] %>% 
  as.data.frame() %>% 
  rownames_to_column("sample") %>% 
  as_tibble() %>% 
  set_colnames(c("sample_id", "logCPM")) %>% 
  left_join(y.hs.DLB.PD$samples) %>% 
  dplyr::filter(disease != "Control") %>% 
  ggscatter(
    x = "DD",
    y = "logCPM",
    shape = "sex",
    size = 3,
    color = "disease",
    add    = "reg.line",
    # facet.by = "cluster", 
  ) + 
  stat_cor(
    method      = "spearman",
    show.legend = F,
    aes(group = disease, colour = disease),  
    label.x.npc = "left",
    label.y.npc = "top", 
  ) +
  facet_wrap(~cluster, ncol = 2) +
  labs(title = "HGSNAT", x = "Disease duration") +
  scale_x_continuous(
    limits = c(-8,20), 
    breaks = seq(0,20, by = 5)
    ) +
  scale_color_manual(values = c(
    Control = "#999999",
    DLB = "#1B9E77",
    PD = "#E69F00",
    PDD = "#A67000"))+
  theme(
    text = element_text(size = 16)
  )

ggarrange(
  f5.e, f5.f, 
  common.legend = T,
  ncol = 1,
  heights = c(1.7,2)
) +
  ggsave(
    "output/plots4paperv2/fig5/Fig5ef.pdf", 
    width = 8, height = 16, units = "cm", 
    #dpi = 800,
    scale = 2.4
  )
```




### B,C DE communiities 
```{r}
ggarrange(
  fry.input.PD %>% 
    dplyr::filter(pathway %in% keep.PD.fry) %>%
    mutate(
      source = case_when(
        grepl(coef, pattern = "PD_") ~ "Human PD", 
        grepl(coef, pattern = "PDD_") ~ "Human PDD", 
        grepl(coef, pattern = "3week") ~ "Mouse Npc1\npre-symptomatic" ,
        grepl(coef, pattern = "7week") ~ "Mouse Npc1\nsymptomatic" ,
      ),
      celltype = str_remove(coef, pattern = "^PDD_"),
      celltype = str_remove(celltype, pattern = "^PD_"),
      celltype = str_remove(celltype, pattern = " - Contro.+"), 
      celltype = str_remove(celltype, pattern = "_[3|7]week"), 
      celltype = str_remove_all(celltype, pattern = "_"), 
    ) %>% 
    ggplot(
      aes(y = reorder(pathway, -PValue.Mixed), x = celltype)
    ) +
    geom_tile(
      aes(fill = -log10(PValue.Mixed))
    ) +
    geom_label(
      aes(label = signif(FDR.Mixed, 1)), 
      data = . %>% 
        dplyr::filter(FDR.Mixed < 0.1)
    ) +
    easy_rotate_x_labels(angle = -45) +
    scale_fill_viridis_c() +
     labs(
    y = "Netcoloc community", 
    title = "LysoPD communities",
    fill = "-log10(Pvalue)"
  ) +
    theme(
      text = element_text(size = 12),
      plot.title = element_text(size = 16, face = "bold"),
      plot.background = element_blank(),
      panel.grid = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank(),
      plot.margin = margin(r = 1, unit = "cm")
    ) +
    facet_wrap(~source, scales = "free_x", space = "free_x"), 
  
  fry.input.DLB %>% 
    dplyr::filter(pathway %in% keep.DLB.fry) %>%
    mutate(
      source = case_when(
        grepl(source, pattern = "human") ~ "Human DLB", 
        grepl(coef, pattern = "3week") ~ "Mouse Npc1\npre-symptomatic" ,
        grepl(coef, pattern = "7week") ~ "Mouse Npc1\nsymptomatic" ,
      ), 
      celltype = str_remove(coef, pattern = "^DLB_"), 
      celltype = str_remove(celltype, pattern = " - Contro.+"), 
      celltype = str_remove(celltype, pattern = "_[3|7]week"), 
      celltype = str_remove(celltype, pattern = "_"), 
    ) %>% 
    ggplot(
      aes(y = reorder(pathway, -PValue.Mixed), x = celltype)
    ) +
    geom_tile(
      aes(fill = -log10(PValue.Mixed))
    ) +
    geom_label(
      aes(label = signif(FDR.Mixed, 1)), 
      data = . %>% 
        dplyr::filter(FDR.Mixed < 0.1)
    ) +
    easy_rotate_x_labels(angle = -45) +
    scale_fill_viridis_c() +
    theme(
      text = element_text(size = 12),
      plot.title = element_text(size = 16, face = "bold"),
      plot.background = element_blank(),
      panel.grid = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank(),
      plot.margin = margin(r = 1.5, unit = "cm")
    ) +
  labs(
    y = "Netcoloc community", 
    title = "LysoDLB communities",
    fill = "-log10(Pvalue)"
  ) +
    facet_wrap(~source, scales = "free_x", space = "free_x"), 
  
  common.legend = T, 
  ncol = 1
) 
    ggsave(
    "output/plots4paperv2/Fig6bc.pdf", 
    #"output/plots4paperv2/Fig3.png", 
    width = 20, height = 15, units = "cm", 
    #dpi = 800,
    scale = 1.5
  )
```
## Figure 6: LysoDLB communiites
```{r}
kegg2plot = c("Mineral absorption", 
  "Antigen processing and presentation", 
  "Lysosome biogenesis", 
  "Ferroptosis",
  "Phagosome",
  "Cell adhesion molecule (CAM) interaction",
  "Cholesterol metabolism",
  "Porphyrin metabolism",
  "PPAR signaling pathway",                      
  "Complement and coagulation cascades",         
  "Osteoclast differentiation"                  
)

ggarrange(
  ORA.communities$DLB$hs$C437 %>% 
    setReadable(OrgDb = org.Hs.eg.db, keyType = "ENTREZID") %>% 
    cnetplot(
      showCategory    = 6, 
      layout          = "nicely",
      cex.params      = list(
        category_label = 2.0,
        gene_label     = 0.8,
        category_node  = 1.3,
        gene_node      = 0.8
      )
    ) +
    labs(
      title = "LysoDLB - C437"
    ) +
    theme(
       text = element_text(size = 18),
      plot.title = element_text(hjust = 0.5, size = 20, face = "bold")
    ), 
  ORA.communities$DLB$hs$C438 %>% 
    setReadable(OrgDb = org.Hs.eg.db, keyType = "ENTREZID") %>% 
    cnetplot(showCategory = kegg2plot, layout = "nicely") + 
    labs(
      title = "LysoDLB - C438"
    ) +
    theme(
      text = element_text(size = 18),
      plot.title = element_text(hjust = 0.5, size = 20, face = "bold")
      
    ), 
  common.legend = T
) +
  ggsave(
    "output/plots4paperv2/fig6/Fig6ab.pdf", 
    #"output/plots4paperv2/Fig3.png", 
    width = 20, height = 7, units = "cm", 
    #dpi = 800,
    scale = 2
  )
```


```{r}
plot.hmv2 = function(toptable, genes2plot, species = "mouse", mm2hs = NULL,
                   communities = NULL) {
  
  prepare_mat = function(value.col) {
    df = toptable %>% 
      bind_rows()
    
    if (species == "mouse") {
      df = df %>%
        left_join(mm2hs, by = c("gene" = "mm_symbol")) %>%
        dplyr::filter(hs_symbol %in% genes2plot) %>%
        dplyr::distinct(coef, hs_symbol, .keep_all = TRUE)
      gene.col = "hs_symbol"
    } else {
      df = df %>%
        dplyr::filter(gene %in% genes2plot) %>%
        dplyr::distinct(coef, gene, .keep_all = TRUE)
      gene.col = "gene"
    }
    
    sig.genes = df %>%
      dplyr::filter(FDR < 0.1) %>%
      dplyr::pull(!!sym(gene.col)) %>%
      unique()
    
    df %>%
      dplyr::filter(!!sym(gene.col) %in% sig.genes) %>%
      dplyr::select(coef, all_of(gene.col), all_of(value.col)) %>%
      pivot_wider(names_from = all_of(gene.col), values_from = all_of(value.col)) %>%
      dplyr::select(coef, any_of(sig.genes)) %>%
      column_to_rownames("coef") %>%
      as.matrix()
  }
  
  logFC_mat = prepare_mat("logFC")
  FDR_mat   = prepare_mat("FDR")
  
  
  col_ann = NULL
  if (!is.null(communities)) {
    
    # communities is a named list of character vectors e.g. list(C437 = c(...), C438 = c(...))
    gene_community = imap_dfr(communities, function(genes, community_name) {
      tibble(gene = genes, community = community_name)
    }) %>%
      dplyr::filter(gene %in% colnames(logFC_mat)) %>%
      dplyr::distinct(gene, .keep_all = TRUE)
    
    ann_vec = gene_community$community[match(colnames(logFC_mat), gene_community$gene)]
    ann_vec[is.na(ann_vec)] = "C289"  # genes in parent but not in children
    
    col_ann = HeatmapAnnotation(
      Community = ann_vec,
      col = list(Community = setNames(
        c("#999999", "#E69F00", "#56B4E9"),
        c("C436", names(communities))
      ))
    )
  }
  
  Heatmap(
    matrix = logFC_mat,
    name   = "logFC",
    col    = colorRamp2(c(-5, 0, 5), c("#2166AC", "#F7F7F7", "#B2182B")),
    
    top_annotation = col_ann,
    
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (!is.na(FDR_mat[i, j]) && FDR_mat[i, j] < 0.05) {
        grid.text("*", x, y, gp = gpar(fontsize = 15))
      }
    },
    
    column_split   = ann_vec,     
    column_names_gp  = gpar(fontsize = 10),
    border           = TRUE,
    show_row_names   = TRUE,
    show_column_names = TRUE,
    cluster_rows     = TRUE,
    cluster_columns  = FALSE
  )
}
```


```{r}
pdf("output/plots4paperv2/fig6/Fig6c.pdf", width = 18, height = 3)
plot.hmv2(
  toptable = toptables$hs.ACC[grepl(names(toptables$hs.ACC), pattern = "^DLB")] %>% 
    lapply(function(x) {
      x %>% 
        mutate(coef = str_remove(coef, pattern = "DLB_"))
               }),
  genes2plot = communities.genes.hs$lysoDLB$C436, 
  species = "hs", 
  mm2hs = mm2hs,
  communities = list(
    C437 = communities.genes.hs$lysoDLB$C437,
    C438 = communities.genes.hs$lysoDLB$C438
  )
)
dev.off()
```


```{r}
pdf("output/plots4paperv2/fig6/Fig6d.pdf", width = 18, height = 4)
plot.hmv2(
  toptable = toptables$npc1,
  genes2plot = communities.mm$DLB$C436, 
  species = "hs", 
  mm2hs = mm2hs,
  communities = list(
    C437 = communities.mm$DLB$C437,
    C438 = communities.mm$DLB$C438
  )
)
dev.off()
```

## Figure 7 : LysoPD 
### B - cnet plots
```{r}
kegg2plot.7b= c("Lysosome biogenesis", 
  "Sphingolipid metabolism", 
  "Fatty acid elongation", 
  "Other glycan degradation", 
  "Fatty acid metabolism", 
  "Phagosome", 
  "Glycerophospholipid metabolism", 
  "Autophagy - animal", 
  "Endocrine and other factor-regulated calcium reabsorption", 
  "Synaptic vesicle cycle", 
  "SNARE interactions in vesicular transport"
)

ggarrange(
  ORA.communities$PD$hs$C289 %>% 
    setReadable(OrgDb = org.Hs.eg.db, keyType = "ENTREZID") %>% 
    cnetplot(
      showCategory    = kegg2plot.7b, 
      layout          = "nicely"
    ) +
    labs(
      title = "LysoPD - C289"
    ) +
    theme(
       text = element_text(size = 18),
      plot.title = element_text(hjust = 0.5, size = 20, face = "bold")
    ), 
  ORA.communities$PD$hs$C290 %>% 
    setReadable(OrgDb = org.Hs.eg.db, keyType = "ENTREZID") %>% 
    cnetplot(showCategory = 6, layout = "nicely") + 
    labs(
      title = "LysoPD - C290"
    ) +
    theme(
      text = element_text(size = 18),
      plot.title = element_text(hjust = 0.5, size = 20, face = "bold")
      
    ), 
  common.legend = T, 
  widths = c(1.5,1)
) +
  ggsave(
    "output/plots4paperv2/fig7/Fig7bc.pdf", 
    #"output/plots4paperv2/Fig3.png", 
    width = 20, height = 10, units = "cm", 
    #dpi = 800,
    scale = 2.5
  )
```
### HM - Human DLB
```{r}
pdf("output/plots4paperv2/fig7/Fig7d.pdf", width = 11, height = 4 )
plot.hmv2(
  toptable = toptables$hs.ACC[grepl(names(toptables$hs.ACC), pattern = "^PD")],
  genes2plot = communities.genes.hs$lysoPD$C289, 
  species = "hs", 
  mm2hs = mm2hs,
  communities = list(
    C290 = communities.genes.hs$lysoPD$C290,
    C289 = communities.genes.hs$lysoPD$C289
    
  )
)
dev.off()
```

```{r}
pdf("output/plots4paperv2/fig7/Fig7e.pdf", width = 18, height = 4)
plot.hmv2(
  toptable = toptables$npc1,
  genes2plot = communities.mm$PD$C289, 
  species = "hs", 
  mm2hs = mm2hs,
  communities = list(
    C290 = communities.mm$PD$C290,
    C289 = communities.mm$PD$C289
  )
)
dev.off()
```




# Supp figs
## S3 - Non neuronoptahic IEMs
```{r}
enrichment_dfs.non.neuro.IEMS %>% 
  dplyr::filter(
    z_comb == 3, z_12 == 1.5, 
    ) %>% 
  mutate(
    # use the childhood vector to extract the CD seed gene name
    adult = str_extract(comparison, "(AD|ALS|PD|DLB|FTLD|brain_malform)"), 
    IEM = str_remove(comparison, pattern = "^(AD|ALS|PD|DLB|FTLD|brain_malform)_"), 
    IEM = case_when(
      grepl(IEM, pattern = "aminoacidopathy") ~  "Aminoacidopathies", 
      grepl(IEM, pattern = "FA_ox") ~  "Fatty acid oxidation disorders", 
      grepl(IEM, pattern = "generalIEM") ~  "General inborn errors of metabolism", 
      grepl(IEM, pattern = "glycosy") ~  "Congenital disorders of glycosylation", 
      grepl(IEM, pattern = "mitochondrial") ~  "Mitochondrial disorders", 
      grepl(IEM, pattern = "nonCD_IEM") ~  "All non-neuronopathic IEMs", 
      grepl(IEM, pattern = "nonCD_lyso") ~  "Non-neurononpathic lysosomal disorders", 
      grepl(IEM, pattern = "peroxisomal") ~  "Non-neurononpathic peroxisomal disorders", 
    ) %>% str_wrap(width = 25),
    FDR = p.adjust(empirical_p, "fdr"),
    facet.order = case_when(
      grepl(comparison, pattern = "nonCD_IEM") ~ 1, 
      grepl(comparison, pattern = "lyso") ~ 2, 
      grepl(comparison, pattern = "general") ~ 9, 
      TRUE ~ 8
    )
    
    ) %>%  
  dplyr::filter(type == "observed") %>% 
  
  ggplot(
    aes(x = obs_exp, y = reorder(adult, obs_exp))
  ) +
  geom_col(
    aes(fill = -log10(FDR), alpha = FDR < 0.05), 
    colour = "grey50", position = "dodge", 
    width = 0.75
  ) +
  scale_fill_viridis_c() +
  scale_alpha_manual(values = c(0.1, 1)) +
  labs(
    x = "observed / expected", 
    y = "Disease"
  ) +
  facet_wrap(~reorder(IEM, facet.order), ncol = 3) +
  ggsave(
    "output/plots4paperv2/supp/S3.png", 
    #"output/plots4paperv2/supp/S3.pdf", 
    width = 20, height = 15, units = "cm", 
    #dpi = 800,
    scale = 1
  )
```

## S9 - marker gene expression
```{r}
fig.5c = Seurat::DotPlot(
    data.seurat.hs.DLB.PD,
    features = markergenes2plot$hs,
    cols = c("#440154FF", "#22A884FF"), 
    scale = TRUE, 
    dot.scale = 10
  ) +
  labs(
    y = "Cell type", 
    x = "Genes"
  ) +
  scale_fill_viridis_c() +
  theme(legend.position = "top") +
  labs(
    title = "Marker genes in human ACC"
  )
  
fig.5d = Seurat::DotPlot(
  data.seurat.mm.Npc1,
  features = markergenes2plot$mm,
  cols = c("#440154FF", "#22A884FF"), 
  scale = TRUE, 
  dot.scale = 10
) +
  labs(
    y = "Cell type", 
    x = "Genes"
  ) +
  scale_fill_viridis_c() +
  theme(legend.position = "none") +
  labs(
    title = "Marker genes in mouse cerebellum"
  )

  fig.5c /
  fig.5d +
  plot_layout(
    heights = c(1,1), 
    ) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(size = 24, face = "bold")) +
   ggsave(
    "output/plots4paperv2/supp/S9.pdf", 
    width = 24, height = 20, units = "cm", 
    #dpi = 800,
    scale = 1.6
  )

```
## S10 - HGSNAT with Tau
```{r}
cpm(y.hs.DLB.PD, log = T) %>% 
  .["HGSNAT",] %>% 
  as.data.frame() %>% 
  rownames_to_column("sample") %>% 
  as_tibble() %>% 
  set_colnames(c("sample_id", "logCPM")) %>% 
  left_join(y.hs.DLB.PD$samples) %>% 
  mutate(TAU = as.numeric(TAU)) %>% 
  ggboxplot(
    x = "TAU",
    y = "logCPM",
    # add = "jitter",
    outlier.shape = NA,
  ) + 
  geom_jitter(
    aes(colour = disease, shape = sex), 
    size = 3
  ) +
  stat_compare_means(
    method = "kruskal.test",
    label = "p.format",
    hide.ns = TRUE,
    label.y.npc = 0.95          # overall p at top
  ) +
  stat_compare_means(
    ref.group = "0",
    vjust = 0.7, 
    size = 10,
    label = "p.signif",     # stars on bars vs stage 0
    hide.ns = TRUE,            # hide ns to keep it cleany
  )+
  labs(
    title = "HGSNAT", 
    x = "Braak Tau stage"
  ) +
  facet_wrap(~cluster, ncol = 2) +
  scale_color_manual(values = c(
    Control = "#999999",
    DLB = "#1B9E77",
    PD = "#E69F00",
    PDD = "#A67000")) +
  theme(
    text = element_text(size = 16)
  ) 
  ggsave(
    "output/plots4paperv2/supp/S10.png", 
    width = 10, height = 10, units = "cm", dpi = 400, 
    scale = 2
  )
```




## S11 - Number of cells
```{r}
data.seurat.hs.DLB.PD@meta.data %>% 
  group_by(celltype, disease) %>% 
  summarise(ncells = n()) %>% 
  ggplot(
    aes(
      x = celltype, y = ncells, fill = disease
    ))+
  geom_col(
    colour = "black", 
    position = "dodge"
  ) +
  coord_flip() +
  scale_y_log10() +
  annotation_logticks(sides = "b") +
  theme(
    panel.grid = element_blank()
  ) +
  scale_fill_manual(values = c(
    control = "#999999",
    DLB = "#1B9E77",
    PD = "#E69F00",
    PDD = "#A67000")) +
  labs(
    title = "Human ACC"
    ) +
  

data.seurat.mm.Npc1@meta.data %>% 
  group_by(celltype, age, genotype) %>% 
  summarise(ncells = n()) %>% 
  ggplot(
    aes(
      x = celltype, y = ncells, fill = genotype
    )
  ) +
  geom_col(
    colour = "black", 
    position = "dodge"
  ) +
  coord_flip() +
  scale_y_log10() +
  annotation_logticks(sides = "b") +
  facet_wrap(~age) +
  scale_fill_manual(
    values = c("#999999", "#742880")
  ) +
  theme(
    panel.grid = element_blank()
  ) +
  labs(
    title = "NPC1 mouse cerebellum"
    ) +
  plot_layout(ncol = 1) +
  plot_annotation(tag_levels = c("A", "B")) +
  ggsave(
    "output/plots4paperv2/supp/S11.png", 
    #"output/plots4paperv2/supp/S3.pdf", 
    width = 20, height = 15, units = "cm", 
    #dpi = 800,
    scale = 1.5
  )
```


## S** DLB & PD KEGG /MPO
```{r}
run.ORA.MPO(node_metadata$CD_1A_lyso_PD$X) %>% 
  as_tibble() %>% 
  mutate(network = "LysoPD") %>% 
  bind_rows(
    run.ORA.MPO(node_metadata$CD_1A_lyso_DLB$X) %>% 
      as_tibble %>% 
      mutate(
        network = "LysoDLB"
      )
  ) %>% 
  mutate(
    source = "MPO"
  ) %>% 
  bind_rows(
    runKEGG(node_metadata$CD_1A_lyso_PD$X) %>% 
      as_tibble() %>% 
      mutate(network = "LysoPD") %>% 
      bind_rows(
        runKEGG(node_metadata$CD_1A_lyso_DLB$X) %>% 
          as_tibble %>% 
          mutate(
            network = "LysoDLB"
          )
      ) %>% 
      mutate(source = "KEGG")
  ) %>% 
  group_by(Description, source) %>% 
  mutate(
    category = case_when(
      n_distinct(network) == 2 ~ 1,
      network == "LysoPD" ~ 2,
      network == "LysoDLB" ~ 3
    ), 
    source = factor(source, levels = c("MPO", "KEGG"))
  ) %>% 
  ungroup() %>% 
  ggplot(
    aes(
      x = network,
      y = reorder(Description, -category)
    )
  ) +
  geom_tile(
    aes(fill = FoldEnrichment)
  ) +
    geom_label(
          aes(
      label = p.adjust %>% signif(2)
      
      )
    ) +
  scale_fill_viridis_c() +
  facet_wrap(~source, 
             nrow = 1, ncol = 2, scales = "free_y") +
  labs(
    x = "Subnetwork",
    y = NULL
  ) +
   theme(
    strip.text = element_text(size = 12),
    text = element_text(size = 14),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom",
    #axis.title.x = element_blank(),
    plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) +
    ggsave(
    "output/plots4paperv2/supp/STBC.png", 
    #"output/plots4paperv2/supp/S3.pdf", 
    width = 20, height = 15, units = "cm", 
    #dpi = 800,
    scale = 1.5
  ) 
```



```{r}

  
```






## export for pub
```{r}
node_metadata[keep.subnets] %>% 
  lapply(function(x) {
  x %>% 
      dplyr::rename(gene_name = X) %>% 
      mutate(d1_name = str_remove(d1_name, pattern = "CD_.{1,2}_")) %>% 
      dplyr::select(-comparison) %>% 
      arrange(z_both)
  }) %>% 
  set_names(names(.) %>% str_remove(pattern = "CD_.{1,2}_")) %>% 
      writexl::write_xlsx(
        "output/plots4paperv2/supp/STBC.xlsx"
  )
  
```

