---
title: "Pan neurodegen disease analysis"
author: "Karissa Barthelson"
date: "2026-02-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


## Introduction
```{r}
knitr::opts_chunk$set(
  message = F, echo = T, warning = F
)
```

```{r}
library(tidyverse)
library(magrittr)
library(readxl)

# vis
library(ComplexUpset)
library(UpSetR)
library(pander)
library(ggeasy)
library(ggfortify)
library(ggpubr)
library(ggrepel)
library(plotly)
library(tidyHeatmap)
library(grid)
library(gridExtra)
library(gridGraphics)
library(viridis)
library(tidytext)
library(pheatmap)
library(patchwork)

# annotations
library(org.Hs.eg.db)
library(ndexr)

# enrichemnt
library(clusterProfiler)
library(gprofiler2)
library(ReactomePA)

theme_set(theme_light() +
            theme(plot.title = element_text(hjust = 0.5), 
                  strip.text = element_text(face = "bold"),
                  strip.background = element_rect(fill = "grey20")
            )
)

knitr::opts_chunk$set(
  message = F, echo = F, warning = F
)
```

# import 
```{r}
## the nodes in PCNet
# the background gene set will be all genes found in the PCNet, as these are the ones I am analysing. The node info was obtained from NDex. 
nodes <- readRDS("data/R_objects/network/nodes.rds")


### obtain the genes in the subnetworks
node_metadata <- 
  list.files(
    path = "data/results/pan neurodegen analysis - adult v2", 
    pattern = "overlapping_nodes_metadata", 
    recursive = TRUE, full.names = TRUE
  ) %>% 
  set_names(
    str_replace(
      ., 
      pattern = "/overlapping_nodes_metadata.csv",
      replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/pan.+v2", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))

## gene sets
# new adult-onset genes
genesets <- list.files(
  "data/genesets/v2/adultchild", 
  pattern = ".csv",
  recursive = T,
  full.names = TRUE
) %>% 
  sapply(function(x) {
    x %>% 
      read_delim(delim = ",") %>% 
      pull(1)
  }, simplify = F) %>% 
  setNames(names(.) %>% 
             str_remove(pattern = "data/genesets/.+/") %>% 
             str_remove(pattern = ".csv$") %>% 
             str_remove(pattern = "_v2")
  )

## enrichment netcoloc
enrichment_dfs <- 
  list.files(
    "data/results/pan neurodegen analysis - adult v2", 
    recursive = TRUE, full.names = TRUE) %>%
  grep(pattern = "network_colocalization_enrichment", 
       value = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/network_colocalization_enrichment.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0(
          "data/results/pan neurodegen analysis - adult v2", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y)) %>% 
  bind_rows() %>%
  dplyr::filter(z_comb == 3, z_12 == 1.5) %>%
  pivot_longer(
    cols = c(observed_overlap, expected_overlap_mean),
    names_to = "type",
    values_to = "value"
  ) %>%
  mutate(
    type = recode(type,
                  "observed_overlap" = "observed",
                  "expected_overlap_mean" = "expected")
  )

```

here is the number of genes per gene set.  

```{r}
genesets %>% 
  unlist %>% 
  as.data.frame(col.names = "gene") %>% 
  rownames_to_column("geneset") %>%
  set_colnames(c("geneset", "gene")) %>% 
  mutate(
    geneset = str_remove_all(geneset, pattern = "\\d+$"), # remove the number generated by unlist
    geneset = fct_infreq(geneset), # Reorder by frequency, 
   ) %>% 
  ggplot(
    aes(x = geneset)
  ) +
  geom_bar() +
  easy_rotate_x_labels(angle = -45) +
  labs(
    title = "Number of genes per gene set", 
    y = "# genes"
  ) 
```


## overlap of genes across gene sets

There is minimal overlap of genes across the adult and childhod dementia gene sets. Only 13 genes are shared. 

```{r}
genesets[grepl(names(genesets), pattern = "_")] %>% 
  fromList %>% 
  UpSetR::upset(
    nsets = length(.), 
    order.by = "freq", 
    )
```

# over-representation analysis of gene sets

Here I will perform pathway/enrichment analysis on these gene sets to see what pathways are enriched, and how much overlap there is. 


<!-- ## GO terms -->

<!-- Here, i performed over-represetation analysis (hypergeometric test) of GO terms (all ontologies) of each gene set using clusterprofiler -->

<!-- ```{r} -->
<!-- runGO = function(geneset) { -->
<!--   enrichGO( -->
<!--     gene = geneset,  -->
<!--     universe = nodes,  -->
<!--     OrgDb = org.Hs.eg.db,  -->
<!--     keyType = "SYMBOL",  -->
<!--     ont = "ALL" -->
<!--   ) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r} -->
<!-- GO <- genesets %>%  -->
<!--   sapply(runGO, simplify = F) %>%  -->
<!--   sapply(as_tibble, simplify = F) -->

<!-- GO %>%  -->
<!--   lapply(function(x) { -->
<!--     x$Description -->
<!--   }) %>%  -->
<!--   fromList() %>%  -->
<!--   UpSetR::upset( -->
<!--     nsets = length(.),  -->
<!--     order.by = "freq" -->
<!--   ) -->

<!-- GO.terms.pergeneset <- GO %>%  -->
<!--   lapply(function(x) { -->
<!--     x$Description -->
<!--   })  -->
<!-- ``` -->

## KEGG pathways

```{r}
runKEGG = function(geneset) {

  uv =   bitr(nodes, 
                fromType = "SYMBOL", toType = "ENTREZID", 
                OrgDb ="org.Hs.eg.db") %>% 
    .$ENTREZID
  
  bitr(geneset, 
       fromType = "SYMBOL", toType = "ENTREZID", 
     OrgDb ="org.Hs.eg.db") %>% 
  .$ENTREZID %>%
  enrichKEGG(
    organism = 'hsa',
    universe = uv,
    pvalueCutoff = 0.05) %>% 
  as_tibble 
  
}
```

```{r}
kegg <- genesets %>% 
  sapply(function(x) {runKEGG(x)}, simplify = F) 

kegg %>% 
  lapply(function(x) {x$Description}) %>% 
  fromList() %>% 
  UpSetR::upset(
    nsets = length(.), 
    order.by = "freq"
  )
```



# Is there significant colocalisation?

Netcoloc was performed and below shows the expected vs observed ratio of the genes identified by dual network propogation. 


```{r}
# enrichment_dfs %<>% 
#   mutate(
#     comparison = str_replace(
#       comparison, 
#       pattern = "CD_all_adult_allv2", 
#       replacement = "childhood dementia\n and \n adult-onset neurodegen"
#     ), 
#     comparison = str_replace(
#       comparison, 
#       pattern = "CD_all_brain_malform", 
#       replacement = "childhood dementia\n and \nbrain malformations"
#     ),
#     
#     comparison = str_replace(
#       comparison, 
#       pattern = "CD_all_nonCD_IEMs", 
#       replacement = "childhood dementia\n and \n non-neuronopathic IEMs"
#     ),
#     
#     comparison = str_replace(
#       comparison,
#       pattern = "AD_PD",
#       replacement = "Alzheimer's disease\n and \nParkinson's disease"
#     ) %>% factor(levels = c(
#       "childhood dementia\n and \nbrain malformations",
#       "Alzheimer's disease\n and \nParkinson's disease",
#       "childhood dementia\n and \n adult-onset neurodegen", 
#       "childhood dementia\n and \n non-neuronopathic IEMs"
#     )),
#     
#   )

enrichment_dfs %>% 
  mutate(
    FDR = p.adjust(empirical_p), 
    comparison = case_when(
      grepl(comparison, pattern = "aminoaci") ~ "CD_all\n&\naminoacidopathies", 
      grepl(comparison, pattern = "v2") ~ "CD_all\n&\nadult-onset v2", 
      grepl(comparison, pattern = "brain_malf") ~ "CD_all\n&\nbrain malformations", 
      grepl(comparison, pattern = "FA") ~ "CD_all\n&\nFA oxidation disorders", 
      grepl(comparison, pattern = "general") ~ "CD_all\n&\ngeneral IEMs", 
      grepl(comparison, pattern = "glyc") ~ "CD_all\n&\ncongenital disorders\nof glycosylation", 
      grepl(comparison, pattern = "mito") ~ "CD_all\n&\n nonCD mitochondrial disorders", 
      grepl(comparison, pattern = "nonCD_IE") ~ "CD_all\n&\nall nonCD IEMs", 
      grepl(comparison, pattern = "nonCD_lyso") ~ "CD_all\n&\nnonCD Lysosomal disorders", 
      grepl(comparison, pattern = "perox") ~ "CD_all\n&\nnonCD peroxisomal disorders", 
      TRUE ~ comparison
    )
  ) %>% 
  ggplot(
    aes(x = type, y = value, fill = type)
    ) +
  geom_bar(
    stat = "identity", position = "dodge", width = 0.8
  ) +
  geom_errorbar(
    data = . %>% 
      dplyr::filter(type == "expected"),
    aes(
      ymin = value - 2 * expected_overlap_std,
      ymax = value + 2 * expected_overlap_std),
    width = 0.25
  ) + 
  geom_label(
    data =  . %>%
      dplyr::filter(type == "expected"),
    aes(
      x = 1.5, y = 1000,
        label = sprintf("FDR = %.1e", empirical_p)),
    inherit.aes = FALSE,
    fill = NA,
    border.colour = NA,
    size = 4,
    vjust = 0,
    fontface = "bold",
    show.legend = FALSE
  )  +
  labs(
    y = "Number of genes", 
    x = ""
  ) +
  facet_wrap(~comparison, 
             nrow = 1) +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("grey50", "skyblue"))+
   theme(
    
    legend.position = "none",
    axis.title.x = element_blank(),
    plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) 
```

## plots4pub
### Fig.1 pan neurodegen
```{r}


enrichment_dfs %>% 
  dplyr::filter(
    comparison %in% c("AD_PD", "CD_all_adult_allv2", "CD_all_brain_malform")
  ) %>% 
  mutate(
    FDR = p.adjust(empirical_p), 
    comparison = case_when(
      grepl(comparison, pattern = "v2") ~ "CD \n&\nadult-onset NDD", 
      grepl(comparison, pattern = "AD") ~ "AD\n&\nPD", 
      grepl(comparison, pattern = "brain_malf") ~ "CD\n&\nbrain malformations",
      TRUE ~ comparison
    )
  ) %>% 
  ggplot(
    aes(x = type, y = value, fill = type)
    ) +
  geom_bar(
    stat = "identity", position = "dodge", width = 0.8
  ) +
  geom_errorbar(
    data = . %>% 
      dplyr::filter(type == "expected"),
    aes(
      ymin = value - 2 * expected_overlap_std,
      ymax = value + 2 * expected_overlap_std),
    width = 0.25
  ) + 
  geom_label(
    data =  . %>%
      dplyr::filter(type == "expected"),
    aes(
      x = 1.5, y = 400,
        label = sprintf("FDR = %.1e", empirical_p)),
    inherit.aes = FALSE,
    fill = NA,
    border.colour = NA,
    size = 4,
    vjust = 0,
    fontface = "bold",
    show.legend = FALSE
  )  +
  labs(
    y = "Number of genes", 
    x = ""
  ) +
  facet_wrap(~comparison, 
             nrow = 1) +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("grey50", "skyblue"))+
   theme(
    
    legend.position = "none",
    axis.title.x = element_blank(),
    plot.background = element_blank(),
    panel.grid = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(), 
   ) 
```




<!-- # Are the subnetworks biologically meaningful? -->

<!-- ## enriched pathways witin the subnetworks -->
<!-- ### using gprofiler -->
<!-- ```{r} -->
<!-- # this was provided by William -->
<!-- # the latest canonical set from MSigDB -->
<!-- upload_GMT_file(gmtfile = "data/genesets/c2.cp.v2025.1.Hs.symbols.gmt") -->

<!-- gprofiler.wholenet.genesets <- node_metadata %>%  -->
<!--   lapply(magrittr::extract2, "X") %>%  -->
<!--   sapply(function(x) { -->
<!--     gost(query = x,  -->
<!--          organism = "gp__XoLn_Cg43_BEg",  -->
<!--          ordered_query = FALSE,  -->
<!--          significant = TRUE, # only return sig results -->
<!--          user_threshold = 0.05, -->
<!--          correction_method = "fdr", -->
<!--          domain_scope = "annotated", # only analyse genes with the annotation.  -->
<!--          evcodes = TRUE, -->
<!--          custom_bg = nodes)  -->
<!--   }, simplify = F) -->
<!-- ``` -->

<!-- ## vis the top 10 genesets enriched in each network -->
<!-- ```{r} -->
<!-- gprofiler.wholenet.genesets$CD_all_adult_allv2$result %>%  -->
<!--   as_tibble %>%  -->
<!--   dplyr::select(term_id, source, FDR = p_value, everything()) %>%  -->
<!--   mutate( -->
<!--     source = "",  -->
<!--     source = str_extract(term_id, pattern = "KEGG|REACTOME|WP|BIOCARTA"),  -->
<!--     term_id = str_remove( -->
<!--       term_id,  -->
<!--       pattern = "KEGG_|REACTOME_|WP_|BIOCARTA_"),  -->
<!--     term_id = str_remove( -->
<!--       term_id, pattern = "MEDICUS_REFERENCE_|MEDICUS_VARIANT_") -->
<!--   ) %>%  -->
<!--   ggplot( -->
<!--     aes( -->
<!--     x = -log10(FDR),  -->
<!--     y = reorder(term_id, -FDR),  -->
<!--     fill = intersection_size/term_size -->
<!--     ) -->
<!--   ) + -->
<!--   geom_col() + -->
<!--   facet_wrap(~source,  -->
<!--              scales = "free_y", space = "free_y",strip.position = "right" -->
<!--              ) + -->
<!--   scale_fill_viridis_c() -->


<!-- gprofiler.wholenet.genesets$AD_PD$result %>%  -->
<!--   as_tibble %>%  -->
<!--   dplyr::filter( -->
<!--     grepl(term_id, pattern = "KEGG") -->
<!--   ) %>%  -->
<!--   dplyr::select(term_id, source, FDR = p_value, everything()) %>%  -->
<!--   mutate( -->
<!--     source = "",  -->
<!--     source = str_extract(term_id, pattern = "KEGG|REACTOME|WP|BIOCARTA|PID|SIG"),  -->
<!--     term_id = str_remove( -->
<!--       term_id,  -->
<!--       pattern = "KEGG_|REACTOME_|WP_|BIOCARTA_"),  -->
<!--     term_id = str_remove( -->
<!--       term_id, pattern = "MEDICUS_REFERENCE_|MEDICUS_VARIANT_") -->
<!--   ) %>%  -->
<!--   ggplot( -->
<!--     aes( -->
<!--     x = -log10(FDR),  -->
<!--     y = reorder(term_id, -FDR),  -->
<!--     fill = intersection_size/term_size -->
<!--     ) -->
<!--   ) + -->
<!--   geom_col() + -->
<!--   facet_wrap(~source,  -->
<!--              scales = "free_y",  -->
<!--              space = "free_y",strip.position = "right" -->
<!--              ) + -->
<!--   scale_fill_viridis_c() -->

<!-- ``` -->


<!-- ## check the genes in the subnetwork are enriched with mouse brain phenotypes -->

<!-- I tested the enrichment of genes associated with mouse brain/neural phenotypes in the NetColoc subnetwork as a *whole*. I also included here a negative control phenotype -  abnormal digit morphology. This is not expected to be associated with the subnetworks.  -->

<!-- ```{r} -->
<!-- mouse.variant.wholesubnetworks <-  -->
<!--   list.files( -->
<!--     "data/results/results_3", recursive = TRUE, full.names = TRUE) %>%  -->
<!--   grep(pattern = "/mousevariant_KO_enrichment.csv",  -->
<!--        value = TRUE) %>%  -->
<!--   set_names( -->
<!--     str_replace(.,  -->
<!--                pattern = "/mousevariant_KO_enrichment.csv", -->
<!--                replacement = "" -->
<!--   ) %>%  -->
<!--     str_remove( -->
<!--       pattern = paste0("data/results/adultchildsubtypes", "/") -->
<!--     ) -->
<!--   ) %>%    -->
<!--   map(read.csv) %>% -->
<!--   imap(~ mutate(.x, comparison = .y)) %>%  -->
<!--   bind_rows() %>%  -->
<!--     mutate( -->
<!--     comparison = str_remove(comparison,  -->
<!--                             pattern = "data/results/results_3/") -->
<!--   ) %>%  -->
<!--   dplyr::filter(comparison %in% sig.networks) %>% -->
<!--   as_tibble() %>% -->
<!--   dplyr::select(comparison, everything()) %>%  -->
<!--   group_by(comparison) %>%  -->
<!--   mutate( -->
<!--         # use the childhood vector to extract the CD seed gene name -->
<!--     CD = purrr::map_chr(comparison, ~ { -->
<!--       match <- childhood[startsWith(.x, childhood)] -->
<!--       if (length(match) > 0) match[1] else NA_character_ -->
<!--     }), -->

<!--     adult = str_remove(comparison, pattern = "CD_.+_.+_"), -->
<!--     adult = case_when( -->
<!--       adult == "malform" ~ "brain_malform",  -->
<!--       TRUE ~ adult -->
<!--     ), -->
<!--     FDR = p.adjust(OR_p), -->
<!--     .after = OR_p -->
<!--   )  -->
<!-- # a function for plotting by CD subgtype: -->
<!-- plot.mouse.variant.subtype.forestplot.by.CD.subtype <- function(data, CD_subtype) { -->
<!--   data %>% -->
<!--     dplyr::filter(CD == CD_subtype) %>%  -->
<!--     group_by(comparison, MPO_term_type) %>% -->
<!--     slice_min(order_by = OR_p, n = 10) %>% -->
<!--     ungroup() %>% -->
<!--     mutate( -->
<!--       MP_description = reorder_within(MP_description, log_OR, comparison) -->
<!--     ) %>% -->
<!--     ggplot( -->
<!--       aes(x = log_OR, y = MP_description) -->
<!--     ) + -->
<!--     geom_errorbarh( -->
<!--       aes(xmin = log_OR_CI_lower, xmax = log_OR_CI_upper), -->
<!--       height = 0.2, color = "black" -->
<!--     ) + -->
<!--     geom_point( -->
<!--       aes(size = num_genes_in_term, color = -log10(FDR)), -->
<!--       alpha = 0.6 -->
<!--     ) + -->
<!--     geom_vline(xintercept = 0, lty = 2, color = "grey50") + -->
<!--     geom_label( -->
<!--       aes(label = signif(OR_p, 2),  -->
<!--           fill = -log10(FDR)), -->
<!--       nudge_x = 1 -->
<!--     ) + -->
<!--     scale_y_reordered() +  # This removes the `___comparison` suffix added by reorder_within() -->
<!--     scale_fill_viridis_c() + -->
<!--     labs( -->
<!--       x = "log(OR) ± 95% CI", -->
<!--       y = NULL, -->
<!--       size = "# Genes in MPO term", -->
<!--       color = "Term Type" -->
<!--     ) + -->
<!--   facet_wrap(~comparison, nrow = 1, scales = "free_y") -->
<!-- } -->

<!-- mouse.variants.plots <- childhood %>%  -->
<!--   lapply( -->
<!--     function(x) { -->
<!--       plot.mouse.variant.subtype.forestplot.by.CD.subtype( -->
<!--         CD_subtype = x, data = mouse.variant.wholesubnetworks) -->
<!--     }) -->

<!-- mouse.variants.plots[1] -->
<!-- mouse.variants.plots[2] -->
<!-- mouse.variants.plots[3] -->
<!-- mouse.variants.plots[4] -->
<!-- mouse.variants.plots[5] -->
<!-- mouse.variants.plots[6] -->
<!-- mouse.variants.plots[7] -->
<!-- mouse.variants.plots[8] -->
<!-- ``` -->

<!-- ## are the genes in the subnetwork human brain expressed? -->

<!-- I next want to see whether the genes in the subnetwork are expressed in the brain. To do this, I will assess it in the GTEx dataset. I will use the GTEx v10 data, which is the most recent version.  -->

<!-- ```{r} -->
<!-- # plot number of genes per subnetworks -->
<!-- node_metadata %>%  -->
<!--   bind_rows() %>%  -->
<!--   dplyr::filter(comparison %in% sig.networks) %>%  -->
<!--   dplyr::rename(gene = X) %>%  -->
<!--   as_tibble() %>% -->
<!--   mutate( -->
<!--     seed_type = case_when( -->
<!--       d1_seeds == 1 & d2_seeds == 1 ~ "both", -->
<!--       d1_seeds == 1 & d2_seeds == 0 ~ "D1 - childhood", -->
<!--       d1_seeds == 0 & d2_seeds == 1 ~ "D2 - adult", -->
<!--       d1_seeds == 0 & d2_seeds == 0 ~ "network proximal" -->
<!--     ) -->
<!--   ) %>%  -->
<!--   ggplot( -->
<!--     aes(y = comparison, fill = seed_type) -->
<!--   ) + -->
<!--   geom_bar() + -->
<!--   scale_fill_manual(values = c( -->
<!--     "#7570b3", "#1b9e77", "#d95f02", "grey70" -->
<!--   )) + -->
<!--   labs( -->
<!--     x = "Number of genes in subnetwork",  -->
<!--     y = "Seed gene sets",  -->
<!--     title = "Number of genes in the subnetworks" -->
<!--   )  -->
<!-- ``` -->

<!-- ## assessing which genes are driving the DLB/Lyso -->
<!-- about half of the DLB_lyso gene set is unique to that comparsion -->
<!-- ```{r} -->
<!-- list( -->
<!--   DLB_lyso_genes = node_metadata$CD_1A_lyso_DLB$X,  -->
<!--   AD_lyso_genes = node_metadata$CD_1A_lyso_AD$X, -->
<!--   PD_lyso_genes = node_metadata$CD_1A_lyso_PD$X, -->
<!--   ALS_lyso_genes = node_metadata$CD_1A_lyso_ALS$X, -->
<!--   FTLD_lyso_genes = node_metadata$CD_1A_lyso_FTLD$X) %>%  -->
<!--   fromList() %>%  -->
<!--   upset() -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # what genes are the unique ones -->
<!-- unique_genes <- node_metadata$CD_1A_lyso_DLB %>%  -->
<!--   dplyr::filter(!(X %in% node_metadata$CD_1A_lyso_AD$X)) %>%  -->
<!--   dplyr::filter(!(X %in% node_metadata$CD_1A_lyso_PD$X)) %>%  -->
<!--   dplyr::filter(!(X %in% node_metadata$CD_1A_lyso_ALS$X)) %>%  -->
<!--   dplyr::filter(!(X %in% node_metadata$CD_1A_lyso_FTLD$X)) %>%  -->
<!--   .$X -->

<!-- node_metadata$CD_1A_lyso_DLB %>%  -->
<!--   dplyr::filter(X %in% unique_genes) %>%  -->
<!--   .$X %>%  -->
<!--   enrichGO(OrgDb = org.Hs.eg.db, keyType = "SYMBOL") %>%  -->
<!--   as_tibble  -->
<!-- ``` -->




<!-- <!-- ### obtain GTEx expression data --> -->
<!-- <!-- This file was downloaded from here : --> -->
<!-- <!-- https://www.gtexportal.org/home/downloads/adult-gtex/bulk_tissue_expression --> -->

<!-- <!-- The file contains the median gene-level TPM by tissue.  --> -->

<!-- <!-- ```{r} --> -->
<!-- <!-- # this was downloaded from gtex websiote --> -->
<!-- <!-- # version 10  --> -->
<!-- <!-- gtex <- read.delim("data/gtex/GTEx_Analysis_v10_RNASeQCv2.4.2_gene_median_tpm.gct.gz", skip =2) --> -->

<!-- <!-- gtex %<>%  --> -->
<!-- <!--   pivot_longer(names_to = "tissue",  --> -->
<!-- <!--                cols = 3:56,  --> -->
<!-- <!--                values_to = "tpm" --> -->
<!-- <!--   ) %>%  --> -->
<!-- <!--   dplyr::select( --> -->
<!-- <!--     gene = Description, tissue, tpm --> -->
<!-- <!--   ) %>%  --> -->
<!-- <!--   mutate( --> -->
<!-- <!--     logtpm = log(tpm)) --> -->
<!-- <!-- ``` --> -->

<!-- <!-- ```{r} --> -->
<!-- <!-- # define a threshold for expressed or not   --> -->
<!-- <!-- gtex.threshold = 0.5 --> -->

<!-- <!-- # Fix: safe join and summarize per gene per tissue --> -->
<!-- <!-- node_metadata %>% --> -->
<!-- <!--   bind_rows(.id = "comparison") %>%  # Add comparison info --> -->
<!-- <!--   dplyr::rename(gene = X) %>% --> -->
<!-- <!--   as_tibble() %>% --> -->
<!-- <!--   dplyr::select(comparison, gene) %>%  --> -->
<!-- <!--   left_join( --> -->
<!-- <!--     gtex %>%  --> -->
<!-- <!--       dplyr::filter(grepl("Brain", tissue)) %>%  --> -->
<!-- <!--       dplyr::distinct(gene, tissue, logtpm) --> -->
<!-- <!--   ) %>%  --> -->
<!-- <!--   mutate( --> -->
<!-- <!--     expressed = ifelse(logtpm > gtex.threshold, 1, 0) --> -->
<!-- <!--     ) %>%  --> -->
<!-- <!--   group_by(tissue, comparison) %>% --> -->
<!-- <!--   dplyr::filter( --> -->
<!-- <!--     grepl("Brain", tissue),  --> -->
<!-- <!--     comparison %in% sig.networks --> -->
<!-- <!--     ) %>%  --> -->
<!-- <!--   mutate( --> -->
<!-- <!--     n_genes = n(), --> -->
<!-- <!--     n_expressed = sum(expressed), --> -->
<!-- <!--     proportion_expressed = n_expressed / n_genes,  --> -->
<!-- <!--            # use the childhood vector to extract the CD seed gene name --> -->
<!-- <!--     CD = purrr::map_chr(comparison, ~ { --> -->
<!-- <!--       match <- childhood[startsWith(.x, childhood)] --> -->
<!-- <!--       if (length(match) > 0) match[1] else NA_character_ --> -->
<!-- <!--     }),  --> -->

<!-- <!--     adult = str_remove(comparison, pattern = "CD_.+_.+_"), --> -->
<!-- <!--   ) %>%  --> -->
<!-- <!--   ungroup %>%  --> -->
<!-- <!--   group_by(CD) %>%  --> -->
<!-- <!--    tidyHeatmap::heatmap( --> -->
<!-- <!--     .row = tissue, --> -->
<!-- <!--     .column = comparison, --> -->
<!-- <!--     .value = proportion_expressed,  --> -->

<!-- <!--     # turn off clustering --> -->
<!-- <!--     cluster_rows = FALSE,  --> -->
<!-- <!--     column_dend_height = unit(2, "cm"), --> -->

<!-- <!--    ) %>%  --> -->
<!-- <!--   annotation_tile(CD) --> -->
<!-- <!-- ``` --> -->

<!-- <!-- ## what cell types are they primarily expressed in human brains? --> -->

<!-- <!-- I have requested access for a human snRNAseq dataset from ROSMAP [Mathys et al.](https://www.nature.com/articles/s41586-024-07606-7) synapse ID 	 --> -->
<!-- <!-- [syn52293442](https://www.synapse.org/Synapse:syn52293442) --> -->

<!-- <!-- Also getting a dementia with lewy body dataset from GEO:  --> -->
<!-- <!-- https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE178146 --> -->
<!-- <!-- https://pmc.ncbi.nlm.nih.gov/articles/PMC8357687/ --> -->

<!-- <!-- A LSD dataset: NPC1 single cell mouse brain: --> -->
<!-- <!-- sra: PRJNA606815 --> -->
<!-- <!-- https://pmc.ncbi.nlm.nih.gov/articles/PMC7432835/ --> -->
<!-- <!-- only n = 2 tho --> -->



<!-- # what pathways are in the subnetwork -->

<!-- To interrogate the biological context of the sub networks, I will perform some enrichment analyses. I will tackle this using two approaches:  -->

<!-- 1- Perform hypergeometric tests using KEGG pathways on the entire subnetwork (background is the list of genes in PCNet) -->
<!-- 2- use the annotations of the clusters within the subnetworks (performed using gprofiler) of KEGG, Reactome and GO terms.  -->

<!-- ## 1- whole subnetowrks -->
<!-- ```{r} -->
<!-- enrichment.wholesubnet.kegg <-  -->
<!--   node_metadata[sig.networks] %>%  -->
<!--   lapply(magrittr::extract2, "X") %>%  -->
<!--   sapply(runKEGG, simplify = F) -->

<!-- enrichment.wholesubnet.kegg %>%  -->
<!--   lapply(function(x) { -->
<!--     x$Description -->
<!--   }) %>%  -->
<!--   fromList() %>%  -->
<!--   UpSetR::upset( -->
<!--     nsets = length(.),  -->
<!--     order.by = "freq" -->
<!--   ) -->
<!-- ``` -->







<!-- ## 2- cluster annotations -->
<!-- I will here show the top pathways identified in the clusters within the subnetworks.  -->

<!-- ```{r} -->
<!-- cluster.gprofiler.results <-  -->
<!--   list.files(path = "data/results/adultchildsubtypes",  -->
<!--              pattern = "gprofiler", recursive = TRUE, full.names = TRUE) %>%  -->
<!--   set_names( -->
<!--     str_replace(.,  -->
<!--                 pattern = "/gprofiler_analysis_results.csv", -->
<!--                 replacement = "" -->
<!--     ) %>%  -->
<!--       str_remove( -->
<!--         pattern = paste0("data/results/adultchildsubtypes", "/") -->
<!--       ) -->
<!--   ) %>%    -->
<!--   map(read.csv) %>% -->
<!--   imap(~ mutate(.x, comparison = .y)) -->

<!-- # only keep the signigicant subnets -->
<!-- cluster.gprofiler.results <- cluster.gprofiler.results[sig.networks] -->
<!-- ``` -->




<!-- ```{r} -->
<!-- cluster.gprofiler.results %>%  -->
<!--   lapply(function(x) { -->
<!--     x %>%  -->
<!--       mutate( -->
<!--         FDR = p.adjust(p_value, method = "fdr"),  -->
<!--         CD = purrr::map_chr(comparison, ~ { -->
<!--           match <- childhood[startsWith(.x, childhood)] -->
<!--           if (length(match) > 0) match[1] else NA_character_ -->
<!--         })  -->
<!--       ) %>%  -->
<!--       dplyr::filter(p_value != "NA") %>%  -->
<!--       dplyr::distinct(Top.pathway, .keep_all = TRUE) %>%  -->
<!--       head(10) %>%  -->
<!--       mutate( -->
<!--         Top.pathway = str_wrap(Top.pathway, width = 30) -->
<!--       ) -->
<!--   }) %>%  -->
<!--   bind_rows() %>%  -->
<!--   dplyr::filter( -->
<!--     comparison %in% c("CD_1A_lyso_PD", "CD_1A_lyso_DLB") -->
<!--   ) %>%  -->
<!--   ggplot( -->
<!--     aes( -->
<!--       x = -log10(FDR), -->
<!--       y = reorder(Top.pathway, -p_value), -->
<!--       fill = ) -->
<!--   ) + -->
<!--   geom_col() + -->
<!--   facet_wrap(~comparison, scales = "free_y") -->


<!-- plot_top_pathways_for_CD <- function(cluster_results, CD_subtype, top_n = 10) { -->
<!--   cluster_results %>% -->
<!--     lapply(function(x) { -->
<!--       x %>% -->
<!--         mutate( -->
<!--           FDR = p.adjust(p_value, method = "fdr"), -->
<!--           CD = purrr::map_chr(comparison, ~ { -->
<!--             match <- childhood[startsWith(.x, childhood)] -->
<!--             if (length(match) > 0) match[1] else NA_character_ -->
<!--           }) -->
<!--         ) %>% -->
<!--         dplyr::filter(!is.na(p_value), CD == CD_subtype) %>% -->
<!--         group_by(comparison) %>% -->
<!--         slice_min(order_by = p_value, n = top_n) %>% -->
<!--         ungroup() %>% -->
<!--         distinct(Top.pathway, comparison, .keep_all = TRUE) %>% -->
<!--         mutate(Top.pathway = str_wrap(Top.pathway, width = 30)) -->
<!--     }) %>% -->
<!--     bind_rows() %>% -->
<!--       mutate( -->
<!--         Top.pathway = reorder_within(Top.pathway, -FDR, comparison) -->
<!--         ) %>% -->
<!--     ggplot(aes( -->
<!--       x = -log10(FDR), -->
<!--       y = reorder(Top.pathway, -FDR) -->
<!--     )) + -->
<!--     geom_col() + -->
<!--      scale_y_reordered() + -->
<!--     facet_wrap(~comparison, scales = "free_y") + -->
<!--     labs( -->
<!--       x = "-log10(FDR)", -->
<!--       y = NULL, -->
<!--       #title = paste0("Top ", top_n, " Pathways for ", CD_subtype) -->
<!--     ) + -->
<!--     theme(strip.text = element_text(face = "bold")) -->
<!-- } -->

<!-- plots_list <- lapply(childhood, function(cd) { -->
<!--   plot_top_pathways_for_CD(cluster.gprofiler.results, CD_subtype = cd) -->
<!-- }) -->


<!-- for (p in plots_list) { -->
<!--   print(p) -->
<!-- } -->

<!-- plots_list[1] -->
<!-- plots_list[2] -->
<!-- plots_list[3] -->
<!-- plots_list[4] -->
<!-- plots_list[5] -->
<!-- plots_list[6] -->
<!-- plots_list[7] -->
<!-- plots_list[8] -->
<!-- plots_list[9] -->
<!-- plots_list[10] -->
<!-- plots_list[11] -->
<!-- ``` -->

<!-- ```{r} -->
<!-- plots_list2 <- lapply(childhood, function(cd) { -->
<!--   plot_top_pathways_for_CD(cluster.gprofiler.results[top.sig.networks], CD_subtype = cd) -->
<!-- }) -->

<!-- ggarrange(plotlist = plots_list2) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- cluster.gprofiler.results[top.sig.networks] %>%  -->
<!--   bind_rows() %>%  -->
<!--   group_by(  Top.pathway) %>%  -->
<!--   summarise(n = n()) %>%  -->
<!--   arrange(-n) %>% view -->
<!-- ``` -->



<!-- # what genes are in the subnetworks which are drug targetable? -->

<!-- Here, i obtained the drug targets data from [omicsynth](https://doi.org/10.1016/j.ajhg.2023.12.006) to see what genes in the subnetwork have a known drug target.  -->

<!-- The plots below show the network proximity score (higher means more proximal to network) vs the evidence of drug interaction (higher interaction = higher score) -->

<!-- ```{r} -->
<!-- # the genes in the subnetworks -->
<!-- gene.proximitiy.scores <- node_metadata %>%  -->
<!--   lapply(function(x) { -->
<!--     x %>%  -->
<!--       dplyr::select(gene_name = X, z_both) %>%  -->
<!--       as_tibble     -->
<!--   }) -->


<!-- omicsynth <- read.csv("data/omicsynth/drug_genome_dgidb.csv")  -->

<!-- temp <- gene.proximitiy.scores %>%  -->
<!--   lapply(function(x) { -->
<!--     omicsynth %>% # has the drug info  -->
<!--       # only reatin genes in subnet -->
<!--       dplyr::filter(gene_name %in% x$gene_name) %>%  -->
<!--       as_tibble %>%  -->
<!--       left_join(x) -->
<!--   }) %>%  -->
<!--   bind_rows(.id = "comparison") %>%  -->
<!--   mutate( -->
<!--     CD = purrr::map_chr(comparison, ~ { -->
<!--       match <- childhood[startsWith(.x, childhood)] -->
<!--       if (length(match) > 0) match[1] else NA_character_ -->
<!--     }),  -->
<!--     .after = comparison -->
<!--   ) %>%  -->
<!--   dplyr::filter(comparison %in% sig.networks)  -->

<!-- temp %>%  -->
<!--   dplyr::filter( comparison == "CD_1A_lyso_AD") -->

<!-- plot_drug_vs_proximity_by_CD <- function(gene_proximity_list, omicsynth, CD_subtype, sig.networks, childhood) { -->
<!--   gene_proximity_list %>%  -->
<!--     lapply(function(x) { -->
<!--       omicsynth %>% -->
<!--         dplyr::filter(gene_name %in% x$gene_name) %>% -->
<!--         as_tibble() %>% -->
<!--         left_join(x) -->
<!--     }) %>% -->
<!--     bind_rows(.id = "comparison") %>% -->
<!--     mutate( -->
<!--       CD = purrr::map_chr(comparison, ~ { -->
<!--         match <- childhood[startsWith(.x, childhood)] -->
<!--         if (length(match) > 0) match[1] else NA_character_ -->
<!--       }), -->
<!--       .after = comparison -->
<!--     ) %>% -->
<!--     dplyr::filter(CD == CD_subtype, comparison %in% sig.networks) %>% -->
<!--     ggplot(aes(x = interaction_group_score, y = z_both)) + -->
<!--     geom_point() + -->
<!--     geom_label_repel( -->
<!--       aes(label = drug_claim_name), -->
<!--       data = . %>% dplyr::filter(z_both > 5 & interaction_group_score > 5) -->
<!--     ) + -->
<!--     labs( -->
<!--       x = "Interaction group score",  -->
<!--       y = "Network proximity z score", -->
<!--       title = paste("Drug proximity scores -", CD_subtype) -->
<!--     ) + -->
<!--     facet_wrap(~comparison, scales = "free") -->
<!-- } -->

<!-- for (cd in childhood) { -->
<!--   p <- plot_drug_vs_proximity_by_CD( -->
<!--     gene_proximity_list = gene.proximitiy.scores, -->
<!--     omicsynth = omicsynth, -->
<!--     CD_subtype = cd, -->
<!--     sig.networks = sig.networks, -->
<!--     childhood = childhood -->
<!--   ) -->
<!--   print(p) -->
<!-- } -->
<!-- ``` -->

<!-- # LSDs and adult onset -->
<!-- ## heatmap of z scores -->
<!-- note i converted NA to 0 -->

<!-- ```{r} -->
<!-- library(RColorBrewer) -->
<!-- library(pheatmap) -->
<!-- # heatmap on the genes in the subnetworks -->
<!-- node_metadata[grepl(names(node_metadata), pattern = "CD_1A")] %>%  -->
<!--   bind_rows(.id = "comparison") %>%  -->
<!--   dplyr::filter(comparison != "CD_1A_lyso_brain_malform") %>%  -->
<!--   dplyr::rename(gene = X) %>%  -->
<!--   as_tibble() %>% -->
<!--   dplyr::select(comparison, gene, z_both) %>%  -->
<!--   pivot_wider(values_from = z_both, names_from = comparison) %>%  -->
<!--   column_to_rownames("gene") %>%  -->
<!--   # set NA values to 0 -->
<!--   mutate(across(everything(), ~ replace_na(., 0))) %>%  -->
<!--   as.matrix() %>%  -->
<!--   pheatmap( -->
<!--     color = colorRampPalette(rev(viridis_pal(option = "turbo")(100)))(100), -->
<!--   ) -->
<!-- ``` -->

<!-- ## PCA -->

<!-- The z-scores of the genes are quite different  -->
<!-- ```{r} -->
<!-- node_metadata[grepl(names(node_metadata), pattern = "CD_1A")] %>%  -->
<!--   bind_rows(.id = "comparison") %>%  -->
<!--   dplyr::filter(comparison != "CD_1A_lyso_brain_malform") %>%  -->
<!--   dplyr::rename(gene = X) %>%  -->
<!--   as_tibble() %>% -->
<!--   dplyr::select(comparison, gene, z_both) %>%  -->
<!--   pivot_wider(values_from = z_both, names_from = comparison) %>%  -->
<!--   column_to_rownames("gene") %>%  -->
<!--   # set NA values to 0 -->
<!--   mutate(across(everything(), ~ replace_na(., 0)))  %>%  -->
<!--   t %>%  -->
<!--   prcomp %>%  -->
<!--   autoplot( -->
<!--     data = tibble( -->
<!--       comparison = rownames(.$x) -->
<!--     ), -->
<!--     alpha = 0.5, -->
<!--     size = 6 -->
<!--   ) + -->
<!--   geom_label_repel( -->
<!--     aes(label = comparison) -->
<!--   ) -->
<!-- ``` -->


<!-- DLB has a unique lysosomal disease overlap, potentially suggesting stronger involvement of lysosomal pathways than in AD/FTLD/ALS. -->

<!-- CD_1A_lyso_PD is in the bottom-left, far from DLB and the AD/FTLD/ALS group. Suggests its LSD colocalization profile is distinct from DLB, despite both being synucleinopathies. -->

<!-- What genes are driving this? -->

<!-- ```{r} -->
<!-- node_metadata[grepl(names(node_metadata), pattern = "CD_1A")] %>%  -->
<!--   bind_rows(.id = "comparison") %>%  -->
<!--   dplyr::filter(comparison %in% c("CD_1A_lyso_AD", "CD_1A_lyso_DLB")) %>%  -->
<!--   dplyr::rename(gene = X) %>%  -->
<!--   as_tibble() %>%  -->
<!--   dplyr::select(gene, comparison, d2_name, d1_seeds, d2_seeds, z_both) %>%  -->
<!--   arrange(z_both) %>%  -->
<!--   mutate( -->
<!--     AD_lyso_seed_both = case_when( -->
<!--       comparison == "CD_1A_lyso_AD" & d1_seeds == 1 & d2_seeds == 1 ~ 1, -->
<!--       TRUE ~ 0 -->
<!--     ), -->
<!--     AD_lyso_seed_AD = case_when( -->
<!--       comparison == "CD_1A_lyso_AD" & d1_seeds == 0 & d2_seeds == 1 ~ 1, -->
<!--       TRUE ~ 0), -->

<!--     AD_lyso_seed_lyso = case_when( -->
<!--       comparison == "CD_1A_lyso_AD" & d1_seeds == 1 & d2_seeds == 0 ~ 1, -->
<!--       TRUE ~ 0), -->
<!--     AD_lyso_noseed = case_when( -->
<!--       comparison == "CD_1A_lyso_AD" & d1_seeds == 0 & d2_seeds == 0 ~ 1, -->
<!--       TRUE ~ 0), -->

<!--     DLB_lyso_seed_both = case_when( -->
<!--       comparison == "CD_1A_lyso_DLB" & d1_seeds == 1 & d2_seeds == 1 ~ 1, -->
<!--       TRUE ~ 0 -->
<!--     ), -->
<!--     DLB_lyso_seed_AD = case_when( -->
<!--       comparison == "CD_1A_lyso_DLB" & d1_seeds == 0 & d2_seeds == 1 ~ 1, -->
<!--       TRUE ~ 0), -->

<!--     DLB_lyso_seed_lyso = case_when( -->
<!--       comparison == "CD_1A_lyso_DLB" & d1_seeds == 1 & d2_seeds == 0 ~ 1, -->
<!--       TRUE ~ 0), -->

<!--     DLB_lyso_noseed = case_when( -->
<!--       comparison == "CD_1A_lyso_DLB" & d1_seeds == 0 & d2_seeds == 0 ~ 1, -->
<!--       TRUE ~ 0) -->
<!--   ) %>%  -->
<!--   dplyr::select(shared_name = gene, everything())  -->
<!--   #write.csv("output/adult child subtypes/lyso_AD_DLB_genes.csv") -->
<!-- ``` -->

<!-- ```{r} -->

<!-- temp <- runKEGG(geneset = node_metadata$CD_1A_lyso_DLB$X) -->

<!-- temp %>%  -->
<!--   dplyr::filter(Count > 5) %>% .$Description -->
<!--   ggplot( -->
<!--     aes( -->
<!--       x = -log10(p.adjust),  -->
<!--       y = reorder(Description, -pvalue), -->
<!--       fill = Count) -->
<!--   ) + -->
<!--   geom_col() + -->
<!--   scale_fill_viridis_c() + -->
<!--   labs( -->
<!--     y = "KEGG pathway",  -->
<!--     fill = "Number of genes" -->
<!--   )  -->
<!--   ggsave( -->
<!--     "output/adult child subtypes/figs/talk/overrep_KEGG.svg",  -->
<!--     width = 10, height = 5, -->
<!--     scale = 0.8 -->
<!--     ) -->

<!-- ``` -->

<!-- # New genesets -->

<!-- Netcoloc was performed on new genesets identified by William. below shows the expected vs observed ratio of the genes identified by dual network propogation.  -->

<!-- ## number of genes per geneset -->
<!-- ```{r} -->
<!-- genesets_v2 <-  -->
<!--   list.files("data/genesets/Alternate_gene_lists_for_NetColoc_senstivity_analyses",  -->
<!--              full.names = T, pattern = "_.*txt") %>%  -->
<!--   sapply(function(x) { -->
<!--     x %>%  -->
<!--       read.delim() %>%  -->
<!--       .$Gene -->
<!--   }, simplify = F) %>%  -->
<!--   set_names( -->
<!--     str_replace( -->
<!--       list.files("data/genesets/Alternate_gene_lists_for_NetColoc_senstivity_analyses",  -->
<!--                  full.names = F, pattern = "_.*txt"),  -->
<!--       pattern = ".txt",  -->
<!--       replacement = "" -->
<!--     ) -->
<!--   ) %>%  -->
<!--   set_names( -->
<!--     str_remove(names(.), pattern = "_v2_list") -->
<!--   ) -->

<!-- genesets %>%  -->
<!--   unlist %>%  -->
<!--   as.data.frame(col.names = "gene") %>%  -->
<!--   rownames_to_column("geneset") %>% -->
<!--   set_colnames(c("geneset", "gene")) %>%  -->
<!--   mutate( -->
<!--     geneset = str_remove(geneset, pattern = "\\d{1,3}$"),  -->
<!--     type = case_when( -->
<!--       grepl(geneset, pattern = "^CD") ~ "childhood",  -->
<!--       TRUE ~ "adult" -->
<!--     ) -->
<!--    ) %>%  -->
<!--   dplyr::filter(type == "childhood") %>%  -->
<!--   bind_rows( -->
<!--     genesets_v2 %>%  -->
<!--       unlist %>%  -->
<!--       as.data.frame(col.names = "gene") %>%  -->
<!--       rownames_to_column("geneset") %>% -->
<!--       set_colnames(c("geneset", "gene")) %>%  -->
<!--       mutate( -->
<!--         geneset = str_remove(geneset, pattern = "\\d{1,3}$"),  -->

<!--         type = case_when( -->
<!--           grepl(geneset, pattern = "^CD") ~ "childhood",  -->
<!--           TRUE ~ "adult" -->
<!--     ) -->
<!--    )  -->
<!--   ) %>%  -->
<!--   mutate( -->
<!--       geneset = case_when( -->
<!--       grepl(geneset, pattern = "1A") ~ "Lysosomal disorders",  -->
<!--       grepl(geneset, pattern = "1B") ~ "Other disorders of lipid metabolism & transport",  -->
<!--       grepl(geneset, pattern = "1C") ~ "Disorders of amino acid & other organic acid metabolism",  -->
<!--       grepl(geneset, pattern = "1D") ~ "Vitamin-responsive inborn errors of metabolism",  -->
<!--       grepl(geneset, pattern = "1E") ~ "Disorders of mineral absorption & transport",  -->
<!--       grepl(geneset, pattern = "1F") ~ "Peroxisomal diseases",  -->
<!--       grepl(geneset, pattern = "1G") ~ "Mitochondrial disorders",  -->
<!--       grepl(geneset, pattern = "1H") ~ "Other inborn errors of metabolism",  -->
<!--       grepl(geneset, pattern = "2_") ~ "Leukodystrophies not notherwise categorised",  -->
<!--       grepl(geneset, pattern = "3_") ~ "Neurodegeneration with brain iron accumulation",  -->
<!--       grepl(geneset, pattern = "4_") ~ "Diseases not otherwise categorised", -->
<!--       TRUE ~ geneset -->

<!--     ) %>%  -->
<!--       str_wrap(width = 30), -->
<!--       geneset = fct_infreq(geneset),  # Reorder by frequency,  -->
<!--   ) %>%  -->
<!--   ggplot( -->
<!--     aes(x = geneset, fill = type) -->
<!--     )+ -->
<!--   geom_bar(size = 1) + -->
<!--   #easy_rotate_x_labels(angle =-90) + -->
<!--   labs( -->
<!--     y = "# genes" -->
<!--   ) + -->
<!--   facet_wrap( -->
<!--     ~type,  -->
<!--     ncol = 1,  -->
<!--     scales = "free_y", -->
<!--     space = "free_y",  -->
<!--     strip.position = "right") + -->
<!--   coord_flip() + -->
<!--   scale_fill_manual( -->
<!--     values = c("#FF6666","#3399FF"))  -->
<!--   # theme( -->
<!--   #   plot.margin = margin(r = 2, unit = "cm"),  -->
<!--   #   legend.position = "none", -->
<!--   #   panel.grid = element_blank(), -->
<!--   #  -->
<!--   #   text = element_text(colour = "white"), -->
<!--   #   axis.title = element_text(colour = "white"), -->
<!--   #   axis.text = element_text(colour = "white"), -->
<!--   #    -->
<!--   #   #strip.background = element_blank(),  -->
<!--   #   plot.background = element_rect(fill = "#25283D"), -->
<!--   #   plot.background = element_blank(), -->
<!--   #   panel.grid.minor = element_blank(), -->
<!--   #   panel.background = element_blank(),  -->
<!--   # ) -->
<!--   # ggsave("output/plots4poster/numGenes.svg",  -->
<!--   #        width = 20, height = 18, units = "cm", scale= 0.8) -->

<!-- ``` -->


<!-- ```{r} -->
<!-- # Define path -->
<!-- base_path <- "data/results/results_3" -->

<!-- # Read enrichment files -->
<!-- enrichment_dfs.2 <-  -->
<!--   list.files(base_path, recursive = TRUE, full.names = TRUE) %>% -->
<!--   grep(pattern = "network_colocalization_enrichment",  -->
<!--        value = TRUE) %>%  -->
<!--   set_names( -->
<!--     str_replace(.,  -->
<!--                 pattern = "/network_colocalization_enrichment.csv", -->
<!--                 replacement = "" -->
<!--     ) %>%  -->
<!--       str_remove( -->
<!--         pattern = paste0(base_path, "/") -->
<!--       ) -->
<!--   ) %>%    -->
<!--   map(read.csv) %>% -->
<!--   imap(~ mutate(.x, comparison = .y)) %>%  -->
<!--   bind_rows() %>% -->
<!--   dplyr::filter(z_comb == 3, z_12 == 1.5) %>% -->
<!--   pivot_longer( -->
<!--     cols = c(observed_overlap, expected_overlap_mean), -->
<!--     names_to = "type", -->
<!--     values_to = "value" -->
<!--   ) %>% -->
<!--   mutate( -->
<!--     type = recode(type, -->
<!--                   "observed_overlap" = "observed", -->
<!--                   "expected_overlap_mean" = "expected") -->
<!--   ) %>%  -->
<!--   mutate(source = "GWAS only (no parental proxy)") -->
<!-- ``` -->

<!-- ## plot observed v expected ratio -->

<!-- ```{r} -->
<!-- enrichment_dfs.2 %>%  -->
<!--   bind_rows(enrichment_dfs %>% mutate(source = "Clingen, NDKP, familial")) %>%  -->
<!--   dplyr::filter(z_comb == 3, z_12 == 1.5) %>%  -->
<!--   mutate( -->
<!--     # use the childhood vector to extract the CD seed gene name -->
<!--     CD = purrr::map_chr(comparison, ~ { -->
<!--       match <- childhood[startsWith(.x, childhood)] -->
<!--       if (length(match) > 0) match[1] else NA_character_ -->
<!--     }), -->

<!--     adult = str_remove(comparison, pattern = "CD_.+_.+_"), -->
<!--     adult = case_when( -->
<!--       adult == "malform" ~ "brain_malform",  -->
<!--       TRUE ~ adult -->
<!--     ), -->
<!--     .after = obs_exp,  -->
<!--     FDR = p.adjust(empirical_p, "fdr") -->
<!--   ) %>%  -->
<!--     mutate( -->
<!--       CD = case_when( -->
<!--       grepl(CD, pattern = "1A") ~ "Lysosomal disorders",  -->
<!--       grepl(CD, pattern = "1B") ~ "Other disorders of lipid metabolism & transport",  -->
<!--       grepl(CD, pattern = "1C") ~ "Disorders of amino acid & other organic acid metabolism",  -->
<!--       grepl(CD, pattern = "1D") ~ "Vitamin-responsive inborn errors of metabolism",  -->
<!--       grepl(CD, pattern = "1E") ~ "Disorders of mineral absorption & transport",  -->
<!--       grepl(CD, pattern = "1F") ~ "Peroxisomal diseases",  -->
<!--       grepl(CD, pattern = "1G") ~ "Mitochondrial disorders",  -->
<!--       grepl(CD, pattern = "1H") ~ "Other inborn errors of metabolism",  -->
<!--       grepl(CD, pattern = "2_") ~ "Leukodystrophies not notherwise categorised",  -->
<!--       grepl(CD, pattern = "3_") ~ "Neurodegeneration with brain iron accumulation",  -->
<!--       grepl(CD, pattern = "4_") ~ "Diseases not otherwise categorised", -->
<!--       TRUE ~ CD) %>% str_wrap(width = 20) -->
<!--     ) %>%   -->
<!--   dplyr::filter(type == "observed") %>%  -->

<!--   ggplot( -->
<!--     aes(x = obs_exp, y = reorder(adult, obs_exp)) -->
<!--   ) + -->
<!--   geom_col( -->
<!--     aes(fill = source, alpha = FDR < 0.05),  -->
<!--     colour = "grey50", position = "dodge",  -->
<!--     width = 0.75 -->
<!--   ) + -->
<!--   scale_fill_viridis_d() + -->
<!--   scale_alpha_manual(values = c(0.1, 1)) + -->
<!--   labs( -->
<!--     x = "observed / expected",  -->
<!--     y = "Disease" -->
<!--   ) + -->
<!--   facet_wrap(~CD, ncol = 3)  -->
<!--   # theme(legend.position = "top") + -->
<!--   # ggsave( -->
<!--   #   "output/figs/paper/networkcoloc_v2.svg", -->
<!--   #   width = 20, height = 20, units = "cm", -->
<!--   #   scale = 1 -->
<!--   ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- tony2keep = c("DLB", "brain_malform") -->

<!-- enrichment_dfs.2 %>%  -->
<!--   bind_rows( -->
<!--     enrichment_dfs %>%  -->
<!--       mutate(source = "Tony") %>%  -->
<!--       dplyr::filter( -->
<!--         grepl(comparison, pattern = "DLB|brain_malf") -->
<!--       ) -->
<!--   ) %>%  -->
<!--   dplyr::filter( -->
<!--     z_comb == 3, z_12 == 1.5 -->
<!--   ) %>%  -->
<!--   mutate( -->
<!--     # use the childhood vector to extract the CD seed gene name -->
<!--     CD = purrr::map_chr(comparison, ~ { -->
<!--       match <- childhood[startsWith(.x, childhood)] -->
<!--       if (length(match) > 0) match[1] else NA_character_ -->
<!--     }), -->
<!--     CD = case_when( -->
<!--       grepl(CD, pattern = "1A") ~ "Lysosomal disorders",  -->
<!--       grepl(CD, pattern = "1B") ~ "Other disorders of lipid metabolism & transport",  -->
<!--       grepl(CD, pattern = "1C") ~ "Disorders of amino acid & other organic acid metabolism",  -->
<!--       grepl(CD, pattern = "1D") ~ "Vitamin-responsive inborn errors of metabolism",  -->
<!--       grepl(CD, pattern = "1E") ~ "Disorders of mineral absorption & transport",  -->
<!--       grepl(CD, pattern = "1F") ~ "Peroxisomal diseases",  -->
<!--       grepl(CD, pattern = "1G") ~ "Mitochondrial disorders",  -->
<!--       grepl(CD, pattern = "1H") ~ "Other inborn errors of metabolism",  -->
<!--       grepl(CD, pattern = "2_") ~ "Leukodystrophies not notherwise categorised",  -->
<!--       grepl(CD, pattern = "3_") ~ "Neurodegeneration with brain iron accumulation",  -->
<!--       grepl(CD, pattern = "4_") ~ "Diseases not otherwise categorised" -->
<!--     ) %>%  -->
<!--       str_wrap(width = 22), -->
<!--     adult = str_remove(comparison, pattern = "CD_.+_.+_"), -->
<!--     adult = case_when( -->
<!--       adult == "malform" ~ "brain_malform",  -->
<!--       TRUE ~ adult -->
<!--     ), -->
<!--     .after = obs_exp,  -->
<!--     FDR = p.adjust(empirical_p, "fdr") -->
<!--   ) %>%  -->
<!--   dplyr::filter( -->
<!--     type == "observed", -->
<!--     CD != "Other inborn errors\nof metabolism" -->
<!--   ) %>%  -->
<!--   ggplot( -->
<!--     aes(x = obs_exp, y = reorder(adult, obs_exp)) -->
<!--   ) + -->
<!--   geom_col( -->
<!--     aes(fill = -log10(FDR), alpha = FDR < 0.05),  -->
<!--     colour = "grey50",  -->
<!--     position = "dodge", width = 0.7 -->
<!--   ) + -->
<!--   scale_fill_viridis_c(option = "viridis", begin = 0.6) + -->
<!--   scale_alpha_manual(values = c(0.1, 1)) + -->
<!--   labs( -->
<!--     x = "observed / expected",  -->
<!--     y = "adult-onset NDD",  -->
<!--   ) + -->
<!--   facet_wrap(~CD, nrow = 3) + -->
<!--   theme( -->

<!--     legend.position = "top", -->

<!--     text = element_text(colour = "white"), -->
<!--     axis.title = element_text(colour = "white"), -->
<!--     axis.text = element_text(colour = "white"), -->

<!--     strip.background = element_blank(),  -->
<!--     legend.background = element_blank(), -->
<!--     plot.background = element_rect(fill = "#25283D"), -->
<!--     panel.grid.minor = element_blank(), -->
<!--     panel.background = element_blank(),  -->

<!--   )  -->
<!--   # ggsave("output/plots4poster/networkcoloc_v2.svg",  -->
<!--   #        width = 45, height = 40, units = "cm",  -->
<!--   #        scale = 0.4) -->
<!--   #  -->
<!-- ``` -->


<!-- ## pathways -->
<!-- ```{r} -->
<!-- sig.networks.2 <-  -->
<!--   enrichment_dfs.2 %>%  -->
<!--   dplyr::filter(z_comb == 3, z_12 == 1.5) %>%  -->
<!--   mutate(FDR = p.adjust(empirical_p, method ="fdr")) %>%  -->
<!--   dplyr::filter(FDR < 0.05) %>%  -->
<!--   .$comparison %>%  -->
<!--   unique -->
<!-- ``` -->

<!-- ### whole subnetworks -->
<!-- ```{r} -->
<!-- node_metadata.2 <-  -->
<!--   list.files(path = "data/results/results_3",  -->
<!--              pattern = "overlapping_nodes_metadata", recursive = TRUE, full.names = TRUE) %>%  -->
<!--   set_names( -->
<!--     str_replace(.,  -->
<!--                 pattern = "/overlapping_nodes_metadata.csv", -->
<!--                 replacement = "" -->
<!--     ) %>%  -->
<!--       str_remove( -->
<!--         pattern = paste0("data/results/results_3", "/") -->
<!--       ) -->
<!--   ) %>%    -->
<!--   map(read.csv) %>% -->
<!--   imap(~ mutate(.x, comparison = .y)) -->

<!-- enrichment.wholesubnet.kegg.2 <-  -->
<!--   node_metadata.2[sig.networks.2] %>%  -->
<!--   lapply(magrittr::extract2, "X") %>%  -->
<!--   sapply(runKEGG, simplify = F) -->

<!-- enrichment.wholesubnet.kegg.2 %>%  -->
<!--   lapply(function(x) { -->
<!--     x$Description -->
<!--   }) %>%  -->
<!--   fromList() %>%  -->
<!--   UpSetR::upset( -->
<!--     nsets = length(.),  -->
<!--     order.by = "freq" -->
<!--   ) -->
<!-- ``` -->

<!-- ```{r} -->
<!--   gprofiler.wholenet.genesetsv2 <- node_metadata.2[sig.networks.2] %>%  -->
<!--   lapply(magrittr::extract2, "X") %>%  -->
<!--   sapply(function(x) { -->
<!--     gost(query = x,  -->
<!--          organism = "gp__XoLn_Cg43_BEg",  -->
<!--          ordered_query = FALSE,  -->
<!--          significant = TRUE, # only return sig results -->
<!--          user_threshold = 0.05, -->
<!--          correction_method = "fdr", -->
<!--          domain_scope = "annotated", # only analyse genes with the annotation.  -->
<!--          evcodes = TRUE, -->
<!--          custom_bg = nodes) %>%  -->
<!--       .$result %>%  -->
<!--       as_tibble -->
<!--   }, simplify = F) -->

<!-- gprofiler.wholenet.genesetsv2 %>%  -->
<!--   imap(~ mutate(.x, comparison = .y)) %>%  -->
<!--   bind_rows() %>%  -->
<!--   group_by(term_id) %>%  -->
<!--   summarise( -->
<!--     n = n() -->
<!--     ) %>%  -->
<!--   arrange(-n) -->

<!-- gprofiler.wholenet.genesetsv2 %>%  -->
<!--   lapply(extract2, "term_id") %>%  -->
<!--   fromList %>%  -->
<!--   upset( -->
<!--     nsets = length(.),  -->
<!--     order.by = "freq",  -->
<!--     mb.ratio = c(0.5,0.5) -->
<!--   ) -->
<!-- ``` -->


<!-- ### clusters -->
<!-- ```{r} -->
<!-- cluster.gprofiler.results.2 <-  -->
<!--   list.files(path = "data/results/results_3",  -->
<!--              pattern = "gprofiler", recursive = TRUE, full.names = TRUE) %>%  -->
<!--   set_names( -->
<!--     str_replace(.,  -->
<!--                 pattern = "/gprofiler_analysis_results.csv", -->
<!--                 replacement = "" -->
<!--     ) %>%  -->
<!--       str_remove( -->
<!--         pattern = paste0("data/results/results_3", "/") -->
<!--       ) -->
<!--   ) %>%    -->
<!--   map(read.csv) %>% -->
<!--   imap(~ mutate(.x, comparison = .y)) -->

<!-- # only keep the significant subnets -->
<!-- cluster.gprofiler.results.2 <- cluster.gprofiler.results.2[sig.networks.2] -->


<!-- ``` -->

<!-- # output -->
<!-- ## output files some for william -->
<!-- ```{r} -->
<!-- cluster.gprofiler.results %>%  -->
<!--   lapply(function(x) { -->
<!--     x %>%  -->
<!--       write_csv(paste0("output/network_pathways/genesets_v1/clusters/", unique(x$comparison), ".csv")) -->
<!--   }) -->

<!-- cluster.gprofiler.results.2 %>%  -->
<!--   lapply(function(x) { -->
<!--     x %>%  -->
<!--       write_csv(paste0("output/network_pathways/genesets_v2/clusters/", unique(x$comparison), ".csv")) -->
<!--   }) -->

<!-- enrichment.wholesubnet.kegg %>%  -->
<!--    imap(~ mutate(.x, comparison = .y)) %>%  -->
<!--     lapply(function(x) { -->
<!--     x %>%  -->
<!--       write_csv(paste0("output/network_pathways/genesets_v1/wholenetwork/", unique(x$comparison), ".csv")) -->
<!--     }) -->

<!-- enrichment.wholesubnet.kegg.2 %>%  -->
<!--    imap(~ mutate(.x, comparison = .y)) %>%  -->
<!--     lapply(function(x) { -->
<!--     x %>%  -->
<!--       write_csv(paste0("output/network_pathways/genesets_v2/wholenetwork/", unique(x$comparison), ".csv")) -->
<!--     }) -->

<!-- gprofiler.wholenet.genesetsv1 %>%  -->
<!--    imap(~ mutate(.x, comparison = .y)) %>%  -->
<!--     lapply(function(x) { -->
<!--     x %>%  -->
<!--       write_csv(paste0("output/network_pathways/genesets_v1/wholenetwork_gprofiler/", unique(x$comparison), ".csv")) -->
<!--     }) -->

<!-- gprofiler.wholenet.genesetsv2 %>%  -->
<!--    imap(~ mutate(.x, comparison = .y)) %>%  -->
<!--     lapply(function(x) { -->
<!--     x %>%  -->
<!--       write_csv(paste0("output/network_pathways/genesets_v2/wholenetwork_gprofiler/", unique(x$comparison), ".csv")) -->
<!--     }) -->

<!-- ``` -->
<!-- ## Figs -->
<!-- ### n genes per gs -->
<!-- ```{r} -->
<!-- genesets %>%  -->
<!--   unlist %>%  -->
<!--   as.data.frame(col.names = "gene") %>%  -->
<!--   rownames_to_column("geneset") %>% -->
<!--   set_colnames(c("geneset", "gene")) %>%  -->
<!--   mutate( -->
<!--     geneset = str_remove(geneset, pattern = "\\d{1,3}$"),  -->
<!--     geneset = fct_infreq(geneset), # Reorder by frequency,  -->
<!--     type = case_when( -->
<!--       grepl(geneset, pattern = "^CD") ~ "childhood",  -->
<!--       TRUE ~ "adult" -->
<!--     ) -->

<!--    ) %>%  -->
<!--   ggplot( -->
<!--     aes(x = geneset) -->
<!--   ) + -->
<!--   geom_bar() + -->
<!--   easy_rotate_x_labels(angle = -45) + -->
<!--   labs( -->
<!--     title = "Number of genes per gene set",  -->
<!--     y = "# genes" -->
<!--   ) + -->
<!--   facet_wrap(~type, scales = "free_x") + -->
<!--   theme( -->
<!--     plot.margin = margin(r = 2, unit = "cm") -->
<!--   ) + -->
<!--     ggsave( -->
<!--     "output/adult child subtypes/figs/talk/numgenespergs.svg",  -->
<!--     width = 8, height = 4, units = "cm",  -->
<!--     scale = 2 -->
<!--   ) -->
<!-- ``` -->

<!-- ### netcoloc -->
<!-- ```{r} -->
<!-- enrichment_dfs %>%  -->
<!--   bind_rows() %>%  -->
<!--   as_tibble %>%  -->
<!--   dplyr::filter(z_comb == 3, z_12 == 1.5) %>%  -->
<!--  mutate( -->
<!--    comparison = str_remove(comparison, pattern = "^CD_\\d{1,2}\\w{0,1}_"), -->
<!--    comparison = str_replace(comparison,  "other_IEMs", "otherIEMs"),  -->
<!--    comparison = str_replace(comparison,  "mineral_abs", "mineralabs"), -->
<!--    comparison = str_replace(comparison,  "brain_malform", "brainmalform") -->
<!--  ) %>%  -->
<!--    separate(comparison, into = c("CD", "adult"), sep = "_", remove = F) %>%  -->
<!--   arrange(empirical_p) %>%  -->
<!--   mutate( -->

<!--     # use the childhood vector to extract the CD seed gene name -->
<!--     CD = purrr::map_chr(comparison, ~ { -->
<!--       match <- childhood[startsWith(.x, childhood)] -->
<!--       if (length(match) > 0) match[1] else NA_character_ -->
<!--     }), -->
<!--     CD = case_when( -->

<!--       grepl(CD, pattern = "lyso") ~ "Lysosomal disorders",  -->
<!--       grepl(CD, pattern = "otherlip") ~ "Other disorders of lipid metabolism & transport",  -->
<!--       grepl(CD, pattern = "aminoa") ~ "Disorders of amino acid & other organic acid metabolism",  -->
<!--       grepl(CD, pattern = "vitamin") ~ "Vitamin-responsive inborn errors of metabolism",  -->
<!--       grepl(CD, pattern = "mineral") ~ "Disorders of mineral absorption & transport",  -->
<!--       grepl(CD, pattern = "perox") ~ "Peroxisomal diseases",  -->
<!--       grepl(CD, pattern = "mito") ~ "Mitochondrial disorders",  -->
<!--       grepl(CD, pattern = "otherIE") ~ "Other inborn errors of metabolism",  -->
<!--       grepl(CD, pattern = "leukod") ~ "Leukodystrophies not otherwise categorised",  -->
<!--       grepl(CD, pattern = "NBIA") ~ "Neurodegeneration with brain iron accumulation",  -->
<!--       grepl(CD, pattern = "notcat") ~ "Diseases not otherwise categorised",  -->
<!--       TRUE ~ CD -->
<!--     ) %>%  -->
<!--       str_wrap(width = 22), -->

<!--     facet_order = case_when( -->
<!--       grepl(CD, pattern = "Diseases not") ~ "2",  -->
<!--       grepl(CD, pattern = "Other inborn") ~ "2",  -->
<!--       TRUE ~ "1" -->
<!--     ), -->
<!--     adult = str_remove(comparison, pattern = "^.+_"), -->
<!--     adult = case_when( -->
<!--       adult == "malform" ~ "brain_malform",  -->
<!--       TRUE ~ adult -->
<!--     ), -->
<!--     FDR = p.adjust(empirical_p, "fdr") -->
<!--   ) %>%  -->
<!--   dplyr::filter(type == "observed") %>%  -->
<!--   arrange(facet_order, FDR) %>%  -->
<!--   mutate( -->
<!--     CD = fct_inorder(CD),  -->
<!--     adult = case_when( -->
<!--       adult == "brainmalform" ~ "brain_malform",  -->
<!--       TRUE ~ adult -->
<!--     ) -->
<!--   ) %>%  -->
<!--   ggplot( -->
<!--     aes(x = obs_exp, y = reorder(adult, obs_exp)) -->
<!--   ) + -->
<!--   geom_col( -->
<!--     aes(fill = -log10(FDR), alpha = FDR < 0.05),  -->
<!--     colour = "grey50" -->
<!--   ) + -->
<!--   scale_fill_viridis_c(option = "viridis") + -->
<!--   scale_alpha_manual(values = c(0.1, 1)) + -->
<!--   labs( -->
<!--     x = "observed / expected",  -->
<!--     y = "Disease" -->
<!--   ) + -->
<!--   facet_wrap(~CD, ncol = 3) + -->



<!--     ggsave( -->
<!--     "output/figs/paper/networkcoloc.svg",  -->
<!--     width = 20, height = 15, units = "cm",  -->
<!--     scale = 1 -->
<!--   ) -->
<!-- ``` -->

<!-- ### mouse variants -->
<!-- ```{r} -->
<!-- mouse.variant.wholesubnetworks %>% -->
<!--   dplyr::filter( -->
<!--     CD == "CD_1A_lyso", -->
<!--     adult %in% c("DLB", "PD") -->
<!--   ) %>%  -->
<!--   group_by(comparison, MPO_term_type) %>% -->
<!--   slice_min(order_by = OR_p, n = 10) %>% -->
<!--   ungroup() %>%  -->
<!--   mutate( -->
<!--     MP_description = reorder_within( -->
<!--       MP_description, log_OR, comparison),  -->
<!--     comparison = case_when( -->
<!--       grepl(comparison, pattern = "DLB") ~ "Lysosomal disorders\n&\ndementia with Lewy bodies",  -->
<!--       grepl(comparison, pattern = "PD") ~ "Lysosomal disorders\n&\nParkinson's disease",  -->
<!--     ) -->

<!--   ) %>% -->
<!--   ggplot( -->
<!--     aes(x = log_OR, y = MP_description) -->
<!--   ) + -->
<!--   geom_errorbarh( -->
<!--     aes(xmin = log_OR_CI_lower, xmax = log_OR_CI_upper), -->
<!--     height = 0.2, color = "black" -->
<!--   ) + -->
<!--   geom_point( -->
<!--     aes(size = num_genes_in_term, color = MPO_term_type), -->
<!--     alpha = 0.6 -->
<!--   ) + -->
<!--   geom_vline(xintercept = 0, lty = 2, color = "grey50") + -->
<!--   geom_label( -->
<!--     aes(label = signif(OR_p, 2)), -->
<!--     nudge_y = 0.4 -->
<!--   ) + -->
<!--   scale_y_reordered() +  # This removes the `___comparison` suffix added by reorder_within() -->
<!--   labs( -->
<!--     x = "log(OR) ± 95% CI", -->
<!--     y = NULL, -->
<!--     size = "# Genes in MPO term", -->
<!--     color = "Term Type" -->
<!--   ) + -->
<!--   facet_wrap(~comparison, nrow = 1, -->
<!--              scales = "free" -->
<!--   ) + -->
<!--   theme( -->
<!--     text = element_text(size = 18) -->
<!--   ) + -->
<!--   ggsave( -->
<!--     "output/figs/paper/mouse_variant_enrichment_lyso_DLB_PD.svg",  -->
<!--     width = 20, height = 7, scale = 1 -->
<!--   ) -->
<!-- ``` -->

<!-- ### simialrity of DLB and PD clusters -->
<!-- ```{r} -->
<!-- # Define a helper to calculate Jaccard Similarity -->
<!-- calc_jaccard <- function(list_a, list_b) { -->
<!--   intersection <- length(intersect(list_a, list_b)) -->

<!--   union <- length(list_a) + length(list_b) - intersection -->

<!--   if (union == 0) return(0) -->
<!--   return(intersection / union) -->
<!-- } -->

<!-- # extract DLB and PD clusters -->
<!-- dlb_clusters <-  -->
<!--   cluster.gprofiler.results$CD_1A_lyso_DLB %>%  -->
<!--   dplyr::select(Cluster, genes) %>%  -->
<!--   split(f = .$Cluster) %>%  -->
<!--   lapply(function(x) { -->
<!--      x$genes %>%  -->
<!--       str_split(" ") %>%  -->
<!--       unlist() -->
<!--   }) -->

<!-- pd_clusters <- cluster.gprofiler.results$CD_1A_lyso_PD %>%  -->
<!--   dplyr::select(Cluster, genes) %>%  -->
<!--   split(f = .$Cluster) %>%  -->
<!--   lapply(function(x) { -->
<!--      x$genes %>%  -->
<!--       str_split(" ") %>%  -->
<!--       unlist -->
<!--   }) -->

<!-- # Create an empty matrix -->
<!-- sim_matrix <- matrix(NA,  -->
<!--                      nrow = length(dlb_clusters),  -->
<!--                      ncol = length(pd_clusters), -->
<!--                      dimnames = list(names(dlb_clusters), names(pd_clusters))) -->

<!-- # Fill the matrix -->
<!-- for (d_name in names(dlb_clusters)) { -->
<!--   for (p_name in names(pd_clusters)) { -->
<!--     sim_matrix[d_name, p_name] <-  -->
<!--       calc_jaccard(dlb_clusters[[d_name]], pd_clusters[[p_name]]) -->
<!--   } -->
<!-- } -->


<!-- # 1. Create heatmap (DO NOT print) -->
<!-- ph <- pheatmap( -->
<!--   sim_matrix, -->
<!--   display_numbers = TRUE, -->
<!--   main = "Jaccard index of LysoPD and LysoDLB network communities", -->
<!--   color = colorRampPalette(c("white", "red"))(100), -->
<!--   silent = TRUE -->
<!-- ) -->

<!-- # Convert pheatmap gtable to a patchwork-compatible object -->
<!-- heatmap_plot <- wrap_elements(full = ph$gtable) -->

<!-- # 2. Create Venn diagram -->
<!-- venn_plot <- ggVennDiagram::ggVennDiagram( -->
<!--   list( -->
<!--     lysoPD  = node_metadata$CD_1A_lyso_PD$X, -->
<!--     lysoDLB = node_metadata$CD_1A_lyso_DLB$X -->
<!--   ), -->
<!--   label = "count", -->
<!--   label_color = "white", -->
<!--   label_size = 6,  -->
<!-- ) + -->
<!--   scale_fill_viridis_c(begin = 0.2) + -->
<!--   theme_void() -->

<!-- # 3. Combine into a single image -->
<!-- venn_plot + heatmap_plot + -->
<!--   plot_layout(widths = c(1, 2), ) + -->
<!--   ggsave("output/figs/venndiagramandjaccard.svg") -->

<!-- ``` -->



