---
title: "05_twimicemyeloid"
author: "Karissa Barthelson"
date: "2025-08-04"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r}
library(tidyverse)
library(forcats)
library(magrittr)
library(readxl)
library(qs) # for fast save and reading of objects into memory (alternative to rds)
library(parallel) 
library(Matrix)

library(GEOquery)
library(Seurat)
library(org.Mm.eg.db)
library(edgeR)
library(clusterProfiler)

# vis
library(pander)
library(patchwork)
library(scales)
library(ggpubr)
library(ggeasy)
library(ggfortify)
library(ComplexHeatmap)

theme_set(theme_light() +
            theme(plot.title = element_text(hjust = 0.5), 
                  strip.text = element_text(face = "bold"),
                  strip.background = element_rect(fill = "grey20")
            )
)
```

This page details the processing for the single cell RNAseq dataset of myeloid cells isolated from the brains of the twitcher mouse model, a model of Krabbe disease. 

This dataset was generated by [Aisenberg et al. (2025)](https://doi.org/10.1016/j.immuni.2025.03.019) and it contains single cell RNAseq data from twitcher mice (Galc W355*) and wild type mice at two time points: P15 (presymptomatic) and P35 (symptomatic). 

The scRNAseq data was processed using cellranger. Here, it will be analysed using Seurat. 
## read in metadata
```{r}
# get the metadaya from GEO
gse <- getGEO("GSE288018", GSEMatrix = TRUE)

meta <- phenoData(gse[[1]]) %>% 
  pData %>% 
  as_tibble %>% 
  dplyr::select(title, geo_accession) %>% 
  separate(title,
            into = c("cell_type", "genotype", "treatment", "age", "sample"), 
           sep = ", "
           ) %>% 
  left_join(
    read_csv("data/public data/PRJNA1215700_twiMiceMyeloid/SraRunTable.csv") %>%
  dplyr::select(
    sra_accession = Run, geo_accession =  `Library Name`, batch
  ) 
  ) %>% 
  dplyr::filter(treatment == "untreated") %>% 
  mutate(
    genotype = as.factor(genotype), 
    age = as.factor(age)
  )

meta %>% 
  ggplot(
    aes(x = age, fill = genotype)
  ) +
  geom_bar(
    position = "dodge"
  ) 
```

## read in seurat object
```{r}
cellranger.parent.dir = "data/public data/PRJNA1215700_twiMiceMyeloid/cellranger"

# create named objects of the path and sample id for each filtered dataset. 
# this largely removes empty droplets
sample.paths.filt = 
  grep(list.dirs(cellranger.parent.dir), pattern = "filter", value = TRUE)

sample.ids <- str_extract(sample.paths.filt, pattern = "SRR\\d{6,8}") 
sample.paths.filt <- setNames(sample.paths.filt, sample.ids)

# read in filtered data and convert to seurat object
data.10x.filt <-  Read10X(
  # get the filtered data from cellranger  
    data.dir = sample.paths.filt, 
  )

data.seurat <- CreateSeuratObject(
  counts = data.10x.filt, 
  project = "twi_myeloid", 
  min.features = 200, # include cells with at least this many genes. 
  min.cells = 3 # only retain genes expressed in at least this many nuclei
  )

# join the metadata to the seurat object
data.seurat@meta.data %<>% 
  as_tibble(rownames = "umi") %>% 
  left_join(
    meta %>% 
      dplyr::rename(orig.ident = sra_accession)
  ) %>% 
  column_to_rownames("umi")
```

# QC
## histogram of lib sizes

```{r}
data.seurat@meta.data %>% 
  ggplot(
    aes(x = nCount_RNA, fill = orig.ident)
  ) +
  geom_histogram(binwidth = 10) +
  coord_cartesian(xlim = c(0,10000)) + 
  facet_wrap(~orig.ident+genotype) +
  scale_y_log10() +
  labs(
    y = "log10(counts)"
  )
```

## percent mt transcripts

When a cell has a low number of detected genes, a low count depth and a high fraction of mitochondrial counts it might have a broken membrane which can indicate a dying cell. As these cells are usually not the main target of our analysis and might distort our downstream analysis, we are removing them during quality control. 

```{r}
data.seurat[["percent.mt"]] <- 
  PercentageFeatureSet(data.seurat, pattern = "^mt-")

# Visualize QC metrics as a violin plot
VlnPlot(data.seurat, 
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), 
        split.by = "genotype",
        pt.size = 0,
        ncol = 3, alpha = 0.25) +
  theme(aspect.ratio = 1)
```

## filtering
I will filter the data based on the number of features (genes) detected per cell, the number of counts per cell and the percentage of mitochondrial genes expressed.

```{r}
data.seurat %<>% 
  subset(
    subset = nCount_RNA >= 1900 & nFeature_RNA > 200 & nFeature_RNA < 3000 & percent.mt < 10
  ) 

VlnPlot(data.seurat, 
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), 
        split.by = "genotype",
        pt.size = 0,
        ncol = 3, alpha = 0.25) +
  theme(aspect.ratio = 1)
```

# seurat pipeline
```{r}
temp <- data.seurat %>% 
   NormalizeData() %>% 
   FindVariableFeatures() %>% 
   ScaleData() %>% 
   RunPCA()

ElbowPlot(temp, ndims = 30)
remove(temp)
nPCs = 20

data.seurat %<>% 
  NormalizeData() %>% 
  FindVariableFeatures() %>% 
  ScaleData() %>% 
  RunPCA() %>% 
  FindNeighbors(dims = 1:nPCs) %>% 
  FindClusters(resolution = 0.1) %>% # less clusters
  RunUMAP(dims = 1:nPCs)

DimPlot(
  data.seurat, 
  split.by = "age", 
  label = TRUE, 
  label.box = TRUE,
  label.color = "white", 
  label.size = 8
  )
  theme_light()

DimPlot(data.seurat, 
         group.by = "genotype", 
        split.by = "age",
        shuffle = TRUE
         ) +
  ggtitle("UMAP of all cells by genotype and age") +
  theme_light()
```

## marker genes
The original authors detected 5 classes of microglia, a border-associated macrophage-like (BAM+, neutrophils and a mixed glia cluster

Here, I will look at their top marker genes to assess whether i am detecting similar clusters. 
```{r}
runFeatureplot = function(seurat.object, genes, celltype) {
  DimPlot(seurat.object,label = T) +
  FeaturePlot(
    seurat.object,
    features= genes
  ) + 
  # add the title to the plot
    plot_annotation(
    title = celltype,  
    theme = theme(plot.title = element_text(size = 18, face = "bold"))
    )
}
```


```{r}

runFeatureplot(
  data.seurat, 
  genes = c("P2ry12", "Ccl12", "Ccl4", "Apoe", "Mki67", "Gpr34"),
  celltype = "Microglia (all clusters)"
)

runFeatureplot(
  data.seurat, 
  genes = c("S100a9", "S100a8", "Camp", "Ngp"),
  celltype = "Neutrophils"
)

runFeatureplot(
  data.seurat, 
  genes = c("Lyz2", "Cd74", "S100a4", "Crip1"),
  celltype = "BAM+"
)

runFeatureplot(
  data.seurat, 
  genes = c("Slc1a2", "Mt3", "Mobp", "Clu"),
  celltype = "Mixed glia"
)
```

```{r}
data.seurat@meta.data %<>% 
  left_join(
    tibble(
      RNA_snn_res.0.1 = as.factor(0:8)
    ) %>% 
      mutate(
        celltype = case_when(
          RNA_snn_res.0.1 == 0 ~ "Microglia", 
          RNA_snn_res.0.1 == 1 ~ "Microglia", 
          RNA_snn_res.0.1 == 2 ~ "Microglia",
          RNA_snn_res.0.1 == 3 ~ "Neutrophil",
          RNA_snn_res.0.1 == 4 ~ "Microglia",
          RNA_snn_res.0.1 == 5 ~ "BAM",
          RNA_snn_res.0.1 == 6 ~ "BAM",
          RNA_snn_res.0.1 == 7 ~ "Microglia",
          RNA_snn_res.0.1 == 8 ~ "Mixedglia",
        )
      )
  )
```

# pseudobulk
```{r}
## prepare DGE object
y <- Seurat2PB(data.seurat, sample="orig.ident", cluster="celltype")

y$samples %<>% 
  as_tibble(rownames = "sample_cluster") %>% 
  dplyr::rename("sra_accession" = "sample") %>% 
  left_join(meta) %>% 
  mutate(
    sample_cluster = str_remove(sample_cluster, "cluster"), 
    celltype = str_remove(sample_cluster, pattern = "SRR.+_"),
    group = paste0(celltype, "_", age, "_", genotype) %>% as.factor(),
    age_geno = paste0(age, "_", genotype) %>% 
      factor(levels = c("P15_WT", "P15_Twi", "P35_WT", "P35_Twi"))
  ) %>% 
  column_to_rownames("sample_cluster")

colnames(y$counts) %<>% str_remove("cluster")
```

## lib.sizes
```{r}
y$samples %>% 
  ggplot(
    aes(x = sra_accession, y = lib.size, fill = age_geno, alpha = lib.size > 20000)
  ) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  scale_fill_manual(values = c("grey70", "#ffa694", "grey20", "#ff5736")) +
  scale_alpha_manual(values = c(0.1, 1)) +
  facet_wrap(~celltype, scales = "free")
```

## filtering samples
Only retained cell type clusters per age if the cluster contained at least 20k reads, and had two samples per genotype. 

```{r}
# retain samples with more than 20k counts
keep.samples <- y$samples$lib.size > 2e4
y <- y[,keep.samples]

# check which samples only one sample per genotype age
y$samples %>% 
  ggplot(
    aes(x = sra_accession, y = lib.size, fill = age_geno)
  ) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +
  facet_wrap(~celltype, scales = "free")

y$samples %>% 
  group_by(age, genotype, celltype) %>% 
  summarise(n = n()) %>% 
  dplyr::filter(n < 2) 

# only retain samples where the group has at least 3 samples
y <- y[,y$samples$celltype != "Mixedglia"]

# check which samples only one sample per genotype age
y$samples %>% 
  ggplot(
    aes(x = sra_accession, y = lib.size, fill = age_geno)
  ) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = comma) +
   scale_fill_manual(values = c("grey70", "#ffa694", "grey20", "#ff5736")) +
  facet_wrap(~celltype+age, scales = "free", nrow = 3)
```

## filtering genes
```{r}
design <- model.matrix(~0 + group, data = y$samples %>% 
                         droplevels()) 
colnames(design) <- str_remove(colnames(design), 
                               "group")
```

```{r}
keep.genes <- filterByExpr(y, group=y$samples$cluster, design = design)
table(keep.genes)

y <- y[keep.genes, , keep=FALSE]

y <- normLibSizes(y, method = "TMM")

y$samples %>% 
  ggplot(
    aes(y = lib.size*norm.factors, x = cluster, colour = age_geno )
  ) +
  geom_boxplot() +
  geom_point(position = position_jitterdodge()) +
  stat_compare_means() +
  facet_wrap(~cluster, scales = "free_x")

y %>% 
  cpm(log = T) %>% 
  as.data.frame() %>% 
  pivot_longer(cols = everything(), 
               names_to = "sample", values_to = "logCPM") %>% 
  left_join(y$samples %>% 
              mutate(sample = rownames(y$samples))) %>% 
  ggplot(
    aes(x = sample, y = logCPM)
    ) +
  geom_boxplot(
    aes(fill = age_geno),
    outlier.size = 0.5
               ) +
  facet_wrap(~cluster, scales = "free") +
  easy_rotate_x_labels(angle = -45) +
  labs(title = "Distribution of logCPM values per sample",
       x = "Sample", y = "log2 CPM")
```

## PCA
```{r}
y$counts %>% 
  cpm(log = TRUE) %>% 
  t %>% 
  prcomp %>% 
  autoplot(
    data = tibble(
      sample = rownames(.$x)
    ) %>% 
      left_join(y$samples %>% 
              mutate(sample = rownames(y$samples))
              ),
    colour = "genotype", 
    shape = "age",
    alpha = 0.5,
    size = 6
  ) +
  theme(aspect.ratio = 1) +
  facet_wrap(~cluster, scales = "free")
```

## DE analysis

### dispersion estimate
```{r}
y <- estimateDisp(y, design, robust = TRUE)
plotBCV(y)
```

## glmFit
```{r}
fit <- glmQLFit(y, design, robust = TRUE)
plotQLDisp(fit)
```

## contrasts matrix
```{r}
# Extract cell types shared by both Control and DLB
design_cols <- colnames(design)

design_df <- tibble(colname = colnames(design)) %>%
  separate(colname, into = c("celltype", "age", "genotype"), sep = "_") %>%
  unite("group_id", celltype, age, remove = FALSE)

# Filter to groups that contain both WT and Npc_hom
contrast_df <- 
  design_df %>%
  mutate(colname = paste0(group_id, "_", genotype)) %>% 
  group_by(group_id) %>% 
  pivot_wider(names_from = genotype, values_from = colname) %>% 
  mutate(contrast = paste0(Twi, " - ", WT))

# Build contrast matrix
contrast_matrix <- makeContrasts(contrasts = contrast_df$contrast, levels = design)

# rename columns for readability
colnames(contrast_matrix) <- gsub("group", "", colnames(contrast_matrix))

```

## toptables
```{r}
toptables <- colnames(contrast_matrix) %>% 
  sapply(function(x) {
      glmQLFTest(fit, design, contrast=contrast_matrix[,x]) %>% 
      topTags(n = Inf, sort.by = "PValue") %>% 
      .$table %>% 
      as_tibble %>% 
      dplyr::select(
        gene, logCPM, logFC, PValue, FDR, everything()
      ) %>% 
      mutate(
        coef = x, 
        DE = FDR < 0.05
      )
  }, simplify = F)
```
## volcanoplots
```{r}
volcano_data <- 
  toptables %>%
  bind_rows() %>% 
  mutate(
  coef = str_remove(coef, pattern = "_Npc1.+_WT$"),
    DE = FDR < 0.05
  ) 

# Count number of DE up/down per celltype (coef)
volcano_labels <- volcano_data %>%
  dplyr::filter(DE) %>%
  mutate(direction = ifelse(logFC > 0, "Up", "Down")) %>%
  group_by(coef, direction) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(
    x = ifelse(direction == "Up", 3, -3),  # approximate x positions
    y = max(-log10(volcano_data$PValue), na.rm = TRUE) * 0.9,  # high y
    label = paste0(direction, ": ", n)
  )

# Plot
ggplot(volcano_data, aes(x = logFC, y = -log10(PValue))) +
  geom_point(aes(colour = DE)) +
  scale_color_manual(values = c("grey50", "red")) +
  facet_wrap(~coef) +
  geom_text(
    data = volcano_labels,
    aes(x = x, y = y, label = label),
    inherit.aes = FALSE,
    size = 4,
    hjust = 0.5
  ) +
  labs(
    x = "log2 Fold Change",
    y = "-log10(P value)"
  ) 
```

## MD plots

```{r}
toptables %>% 
  bind_rows() %>% 
  ggplot(
    aes(x = logCPM, y = logFC)
  ) +
  geom_point(
    aes(colour = DE )
  ) +
  facet_wrap(~coef) +
  scale_color_manual(values = c("grey50", "red"))
```

# enrichmenmt andlysis 
## CD pathways
Here, I need to convert the human gene id to mouse id. Network genes were only retained if they were detected in any cell type and had a 1 to 1 mouse orthologue. 

```{r}
# import clusters
cluster.gprofiler.results <- 
  list.files(path = "data/results/adultchildsubtypes", 
             pattern = "gprofiler", recursive = TRUE, full.names = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/gprofiler_analysis_results.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/adultchildsubtypes", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))

network.genes.cluster <- 
  cluster.gprofiler.results$CD_1A_lyso_DLB %>% 
  dplyr::select(Cluster, genes) %>% 
  mutate(genes = str_split(genes, " ")) %>% 
  deframe() %>% 
  lapply(function(x) {
    x %>% 
      as_tibble %>% 
      set_colnames("human")
  })

# download the mouse2human id conversion
mouse_human_genes = read.csv("http://www.informatics.jax.org/downloads/reports/HOM_MouseHumanSequence.rpt",sep="\t")

convert_mouse_to_human <- function(gene_list){

# Filter to mouse genes in gene_list
  mouse_subset <- mouse_human_genes %>%
    dplyr::filter(Common.Organism.Name == "mouse, laboratory", Symbol %in% gene_list)

  # Join with human genes via DB.Class.Key
  result <- mouse_subset %>%
    left_join(
      mouse_human_genes %>%
        dplyr::filter(Common.Organism.Name == "human") %>%
        dplyr::select(DB.Class.Key, human_symbol = Symbol),
      by = "DB.Class.Key"
    ) %>%
    dplyr::select(mouse = Symbol, human = human_symbol) %>%
    dplyr::distinct() %>% 
    as_tibble

  return(result)
}

network.genes.cluster %<>% 
  lapply(function(x) {
    x %>% 
      left_join(convert_mouse_to_human(rownames(y))) %>% 
      na.omit() %>% 
      .$mouse %>% 
      unique()
  })

# for gene set testing, only retain clusters with more than 3 genes
network.genes.cluster <- network.genes.cluster[lapply(network.genes.cluster, length) > 3]

  
### obtain the genes in the subnetworks
node_metadata <- 
  list.files(path = "data/results/adultchildsubtypes", 
             pattern = "overlapping_nodes_metadata", recursive = TRUE, full.names = TRUE) %>% 
  set_names(
    str_replace(., 
                pattern = "/overlapping_nodes_metadata.csv",
                replacement = ""
    ) %>% 
      str_remove(
        pattern = paste0("data/results/adultchildsubtypes", "/")
      )
  ) %>%   
  map(read.csv) %>%
  imap(~ mutate(.x, comparison = .y))

network.genes <- node_metadata$CD_1A_lyso_DLB

network.genes %<>% 
  dplyr::select(human = X, everything()) %>% 
  left_join(convert_mouse_to_human(rownames(y))) %>% 
  na.omit %>% 
  dplyr::select(gene_name = mouse, everything())
```

### ORA

#### ORA of pathways in the clusters
first, I perfored ORA analysis on the clusters themselves. this is to give biological context to the genes in each cluster. 
```{r}
ORA.clusters <- network.genes.cluster %>% 
  lapply(function(x) {
    clusterProfiler::enrichKEGG(
      gene = bitr(x, 
                  OrgDb = org.Mm.eg.db,
                  fromType="SYMBOL", toType="ENTREZID") %>% .$ENTREZID,
      keyType = "ncbi-geneid", 
      organism = "mmu"
    ) 
    })
```

```{r}
ORA.clusters %>% 
  lapply(function(x) {
    x %>% 
      as_tibble 
  }) %>% 
  bind_rows(.id = "cluster") %>% 
  ggplot(
    aes(x = -log10(p.adjust), y = Description, fill = cluster)
  ) +
  geom_col() +
  facet_wrap(~cluster, scales = "free_y")
```


#### number of DE genes per cluster

I then tested whether each of the netcoloc clusters are significantly overrepresented with DE genes within each cell type and age. 
```{r}
ora.DEG.in.clusters <- toptables %>%
  lapply(function(x) {
    genes <- x %>%
      dplyr::filter(DE == TRUE) %>%
      pull(gene)
    
    if (length(genes) < 3) return(NULL)  # skip if too few genes
    
    clusterProfiler::enricher(
      gene = genes,
      TERM2GENE = network.genes.cluster %>%
        enframe(name = "TERM", value = "GENE") %>%
        unnest(cols = "GENE"), 
      universe = x$gene
    )
  }) %>%
  purrr::compact()
```

#### pvals
```{r}
ora.DEG.in.clusters %>% 
  lapply(as_tibble) %>% 
  bind_rows(.id = "comparison") %>% 
  mutate(
    comparison = str_extract(comparison, 
                             pattern = "P.+")
  ) %>% 
  ggplot(
    aes(x = -log10(p.adjust), y = ID, fill = comparison)
  ) +
  geom_col(position = "dodge")
```

Clusters 438 is significantly enriched with DEGs in microglia. Both ages. 
```{r}
keep = toptables %>% 
  bind_rows() %>% 
  dplyr::filter(
    gene %in% network.genes.cluster$C438
  ) %>% 
  dplyr::filter(DE == T) %>% 
  .$gene

  # Build the FDR matrix (for asterisk overlay)
  FDR_mat <- toptables %>%
    bind_rows() %>%
    dplyr::filter(gene %in% keep) %>%
    dplyr::select(coef, gene, FDR) %>%
    mutate(
      coef = str_extract(coef, 
                             pattern = "P.+")
    ) %>% 
    pivot_wider(names_from = gene, values_from = FDR) %>%
    column_to_rownames("coef") %>%
    as.matrix()

Heatmap(
  matrix = toptables %>%
    bind_rows() %>%
    dplyr::filter(gene %in% keep) %>%
    dplyr::select(coef, gene, logFC) %>%
      mutate(
      coef = str_extract(coef, 
                             pattern = "P.+")
    ) %>% 
    pivot_wider(names_from = gene, values_from = logFC) %>%
    column_to_rownames("coef") %>%
    as.matrix(),
  name = "logFC",
  col = brewer_pal(palette = "RdBu")(9), 
  
  cell_fun =function(j, i, x, y, width, height, fill) {
      # j = column index, i = row index
      # x, y = coordinates of the center of the cell
      # width, height = size of the cell
      # fill = background color (not used here)

      # Check if the FDR value for this cell exists and is below 0.05
      if (!is.na(FDR_mat[i, j]) && FDR_mat[i, j] < 0.05) {
        # If yes, overlay a text label "*" on that cell
        grid.text("*", x, y, gp = gpar(fontsize = 15))
      }
    },
  column_names_gp = gpar(fontsize = 10),
  border = TRUE,
  show_row_names = TRUE,
  show_column_names = TRUE,
  cluster_rows = TRUE,
  cluster_columns = TRUE
)

```



```{r}
plot.cnet = function(toptable, plot.title, ORA.cluster.object) {
  # 1. Get toptable for the cluster you're plotting
top.genes <- toptable %>%
   dplyr::filter(PValue < 0.05) %>% 
  dplyr::select(gene, logFC)

# 2. Map gene symbols to ENTREZID
fc.entrez <- bitr(
  top.genes$gene,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Mm.eg.db
) %>%
  left_join(top.genes, by = c("SYMBOL" = "gene")) %>%
  dplyr::distinct(ENTREZID, .keep_all = TRUE) %>%
  dplyr::select(-SYMBOL) %>% 
  deframe()  # named vector: names are ENTREZID, values are logFC

cnetplot(
  x = setReadable(ORA.cluster.object, 
                  OrgDb = org.Mm.eg.db, keyType = "ENTREZID"),
  foldChange = fc.entrez,
  showCategory = 20,
  circular = FALSE, 
  color_category='grey20', 
  size_category = 0.5,
  color_item = "grey90", 
  colorEdge = F
) +
  scale_color_gradient2(name='logFC', low='blue4', high='firebrick4') + 
  ggtitle(plot.title)
  
}

```


## fry gene set testing
```{r}
fry <- colnames(contrast_matrix) %>% 
  sapply(function(x) {
   cpm(y, log = T)  %>% 
      fry(
        index = network.genes.cluster,
        design = design, 
        contrast = contrast_matrix[,x], 
        sort = "directional"
      ) %>% 
      rownames_to_column("pathway") %>% 
      as_tibble() %>% 
      mutate(
        coef = x
        )
  }, simplify = FALSE)

fry
sig.clusters.fry <- fry %>% 
  bind_rows() %>% 
  dplyr::filter(
    FDR < 0.05 | FDR.Mixed < 0.05,
    NGenes > 1
  ) %>% 
  .$pathway %>% 
  unique
  
fry %>% 
  bind_rows() %>% 
  dplyr::filter(pathway %in% sig.clusters.fry) %>% 
  pivot_longer(
    cols = c(PValue, PValue.Mixed),
    names_to = "pval_type",
    values_to = "pval"
  ) %>%
  mutate(
    neglog10_p = -log10(pval),
    pval_type = recode(pval_type, "PValue" = "Directional", "PValue.Mixed" = "Mixed"),
    fdr = case_when(
      pval_type == "Directional" ~ FDR,
      pval_type == "Mixed" ~ FDR.Mixed
    ),
    sig = fdr < 0.05,
    alpha_val = if_else(sig, 1, 0.3),
    pathway = factor(pathway, levels = unique(pathway))
  ) %>% 
  ggplot(aes(x = pval_type,
             y = pathway, fill = neglog10_p, alpha = alpha_val)) +
  geom_tile() +
  facet_wrap(~coef) +
  scale_fill_viridis_c(option = "plasma", name = "-log10(P)") +
  scale_alpha(range = c(0.3, 1), guide = "none") +
  labs(
    x = "Hypothesis test",
    y = "Netcoloc cluster"
  ) 
```

```{r}
sig.clusters.fry <- fry %>% 
  bind_rows() %>% 
  dplyr::filter(
    grepl(coef, pattern = "^Microglia"),
    FDR.Mixed < 0.05
  ) %>% 
  .$pathway %>% 
  unique

toptables$`Microglia_3week_Npc1 - Microglia_3week_WT` %>% 
  dplyr::filter(
    gene %in% network.genes.cluster$C439
  ) %>% 
  dplyr::select(gene, logFC, FDR) %>%
  mutate(age = "3week") %>% 
  rbind(
    toptables$`Microglia_7week_Npc1 - Microglia_7week_WT` %>% 
      dplyr::filter(
        gene %in% network.genes.cluster$C439
      ) %>% 
      dplyr::select(gene, logFC, FDR) %>%
      mutate(age = "7week") 
  ) %>% 
  ggplot(
    aes(x = age, y = reorder(gene, logFC), fill = logFC)
  ) + 
  geom_tile() +
  geom_label(
    aes(label = signif(FDR,1)), 
    data = . %>% 
      dplyr::filter(FDR < 0.05)
    ) +
  scale_fill_distiller(palette  = "RdBu") +
  labs(
    y = "gene", 
    title = "Netcoloc cluster 439"
  )
```





# export data
```{r}
data.seurat %>% 
  qsave("data/public data/PRJNA1215700_twiMiceMyeloid/R_objects/dataseurat.qs")
    
y %>% 
  qsave("data/public data/PRJNA1215700_twiMiceMyeloid/R_objects/pseudobulk_dgeObject.qs")

ORA.clusters %>% 
  qsave("data/public data/PRJNA1215700_twiMiceMyeloid/R_objects/ORA_clusters.qs")

ora.DEG.in.clusters %>% 
  qsave("data/public data/PRJNA1215700_twiMiceMyeloid/R_objects/ORA_DEGs_clusters.qs")

fry %>% 
  qsave("data/public data/PRJNA1215700_twiMiceMyeloid/R_objects/fry_results.qs")

toptables %>% 
  qsave("data/public data/PRJNA1215700_twiMiceMyeloid/R_objects/toptables.qs")
```














